### 2013-09-06 inst/doc/README.Rmd
started knitr-based readme #17

RNeXML: The next-generation phylogenetics format comes to R
=====================================================

An R package for reading, writing, integrating and publishing data using the Ecological Metadata Language (EML) format.

* **Note:** This package is still in active development and not yet submitted to CRAN.  Functions and documentation may be incomplete and subject to change.  
* Maintainer: Carl Boettiger
* Authors: Carl Boettiger, Scott Chamberlain, Hilmar Lapp, Kseniia Shumelchyk, Rutger Vos
* License: BSD-3 
* [Issues](https://github.com/ropensci/RNeXML/issues): Bug reports, feature requests, and development discussion.

An extensive and rapidly growing collection of richly annotated phylogenetics data is now available in the NeXML format. NeXML relies on state-of-the-art data exchange technology to provide a format that can be both validated and extended, providing a data quality assurance and and adaptability to the future that is lacking in other formats [Vos et al 2012](http://doi.org/10.1093/sysbio/sys025 "NeXML: Rich, Extensible, and Verifiable Representation of Comparative Data and Metadata.").

```{r include=FALSE}
opts_chunk$set(tidy=FALSE)
```

Getting Started
---------------

The development version of RNeXML is [available on Github](https://github.com/ropensci/RNeXML).  With the `devtools` package installed on your system, RNeXML can be installed using:

```{r eval=FALSE}
library(devtools)
install_github("RNeXML", "ropensci")
library(RNeXML)
```

Read in a `nexml` file into the `ape::phylo` format:

```{r}
f <- system.file("examples", "trees.xml", package="RNeXML")
tr <- nexml_read(f, "phylo")
plot(tr[[1]])
```

Write an `ape::phylo` tree into the `nexml` format:

```{r}
data(bird.orders)
nexml_write(bird.orders, "test.xml")
```

<!---
NOT IMPLEMENTED YET

Extract metadata from the NeXML file:

```{r eval=FALSE} 
ex2 <- nexml_read("treebase1.xml", "nexml")
metadata(ex2)
```

Add metadata to a NeXML tree:

```{r eval=FALSE}
nexml_write(bird.orders, "birds.xml", 
            author="Carl Boettiger <cboettig@ropensci.org>", 
            title = "example NeXML file for bird orders", 
            description = "Example phylogeny taken from the ape documentation showing the major bird orders", 
            in_publication = "doi:10.1010/fakedoi.1234",
            add_itis = TRUE)
```

-->

### 2013-09-08 inst/doc/manuscript.Rmd
start outlining a manuscript

## Abstract

Why RNeXML?

* Why NeXML -- because Nexus standard isn't a standard.  Developers extend it arbitrarily to meet their needs, making it incompatible.  Cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, no validator.  NeXML solves all of these problems [Vos et al] with a state-of-the-art data management format.

* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read nexml.  All would benefit by being able to read NeXML, both as a robust standard free of the pitfalls of Nexus, and more importantly, as a much richer source of data thanks to its extendable annotation model.  (2) Write nexml.  R packages and R users add rich information to the phylogenies they work with, but need a platform independent data exchange format to share and archive this. (3) Annotate and extend nexml.

* as your own personal treebase

* Adding metadata to capture workflow

## Background

### Challenges with existing formats
See [Vos _et al._ 2012].

### Challenges with R formats

See [phylobase vignette, 2010]

### NeXML

See [Vos _et al._ 2012].

Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)

## Introduction to RNeXML

### Reading NeXML phylogenies

```{r}
phy = read.nexml("trees.xml", as="phylo")
plot(phy)
```

### Reading NeXML metadata

Replicate examples from [treebase] metadata section: publication information, tree size, taxanomic coverage

### Writing NeXML phylogenies

```{r}
write.nexml(phy, "example.xml")
```

### Writing NeXML metadata

Writing in authorship/publication information, copyright.
Writing in methodological information

Writing annotations on otus.  Annotations on nodes.

### Implementation and architecture

The `RNeXML` package is built for the NeXML Schema, [version 2009], using the R software environment and the XML package for parsing and generating XML [Temple Lang 2013, XML].

<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->

[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/

### Quality control

The `RNeXML` package can performs two levels of validation on NeXML files written from or read into R: validating the XML, and validating against the current NeXML schema.

Several mechanisms are in place to evaluate the quality of the `RNeXML` package functions themselves.  In addition to the examples from the documentation and the automated package checks provided for the R system, the `RNeXML` package includes a complete test suite using the testthat package [Wickham 2011].

[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html

### Software reuse and support

The `RNeXML` package is available through the CRAN repository under a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. The current development branch can be found on Github at [https://github.com/ropensci/RNeXML].  See the NEWS file for changes in recent versions. Users are encouraged to submit bug reports or feature requests in the issues log.  Contact ropensci-discuss@googlegroups.com or the package maintainer for help.

There are many promising directions for further development of the `RNeXML` package.  The project's Milestones page provides a list of steps currently planned or under consideration.  In particular, these milestones include increased support for interactive mode of entering data (wizards), and support for integration of semantics defined through the Web Ontology language (OWL), as discussed in the Challenges section.

### Acknowledgements

This project was supported in part by a grant from the Alfred P Sloan Foundation (CB and KR), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part of the rOpenSci project, [http://ropensci.org].

phylobase::phylo4 was an attempt to solve the problems associated with the popular ape::phylo format (unpredictable errors, loose and multiple conflicting specifications).

### 2013-09-20 inst/doc/manuscript.Rmd
writing

@@
 
+### Reading NeXML metadata
+
+One of the most powerful features of the NeXML format is the ability add arbitrary metadata information using the RDFa annotation standard [cite W3C]().
+
+Metadata from treebase NeXML
+- article citation information
+- Summary data: taxa, number of taxa, number of trees, number of tips.  
+- Metadata of nodes, complete RDF extraction
+
 ### Writing NeXML metadata
 
-Writing in authorship/publication information, copyright.
-Writing in methodological information
+`RNeXML` also helps users add richer metadata when producing NeXML files. This occurs on several levels:
+
+- Automatic annotation.
+`RNeXML` automatically adds basic metadata by default such as a timestamp of when the tree was produced and an open data license (CC0, compatible with deposition on Dryad data repository), which can be configured as needed.
+
+- Utilities to add common metadata
+
+  - Add a citation to any papers associated with the phylogeny data.  
+  - Writing in methodological information on how tree has been generated or analyzed (including R code)
+  - Add annotations to any taxon.
 

### 2013-10-15 inst/doc/README.Rmd
Add support for arbitrary rdfa/xml children of meta nodes
Closes #23
See the last test case in test_meta.R for an example.

@@
 
-<!---
-NOT IMPLEMENTED YET
-
 Extract metadata from the NeXML file:
 
-```{r eval=FALSE} 
+```{r } 
 ex2 <- nexml_read("treebase1.xml", "nexml")
@@
 ```{r eval=FALSE}
+  history <- new("meta", 
+      content = "Mapped from the bird.orders data in the ape package using RNeXML",
+      datatype = "xsd:string", id = "meta5144", property = "skos:historyNote", 
+      'xsi:type' = "LiteralMeta")
+  modified <- new("meta",
+                  content = "2013-10-04", datatype = "xsd:string", id = "meta5128",
+                  property = "prism:modificationDate", 'xsi:type' = "LiteralMeta")
+  website <- new("meta", 
+                 href = "http://carlboettiger.info", 
+                 rel = "foaf:homepage", 'xsi:type' = "ResourceMeta")
+  nexml_write(bird.orders, 
+              file = "example.xml", 
+              additional_metadata = list(history, modified, website), 
+              additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
+                                        prism = "http://prismstandard.org/namespaces/1.2/basic/",
+                                        foaf = "http://xmlns.com/foaf/0.1/"))
+
  nexml_write(bird.orders, "birds.xml", 
@@
 ```
-
--->

### 2013-10-15 R/classes.R

@@ -39,7 +39,9 @@ setMethod("fromNeXML",
 
 #########################
 
-setClass("Meta", contains = "Base")
+setClass("Meta",
+         representation(children = "list"), 
+         contains = "Base")
 setMethod("fromNeXML", 
           signature("Meta", "XMLInternalElementNode"),
           function(obj, from){
@@ -128,19 +130,24 @@ setAs("ResourceMeta", "XMLInternalNode", function(from) toNeXML(from, newXMLNode
 
 ##############################################
 
-setClass("meta", contains=c("LiteralMeta", "ResourceMeta"))
+setClass("meta", 
+         contains=c("LiteralMeta", "ResourceMeta"))
 setAs("XMLInternalElementNode", "meta", function(from){ 
       type <- xmlAttrs(from)["type"]
       if(is.na(type)) ## FIXME This is CRUDE
         type <- xmlAttrs(from)["xsi:type"]
-      type <- gsub("nex:", "", type) ## FIXME This is CRUDE
-      fromNeXML(new(type[1]), from)
+      if(is.na(type)) # if still not defined...
+        fromNeXML(new("meta", from))
+      else {
+        type <- gsub("nex:", "", type) ## FIXME This is CRUDE
+        fromNeXML(new(type[1]), from)
+      }
 }) 
 setAs("meta", "XMLInternalElementNode", function(from){ 
-      toNeXML(as(from, slot(from, "xsi:type")), newXMLNode("meta"))
+      toNeXML(as(from, slot(from, "xsi:type")), newXMLNode("meta", .children = from@children))
 })
 setAs("meta", "XMLInternalNode", function(from)  ## Really, do we need this?
-      toNeXML(as(from, slot(from, "xsi:type")), newXMLNode("meta")))
+      toNeXML(as(from, slot(from, "xsi:type")), newXMLNode("meta", .children = from@children)))
 # Methods inherited automatically?
 
 

### 2013-10-15 R/meta.R

@@ -29,7 +29,8 @@ meta <- function(property = character(0),
                  rel = character(0), 
                  href = character(0), 
                  id = character(0),
-                 type = character(0)){
+                 type = character(0),
+                 children = NULL){
   if(is.logical(content))
     datatype <- "xsd:boolean"
   else if(is(content, "Date"))

### 2013-10-15 inst/tests/test_meta.R

@@ -1,11 +1,13 @@
+context("meta")
+
+
+library(ape)
+library(RNeXML)
+library(XML)
+data(bird.orders)
 
 test_that("We can add license and pubdate by default in metadata", {
   ## The short version using an RNeXML API
-  library(ape)
-  library(RNeXML)
-  library(XML)
-
-  data(bird.orders)
   nexml_write(bird.orders, file="example.xml")
   nex <- nexml_read("example.xml", "nexml")
   
@@ -17,11 +19,7 @@ test_that("We can add license and pubdate by default in metadata", {
 
 test_that("We can add additional metadata", {
   ## The short version using an RNeXML API
-  library(ape)
-  library(RNeXML)
-  library(XML)
 
-  data(bird.orders)
   nexml_write(bird.orders, file="meta_example.xml",
               title = "My test title",
               description = "A description of my test",
@@ -43,9 +41,6 @@ test_that("We can add additional metadata", {
 
 test_that("We can add R bibentry type metadata", {
   ## The short version using an RNeXML API
-  library(ape)
-  library(RNeXML)
-  data(bird.orders)
 
   nexml_write(bird.orders, file="example.xml", citation=citation("ape")) 
   nex <- nexml_read("example.xml", "nexml")
@@ -64,9 +59,6 @@ test_that("We can add R bibentry type metadata", {
 
 test_that("We can add arbitrary metadata", {
   ## The short version using an RNeXML API
-  library(ape)
-  library(RNeXML)
-  data(bird.orders)
 
   history <- new("meta", 
       content = "Mapped from the bird.orders data in the ape package using RNeXML",
@@ -97,5 +89,54 @@ test_that("We can add arbitrary metadata", {
   unlink("example.xml") # cleanup
 
 })
+
+
+
+
+
+test_that("We can add arbitrary metadata", {
+  ## The short version using an RNeXML API
+
+  rdfa <- '<div typeof="foaf:Person" about="">
+             <a rel="foaf:account" href="https://twitter.com/cboettig">twitter</a> 
+             <a rel="foaf:account" href="https://github.com/cboettig">github</a>
+           </div>'
+  parsed <- xmlRoot(xmlParse(rdfa))
+  arbitrary_rdfa <- meta(property="eml:additionalMetadata", content="additional metadata", children = parsed)
+  nexml_write(bird.orders, 
+            file = "example.xml", 
[…]

### 2013-10-15 inst/doc/README.Rmd
update readme with metadata example

--- 2013-09-06 inst/doc/README.Rmd
@@
 
-```{r include=FALSE}
-opts_chunk$set(tidy=FALSE)
+```{r echo=FALSE}
+opts_knit$set(upload.fun = socialR::flickr.url)
+opts_chunk$set(tidy=FALSE, warning=FALSE, cache=FALSE)
 ```
@@
 install_github("RNeXML", "ropensci")
-library(RNeXML)
 ```
@@
 ```{r}
+library(RNeXML)
 f <- system.file("examples", "trees.xml", package="RNeXML")
@@
 
-<!---
-NOT IMPLEMENTED YET
-
+<!--
 Extract metadata from the NeXML file:
@@
 ```{r eval=FALSE} 
-ex2 <- nexml_read("treebase1.xml", "nexml")
+ex2 <- nexml_read(f, "nexml")
 metadata(ex2)
 ```
+-->
+
+--------------------------------------------
+
+### Advanced annotation
 
@@
 
+```{r }
+  history <- new("meta", 
+      content = "Mapped from the bird.orders data in the ape package using RNeXML",
+      datatype = "xsd:string", id = "meta5144", property = "skos:historyNote", 
+      'xsi:type' = "LiteralMeta")
+  modified <- new("meta",
+                  content = "2013-10-04", datatype = "xsd:string", id = "meta5128",
+                  property = "prism:modificationDate", 'xsi:type' = "LiteralMeta")
+  website <- new("meta", 
+                 href = "http://carlboettiger.info", 
+                 rel = "foaf:homepage", 'xsi:type' = "ResourceMeta")
+  nexml_write(bird.orders, 
+              file = "example.xml", 
+              additional_metadata = list(history, modified, website), 
+              additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
+                                        prism = "http://prismstandard.org/namespaces/1.2/basic/",
+                                        foaf = "http://xmlns.com/foaf/0.1/"))
+
+```
+
+<!--
 ```{r eval=FALSE}
@@
 ```
-
 -->

### 2013-11-12 inst/doc/manuscript.Rmd
nexml_read always reads in nexml format now

@@
 
+Using the nexml and nexmlTree class for metadata.
+
 ### Writing NeXML metadata
@@
 
+Temporal, spatial, taxonomic coverage for dryad/knb?
+
 ### Implementation and architecture
 
-The `RNeXML` package is built for the NeXML Schema, [version 2009], using the R software environment and the XML package for parsing and generating XML [Temple Lang 2013, XML].
+The `RNeXML` package is built for the NeXML Schema, [version 2009], using the R software environment and the XML package for parsing and generating XML [Temple Lang 2013, XML], and provides conversions into the popular `phylo` format provided by the `ape` package [Paradis 2004].
+`RNeXML` can automatically check taxonomic names and add taxonomic identifiers when generating NeXML using methods from the `taxize` package ([Chamberlain 2013]).
 
@@
 
-The `RNeXML` package can performs two levels of validation on NeXML files written from or read into R: validating the XML, and validating against the current NeXML schema.
+The `RNeXML` package performs two levels of validation on NeXML files written from or read into R: validating the XML, and validating against the current NeXML schema.
 

### 2013-11-12 R/metadata_methods.R

@@ -7,6 +7,18 @@ setGeneric("get_license", function(object) standardGeneric("get_license"))
 #' @export
 setGeneric("get_citation", function(object) standardGeneric("get_citation"))
 
+#' @export
+setGeneric("get_taxa", function(object) standardGeneric("get_taxa"))
+
+setMethod("get_taxa", 
+          signature("nexml"), 
+          function(object) 
+           sapply(object@otus@otu, function(otu) otu@label)
+          )
+
+
+
+
 
 setMethod("summary", 
           signature("nexml"), 
@@ -33,6 +45,40 @@ setxpath <- function(object){
             doc
 }
 
+
+#' get all top-level metadata  More extensible than hardwired functions
+setMethod("get_metadata", signature("nexml"), function(object, level){
+
+            level <- match.arg(level) 
+            string <- paste0("//nex:", level, "/nex:meta" )
+            b <- setxpath(as(object, "XMLInternalElementNode"))
+
+            references <- getNodeSet(b, 
+                                     paste0(string, "[@rel]"),
+                                     namespaces = nexml_namespaces)
+            rel = sapply(references, 
+                              function(x) 
+                                xmlAttrs(x)['rel'])
+            href = sapply(references, 
+                             function(x) 
+                               xmlAttrs(x)['href'])
+            names(href) = rel
+            literals <- getNodeSet(b, 
+                                   paste0(string, "[@property]"), 
+                                   namespaces = nexml_namespaces)
+            property = sapply(literals, 
+                              function(x) 
+                                xmlAttrs(x)['property'])
+            content = sapply(literals, 
+                             function(x) 
+                               xmlAttrs(x)['content'])
+            names(content) = property
+            c(content, href)
+          })
+
+
+
+
 ## Note that we define our namespace prefixes explicitly, so that should the NeXML use a different abberivation, this should still work.  
 setMethod("get_citation", 
           signature("nexml"), 
@@ -53,36 +99,6 @@ setMethod("get_license",
           else
             cc_license
           })
-
-#' get all top-level metadata
-setMethod("get_metadata", signature("nexml"), function(object, level){
-
-            level <- match.arg(level) 
-            string <- paste0("//nex:", level, "/nex:meta" )
-            b <- setxpath(as(object, "XMLInternalElementNode"))
-            references <- getNodeSet(b, paste0(string, "[@property]"), namespaces = nexml_namespaces)
-
-
-            rel = sapply(references, 
-                              function(x) 
-                                xmlAttrs(x)['rel'])
[…]

### 2013-11-12 R/nexml_read.R

@@ -1,7 +1,7 @@
 #' Read NeXML files into various R formats
 #' 
 #' @param x Path to the file to be read in 
-#' @param type the type of object to be returned.  If the file 
+#' @param as the as of object to be returned.  If the file 
 #' contains multiple trees, all will be read into the appropriate 
 #' multi-tree container, or else returned as a list of such objects.  
 #' @import XML
@@ -11,12 +11,10 @@
 #' @examples
 #' f <- system.file("examples", "trees.xml", package="RNeXML")
 #' nexml_read(f) 
-nexml_read <- function(x, type = c("nexmlTree", "phylo", "phylo4", "ouch",
-                                   "matrix", "nexml")){
-  type <- match.arg(type) 
+nexml_read <- function(x){
   doc <- xmlParse(x) 
   # will return class multiphylo if phylo is asked for and multiple trees are given.  
-  output <- as(xmlRoot(doc), type)
+  output <- as(xmlRoot(doc), "nexml")
   free(doc)
   output
 }

### 2013-11-12 inst/tests/test_toplevel_api.R

@@ -7,8 +7,8 @@ test_that("read.nexml works (to ape::phylo)", {
   library(RNeXML)
 
   f <- system.file("examples", "trees.xml", package="RNeXML")
-  phy <- nexml_read(f, type="phylo")
-#  phy <- read.nexml(f, type="phylo")
+  phy <- nexml_read(f, as="phylo")
+  phy <- read.nexml(f, as="phylo")
   layout(matrix(1:2, 1, 2), c(5, 4))
   plot(phy[[1]])
   plot(phy[[2]])

### 2013-11-13 inst/doc/README.Rmd
updated documentation

--- 2013-09-06 inst/doc/README.Rmd
@@
 ```{r}
+library(RNeXML)
 f <- system.file("examples", "trees.xml", package="RNeXML")
-tr <- nexml_read(f, "phylo")
-plot(tr[[1]])
+nexml <- nexml_read(f)
+tr <- get_tree(nexml) # or: as(nexml, "phylo")
+plot(tr)
 ```
@@
 
-<!---
-NOT IMPLEMENTED YET
-
 Extract metadata from the NeXML file:
 
-```{r eval=FALSE} 
-ex2 <- nexml_read("treebase1.xml", "nexml")
-metadata(ex2)
+```{r } 
+get_taxa(nexml)
+get_metadata(nexml) 
 ```
 
-Add metadata to a NeXML tree:
+--------------------------------------------
 
-```{r eval=FALSE}
-nexml_write(bird.orders, "birds.xml", 
-            author="Carl Boettiger <cboettig@ropensci.org>", 
-            title = "example NeXML file for bird orders", 
-            description = "Example phylogeny taken from the ape documentation showing the major bird orders", 
-            in_publication = "doi:10.1010/fakedoi.1234",
-            add_itis = TRUE)
+Add basic additional metadata:
+
+```{r}
+  nexml_write(bird.orders, file="meta_example.xml",
+              title = "My test title",
+              description = "A description of my test",
+              creator = "Carl Boettiger <cboettig@gmail.com>",
+              publisher = "unpublished data",
+              pubdate = "2012-04-01")
+
 ```
 
--->
+### Advanced annotation
+
+We can also add arbitrary metadata to a NeXML tree by define `meta` objects:
+
+```{r}
+modified <- new("meta",
+                 property = "prism:modificationDate",
+                 content = "2013-10-04")
+```
+
+Advanced use requires specifying the namespace used.
+Metadata follows the RDFa conventions.
+Here we indicate the modification date using the prism vocabulary.
+This namespace is included by default, as it is used for some of the basic metadata shown in the previous example.
+We can see from this list:
+
+```{r}
+RNeXML:::nexml_namespaces
+```
+
+This next block defines a resource (link), described by the `rel` attribute as a homepage, a term in the `foaf` vocabulalry.
+Becuase `foaf` is not a default namespace, we will have to provide its URL in the full definition below.
+
+```{r}
+website <- new("meta", 
+               href = "http://carlboettiger.info", 
+               rel = "foaf:homepage")
+```
+
+Here we create a history node using the `skos` namespace.  We can also add id values to any metadata element to make the element easier to reference externally:
+
+```{r }
+  history <- new("meta", 
+      property = "skos:historyNote", 
+      content = "Mapped from the bird.orders data in the ape package using RNeXML",
+      id = "meta123")
+```
+
+Once we have created the `meta` elements, we can pass them to our `nexml_write` function, along with definitions of the namespaces.
+
+```{r}
+  nexml_write(bird.orders, 
+              file = "example.xml", 
+              additional_metadata = list(history, modified, website), 
+              additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
+                                        prism = "http://prismstandard.org/namespaces/1.2/basic/",
+                                        foaf = "http://xmlns.com/foaf/0.1/"))
+
+```

### 2013-11-13 inst/doc/README.Rmd
updated README

@@
 opts_knit$set(upload.fun = socialR::flickr.url)
-opts_chunk$set(tidy=FALSE, warning=FALSE, cache=FALSE)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
 ```
@@
 tr <- get_tree(nexml) # or: as(nexml, "phylo")
-plot(tr)
+plot(tr[[1]])
 ```
@@
 ```{r } 
-get_taxa(nexml)
-get_metadata(nexml) 
+birds <- nexml_read("test.xml")
+get_taxa(birds)
+get_metadata(birds) 
 ```
@@
 ```
+By default, `RNeXML` adds certain metadata, including the NCBI taxon id numbers for all named taxa.
+This acts a check on the spelling and definitions of the taxa as well as providing a link to additional metadata about each taxonomic unit described in the dataset.
 
@@
 ```{r}
-modified <- new("meta",
-                 property = "prism:modificationDate",
+modified <- meta(property = "prism:modificationDate",
                  content = "2013-10-04")
@@
 ```{r}
-website <- new("meta", 
-               href = "http://carlboettiger.info", 
+website <- meta(href = "http://carlboettiger.info", 
                 rel = "foaf:homepage")
@@
 ```{r }
-  history <- new("meta", 
-      property = "skos:historyNote", 
+  history <- meta(property = "skos:historyNote", 
                   content = "Mapped from the bird.orders data in the ape package using RNeXML",
@@
               additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
-                                        prism = "http://prismstandard.org/namespaces/1.2/basic/",
                                         foaf = "http://xmlns.com/foaf/0.1/"))

### 2013-11-13 inst/tests/test_meta.R

@@ -74,7 +74,7 @@ test_that("We can add arbitrary metadata", {
               file = "example.xml", 
               additional_metadata = list(history, modified, website), 
               additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
-                                        prism = "http://prismstandard.org/namespaces/1.2/basic/",
+#                                        prism = "http://prismstandard.org/namespaces/1.2/basic/", # FIXME check and remove duplicates
                                         foaf = "http://xmlns.com/foaf/0.1/"))
 
   

### 2013-11-18 inst/doc/README.Rmd
update tests, add Makefile for README

@@
 nexml_write(bird.orders, "test.xml")
+```
+
+A key feature of NeXML is the ability to formally validate the construction of the data file against the standard (the lack of such a feature in nexus files had lead to inconsistencies across different software platforms, and some files that cannot be read at all).
+While it is difficult to make an invalid NeXML file from `RNeXML`, it never hurts to validate just to be sure:
+
+```{r}
+nexml_validate("test.xml")
 ```

### 2013-11-18 R/ape.R

@@ -8,7 +8,9 @@ setGeneric("getTaxonNames", function(otus, ids) standardGeneric("getTaxonNames")
 # depending on the approriate number of trees in the nexml.  
 
 
-setAs("nexml", "multiPhylo", function(from){
+
+#' 
+setAs("nexml", "multiPhyloList", function(from){
    unname(lapply(from@trees, 
            function(X){
              out <- unname(lapply(X@tree,  toPhylo, from@otus))
@@ -18,10 +20,22 @@ setAs("nexml", "multiPhylo", function(from){
 })
 
 
+#' Always collapses all trees nodes into a multiphylo
+setAs("nexml", "multiPhylo", function(from){
+   out <- unname(lapply(from@trees, 
+           function(X){
+             out <- unname(lapply(X@tree,  toPhylo, from@otus))
+             class(out) <- "multiPhylo"
+             out
+           }))
+  flatten_multiphylo(out)
+})
+
+
 #' Flatten a multiphylo object
 #' 
 #' NeXML has the concept of multiple <trees> nodes, each with multiple child <tree> nodes.
-#' This maps naturally to a list of multiphylo (heirachical multiphylo) objects.  Sometimes
+#' This maps naturally to a list of multiphylo  objects.  Sometimes
 #' this heirarchy conveys important structural information, so it is not discarded by default. 
 #' Occassionally it is useful to flatten the structure though, hence this function.  Note that this
 #' discards the original structure, and the nexml file must be parsed again to recover it.  
@@ -34,7 +48,6 @@ flatten_multiphylo <- function(object){
 }
 
 
-
 setAs("nexml", "phylo", function(from){ 
     if(length(from@trees[[1]]@tree) == 1){
       out <- toPhylo(from@trees[[1]]@tree[[1]], from@otus)
@@ -42,7 +55,9 @@ setAs("nexml", "phylo", function(from){
       warning("Multiple trees found, Returning multiPhylo object")
       out <- as(from, "multiPhylo") 
     }
-    out 
+    if(length(out) == 1)
+     out <- flatten_multiphylo(out)
+   out 
   })
 
 

### 2013-11-18 R/metadata_methods.R

@@ -64,8 +64,8 @@ get_item <- function(nexml,
 
   switch(element,
          tree = as(nexml, "phylo"), # will warn if more than one tree is available
-         trees = as(nexml, "multiPhylo"),
-         flat_trees = flatten_multiphylo(as(nexml, "multiPhylo")),
+         trees = as(nexml, "multiPhyloList"),
+         flat_trees = flatten_multiphylo(as(nexml, "multiPhyloList")),
          metadata = get_metadata(nexml, level),
          otu = get_taxa(nexml))
 }

### 2013-11-18 inst/tests/test_toplevel_api.R

@@ -7,8 +7,8 @@ test_that("read.nexml works (to ape::phylo)", {
   library(RNeXML)
 
   f <- system.file("examples", "trees.xml", package="RNeXML")
-  phy <- as(nexml_read(f),"phylo")
-  phy <- as(read.nexml(f),"phylo")
+  phy <- as(nexml_read(f), "phylo")
+  phy <- as(read.nexml(f), "phylo")
   layout(matrix(1:2, 1, 2), c(5, 4))
   plot(phy[[1]])
   plot(phy[[2]])

### 2013-11-27 inst/doc/README.Rmd

@@
 ```
+
+## Working with character data
+
+NeXML also provides a standard exchange format for handling character data.
+The R platform is particularly popular in the context of phylogenetic comparative methods, which consider both a given phylogeny and a set of traits.
+NeXML provides an ideal tool for handling this metadata.
+
+### Extracting character data
+
+We can load the library, parse the NeXML file and extract both the characters and the phylogeny.
+
+```{r}
+library(RNeXML)
+nexml <- read.nexml(system.file("examples", "comp_analysis.xml", package="RNeXML"))
+traits <- get_characters_list(nexml)
+tree <- get_tree(nexml)
+```
+
+(Note that `get_characters` would return both discrete and continuous characters together in the same data.frame, but we use `get_characters_list` to get separate data.frames for the continuous `characters` block and the discrete `characters` block).
+
+We can then fire up `geiger` and fit, say, a Brownian motion model the continuous data and a Markov transition matrix to the discrete states:
+
+```{r}
+library(geiger)
+cts_model <- fitContinuous(tree, traits[[1]])
+dct_model <- fitDiscrete(tree, traits[[2]])
+```

### 2013-11-27 inst/doc/README.Rmd
updated documentation

@@
 opts_knit$set(upload.fun = socialR::flickr.url)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=TRUE)
 ```
@@
 install_github("RNeXML", "ropensci")
+library(RNeXML)
+```
+
+```{r echo=FALSE}
+library(RNeXML)
 ```
@@
 ```{r}
-library(RNeXML)
-f <- system.file("examples", "trees.xml", package="RNeXML")
+f <- system.file("examples", "comp_analysis.xml", package="RNeXML")
 nexml <- nexml_read(f)
 tr <- get_tree(nexml) # or: as(nexml, "phylo")
-plot(tr[[1]])
+plot(tr)
 ```

### 2013-12-02 inst/doc/manuscript.Rmd
use slots instead of representation. closes #47

@@
 
+A nexml file can be thought of as having three components: (1) phylogeny data (2) character-trait data, and (3) metadata.
+`RNeXML` seeks to provide a seemless way to convert these to and from their corresponding R object representations, such as `ape::phylo` phylogenies, character matrices (`data.frames`) and metadata lists (or for advanced users, `XMLInternalDocument` and `rrdf` triplestores).
+The RNeXML package functions (the package Application Programming Interface, API)
+
 ## Introduction to RNeXML
 
-### Reading NeXML phylogenies
+<!-- Do we describe installation of the package?  `install.packages("RNeXML")`?-->
+
+### Reading NeXML
+
+```{r echo=FALSE}
+library(RNeXML)
+```
+
+We begin by reading in an example NeXML file.
 
 ```{r}
-phy = read.nexml("trees.xml", as="phylo")
-plot(phy)
+f <- system.file("examples", "trees.xml", package="RNeXML")
+nex <- read.nexml("trees.xml")
 ```
+
+The resulting `nex` object is an R
 

### 2013-12-02 R/character_classes.R

@@ -4,7 +4,7 @@
 
 
 setClass("char",
-         representation(states = "character"),
+         slots = c(states = "character"),
          contains = "IDTagged")
 setMethod("fromNeXML", 
           signature("char", "XMLInternalElementNode"),
@@ -34,9 +34,9 @@ setAs("XMLInternalElementNode", "char",
 
 ###############################################
 
-setClass("ListOfrow", representation(names="character"), contains="list")
+setClass("ListOfrow", slots = c(names="character"), contains="list")
 setClass("obsmatrix",
-         representation(row="ListOfrow"),
+         slots = c(row="ListOfrow"),
          contains = "Annotated")
 setMethod("fromNeXML", 
           signature("obsmatrix", "XMLInternalElementNode"),
@@ -69,11 +69,11 @@ setAs("XMLInternalElementNode", "obsmatrix",
 
 ######################################################
 
-setClass("ListOfcell", representation(names="character"), contains="list")
-setClass("ListOfseq", representation(names="character"), contains="list")
+setClass("ListOfcell", slots = c(names="character"), contains="list")
+setClass("ListOfseq", slots = c(names="character"), contains="list")
 
 setClass("row",
-         representation(cell = "ListOfcell",
+         slots = c(cell = "ListOfcell",
                         seq = "ListOfseq"),
          contains = "OptionalTaxonLinked")
 setMethod("fromNeXML", 
@@ -107,10 +107,10 @@ setAs("XMLInternalElementNode", "row",
       function(from) fromNeXML(new("row"), from))
 
 #######################################################
-setClass("ListOfstate", representation(names="character"), contains="list")
+setClass("ListOfstate", slots = c(names="character"), contains="list")
 
 setClass("states",
-         representation(state="ListOfstate"),
+         slots = c(state="ListOfstate"),
          contains = "IDTagged")
 setMethod("fromNeXML", 
           signature("states", "XMLInternalElementNode"),
@@ -142,7 +142,7 @@ setAs("XMLInternalElementNode", "states",
 
 ####################################################### 
 setClass("state",
-         representation(symbol = "character"), 
+         slots = c(symbol = "character"), 
          contains = "IDTagged")
 setMethod("fromNeXML", 
           signature("state", "XMLInternalElementNode"),
@@ -167,10 +167,10 @@ setAs("XMLInternalElementNode", "state",
 
 ################################################
 
-setClass("ListOfmember", representation(names="character"), contains="list")
[…]

### 2013-12-02 R/classes.R

@@ -15,7 +15,7 @@ setGeneric("toNeXML",
 ##############################
 
 setClass("Base",
-         representation('xsi:type' = "character"))
+         slots = c('xsi:type' = "character"))
 setMethod("toNeXML", 
           signature("Base", "XMLInternalElementNode"), 
           function(object, parent){
@@ -42,7 +42,7 @@ setMethod("fromNeXML",
 #########################
 
 setClass("Meta",
-         representation(children = "list"), 
+         slots = c(children = "list"), 
          contains = "Base")
 setMethod("fromNeXML", 
           signature("Meta", "XMLInternalElementNode"),
@@ -60,7 +60,7 @@ setMethod("toNeXML",
 #########################
 
 setClass("LiteralMeta", 
-         representation(id = "character",
+         slots = c(id = "character",
                         property = "character", 
                         datatype = "character",
                         content = "character"),
@@ -99,7 +99,7 @@ setAs("LiteralMeta", "XMLInternalNode", function(from) toNeXML(from, newXMLNode(
 ##############################################
 
 setClass("ResourceMeta", 
-         representation(id = "character",
+         slots = c(id = "character",
                         rel = "character", 
                         href = "character"),
          contains="Meta")
@@ -157,14 +157,14 @@ setAs("meta", "XMLInternalNode", function(from)  ## Really, do we need this?
 
 ###############################################
 
-setClass("ListOfmeta", representation(names="character"), contains = "list")
+setClass("ListOfmeta", slots = c(names="character"), contains = "list")
       
 
 ###############################################
 
 
 setClass("Annotated",
-         representation(meta = "ListOfmeta",
+         slots = c(meta = "ListOfmeta",
                         about = "character"),
          contains = "Base")
 setMethod("fromNeXML",
@@ -195,7 +195,7 @@ setMethod("toNeXML",
 ######################################################
 
 setClass("Labelled",
-         representation(label = "character"),
+         slots = c(label = "character"),
          contains = "Annotated")
 setMethod("fromNeXML", 
           signature("Labelled", "XMLInternalElementNode"),
@@ -218,7 +218,7 @@ setMethod("toNeXML",
 ##############################
 
 setClass("IDTagged",
-         representation(id = "character"),
+         slots = c(id = "character"),
          contains = "Labelled")
 setMethod("fromNeXML", 
           signature("IDTagged", "XMLInternalElementNode"),
@@ -243,7 +243,7 @@ setMethod("toNeXML",
 
 
[…]

### 2013-12-02 R/extend_phylo.R

@@ -1,5 +1,5 @@
 setClass("phyloS4", 
-         representation(edge = "matrix",
+         slots = c(edge = "matrix",
                         Nnode = "integer",
                         tip.label = "character",
                         edge.length = "numeric"))
@@ -8,7 +8,7 @@ setOldClass("phylo", S4Class="phyloS4")
 selectMethod("show", "phylo")
 removeClass("phyloS4")
 
-setClass("nexmlTree", representation(nexml = "nexml"), contains="phylo")
+setClass("nexmlTree", slots = c(nexml = "nexml"), contains="phylo")
 setMethod("show", "nexmlTree", function(object) print.phylo(object))
 # callNextMethod(object) ## callNextMethod might have been an option, but it looks for 'show' method, not print method?? 
 

### 2013-12-03 R/get_metadata.R
tweak documentation, metadata tests

@@ -4,7 +4,7 @@
 
 #' @import XML
 #' @export
-get_metadata <-  function(nexml, level){
+get_metadata <-  function(nexml, level=c("nexml", "otus", "trees", "characters")){
   level <- match.arg(level) 
   string <- paste0("//nex:", level, "/nex:meta" )
   b <- setxpath(as(nexml, "XMLInternalElementNode"))

### 2013-12-03 R/meta.R

@@ -18,12 +18,11 @@
 #' character - xs:string, 
 #' Date - xs:date,
 #' integer - xs:integer,
-#' numeric - xs:decimal
+#' numeric - xs:decimal,
 #' logical - xs:boolean
 #' 
 #' @export 
 #' @seealso \code{\link{nexml_write}}
-# FIXME generate id elements??
 meta <- function(property = character(0), 
                  content = character(0), 
                  rel = character(0), 
@@ -51,8 +50,8 @@ meta <- function(property = character(0),
   if(is(children, "XMLAbstractNode") || is(children, "XMLInternalNode"))
     children <- list(children)
 
-  if(length(property) > 0){
-    if(is.null(content))
+  if(length(property) > 0){ ## avoid 
+    if(is.null(content) && length(children) == 0) ## Avoid writing when content is missing, e.g. prism:endingpage is blank
       NULL
     else
       new("meta", content = content, datatype = datatype, 

### 2013-12-03 R/nexmlTree.R

@@ -23,11 +23,11 @@ nexmlTree <- function(object){
 ## Coercions between classes
 setAs("XMLInternalElementNode", "nexmlTree", function(from)
       nexmlTree(as(from, "nexml")))
-setAs("nemxmlTree", "XMLInternalElementNode", function(from)
+setAs("nexmlTree", "XMLInternalElementNode", function(from)
       as(from@nexml, "XMLInternalElementNode"))
 setAs("XMLInternalNode", "nexmlTree", function(from)
       nexmlTree(as(from, "nexml")))
-setAs("nemxmlTree", "XMLInternalNode", function(from)
+setAs("nexmlTree", "XMLInternalNode", function(from)
       as(from@nexml, "XMLInternalNode"))
 setAs("phylo", "nexmlTree", function(from)
       nexmlTree(as(from, "nexml")))

### 2013-12-03 R/nexml_methods.R

@@ -1,0 +1,38 @@
+
+#setMethod("head", signature("nexml"), function(x, n=6L, ...){
+#          write.nexml(x, "tmp123.xml")
+#          txt <- readLines("tmp123.xml", n=n)
+#          unlink("tmp123.xml")
+#          cat(txt, "\n")
+#})
+#setMethod("tail", signature("nexml"), function(x, n=6L, ...){
+#          write.nexml(x, "tmp123.xml")
+#          txt <- readLines("tmp123.xml", n=-n)
+#          unlink("tmp123.xml")
+#          cat(txt, "\n")
+#})
+setMethod("show", signature("nexml"), function(object){
+          summary(object)
+})
+
+
+
+setMethod("summary", signature("nexml"), function(object){
+          doc <- xmlParse(write.nexml(object))
+          nmeta <- length(getNodeSet(doc, "//x:meta", namespaces="x"))
+          ntrees <- length(getNodeSet(doc, "//x:tree", namespaces="x"))
+          ncharacters <- length(getNodeSet(doc, "//x:characters", namespaces="x"))
+          notu <- length(getNodeSet(doc, "//x:otu", namespaces="x"))
+
+          cat(paste("A nexml object representing:\n",
+              "\t", ntrees, "phylogenetic trees", "\n",
+              "\t", nmeta, "meta elements", "\n",
+              "\t", ncharacters, "character matrices", "\n",
+              "\t", notu, "taxonomic units", "\n",
+              "Taxa: \t", paste(head(get_taxa(object)), collapse = ", "), "...", "\n\n",
+              "NeXML generated by", object@generator, "using",
+              "schema version:", object@version, "\n", 
+              "size:", capture.output(print(object.size(object), units="auto")), "\n"))
+
+})
+

### 2013-12-03 inst/tests/test_meta.R

@@ -65,21 +65,30 @@ test_that("We can add additional metadata", {
 })
 
 
+test_that("We can add additional metadata at arbitrary level", {
+  nex <- add_trees(bird.orders)
+  modified <- meta(property = "prism:modificationDate",
+                  content = "2013-10-04")
 
+  nex@trees[[1]]@meta <- new("ListOfmeta", list(modified))
+
+  expect_identical(get_metadata(nex, "trees")[[modified@property]], modified@content)
+})
 
 
 test_that("We can add arbitrary metadata", {
-
-  rdfa <- '<meta typeof="foaf:Person" about="http://carlboettiger.info#me">
+  require(XML)
+  rdfa <- '<xhtml:div typeof="foaf:Person" about="http://carlboettiger.info#me">
              <a rel="foaf:account" href="https://twitter.com/cboettig">twitter</a> 
              <a rel="foaf:account" href="https://github.com/cboettig">github</a>
-           </meta>'
+           </xhtml:div>'
   parsed <- xmlRoot(xmlParse(rdfa))
   arbitrary_rdfa <- meta(property="eml:additionalMetadata", content="additional metadata", children = parsed)
 
   nex <- add_meta(arbitrary_rdfa, 
                   namespaces = c(foaf = "http://xmlns.com/foaf/0.1/", 
-                                 eml = "eml://ecoinformatics.org/eml-2.1.1"))
+                                 eml = "eml://ecoinformatics.org/eml-2.1.1", 
+                                 xhtml = "http://www.w3.org/1999/xhtml"))
 
   nexml_write(nex, file = "example.xml")
 

### 2013-12-03 inst/tests/test_meta_extract.R

@@ -9,17 +9,21 @@ nex <- add_basic_meta(
             pubdate = "2012-04-01",
             citation = citation("ape"))
 
-test_that("we can extract various metadata", {
-  write.nexml(nex, file = "example.xml")
+test_that("we can extract metadata using the dedicated functions", {
 
-  ## FIXME add the appropriate expect_that checks here
-  nex <- read.nexml("example.xml")
   get_citation(nex)
   get_license(nex)
   get_metadata(nex)
   summary(nex)
 
   unlink("example.xml")
+})
+
+
+
+test_that("we can extract all available metadata at a specified level of the DOM", {
+ get_metadata(nex) 
+ get_metadata(nex, "trees") 
 })
 
 

### 2013-12-03 inst/doc/manuscript.Rmd
tweak documentation, metadata tests
renamed tree access methods get_trees and get_trees_list, in closer analogy to get_characters and _get_characters_list.  The first provides a smart/integrated representation, whereas the latter provides a consistent but more cumbersome structure that corresponds 1:1 with the NeXML heirarchy

@@
 
-The resulting `nex` object is an R
+The resulting `nex` object is an R object representation of the NeXML.
+
+```{r}
+phy <- get_trees(nex)
+plot(phy)
+```
+
+Where we have used the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.
+
+NeXML can not only contain multiple phylogenetic trees in a single file, but can also group sets of trees into separate blocks.
+The `get_trees` function attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo` object; if multiple trees are available in a single `trees` block, it will return an `ape::multiPhylo` object (a list of `phylo` objects).
+If multiple blocks are available, it will return a list of `multiPhylo` blocks.
+While most users will find this behavior convenient, there are times (such as looping over many NeXML files) when having a variable return type is undisirable.
+To always get back a list of `multiPhylo` objects, even when only one tree is available, use `get_trees_list()`.
+
+##### Character data
 

### 2013-12-03 R/get_item.R

@@ -19,14 +19,14 @@
 #' @seealso \code{\link{get_tree}}
 #' @include classes.R
 get_item <- function(nexml, 
-                     element = c("tree", "trees", "flat_trees", "metadata", "otu", "characters", "characters_list"), 
+                     element = c("trees", "trees_list", "flat_trees", "metadata", "otu", "characters", "characters_list"), 
                      level = c("nexml", "otus", "otu", "trees", "tree")){
   element <- match.arg(element)
   level <- match.arg(level)
 
   switch(element,
-         tree = get_tree(nexml), # will warn if more than one tree is available
-         trees = get_trees(nexml),
+         trees = get_trees(nexml), # will warn if more than one tree is available
+         trees_list = get_trees_list(nexml),
          flat_trees = get_flat_trees(nexml),
          metadata = get_metadata(nexml, level),
          otu = get_taxa(nexml),

### 2013-12-03 R/get_trees.R

@@ -5,17 +5,17 @@
 #' @export
 #' @import plyr 
 #' @seealso \code{\link{get_tree}} \code{\link{get_flat_trees}} \code{\link{get_item}}
-get_trees <- function(nexml) as(nexml, "multiPhyloList")
+get_trees_list <- function(nexml) as(nexml, "multiPhyloList")
 
 
 
 #' extract a phylogenetic tree from the nexml
 #' 
 #' @param object a representation of the nexml object from  which the data is to be retrieved
-#' @return an ape::phylo tree, if only one tree is represented.  Otherwise returns a list of lists of multiphylo trees.  To consistently recieve the list of lists format (preserving the heriarchical nature of the nexml), use \code{\link{get_trees}} instead.  
+#' @return an ape::phylo tree, if only one tree is represented.  Otherwise returns a list of lists of multiphylo trees.  To consistently recieve the list of lists format (preserving the heriarchical nature of the nexml), use \code{\link{get_trees_list}} instead.  
 #' @export
 #' @seealso \code{\link{get_trees}} \code{\link{get_flat_trees}} \code{\link{get_item}}
-get_tree <- function(nexml) as(nexml, "phylo")
+get_trees <- function(nexml) as(nexml, "phylo")
 
 
 #' extract a single multiPhylo object containing all trees in the nexml

### 2013-12-03 inst/tests/test_rdf.R

@@ -21,7 +21,4 @@ test_that("we can perform sparql queries with rrdf", {
  
  ## Perform a SPARQL query:
  out <- sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-
-
-
 })

### 2013-12-12 inst/doc/manuscript.Rmd
a little writing on manuscript

@@
 Cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, no validator.
-NeXML solves all of these problems [Vos et al] with a state-of-the-art data management format.
+NeXML solves all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
 
@@
 
-### Reading NeXML
+## Working with phylogenetic trees
+
+#### Read methods
 
@@
 f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- read.nexml("trees.xml")
+nex <- nexml_read("trees.xml")
 ```
 
+(For convenience `nexml_read` is also aliased as `read.nexml`, for users familiar with the somewhat analogous `read.nexus` function).
 The resulting `nex` object is an R object representation of the NeXML.
+From this object we can extract any phylogenies it contains in the `ape::phylo` format:
 
@@
 phy <- get_trees(nex)
+```
+We can then leverage the rich suite of methods available from the R phylogenetics community (including coercion into other representations, such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny formats).
+For instance, here we use the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.
+
+[@ouch]: http://cran.r-project.org/web/packages/ouch/
+[@phylobase]: http://cran.r-project.org/web/packages/phylobase
+
+``{r}
 plot(phy)
 ```
-
-Where we have used the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.
 
@@
 To always get back a list of `multiPhylo` objects, even when only one tree is available, use `get_trees_list()`.
+
+<!-- have we already said all this above?-->
+The ability to read in phylogenetic trees in the `nexml` format opens up a wide and rapidly growing array of resources to the R user.
+For instance, a user can access all the phylogenies available in TreeBASE through the `nexml` format.
+`RNeXML` can read directly from a URL:
+
+```{r eval=FALSE}
+nex = nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+```
+
+Previously, this was possible using the `treebase` package [@treebase] for R, which relied on the nexus parser provided in the `ape` package.
+Due to inconsistencies present in the Nexus standard [Vos _et al._ 2012], a significant number <!-- State explicitly --> of TreeBASE nexus files could not be read in this way.
+Using the more precisely defined NEXML standard, every valid NeXML file on TreeBASE can now successfully be parsed.
+
+#### Write methods
 

### 2013-12-24 inst/doc/manuscript.Rmd
writing writing

@@
-## Abstract
+---
+title: RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R
+author: 
+  - name: Carl Boettiger
+    affiliation: cstar
+    email: cboettig@gmail.com
+    footnote: Corresponding author
+  - name: Hilmar Lapp
+    affiliation: NESCent
+  - name: Scott Chamberlain
+    affiliation: SimonFraser
+  - name: Rutger Vos
+    affiliation: NCB
+address: 
+  - code: cstar
+    address: | 
+      Center for Stock Assessment Research, 
+      Department of Applied Math and Statistics, 
+      University of California, Mail Stop SOE-2,
+      Santa Cruz, CA 95064, USA
+  - code: NESCent
+    address: | 
+      National Evolutionary Synthesis Center, Durham, NC, USA
+  - code: SimonFraser
+    address: |
+      Department of Biology, Simon Fraser University, Burnaby, Canada
+  - code: NCB
+    address: | 
+      NCB Naturalis, Leiden, the Netherlands
+abstract: | 
+      1. NeXML promises to be the next generation of phylogenetic
+      informatics data and metadata exchange format, offering substantial
+      improvements in reliability, extensibility, and richness over
+      earlier standards such as Nexus.
+
+      2. In this paper we present the RNeXML package which provides
+      a bridge between this rich and reliable data format and the
+      extensive library of phylogenetic tools already available in R,
+      with a particular emphasis on comparative methods.
+
+      3. RNeXML allows R users to both read and write NeXML files. Yet
+      RNeXML is much richer than a basic parser: here we illustrate
+      how the package facilitates adding and manipulating semantic 
+      annotations.
+
+      4. An essential feature of NeXML is to provide both the flexibility
+      to be continually extended to meet the emerging needs of the phylogenetics
+      community while remaining backwards-compatible with software developed
+      for earlier versions.  Here we illustrate how RNeXML makes it easy
+      to extend NeXML in this way.
+
+      5. RNeXML is perfectly suited for users to preserve and maintain their
+      personal archives of phylogenetic and comparative trait data in consistent,
+      query-able, and cross-platform format, or leverage the rapidly growing 
+      public community archives of rich NeXML files.
+
+---
+
+Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient format to exchange and archive comparative methods data.
 
@@
 ### Challenges with R formats
+
+- Most R users rely on the nexus inherits the limitations of nexus file format
 
@@
 ```
+
 We can then leverage the rich suite of methods available from the R phylogenetics community (including coercion into other representations, such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny formats).
@@
 
+#### Character data
+
+Like nexus files, NeXML files can also contain character data.
+While most researchers will be familiar with the use of character matrices in nexus files as a container for molecular sequence data used to infer the phylogenetic tree, we find this ideal for character data used in comparative methods popular in R.
+Methods to get extract character data work much like those for the phylogenetic data:
+
+```{r }
+nex <- nexml_read("comp_analysis.xml")
+get_characters(nex)
+```
+
+Returns a `data.frame` with columns as characters and rows as taxa.
+Continuous and discrete characters are represented as seperate blocks in NeXML files, but will be combined as seperate columns of a single data frame by `get_characters()` if they correspond to the same taxa.
+The `get_characters_list()` function will consistently recieve a list of data frames corresponding to each character matrix block of the NeXML.
+
 #### Write methods
 
-##### Character data
+Generating NeXML files from R is likewise straight forward.  Here we write a phylogeny from the `ape::phylo` format out to a NeXML file:
+
+```{r}
+data(bird.orders)
+nexml_write(bird.orders, file = "birds.xml")
+```
+
+The first argument to `nexml_write` can be a `phylo` phylogeny, `multiPhylo` list of phylogenies, or a `nexml` object -- anything that can be coerced into the `nexml` class.
+As we have seen, NeXML files can contain multiple phylogenetic trees, as well as character matrices.
+We can add both trees  and character data using `nexml_write`,
+
+```{r}
+data(geospiza)
+nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
+```
+
+NeXML files can also be built up bit by bit using the `add_characters`, `add_trees`, and `add_meta`.
+Each of these functions can add data to an existing `nexml` object or create a new nexml object if none is provided.
+Here we first create a `nexml` object containing the phylogeny data, and then add the character data to it:
+
+```{r}
+nex <- add_trees(geospiza$phy)
+nex <- add_characters(geospiza$dat, nex)
+```
+
+The data need not share the same taxa.  We can append additional phylogenies and characters corresponding to different taxa:
+
+```{r}
+data(primates)
+nex <- add_trees(primates$phy, nex)
+nex <- add_characters(primates$dat, nex)
+```
+
+<!-- explain OTU blocks somewhere -->
+
+### Metadata
 
@@
 ### Writing NeXML phylogenies
-
-```{r}
-write.nexml(phy, "example.xml")
-```
-
 ### Reading NeXML metadata

### 2013-12-30 R/add_trees.R
basic simmap functions in place (#48) could still be adjusted in several ways: tweaks to the XML definition to define ordering and direction of ordering.  Could be extended to include the node states defined in phytools.  The R mechanism needs to fill out the remaining (sometimes non-essential or redundant) parts of the phytools simmap slots.  The R mechanism still needs to write the char/states part of the nexml and substitute state names for state ids in the metadata

@@ -73,7 +73,7 @@ add_trees_block <- function(nexml, phy, otus_id){
 }
 
 
-
+# Main routine to generate NeXML from ape:phylo
 fromPhylo <- function(phy, otu_map){
   
   node_ids <- sapply(unique(as.numeric(phy$edge)), 

### 2013-12-30 R/classes.R

@@ -22,9 +22,10 @@ setMethod("toNeXML",
             type <- slot(object, "xsi:type")
             if(length(type) > 0){
 ## FIXME Don't assume type should be `nex:` namespaced
-## FIXME 
+              if(is.na(pmatch("nex:", type)))
+                type <- paste0("nex:", type)
               addAttributes(parent, 
-                           "xsi:type" = paste0("nex:", type),  
+                           "xsi:type" = type,  
                             suppressNamespaceWarning=TRUE) # We always define xsi namespace in the header... 
               }
             parent
@@ -149,12 +150,14 @@ setAs("XMLInternalElementNode", "meta", function(from){
       }
 })
 
-setAs("meta", "XMLInternalElementNode", function(from){ 
-      if(slot(from, "xsi:type") %in% c("LiteralMeta", 
-                                       "ResourceMeta",
-                                       "nex:LiteralMeta", 
-                                       "nex:ResourceMeta"))
-        m <- as(from, slot(from, "xsi:type"))
+setAs("meta", "XMLInternalElementNode", function(from){
+      if(length( slot(from, "xsi:type") ) > 0 ){
+        if(slot(from, "xsi:type") %in% c("LiteralMeta", 
+                                         "ResourceMeta",
+                                         "nex:LiteralMeta", 
+                                         "nex:ResourceMeta"))
+          m <- as(from, slot(from, "xsi:type"))
+        }
       else
         m <- from
       toNeXML(m, newXMLNode("meta", .children = from@children))

### 2013-12-30 R/get_trees.R

@@ -24,7 +24,7 @@ get_trees <- function(nexml) as(nexml, "phylo")
 #' @param object a representation of the nexml object from  which the data is to be retrieved
 #' @export
 #' @seealso \code{\link{get_tree}} \code{\link{get_trees}} \code{\link{get_item}} 
-get_flat_trees <- function(nexml) flatten_multiphylo(get_trees(nexml))
+get_flat_trees <- function(nexml) flatten_multiphylo(get_trees_list(nexml))
 
 
 

### 2013-12-30 R/nexml_read.R

@@ -10,7 +10,7 @@
 #' f <- system.file("examples", "trees.xml", package="RNeXML")
 #' nexml_read(f) 
 nexml_read <- function(x, ...){
-  if(grep("^https?://", x) == 1) # handle remote paths using httr::GET
+  if(length(grep("^https?://", x)) > 0) # handle remote paths using httr::GET
     x <- GET(x)
 
   doc <- xmlParse(x, ...) 

### 2013-12-30 inst/tests/test_geiger.R

@@ -54,6 +54,11 @@ test_that("We can write amphibia multiphylo to nexml. Two of these phylogenies e
   data(amphibia)
   class(amphibia) <- "multiPhylo"
   runtime <- system.time(nexml_write(amphibia, file="tmp.xml")) # Slow! about 12 minutes
+  library(XML)
+  results <- xmlSchemaValidate("http://www.nexml.org/2009/nexml.xsd", "tmp.xml")
+  expect_equal(results$status, 0)
+  expect_equal(length(results$errors), 0)
+
   expect_true(nexml_validate("tmp.xml"))
   unlink("tmp.xml") # cleanup
 })

### 2013-12-30 inst/tests/test_toplevel_api.R

@@ -63,22 +63,24 @@ test_that("We can get the right level of lists of trees ", {
 
   ## identical methods, Collapses length-1 lists 
   phy <- as(nex, "phylo")  ##
-  phy2 <- get_tree(nex)
-  phy3 <- get_item(nex, "tree")
-  expect_identical(phy, phy2)
-  expect_identical(phy3, phy2)
-
-  ## Doesn't collapse the length-1 lists, returns list of multiPhylo always: 
-  phy <- as(nex, "multiPhyloList")  ##
   phy2 <- get_trees(nex)
   phy3 <- get_item(nex, "trees")
   expect_identical(phy, phy2)
   expect_identical(phy3, phy2)
 
+  ## Doesn't collapse the length-1 lists, returns list of multiPhylo always: 
+  phy <- as(nex, "multiPhyloList")  ##
+  phy2 <- get_trees_list(nex)
+  phy3 <- get_item(nex, "trees_list")
+  expect_identical(phy, phy2)
+  expect_identical(phy3, phy2)
+
   ## Collapse to multiPhylo   
   phy <- as(nex, "multiPhylo")  ##
-  phy2 <- get_tree(nex) # same because there are two trees in the same `trees` node.  
+  phy2 <- get_trees(nex) # same because there are two trees in the same `trees` node.  
   expect_identical(phy, phy2)
+  phy3 <- get_item(nex, "flat_trees") ## FIXME SOMETHING WRONG!
+  expect_identical(phy3, phy2)
 })
 
 

### 2014-02-20 inst/doc/manuscript.Rmd

@@
 Cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, no validator.
-NeXML solves all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
+NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
 
@@
 
+### Taxonomic identifiers
+
+Add taxonomic identifier metadata to the OTU elements:
+
+```{r}
+nex <- add_trees(bird.orders)
+nex <- nexml_taxize(nex)
+```
+
 ### Metadata
@@
 
-The `RNeXML` package is built for the NeXML Schema, [version 2009], using the R software environment and the XML package for parsing and generating XML [Temple Lang 2013, XML], and provides conversions into the popular `phylo` format provided by the `ape` package [Paradis 2004].
-`RNeXML` can automatically check taxonomic names and add taxonomic identifiers when generating NeXML using methods from the `taxize` package ([Chamberlain 2013]).
+The `RNeXML` package is built for the NeXML Schema, [version 2009],
+using the R software environment and the XML package for parsing and
+generating XML [Temple Lang 2013, XML], and provides conversions into
+the popular `phylo` format provided by the `ape` package [Paradis 2004].
+`RNeXML` can automatically check taxonomic names and add taxonomic
+identifiers when generating NeXML using methods from the `taxize` package
+([Chamberlain 2013]).
 
@@
 
-The `RNeXML` package performs two levels of validation on NeXML files written from or read into R: validating the XML, and validating against the current NeXML schema.
+The `RNeXML` package performs two levels of validation on NeXML files
+written from or read into R: validating the XML, and validating against
+the current NeXML schema.
 
-Several mechanisms are in place to evaluate the quality of the `RNeXML` package functions themselves.
-In addition to the examples from the documentation and the automated package checks provided for the R system, the `RNeXML` package includes a complete test suite using the testthat package [Wickham 2011].
+Several mechanisms are in place to evaluate the quality of the `RNeXML`
+package functions themselves.  In addition to the examples from the
+documentation and the automated package checks provided for the R system,
+the `RNeXML` package includes a complete test suite using the testthat
+package [Wickham 2011].
 
@@
 
-The `RNeXML` package is available through the CRAN repository under a CC0 license.
-At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`.
-The current development branch can be found on Github at [https://github.com/ropensci/RNeXML].
-See the NEWS file for changes in recent versions.
-Users are encouraged to submit bug reports or feature requests in the issues log.
-Contact ropensci-discuss@googlegroups.com or the package maintainer for help.
+The `RNeXML` package is available through the CRAN repository under
+a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
+The current development
+branch can be found on Github at [https://github.com/ropensci/RNeXML].
+See the NEWS file for changes in recent versions. Users are encouraged
+to submit bug reports or feature requests in the issues log.  Contact
+ropensci-discuss@googlegroups.com or the package maintainer for help.
 
-There are many promising directions for further development of the `RNeXML` package.
-The project's Milestones page provides a list of steps currently planned or under consideration.
-In particular, these milestones include increased support for interactive mode of entering data (wizards), and support for integration of semantics defined through the Web Ontology language (OWL), as discussed in the Challenges section.
+There are many promising directions for further development of the
+`RNeXML` package.  The project's Milestones page provides a list of
+steps currently planned or under consideration.  In particular, these
+milestones include increased support for interactive mode of entering data
+(wizards), and support for integration of semantics defined through the
+Web Ontology language (OWL), as discussed in the Challenges section.
 
@@
 
-This project was supported in part by a grant from the Alfred P Sloan Foundation (CB and KR), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part of the rOpenSci project, [http://ropensci.org].
+This project was supported in part by a grant from the Alfred P Sloan
+Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part
+of the rOpenSci project, [http://ropensci.org].
 
-phylobase::phylo4 was an attempt to solve the problems associated with the popular ape::phylo format (unpredictable errors, loose and multiple conflicting specifications).
+phylobase::phylo4 was an attempt to solve the problems associated with
+the popular ape::phylo format (unpredictable errors, loose and multiple
+conflicting specifications).

### 2014-02-20 inst/doc/pubs/README.Rmd
working on check

--- 2013-11-27 inst/doc/README.Rmd
@@
+<!--
+%\VignetteEngine{knitr::knitr}
+%\VignetteIndexEntry{An Introduction to the RNeXML package}
+-->
+
+[![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
+
 RNeXML: The next-generation phylogenetics format comes to R
@@
 ```{r echo=FALSE}
+library(knitr)
 opts_knit$set(upload.fun = socialR::flickr.url)
@@
 nexml <- nexml_read(f)
-tr <- get_tree(nexml) # or: as(nexml, "phylo")
-plot(tr)
+tr <- get_trees(nexml) # or: as(nexml, "phylo")
+plot(tr[[1]])
 ```
@@
               file = "example.xml", 
-              additional_metadata = list(history, modified, website), 
-              additional_namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
+              meta = list(history, modified, website), 
+              namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
                              foaf = "http://xmlns.com/foaf/0.1/"))
 
+```
+
+### Taxonomic identifiers
+
+Add taxonomic identifier metadata to the OTU elements:
+
+```{r}
+nex <- add_trees(bird.orders)
+nex <- taxize_nexml(nex)
 ```
@@
 nexml <- read.nexml(system.file("examples", "comp_analysis.xml", package="RNeXML"))
-traits <- get_characters_list(nexml)
-tree <- get_tree(nexml)
+traits <- get_characters(nexml)
+tree <- get_trees(nexml)
 ```
@@
 library(geiger)
-cts_model <- fitContinuous(tree, traits[[1]])
-dct_model <- fitDiscrete(tree, traits[[2]])
+fitContinuous(tree, traits[[1]])
+fitDiscrete(tree, traits[[2]])
 ```

### 2014-02-20 R/add_basic_meta.R

@@ -20,6 +20,7 @@
 #'    can be included here.  
 #'    citation can be a plain text object, but is preferably an R `citation` or `bibentry` object (which
 #'    can include multiple citations.  See examples
+#' @param ... additional arguments, currently no effect. 
 #' @details \code{add_basic_meta()} is just a wrapper for \link{\code{add_meta}} to make it easy to 
 #'    provide generic metadata without explicitly providing the namespace.  For instance, 
 #'    \code{add_basic_meta(title="My title", description="a description")} is identical to:
@@ -53,7 +54,8 @@ add_basic_meta <- function(nexml = new("nexml"),
                            pubdate = Sys.Date(),
                            rights = "CC0",
                            publisher = NULL,
-                           citation = NULL){
+                           citation = NULL,
+                           ...){
 
   if(!is.null(title)) 
     nexml <- add_meta(meta("dc:title", title), nexml)

### 2014-02-20 R/add_meta.R

@@ -5,7 +5,7 @@
 #' @param level the level at which the metadata annotation should be added.
 #' @param namespaces named character string for any additional namespaces that should be defined.  
 #' @param i for otus, trees, characters: if there are multiple such blocks, which one should be annotated?  Default is first/only block.  
-#' @param at_id the id of the element to be annotated.  Optional, advanced use only.  
+#' @param at_id the id of the element to be annotated.  Optional, advanced use only. 
 #' @seealso \code{\link{meta}} \code{\link{add_trees}} \code{\link{add_characters}} \code{\link{add_basic_meta}}
 #' @return the updated nexml object
 #' @examples

### 2014-02-20 R/add_trees.R

@@ -30,8 +30,16 @@ standardize_phylo_list <- function(phy){
 }
 
 
-
-
+#' add_trees
+#' 
+#' add_trees
+#' @param phy a phylo object, multiPhylo object, or list of 
+#'  mulitPhylo to be added to the nexml
+#' @param nexml a nexml object to which we should append this phylo.
+#'  By default, a new nexml object will be created.  
+#' @param append_to_existing_otus logical, indicating if we should 
+#'  make a new OTU block (default) or append to the existing one. 
+#' @return a nexml object containing the phy in nexml format. 
 #' @export 
 add_trees <- function(phy, 
                       nexml=new("nexml"), 

### 2014-02-20 R/get_characters.R

@@ -98,6 +98,8 @@ get_characters <- function(input, suffixes=FALSE){
 
 ### The subroutines of "get_characters_list# function
 ## Fixme these could be adapted to use the get_*_maps functions
+
+
 otu_to_label <- function(dat, otu_map){
   rownames(dat) <- otu_map[rownames(dat)]
   dat

### 2014-02-20 R/get_item.R

@@ -18,7 +18,7 @@
 #' and so forth for each element type.  
 #' @return return type depends on the element requested.  See details.  
 #' @export
-#' @seealso \code{\link{get_tree}}
+#' @seealso \code{\link{get_trees}}
 #' @include classes.R
 get_item <- function(nexml, 
                      element = c("trees", "trees_list", "flat_trees", "metadata", "otu", "characters", "characters_list"), 

### 2014-02-20 R/get_trees.R

@@ -1,11 +1,11 @@
 #' extract all phylogenetic trees in ape format
 #' 
 #' extract all phylogenetic trees in ape format
-#' @param object a representation of the nexml object from  which the data is to be retrieved
+#' @param nexml a representation of the nexml object from  which the data is to be retrieved
 #' @return returns a list of lists of multiphylo trees, even if all trees are in the same `trees` node (and hence the outer list will be of length 1) or if there is only a single tree (and hence the inner list will also be of length 1.  This guarentees a consistent return type regardless of the number of trees present in the nexml file, and also preserves any heirarchy/grouping of trees.  
 #' @export
 #' @import plyr 
-#' @seealso \code{\link{get_tree}} \code{\link{get_flat_trees}} \code{\link{get_item}}
+#' @seealso \code{\link{get_trees}} \code{\link{get_flat_trees}} \code{\link{get_item}}
 get_trees_list <- function(nexml) as(nexml, "multiPhyloList")
 
 
@@ -13,7 +13,7 @@ get_trees_list <- function(nexml) as(nexml, "multiPhyloList")
 #' extract a phylogenetic tree from the nexml
 #' 
 #' extract a phylogenetic tree from the nexml
-#' @param object a representation of the nexml object from  which the data is to be retrieved
+#' @param nexml a representation of the nexml object from  which the data is to be retrieved
 #' @return an ape::phylo tree, if only one tree is represented.  Otherwise returns a list of lists of multiphylo trees.  To consistently recieve the list of lists format (preserving the heriarchical nature of the nexml), use \code{\link{get_trees_list}} instead.  
 #' @export
 #' @seealso \code{\link{get_trees}} \code{\link{get_flat_trees}} \code{\link{get_item}}
@@ -25,9 +25,9 @@ get_trees <- function(nexml) as(nexml, "phylo")
 #' extract a single multiPhylo object containing all trees in the nexml
 #' @details Note that this method collapses any heirachical structure that may have been present as multiple `trees` nodes in the original nexml (though such a feature is rarely used).  To preserve that structure, use \code{\link{get_trees}} instead.  
 #' @return a multiPhylo object (list of ape::phylo objects).  See details.  
-#' @param object a representation of the nexml object from  which the data is to be retrieved
[…]

### 2014-02-20 R/internal_get_node_maps.R

@@ -1,8 +1,8 @@
-#' get otus map
-#'
-#' @param nexml nexml object 
-#' @return a list showing the mapping between (internal) otu identifiers and labels (taxonomic names). List is named by the id of the otus block. 
-#' @details largely for internal use   
+# get otus map
+#
+# @param nexml nexml object 
+# @return a list showing the mapping between (internal) otu identifiers and labels (taxonomic names). List is named by the id of the otus block. 
+# @details largely for internal use   
 get_otu_maps <- function(nexml){
   otus <- as.list(nexml@otus)
   names(otus) <- name_by_id(otus)

### 2014-02-20 R/internal_nexml_id.R

@@ -34,6 +34,9 @@ nexml_id <- function(prefix = "",
   uid
 }
 
+#' reset id counter
+#' 
+#' reset the id counter
 #' @export
 reset_id_counter <- function(){
   rm(list=ls(envir=nexml_env), envir=nexml_env)

### 2014-02-20 R/nexml_write.R

@@ -21,7 +21,8 @@
 #'  write.nexml(bird.orders, file="example.xml")
 #' 
 #'  ## Assemble a nexml section by section and then write to file:
-#'  library(geiger); data(geospiza)
+#'  library(geiger)
+#'  data(geospiza)
 #'  nexml <- add_trees(geospiza$phy) # creates new nexml
 #'  nexml <- add_characters(geospiza$dat, nexml) # pass the nexml obj to append character data
 #'  nexml <- add_basic_meta(nexml, title="my title", creator = "Carl Boettiger")
@@ -29,21 +30,26 @@
 #'  write.nexml(nexml, file="example.xml")
 #'
 #'  ## As above, but in one call (except for add_meta() call).  
-#'  write.nexml(trees = geospiza$phy, characters=geospiza$dat, 
-#'              title = "My title", creator="Carl Boettiger",
+#'  write.nexml(trees = geospiza$phy, 
+#'              characters = geospiza$dat, 
+#'              title = "My title", 
+#'              creator = "Carl Boettiger",
 #'              file = "example.xml")
 #'  
 #'  ## Mix and match: identical to the section by section: 
 #'  nexml <- add_meta(meta("prism:modificationDate", format(Sys.Date())))
 #'  write.nexml(x = nexml,
-#'              trees = geospiza$phy, characters=geospiza$dat, 
-#'              title = "My title", creator="Carl Boettiger",
+#'              trees = geospiza$phy, 
+#'              characters = geospiza$dat, 
+#'              title = "My title", 
+#'              creator = "Carl Boettiger",
 #'              file = "example.xml")
 #' 
 nexml_write <- function(x = new("nexml"),
                         file = NULL,
                         trees = NULL,
                         characters = NULL,
+                        meta = NULL, 
                         ...){
   
   nexml <- as(x, "nexml")
@@ -51,6 +57,8 @@ nexml_write <- function(x = new("nexml"),
     nexml <- add_trees(trees, nexml)
   if(!is.null(characters))
     nexml <- add_characters(characters, nexml)
+  if(!is.null(meta))
+    nexml <- add_meta(meta, nexml, ...)
   nexml <- do.call(add_basic_meta, c(list(nexml=nexml), list(...)))
   
   out <- as(nexml, "XMLInternalNode")

### 2014-02-20 R/simmap.R

@@ -5,6 +5,17 @@
 
 ## FIXME support writing multiphylos, list of multiphylos to nexml 
 
+
+#' simmap_to_nexml
+#' 
+#' simmap_to_nexml
+#' @param phy a phy object containing simmap phy$maps element,
+#'  from the phytools pacakge
+#' @param state_ids a named character vector giving the state 
+#'  names corresponding to the ids used to refer to each state
+#'  in nexml.  If null ids will be generated and states taken from 
+#'  the phy$states names. 
+#' @return a nexml representation of the simmap
 #' @export 
 #' @import XML
 simmap_to_nexml <- function(phy, state_ids = NULL){
@@ -84,6 +95,13 @@ simmap_edge_annotations <- function(maps, nexml, state_ids = NULL, char_id = "si
 
 
 ## Returns list of multiPhylo ...
+
+#' nexml_to_simmap
+#'
+#' nexml_to_simmap
+#' @param nexml a nexml object
+#' @return a simmap object (phylo object with a $maps element 
+#'  for use in phytools functions).
 #' @export 
 nexml_to_simmap <- function(nexml){
 

### 2014-02-20 R/taxize_nexml.R

@@ -1,3 +1,12 @@
+
+#' taxize nexml
+#'
+#' Check taxanomic names against the specified service and 
+#' add appropriate semantic metadata to the nexml OTU unit 
+#' containing the corresponding identifier. 
+#' @param nexml a nexml object
+#' @param type the name of the identifier to use
+#' @param ... additional arguments (not implemented yet)
 #' @import taxize
 #' @export 
 taxize_nexml <- function(nexml, type = c("NCBI"), ...){

### 2014-03-06 inst/doc/pubs/README.Rmd
update documentation

@@
 tr <- get_trees(nexml) # or: as(nexml, "phylo")
-plot(tr[[1]])
+plot(tr)
 ```
@@
 library(geiger)
-fitContinuous(tree, traits[[1]])
-fitDiscrete(tree, traits[[2]])
+fitContinuous(tree, traits[1])
+fitDiscrete(tree, traits[2])
 ```

### 2014-03-06 R/add_basic_meta.R

@@ -21,7 +21,7 @@
 #'    citation can be a plain text object, but is preferably an R `citation` or `bibentry` object (which
 #'    can include multiple citations.  See examples
 #' @param ... additional arguments, currently no effect. 
-#' @details \code{add_basic_meta()} is just a wrapper for \link{\code{add_meta}} to make it easy to 
+#' @details \code{add_basic_meta()} is just a wrapper for \code{\link{add_meta}} to make it easy to 
 #'    provide generic metadata without explicitly providing the namespace.  For instance, 
 #'    \code{add_basic_meta(title="My title", description="a description")} is identical to:
 #'    \code{add_meta(list(meta("dc:title", "My title"), meta("dc:description", "a description")))}

### 2014-03-06 R/add_characters.R

@@ -8,6 +8,12 @@
 #'        format. 
 #' @param nexml a nexml object, if appending character table to an existing
 #'        nexml object.  If ommitted will initiate a new nexml object.  
+#' @param append_to_existing_otus logical. If TRUE, will add any new taxa
+#' (taxa not matching any existing otus block) to the existing (first)
+#' otus block.  Otherwise (default), a new otus block is created, even
+#' though it may contain duplicate taxa to those already present.  While
+#' FALSE is the safe option, TRUE may be appropriate when building nexml
+#' files from scratch with both characters and trees.  
 #' @include classes.R
 #' @export 
 add_characters <- function(x, 

### 2014-03-06 R/concatenate_nexml.R

@@ -1,8 +1,13 @@
 
 #' Concatenate nexml files 
-#' @param a nexml object, e.g. from write.nexml() or read.nexml()
-#' @param ... additional nexml objects to be concatenated.  
-#'   must have unique ids on all elements
+#' 
+#' Concatenate nexml files 
+#' @param x,... nexml objects to be concatenated, e.g. from 
+#'  \code{\link{write.nexml}} or \code{\link{read.nexml}}. 
+#'  Must have unique ids on all elements
+#' @param recursive  logical.  If 'recursive = TRUE', the function recursively
+#'        descends through lists (and pairlists) combining all their
+#'        elements into a vector. (Not implemented).  
 #' @return a concatenated nexml file
 #' @examples
 #' f1 <- system.file("examples", "trees.xml", package="RNeXML")
@@ -12,7 +17,7 @@
 #' nex <- c(nex1, nex2)
 setMethod("c", 
           signature("nexml"), 
-          function(x, ...){
+          function(x, ..., recursive = FALSE){
               elements = list(x, ...)
               nexml <- new("nexml")
   ## Check that ids are unique

### 2014-03-06 R/get_characters.R

@@ -3,9 +3,8 @@
 
 #' Extract the character matrix
 #'
-#' @param nexml a nexml object
-#' @return the list of taxa
 #' @param nexml nexml object (e.g. from read.nexml)
+#' @return the list of taxa
 #' @export
 get_characters_list <- function(nexml){
 # extract mapping between otus and taxon ids 

### 2014-03-06 R/get_metadata.R

@@ -5,7 +5,7 @@
 #' 
 #' get_metadata 
 #' @param nexml a nexml object
-#' @param the name of the level of element desired
+#' @param level the name of the level of element desired
 #' @return the requested metadata
 #' @import XML
 #' @export

### 2014-03-06 R/get_trees.R

@@ -97,7 +97,7 @@ setAs("nexml", "phylo", function(from){
 #' nexml to phylo 
 #' 
 #' nexml to phylo coercion 
-#' @param tree an nexml tree element (nexml@trees[[i]]@tree[[j]]
+#' @param tree an nexml tree element 
 #' @param otus a character string of taxonomic labels, named by the otu ids.  
 #' e.g. (from get_otu_maps for the otus set matching the relevant trees node. 
 #' @return phylo object 
@@ -132,7 +132,8 @@ toPhylo <- function(tree, otus){
 
 ## Identifies tip.label based on being named with OTUs while others are NULL
 ## FIXME Should instead decide that these are tips based on the edge labels?
-  nodes <- cbind(plyr::arrange(nodes, otu), id = 1:dim(nodes)[1])
+  nodes <- cbind(plyr::arrange(nodes, otu), id = 1:dim(nodes)[1])  # Also warns because arrange isn't quoting the column name.  
+
 ## NB: these ids are the ape:id numbers by which nodes are identified in ape::phylo
 ## Arbitrary ids are not supported - ape expecs the numbers 1:n, starting with tips. 
 

### 2014-03-06 R/meta.R

@@ -21,6 +21,8 @@
 #' numeric - xs:decimal,
 #' logical - xs:boolean
 #' 
+#' @examples
+#' meta(content="example", property="dc:title")
 #' @export 
 #' @seealso \code{\link{nexml_write}}
 meta <- function(property = character(0), 
@@ -125,10 +127,21 @@ nexml_citation <- function(obj){
 
 
 
-
+#' Concatenate meta elements into a ListOfmeta
+#' 
+#' Concatenate meta elements into a ListOfmeta
+#' @param x,... meta elements to be concatenated, e.g. see \code{\link{meta}}
+#' @param recursive  logical, if 'recursive=TRUE', the function 
+#' descends through lists and combines their elements into a vector.
+#' Currently not implemented. so "c(meta, c(meta))" will not work.  
+#' @return a listOfmeta object containing multiple meta elements. 
+#' @examples 
+#' c(meta(content="example", property="dc:title"),
+#'   meta(content="Carl", property="dc:creator"))
+#' 
 setMethod("c", 
           signature("meta"),
-          function(x, ...){
+          function(x, ..., recursive = FALSE){
             elements <- list(x, ...)
             new("ListOfmeta", elements)
           })

### 2014-03-06 R/nexml_write.R

@@ -3,10 +3,12 @@
 #' @param x a nexml object, or any phylogeny object (e.g. phylo, phylo4) 
 #' that can be coerced into one. Can also be omitted, in which case a new 
 #' nexml object will be constructed with the additional parameters specified.
+#' @param file the name of the file to write out
 #' @param trees phylogenetic trees to add to the nexml file (if not already given in x)
 #' see \code{\link{add_trees}} for details.  
 #' @param characters additional characters
-#' @param file the name of the file to write out
+#' @param meta A meta element or list of meta elements, see \code{\link{add_meta}}
+#' @param ... additional arguments to add_meta, such as the namespaces.  See \code{\link{add_meta}}.   
 #' @return Writes out a nexml file
 #' @import ape
 #' @import XML 

### 2014-03-06 R/simmap.R

@@ -62,7 +62,7 @@ simmap_edge_annotations <- function(maps, nexml, state_ids = NULL, char_id = "si
     mapping <- lapply(1:length(edge_map), function(j){
       ##  A node has an id, a length and a state 
       meta(property = "simmap:stateChange", 
-           children = list(meta(propery = "simmap:order",
+           children = list(meta(property = "simmap:order",
                                 content = j),
                            meta(property = "simmap:length", 
                                 content = edge_map[[j]]),

### 2014-03-06 R/taxize_nexml.R

@@ -12,15 +12,17 @@
 taxize_nexml <- function(nexml, type = c("NCBI"), ...){
           type <- match.arg(type)
           if(type == "NCBI"){
-            for(i in 1:length(nexml@otus@otu)){
-              id <- get_uid(nexml@otus@otu[[i]]@label)
-              if(is.na(id))
-                warning(paste("ID for otu", nexml@otus@otu[[i]]@label, "not found. Consider checking the spelling or alternate classification"))
-              else 
-                nexml@otus@otu[[i]]@meta <- new("ListOfmeta", list(
-                               meta(href = paste0("http://ncbi.nlm.nih.gov/taxonomy/", id),
-                                    rel = "tc:toTaxon")))
+            for(j in 1:length(nexml@otus)){
+              for(i in 1:length(nexml@otus[[j]]@otu)){
+                id <- get_uid(nexml@otus[[j]]@otu[[i]]@label)
+                if(is.na(id))
+                  warning(paste("ID for otu", nexml@otus[[j]]@otu[[i]]@label, "not found. Consider checking the spelling or alternate classification"))
+                else 
+                  nexml@otus[[j]]@otu[[i]]@meta <- new("ListOfmeta", list(
+                                 meta(href = paste0("http://ncbi.nlm.nih.gov/taxonomy/", id),
+                                      rel = "tc:toTaxon")))
 
+              }
             }
           }
   nexml

### 2014-03-06 inst/doc/pubs/README.Rmd
update README

@@
 opts_knit$set(upload.fun = socialR::flickr.url)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=TRUE)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
 ```

### 2014-03-12 inst/doc/pubs/manuscript.Rmd
updated manuscript, add test for new bug

--- 2014-02-20 inst/doc/manuscript.Rmd
@@
 
-Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient format to exchange and archive comparative methods data.
+Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient, extensible format to exchange and archive comparative methods data.
+
+## Background
+
+### Challenges with existing formats
+See [Vos _et al._ 2012].
+
+### Challenges with R formats
+
+- Most R users rely on the nexus inherits the limitations of nexus file format
+
+See [phylobase vignette, 2010]
+
+### NeXML
+
+See [Vos _et al._ 2012].
+
+Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)
 
@@
 
-## Background
-
-### Challenges with existing formats
-See [Vos _et al._ 2012].
-
-### Challenges with R formats
-
-- Most R users rely on the nexus inherits the limitations of nexus file format
-
-See [phylobase vignette, 2010]
-
-### NeXML
-
-See [Vos _et al._ 2012].
-
-Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)
+## Introduction to RNeXML
 
@@
 `RNeXML` seeks to provide a seemless way to convert these to and from their corresponding R object representations, such as `ape::phylo` phylogenies, character matrices (`data.frames`) and metadata lists (or for advanced users, `XMLInternalDocument` and `rrdf` triplestores).
-The RNeXML package functions (the package Application Programming Interface, API)
+Metadata can provide information about the study in which the phylogeny was generated or applied (such as authors, citations to publications, timestamps, and even descriptions of methods), but can also be used to annotate individual data components.
+We illustrate how metadata annotations can be added to the operational taxonomic units of the phylogeny and can be used to extend the data representation to new uses such as stochastic character maps.
 
-## Introduction to RNeXML
+The current stable version of the `RNeXML` package can be installed from the CRAN library <!-- Actually not yet -->
+using the standard
 
-<!-- Do we describe installation of the package?  `install.packages("RNeXML")`?-->
+```r
+install.packages("RNeXML")
+```
+
+The latest stable development version can be installed from Github using the `devtools` package:
+
+```r
+library(devtools)
+install_github("RNeXML", "ropensci")
+```
 
@@
 
-We begin by reading in an example NeXML file.
+We begin by reading in an example NeXML file provided with the package.
 
@@
 
-(For convenience `nexml_read` is also aliased as `read.nexml`, for users familiar with the somewhat analogous `read.nexus` function).
+(For convenience `nexml_read` is also aliased as `read.nexml`, for
+users familiar with the somewhat analogous `read.nexus` function).
 The resulting `nex` object is an R object representation of the NeXML.
-From this object we can extract any phylogenies it contains in the `ape::phylo` format:
+From this object we can extract any phylogenies it contains in the
+`ape::phylo` format:
 
@@
 
-We can then leverage the rich suite of methods available from the R phylogenetics community (including coercion into other representations, such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny formats).
-For instance, here we use the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.
-
-[@ouch]: http://cran.r-project.org/web/packages/ouch/
-[@phylobase]: http://cran.r-project.org/web/packages/phylobase
+We can then leverage the rich suite of methods available from the R
+phylogenetics community (including coercion into other representations,
+such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny
+formats). For instance, here we use the standard plotting function from
+the `ape` package to plot the resulting `ape::phylo` object returned by
+`get_trees`.
 
@@
 
-NeXML can not only contain multiple phylogenetic trees in a single file, but can also group sets of trees into separate blocks.
-The `get_trees` function attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo` object; if multiple trees are available in a single `trees` block, it will return an `ape::multiPhylo` object (a list of `phylo` objects).
-If multiple blocks are available, it will return a list of `multiPhylo` blocks.
-While most users will find this behavior convenient, there are times (such as looping over many NeXML files) when having a variable return type is undisirable.
-To always get back a list of `multiPhylo` objects, even when only one tree is available, use `get_trees_list()`.
+NeXML can not only contain multiple phylogenetic trees in a single file,
+but can also group sets of trees into separate blocks.  The `get_trees`
+function attempts to return the most sensible R object for NeXML: if
+only a single tree is available, it will return an `ape::phylo` object;
+if multiple trees are available in a single `trees` block, it will return
+an `ape::multiPhylo` object (a list of `phylo` objects).  If multiple
+blocks are available, it will return a list of `multiPhylo` blocks.
+While most users will find this behavior convenient, there are times
+(such as looping over many NeXML files) when having a variable return
+type is undisirable.  To always get back a list of `multiPhylo` objects,
+even when only one tree is available, use `get_trees_list()`.
 
-<!-- have we already said all this above?-->
-The ability to read in phylogenetic trees in the `nexml` format opens up a wide and rapidly growing array of resources to the R user.
-For instance, a user can access all the phylogenies available in TreeBASE through the `nexml` format.
+<!-- have we already said all this above?--> The ability to read in
+phylogenetic trees in the `nexml` format opens up a wide and rapidly
+growing array of resources to the R user.  For instance, a user can access
+all the phylogenies available in TreeBASE through the `nexml` format.
 `RNeXML` can read directly from a URL:
@@
 
-Previously, this was possible using the `treebase` package [@treebase] for R, which relied on the nexus parser provided in the `ape` package.
-Due to inconsistencies present in the Nexus standard [Vos _et al._ 2012], a significant number <!-- State explicitly --> of TreeBASE nexus files could not be read in this way.
-Using the more precisely defined NEXML standard, every valid NeXML file on TreeBASE can now successfully be parsed.
+Previously, this was possible using the `treebase` package [@treebase]
+for R, which relied on the nexus parser provided in the `ape` package.
+Due to inconsistencies present in the Nexus standard [Vos _et al._
+2012], a significant number <!-- State explicitly --> of TreeBASE nexus
+files could not be read in this way.  Using the more precisely defined
+NEXML standard, every valid NeXML file on TreeBASE can now successfully
+be parsed.
 
@@
 
-Like nexus files, NeXML files can also contain character data.
-While most researchers will be familiar with the use of character matrices in nexus files as a container for molecular sequence data used to infer the phylogenetic tree, we find this ideal for character data used in comparative methods popular in R.
-Methods to get extract character data work much like those for the phylogenetic data:
+Like nexus files, NeXML files can also contain character data.  While
+most researchers will be familiar with the use of character matrices in
+nexus files as a container for molecular sequence data used to infer
+the phylogenetic tree, we find this ideal for character data used in
+comparative methods popular in R. Methods to get extract character data
+work much like those for the phylogenetic data:
 
@@
 
-Returns a `data.frame` with columns as characters and rows as taxa.
-Continuous and discrete characters are represented as seperate blocks in NeXML files, but will be combined as seperate columns of a single data frame by `get_characters()` if they correspond to the same taxa.
-The `get_characters_list()` function will consistently recieve a list of data frames corresponding to each character matrix block of the NeXML.
+Returns a `data.frame` with columns as characters and rows as
+taxa. Continuous and discrete characters are represented as seperate
+blocks in NeXML files, but will be combined as seperate columns of a
+single data frame by `get_characters()` if they correspond to the same
+taxa. The `get_characters_list()` function will consistently recieve
+a list of data frames corresponding to each character matrix block of
+the NeXML.
 
@@
 
-Generating NeXML files from R is likewise straight forward.  Here we write a phylogeny from the `ape::phylo` format out to a NeXML file:
+Generating NeXML files from R is likewise straight forward.  Here we
+write a phylogeny from the `ape::phylo` format out to a NeXML file:
 
@@
 
-The first argument to `nexml_write` can be a `phylo` phylogeny, `multiPhylo` list of phylogenies, or a `nexml` object -- anything that can be coerced into the `nexml` class.
-As we have seen, NeXML files can contain multiple phylogenetic trees, as well as character matrices.
+The first argument to `nexml_write` can be a `phylo` phylogeny,
+`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
+can be coerced into the `nexml` class.  As we have seen, NeXML files
+can contain multiple phylogenetic trees, as well as character matrices.
 We can add both trees  and character data using `nexml_write`,
@@
 
-NeXML files can also be built up bit by bit using the `add_characters`, `add_trees`, and `add_meta`.
-Each of these functions can add data to an existing `nexml` object or create a new nexml object if none is provided.
-Here we first create a `nexml` object containing the phylogeny data, and then add the character data to it:
+NeXML files can also be built up bit by bit using the `add_characters`,
+`add_trees`, and `add_meta`.  Each of these functions can add data to an
+existing `nexml` object or create a new nexml object if none is provided.
+Here we first create a `nexml` object containing the phylogeny data,
+and then add the character data to it:
 
@@
 
-The data need not share the same taxa.  We can append additional phylogenies and characters corresponding to different taxa:
+The data need not share the same taxa.  We can append additional
+phylogenies and characters corresponding to different taxa:
 
@@
 ```
+
+This allows us to include multiple trees and character matrices to a single NeXML file.
 
@@
 
-[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
-
 ### Quality control
@@
 package [Wickham 2011].
-
-[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
 
@@
 
+[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
+[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
+[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
+[@ouch]: http://cran.r-project.org/web/packages/ouch/
+[@phylobase]: http://cran.r-project.org/web/packages/phylobase
+
 phylobase::phylo4 was an attempt to solve the problems associated with

### 2014-03-12 tests/testthat/test_characters.R

@@ -124,3 +124,13 @@ test_that("we can add characters to a nexml file using a data.frame", {
 
 
 
+
+## based on bug on 2014-03-12 65ae459523c529452adb699c3d5d118c0a207402 
+test_that("we can add multiple character matrices to a nexml file", {
+          library("geiger")
+          data(geospiza)
+          data(primates)
+          nex <- add_characters(geospiza$dat)
+          nex <- add_characters(primates$dat, nex)
+})
+

### 2014-03-12 inst/doc/pubs/manuscript.Rmd
writing writing

@@
 
+### Metadata
+
+At the heart of the NeXML standard is the ability to encode arbitrary metadata in precisely specified, machine-readable formats that maintain backwards compabitibilty with any existing NeXML parser.
+NeXML uses the "Resource Description Framework in Annotations", or RDFa [W3C 2013], as described in [Vos et al 2011].
+
+### Reading NeXML metadata
+
+Metadata from treebase NeXML
+- article citation information
+- Summary data: taxa, number of taxa, number of trees, number of tips.  
+- Metadata of nodes, complete RDF extraction
+
+```{r } 
+birds <- nexml_read("test.xml")
+get_taxa(birds)
+get_metadata(birds) 
+```
+
 This allows us to include multiple trees and character matrices to a single NeXML file.
@@
 
-### Metadata
-
-### Reading NeXML metadata
-
 Replicate examples from [treebase] metadata section: publication information, tree size, taxanomic coverage
-
-### Writing NeXML phylogenies
-### Reading NeXML metadata
-
-One of the most powerful features of the NeXML format is the ability add arbitrary metadata information using the RDFa annotation standard [cite W3C]().
-
-Metadata from treebase NeXML
-- article citation information
-- Summary data: taxa, number of taxa, number of trees, number of tips.  
-- Metadata of nodes, complete RDF extraction
-
-Using the nexml and nexmlTree class for metadata.
 
@@
 
-`RNeXML` also helps users add richer metadata when producing NeXML files. This occurs on several levels:
+Researchers cannot take advantage of NeXML metadata manipulation if there is no metadata there.
+The burden of adding good metadata remains a significant barrier to increased sharing and reuse of scientific research [Tenopir et al.
+2011].
+`RNeXML` seeks to reduce the burden of adding high quality, machine-readable metadata to NeXML files.
+This occurs on several levels:
 
-- Automatic annotation.
-`RNeXML` automatically adds basic metadata by default such as a timestamp of when the tree was produced and an open data license (CC0, compatible with deposition on Dryad data repository), which can be configured as needed.
+- Automatic annotation.  `RNeXML` automatically adds basic metadata by
+default such as a timestamp of when the tree was produced and an open
+data license (CC0, compatible with deposition on Dryad data repository),
+which can be configured as needed.
 
@@
 Temporal, spatial, taxonomic coverage for dryad/knb?
+
+- Extending the NeXML standard through metadata annotation.
+
+No data standard can anticipate all the uses and extensions researchers may one day need.
+Researchers have repeatedly adapted the existing NEXUS file format to accomidate encoding new kinds of information, but in the process break compatibility between different flavors of NEXUS files, with few reliable methods to distinguish between these flavors.
+NeXML provides a solution to this problem through the use of metadata annotation.
+
+Here we illustrate this process using the example of stochastic character mapping [Huelsenbeck et al 2003].
+A stochastic character map is simply an annotation of the branches on a phylogeny, assigning each section of each branch to a particular "state" (typically of a morphological characteristic).
+
+[Bollback 2006] provides a widely used stand-alone software implementation of this method in the software `simmap`, which modified the standard Newick tree format to express this additional information.
+This can break compatibility with other software[^1], and creates a format that cannot be interpreted without additional information describing this convention.
+By contrast, the NeXML extension is not only backwards compatible but contains a precise and machine-readable description of what it is encoding.
+
+[^1]: By using the commenting mechanism of the Newick format, it is possible that other software that doesn't also use the comment mechanism for some other such purpose would be able to successfully parse the tree.
+However there is no way to guarentee that this is the case or for the data format to describe its use.
+
+[Revell 2011] describes the `phytools` package for R, which includes utilities for reading, manipulating, and writing `simmap` files.
+
+NeXML annotation of an edge that switches from state `s2` to state `s1`:
+
+```xml
+<edge about="#e34" id="e34" source="n16" target="n35" length="0.205223">
+        <meta xsi:type="nex:LiteralMeta" id="m262" property="simmap:reconstructions" datatype="xsd:string">
+          <meta xsi:type="nex:LiteralMeta" id="m260" property="simmap:reconstruction" datatype="xsd:string">
+            <meta xsi:type="nex:LiteralMeta" id="m261" property="simmap:char" datatype="xsd:string" content="cr1"/>
+            <meta xsi:type="nex:LiteralMeta" id="m252" property="simmap:stateChange" datatype="xsd:string">
+              <meta xsi:type="nex:LiteralMeta" id="m253" property="simmap:order" datatype="xsd:decimal" content="1"/>
+              <meta xsi:type="nex:LiteralMeta" id="m254" property="simmap:length" datatype="xsd:decimal" content="0.202947550453246"/>
+              <meta xsi:type="nex:LiteralMeta" id="m255" property="simmap:state" datatype="xsd:string" content="s2"/>
+            </meta>
+            <meta xsi:type="nex:LiteralMeta" id="m256" property="simmap:stateChange" datatype="xsd:string">
+              <meta xsi:type="nex:LiteralMeta" id="m257" property="simmap:order" datatype="xsd:decimal" content="2"/>
+              <meta xsi:type="nex:LiteralMeta" id="m258" property="simmap:length" datatype="xsd:decimal" content="0.00227544954675424"/>
+              <meta xsi:type="nex:LiteralMeta" id="m259" property="simmap:state" datatype="xsd:string" content="s1"/>
+            </meta>
+          </meta>
+        </meta>
+      </edge>
+```
+
+[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
+[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
+[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
 
@@
 
-The `RNeXML` package performs two levels of validation on NeXML files
+The `RNeXML` package performs severak levels of validation on NeXML files
 written from or read into R: validating the XML, and validating against
-the current NeXML schema.
+the current NeXML schema, and performing additional checks through the 
+NeXML online validation tool for elements that cannot be enforced by
+schema validation alone.  Validation serves to improve consistency in
+data encoding which facilitates the use of NeXML across a wide variety
+of platforms and software.  A lack of a consistent standard and validation
+mechanism has continually plagued the current NEXUS standard and made 
+interoperability difficult [Vos et al. 2011].
 
@@
 documentation and the automated package checks provided for the R system,
-the `RNeXML` package includes a complete test suite using the testthat
-package [Wickham 2011].
+the `RNeXML` package includes a complete test suite using the `testthat`
+package [Wickham 2011]. Package development on Github includes continuous 
+integration through the `travis` platform, in which the full test suite
+is run every time the code is changed.  This not only gives users confidence
+that the development copy can be successfully installed, but should assist
+other developers seeking to contribute to the package to test that their proposed 
+changes have not negatively impacted any of the existing code base.
 
@@
 [@phylobase]: http://cran.r-project.org/web/packages/phylobase
+[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
+[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
 

### 2014-03-13 inst/doc/pubs/manuscript.Rmd
update documentation

@@
 f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read("trees.xml")
+nex <- nexml_read(f)
 ```
@@
 
-At the heart of the NeXML standard is the ability to encode arbitrary metadata in precisely specified, machine-readable formats that maintain backwards compabitibilty with any existing NeXML parser.
-NeXML uses the "Resource Description Framework in Annotations", or RDFa [W3C 2013], as described in [Vos et al 2011].
+At the heart of the NeXML standard is the ability to encode arbitrary
+metadata in precisely specified, machine-readable formats that maintain
+backwards compabitibilty with any existing NeXML parser. NeXML uses the
+"Resource Description Framework in Annotations", or RDFa [W3C 2013],
+as described in [Vos et al 2011].  These annotations can be used to add 
+information to any particular element of the data, (a tree, node, edge,
+or taxonomic unit; a character matrix or specific trait, or even another 
+metadata annotation or the NeXML data file as a whole). `RNeXML` provides
+a range of tools for accessing metadata embedded in a NeXML file.
 
@@
 
-Metadata from treebase NeXML
-- article citation information
-- Summary data: taxa, number of taxa, number of trees, number of tips.  
-- Metadata of nodes, complete RDF extraction
+To illustrate metadata access, we begin by reading in a nexml file.
 
 ```{r } 
-birds <- nexml_read("test.xml")
-get_taxa(birds)
-get_metadata(birds) 
+S100 <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
 ```
+
+A call to the nexml object prints some metadata summarizing the data structure:
+
+```{r }
+S100
+```
+
+We can extract all metadata pertaining to the NeXML document as a whole
+(annotations of the XML root node, `<nexml>`) with the command
+
+```{r}
+meta <- get_metadata(S100) 
+```
+
+This returns a named list of available metadata. We can see the kinds
+of metadata recorded from the names:
+
+```{r}
+names(meta)
+```
+
+and can ask for a particular element using the standard list subsetting
+mechanism (i.e. either the name of an element or it's numeric position),
+
+```{r}
+meta[["dc:title"]]
+meta[["dc:publisher"]]
+```
+
+All metadata terms must belong to an explicit *namespace* or vocabulary that allows a computer to interpret the term precisely.
+The prefix (before the `:`) indicates to which vocabulary the term belongs, e.g.
+`dc` in this case.
+The `get_namespaces` function tells us the definition of the vocabulary using a link:
+
+```{r}
+prefixes <- get_namespaces(S100)
+prefixes["dc"]
+```
+
+Following the link we can see that `dc` refers to the "Dublin Core"
+vocabulary, and the term `title` refers to the title of a creative work
+(as opposed to the title of a person, a legal title, the verb title, etc).
+Such explicit definitions allow a computer to reason about these objects,
+as we shall see later.
+
+Other metadata is available
+
+```{r}
+get_taxa(S100)
+```
+
+- article citation information
+
+- Summary data: taxa, number of taxa, number of trees, number of tips.
+
+- Metadata of nodes, complete RDF extraction
 
@@
 
-Researchers cannot take advantage of NeXML metadata manipulation if there is no metadata there.
-The burden of adding good metadata remains a significant barrier to increased sharing and reuse of scientific research [Tenopir et al.
-2011].
-`RNeXML` seeks to reduce the burden of adding high quality, machine-readable metadata to NeXML files.
-This occurs on several levels:
+Researchers cannot take advantage of NeXML metadata manipulation if
+there is no metadata there.  The burden of adding good metadata remains a
+significant barrier to increased sharing and reuse of scientific research
+[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
+high quality, machine-readable metadata to NeXML files. This occurs on
+several levels:
 
-- Automatic annotation.  `RNeXML` automatically adds basic metadata by
-default such as a timestamp of when the tree was produced and an open
-data license (CC0, compatible with deposition on Dryad data repository),
-which can be configured as needed.
+- Automatic annotation.
+
+`RNeXML` automatically adds basic metadata by default such as a timestamp
+of when the tree was produced and an open data license (CC0, compatible
+with deposition on Dryad data repository), which can be configured
+as needed.
 
@@
 
-- Extending the NeXML standard through metadata annotation.
+### Extending the NeXML standard through metadata annotation.
 
@@
 [Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
+
+### Subsetting nexml objects directly
+
+The `RNeXML` package provides many convenient functions to add and extract
+information from `nexml` objects in the R environment without requiring
+the reader to understand the details of the NeXML data structure and
+making it less likely that a user will generate invalid NeXML syntax
+that could not be read by other parsers. The `nexml` object we have been using
+in all of the examples is built on R's S4 mechanism. Advanced users may
+sometimes prefer to interact with the data structure more directly using 
+R's S4 class mechanism and subsetting methods. Many R users are more familiar
+with the S3 class mechanism (such as in the `ape` package phylo objects)
+rather than the S4 class mechanism used in phylogenetics packages such as 
+`ouch` and `phylobase`. The `phylobase` vignette provides an excellent introduction
+to these data structures.  Users already familiar with subsetting lists and other
+S3 objects in R are likely familar with the use of the `$` operator, such as
+`phy$edge`. S4 objects simply use an `@` operator instead (but cannot be subset
+using numeric arguments such as `phy[[1]]` or named arguments such as phy[["edge"]]).
+
+The `nexml` object is an S4 object, as are all of its components (slots).  It's 
+heirachical structure corresponds exactly with the XML tree of a NeXML file, with 
+the single exception that both XML attributes and children are represented as slots.  
+S4 objects have constructor functions to initialize them.  We create a new `nexml` 
+object with the command:
+
+```{r}
+nex <- new("nexml")
+```
+
+We can see a list of slots contained in this object with
+
+```{r}
+slotNames(nex)
+```
+
+Some of these slots have already been populated for us, for instance, the schema version and default namespaces:
+
+```{r}
+nex@version
+nex@namespaces
+```
+
+Recognize that `nex@namespaces` serves the same role as `get_namespaces` function, but provides direct access to the slot data.
+For instance, with this syntax we could also overwrite the existing namespaces with `nex@namespaces <- NULL`.
+Changing the namespace in this way is not advised.
+
+Some slots can contain multiple elements of the same type, such as `trees`, `characters`, and `otus`.  For instance, we see that
+
+```{r}
+class(nex@characters)
+```
+
+is an object of class `ListOfcharacters`, and is currently empty,
+
+```{r}
+length(nex@characters)
+```
+
+In order to assign an object to a slot, it must match the class definition
+of the slot.  We can create a new element of any given class with the
+`new` function,
+
+```{r}
+nex@characters <- new("ListOfcharacters", list(new("characters")))
+```
+
+and now we have a length-1 list of character matrices,
+
+```{r}
+length(nex@characters)
+```
+
+and we access the first character matrix using the list notation, `[[1]]`. Here we check the class is a `characters` object.
+
+```{r}
+class(nex@characters[[1]])
+```
+
+Direct subsetting has two primary use cases: (a) useful in looking up (and possibly editing) a specific value of an element, or (b) when adding metadata annotations to specific elements.
+Consider the example file
+
+```{r}
+f <- system.file("examples", "trees.xml", package="RNeXML")
+nex <- nexml_read(f)
+```
+
+We can look up the species label of the first `otu` in the first `otus` block:
+
+```{r}
+nex@otus[[1]]@otu[[1]]@label
+```
+
+We can add metadata to this particular OTU using this subsetting format
+
+```{r}
+nex@otus[[1]]@otu[[1]]@meta <- 
+  c(meta("skos:note", 
+          "This species was incorrectly identified"),
+         nex@otus[[1]]@otu[[1]]@meta)
+```
+
+Here we use the `c` operator to append this element to any existing meta annotations to this otu.
 

### 2014-03-13 R/nexml_get.R

@@ -1,9 +1,10 @@
 #' Get the desired element from the nexml object
 #' 
 #' Get the desired element from the nexml object
+#' @aliases nexml_get get_item
 #' @param nexml a nexml object (from read_nexml)
 #' @param element the kind of object desired, see details.  
-#' @param level metadata argument only.  Define whose metadata we want. See examples for details.  
+#' @param ... additional arguments, if applicable to certain elements 
 #' @details
 #'  
 #' \itemize{
@@ -12,6 +13,7 @@
 #'  \item{"flat_trees"}{ a multiPhylo object (list of ape::phylo objects) Note that this method collapses any heirachical structure that may have been present as multiple `trees` nodes in the original nexml (though such a feature is rarely used).  To preserve that structure, use `trees` instead.}
 #'  \item{"metadata"}{Get metadata from the specified level (default is top/nexml level) }
 #'  \item{"otu"}{ returns a named character vector containing all available metadata.  names indicate \code{property} (or \code{rel} in the case of links/resourceMeta), while values indicate the \code{content} (or \code{href} for links). }
+#'  \item{"taxa"}{ alias for otu }
 #' }
 #' For a slightly cleaner interface, each of these elements is also defined as an S4 method
 #' for a nexml object.  So in place of `get_item(nexml, "tree")`, one could use `get_tree(nexml)`,
@@ -20,21 +22,31 @@
 #' @export
 #' @seealso \code{\link{get_trees}}
 #' @include classes.R
-get_item <- function(nexml, 
-                     element = c("trees", "trees_list", "flat_trees", "metadata", "otu", "characters", "characters_list"), 
-                     level = c("nexml", "otus", "otu", "trees", "tree")){
+nexml_get <- function(nexml, 
+                     element = c("trees", 
+                                 "trees_list", 
+                                 "flat_trees", 
+                                 "metadata", 
+                                 "otu",
+                                 "taxa",
+                                 "characters", 
+                                 "characters_list",
+                                 "namespaces"), 
+                     ...){
   element <- match.arg(element)
-  level <- match.arg(level)
 
   switch(element,
          trees = get_trees(nexml), # will warn if more than one tree is available
          trees_list = get_trees_list(nexml),
          flat_trees = get_flat_trees(nexml),
-         metadata = get_metadata(nexml, level),
+         metadata = get_metadata(nexml, ...),
          otu = get_taxa(nexml),
+         taxa = get_taxa(nexml),
[…]

### 2014-03-13 inst/doc/pubs/manuscript.Rmd
writing writing

@@
 
-Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient, extensible format to exchange and archive comparative methods data.
+<!-- Emphasize: 
+1) storing character data and tree data together, convenient for the comparative methods community
+2) Adding metadata, including automated addition of metadata
+3)
+
+-->
+
+Though R provides the most extensive and widely used collection of
+comparative phylogenetic methods tools, R users lack a convenient,
+extensible format to exchange and archive comparative methods data.
 
@@
 
-Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)
+Mention of data sharing challenges and opportunities.  See [Drew et al 2013]
 
@@
 
-* Why NeXML -- because Nexus standard isn't a standard.
-Developers extend it arbitrarily to meet their needs, making it incompatible.
-Cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, no validator.
-NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
+* Why NeXML -- because Nexus standard isn't a standard.  Developers extend
+it arbitrarily to meet their needs, making it incompatible.  Cannot
+represent horizontal gene transfer, ambiguous use of symbols, cannot be
+extended reliably, no validator.  NeXML addresses all of these problems
+[Vos _et al._ 2012] with a state-of-the-art data management format.
 
-* Why RNeXML: Huge ecosystem of phylogenetics packages.
-(1) Read nexml.
-All would benefit by being able to read NeXML, both as a robust standard free of the pitfalls of Nexus, and more importantly, as a much richer source of data thanks to its extendable annotation model.
-(2) Write nexml.
-R packages and R users add rich information to the phylogenies they work with, but need a platform independent data exchange format to share and archive this.
-(3) Annotate and extend nexml.
+* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read nexml.
+All would benefit by being able to read NeXML, both as a robust standard
+free of the pitfalls of Nexus, and more importantly, as a much richer
+source of data thanks to its extendable annotation model.  (2) Write
+nexml.  R packages and R users add rich information to the phylogenies
+they work with, but need a platform independent data exchange format to
+share and archive this. (3) Annotate and extend nexml.
 
@@
 
-A nexml file can be thought of as having three components: (1) phylogeny data (2) character-trait data, and (3) metadata.
-`RNeXML` seeks to provide a seemless way to convert these to and from their corresponding R object representations, such as `ape::phylo` phylogenies, character matrices (`data.frames`) and metadata lists (or for advanced users, `XMLInternalDocument` and `rrdf` triplestores).
-Metadata can provide information about the study in which the phylogeny was generated or applied (such as authors, citations to publications, timestamps, and even descriptions of methods), but can also be used to annotate individual data components.
-We illustrate how metadata annotations can be added to the operational taxonomic units of the phylogeny and can be used to extend the data representation to new uses such as stochastic character maps.
+A nexml file can be thought of as having three components: (1) phylogeny
+data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
+provide a seemless way to convert these to and from their corresponding
+R object representations, such as `ape::phylo` phylogenies, character
+matrices (`data.frames`) and metadata lists (or for advanced users,
+`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
+information about the study in which the phylogeny was generated or
+applied (such as authors, citations to publications, timestamps, and even
+descriptions of methods), but can also be used to annotate individual
+data components.  We illustrate how metadata annotations can be added to
+the operational taxonomic units of the phylogeny and can be used to extend
+the data representation to new uses such as stochastic character maps.
 
-The current stable version of the `RNeXML` package can be installed from the CRAN library <!-- Actually not yet -->
-using the standard
+The current stable version of the `RNeXML` package can be installed from
+the CRAN library using the standard <!-- Actually not yet -->
 
@@
 
+This allows us to include multiple trees and character matrices to a single NeXML file.  
+<!-- We can add relevant metadata to each tree, such as the citation information 
+     indicating the paper in which it was first described.  -->
+
 ### Metadata
@@
 a range of tools for accessing metadata embedded in a NeXML file.
+
+### Writing NeXML metadata
+
+Researchers cannot take advantage of NeXML metadata manipulation if
+there is no metadata there.  The burden of adding good metadata remains a
+significant barrier to increased sharing and reuse of scientific research
+[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
+high quality, machine-readable metadata to NeXML files. This occurs on
+several levels:
+
+- Automatic annotation.
+
+`RNeXML` automatically adds basic metadata by default such as a timestamp
+of when the tree was produced and an open data license (CC0, compatible
+with deposition on Dryad data repository), which can be configured
+as needed.  To adjust these and other common metadata fields that should 
+be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
+function.  This function does not assume any familiarity with the concepts
+of namespaces or other more advanced metadata tools discussed later.  The
+`add_basic_meta` takes an existing `nexml` object (or will create a new one
+if none is provided, like the other `add_` functions), and any of the following
+fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
+`citation`.  Additional fields may be added to subsequent versions, see
+`?add_basic_metadata` for the most up-to-date information.  Here we create
+an new `nexml` object with the basic metadata for the `bird.orders` data
+provided in the `ape` package:
+
+```{r}
+data(bird.orders)
+nex <- add_trees(bird.orders)
+nex <- add_basic_meta(nex,
+  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
+
+  description =" This data set describes the phylogenetic relationships of the
+     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
+     and Ahlquist inferred this phylogeny from an extensive number of
+     DNA/DNA hybridization experiments. The ``tapestry'' reported by
+     these two authors (more than 1000 species out of the ca. 9000
+     extant bird species) generated a lot of debates.
+
+     The present tree is based on the relationships among orders. The
+     branch lengths were calculated from the values of Delta T50H as
+     found in Sibley and Ahlquist (1990, fig. 353).",
+
+  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
+     classification of birds: a study in molecular evolution. New
+     Haven: Yale University Press.",
+
+  creator = "Sibley, C. G. and Ahlquist, J. E.")
+```
+
+`RNeXML` provides various functions to assist in automatically adding
+other more specific metadata to various elements of the NeXML data.
+
+- Utilities to add common metadata
+
+  - Add a citation to any papers associated with the phylogeny data.  
+  - Writing in methodological information on how tree has been generated or analyzed (including R code)
+  - Add annotations to any taxon.
+
+Writing annotations on otus.  Annotations on nodes.
+
+For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
+data are a common source of frustration in reusing phylogenetic trees. By 
+associating the taxonomic labels against a specific, centralized database
+we can catch typographical errors and indicate more precisely which classification
+we are appealing to in the event of conflicting taxonomic definitions.  In this 
+example, we indicate that we intend the bird orders listed in the tree to correspond
+to the NCBI definitions
+
+```{r}
+nex <- taxize_nexml(nex, "NCBI")
+```
+
+This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
+label against the NCBI database.  If a unique match is found, a metadata annotation
+is added to the taxon providing the NCBI identification number to the taxonomic unit.
+If no match is found, the user is warned to check for possible typographic errors in
+the taxonomic labels provided.
+
+Temporal, spatial, taxonomic coverage for dryad/knb?
+
+Much richer metadata annotation is possible. Later we illustrate how metadata annotation
+can be used to extend the base NeXML format to represent new forms of data while maintaining
+compatability with any NeXML parser. The `RNeXML` package can be easily extended to support
+helper functions such as `taxize_nexml` to add additional metadata without imposing a large
+burden on the user.
 
@@
 
-All metadata terms must belong to an explicit *namespace* or vocabulary that allows a computer to interpret the term precisely.
-The prefix (before the `:`) indicates to which vocabulary the term belongs, e.g.
-`dc` in this case.
-The `get_namespaces` function tells us the definition of the vocabulary using a link:
+All metadata terms must belong to an explicit *namespace* or vocabulary
+that allows a computer to interpret the term precisely. The prefix (before
+the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
+this case.  The `get_namespaces` function tells us the definition of
+the vocabulary using a link:
 
@@
 
-Other metadata is available
+Common metadata can be accessed with a few dedicated functions:
+
+```{r}
+get_citation(S100)
+```
 
@@
 
-- article citation information
+Which returns text from the otu element labels, typically used to define
+taxonomic names, rather than text from explicit meta elements.
 
-- Summary data: taxa, number of taxa, number of trees, number of tips.
-
-- Metadata of nodes, complete RDF extraction
-
-This allows us to include multiple trees and character matrices to a single NeXML file.
-
-<!-- explain OTU blocks somewhere -->
-
-### Taxonomic identifiers
-
-Add taxonomic identifier metadata to the OTU elements:
+We can also access metadata at a specific level (or use `level=all`
+to extract all meta elements in a list).  Here we show only the first
+few results:
 
 ```{r} 
-nex <- add_trees(bird.orders)
-nex <- nexml_taxize(nex)
+otu_meta <- get_metadata(S100, level="otu")
+otu_meta[1:2]
 ```
 
-Replicate examples from [treebase] metadata section: publication information, tree size, taxanomic coverage
+This approach does not preserve the relationship between the metadata element and the 
+`otu` it was annotating, and as such is most useful for quick visual exploration. 
+Metadata values for specific elements or sets of elemetns can always be accessed using
+the S4 subsetting mechanisms (see section below).
 
-### Writing NeXML metadata
+## RDF manipulation and SPARQL queries
 
-Researchers cannot take advantage of NeXML metadata manipulation if
-there is no metadata there.  The burden of adding good metadata remains a
-significant barrier to increased sharing and reuse of scientific research
-[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
-high quality, machine-readable metadata to NeXML files. This occurs on
-several levels:
+Consider a better example than this one:
 
-- Automatic annotation.
+```{r}
+rdf <- get_rdf(S100)
+```
 
-`RNeXML` automatically adds basic metadata by default such as a timestamp
-of when the tree was produced and an open data license (CC0, compatible
-with deposition on Dryad data repository), which can be configured
-as needed.
+Query the rdf using XPath expressions:
 
-- Utilities to add common metadata
+```{r}
+library(XML)
+xpathSApply(rdf, "//dc:title", xmlValue) 
+```
 
-  - Add a citation to any papers associated with the phylogeny data.  
-  - Writing in methodological information on how tree has been generated or analyzed (including R code)
-  - Add annotations to any taxon.
+XPath is an expressive syntax for querying XML-structured data such
+as RDF.  Queries using the XPath syntax are possible in R using the
+XML library [Temple Lang 2013, XML].  (Note that XPath is far more
+expressive for querying the metadata once the RDFa has been extracted,
+since attribute values are now interpreted as namespaced XML nodes and
+not just arbitrary text.)
 
-Writing annotations on otus.  Annotations on nodes.
+The most powerful way to take advantage of the machine-readable nature
+of the metadata involves the construction of SPARQL queries. This
+approach can exploit the reasoning and logical deduction capacity of
+formal vocabularies (ontologies). For a proper introduction to SPARQL
+queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
+necessary to apply such queries to NeXML files from R.  First we load
+the necessary R library and import the RDF-extracted metadata:
 
-Temporal, spatial, taxonomic coverage for dryad/knb?
+```{r}
+library(rrdf)
+saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
+lib <- load.rdf("rdf_meta.xml")
+```
+
+We can then make arbitrary SPARQL queries directly from R.  For instance,
+here is a simple query to extract the value of any `dc:title` element, analogous
+to the XPath expression above:
+
+```{r}
+sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
+```
 
@@
 
-No data standard can anticipate all the uses and extensions researchers may one day need.
-Researchers have repeatedly adapted the existing NEXUS file format to accomidate encoding new kinds of information, but in the process break compatibility between different flavors of NEXUS files, with few reliable methods to distinguish between these flavors.
-NeXML provides a solution to this problem through the use of metadata annotation.
+No data standard can anticipate all the uses and extensions researchers
+may one day need. Researchers have repeatedly adapted the existing NEXUS
+file format to accomidate encoding new kinds of information, but in the
+process break compatibility between different flavors of NEXUS files, with
+few reliable methods to distinguish between these flavors.  NeXML provides
+a solution to this problem through the use of metadata annotation.
 
-Here we illustrate this process using the example of stochastic character mapping [Huelsenbeck et al 2003].
-A stochastic character map is simply an annotation of the branches on a phylogeny, assigning each section of each branch to a particular "state" (typically of a morphological characteristic).
+Here we illustrate this process using the example of stochastic character
+mapping [Huelsenbeck et al 2003]. A stochastic character map is simply
+an annotation of the branches on a phylogeny, assigning each section
+of each branch to a particular "state" (typically of a morphological
+characteristic).
 
-[Bollback 2006] provides a widely used stand-alone software implementation of this method in the software `simmap`, which modified the standard Newick tree format to express this additional information.
-This can break compatibility with other software[^1], and creates a format that cannot be interpreted without additional information describing this convention.
-By contrast, the NeXML extension is not only backwards compatible but contains a precise and machine-readable description of what it is encoding.
+[Bollback 2006] provides a widely used stand-alone software implementation
+of this method in the software `simmap`, which modified the standard
+Newick tree format to express this additional information. This can
+break compatibility with other software[^1], and creates a format that
+cannot be interpreted without additional information describing this
+convention.  By contrast, the NeXML extension is not only backwards
+compatible but contains a precise and machine-readable description of
+what it is encoding.
 
-[^1]: By using the commenting mechanism of the Newick format, it is possible that other software that doesn't also use the comment mechanism for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the data format to describe its use.
+[^1]: By using the commenting mechanism of the Newick format, it is
+possible that other software that doesn't also use the comment mechanism
+for some other such purpose would be able to successfully parse the tree.
+However there is no way to guarentee that this is the case or for the
+data format to describe its use.
 
-[Revell 2011] describes the `phytools` package for R, which includes utilities for reading, manipulating, and writing `simmap` files.
+[Revell 2011] describes the `phytools` package for R, which includes
+utilities for reading, manipulating, and writing `simmap` files.
 
@@
 
+### Publishing NeXML files from R
+
+Data archiving is increasingly required by journals in evolutionary biology and 
+biodiversity [e.g. @Whitlock2011].  The burden of preparing and submitting properly
+annotated data to archives continues to be a significant barrier [Tenopir et al. 2011, Stodden 2013],
+and many phylogenetic trees upon which studies are based are inaccessible or lost 
+to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
+by providing functions to immediately archive NeXML files in scientific repositories.
+
+```{r eval = FALSE}
+id <- nexml_publish(nexml, visibility = "public", repository="figshare")
+```
+
+_Show downloading the nexml file directly from figshare, explain value
+of global paths instead of personal directory paths appearing in scripts, etc_
+
+Robust data archiving need not involve immediate release of the data. The `figshare` 
+repository supports the option of `visibility = "private"`, allowing secure 
+publication that provides a way to backup and share data with collaborators 
+prior to public release.
+
+<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->
+
 ### Implementation and architecture
@@
 [Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
+[Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636
 

### 2014-03-14 inst/doc/pubs/manuscript.Rmd
writing writing

@@
 1) storing character data and tree data together, convenient for the comparative methods community
-2) Adding metadata, including automated addition of metadata
-3)
+2) Adding metadata, including automated addition of metadata and semantic utilities 
+3) Extending the NeXML standard
 
@@
 
-* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read nexml.
+* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
 All would benefit by being able to read NeXML, both as a robust standard
@@
 source of data thanks to its extendable annotation model.  (2) Write
-nexml.  R packages and R users add rich information to the phylogenies
+NeXML.  R packages and R users add rich information to the phylogenies
 they work with, but need a platform independent data exchange format to
-share and archive this. (3) Annotate and extend nexml.
-
-* as your own personal treebase
-
-* Adding metadata to capture workflow
+share and archive this. In particular, we illustrate how NeXML can
+be used to distribute both phylogenies and character trait data used in 
+many popular phylogenetic comparative methods tools in R.  (3) Annotate 
+and extend nexml. `RNeXML` provides convenient tools for adding metadata
+to NeXML files, including the automated generation of metadata and examples
+of how to extend the NeXML standard to represent new data features without
+breaking validation or backwards compatability of the file type.
 

### 2014-03-14 inst/doc/pubs/manuscript.Rmd
working on manuscript

@@
 ```{r}
+library(geiger)
 data(geospiza)
@@
 ```{r}
-nex <- add_trees(geospiza$phy)
-nex <- add_characters(geospiza$dat, nex)
+geiger_nex <- add_trees(geospiza$phy)
+geiger_nex <- add_characters(geospiza$dat, geiger_nex)
 ```
@@
 data(primates)
-nex <- add_trees(primates$phy, nex)
-nex <- add_characters(primates$dat, nex)
+geiger_nex <- add_trees(primates$phy, geiger_nex)
+geiger_nex <- add_characters(primates$dat, geiger_nex)
 ```
@@
 data(bird.orders)
-nex <- add_trees(bird.orders)
-nex <- add_basic_meta(nex,
+birds <- add_trees(bird.orders)
+birds <- add_basic_meta(birds,
   title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
@@
 
-`RNeXML` provides various functions to assist in automatically adding
+Citations can also be added using R's `bibentry` type, which allows
+a user to add citations to R packages,
+
+```{r}
+birds <- add_basic_meta(birds, citation = citation("ape"))
+```
+
+or to papers by simply using the object's DOI:
+
+```{r}
+library("knitcitations")
+geiger_nex <- add_basic_meta(geiger_nex, citation=cite("10.2307/2408428"))
+```
+
+(Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)
+
+`RNeXML` also provides various functions to assist in automatically adding
 other more specific metadata to various elements of the NeXML data.
-
-- Utilities to add common metadata
-
-  - Add a citation to any papers associated with the phylogeny data.  
-  - Writing in methodological information on how tree has been generated or analyzed (including R code)
-  - Add annotations to any taxon.
-
-Writing annotations on otus.  Annotations on nodes.
 
@@
 ```{r}
-nex <- taxize_nexml(nex, "NCBI")
+birds <- taxize_nexml(birds, "NCBI")
 ```
@@
 
-Temporal, spatial, taxonomic coverage for dryad/knb?
+### Custom metadata extensions
+
+So far our examples have relied on functions that have a built-in understanding of the 
+kind of metadata being provided, and can thus generate the appropriate metadata with 
+little further guidence.  In order to be machine-readable and understandable, 
+NeXML requires all metadata annotations follow the RDFa format which explicitly declares 
+the meaning of the annotation in a way a computer can understand.  For instance, the 
+title we gave to the `bird.orders` phylogeny appears in the NeXML as:
+
+```xml
+<meta id="m2" property="dc:title" datatype="xsd:string" content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
+```
+
+We see the title appears as the `content` of this `meta` element.  The `id` attribute
+provides us with a unique way to refer to this meta element in this document.  The `datatype`
+tells us that the content is a text string (as opposed to a numerical or boolean value).  
+The `property` is the attribute that most interests us here; it tells us that the content 
+is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
+the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:
+
+```{r}
+prefixes <- get_namespaces(birds)
+prefixes["dc"]
+```
+
+This URL links us to a webpage providing both a human and machine-readable version of precisely
+what the term `title` means, showing us that it refers to `title` in the sense of a creative work, 
+rather than the title of a person, a legal title, the verb title, etc.  This precision can allow
+us to perform precise computer reasoning on metadata, without relying on heuristics or assumptions,
+as we illustrate later.  For the moment though, the need to refer to an explicit controlled vocabulary (or
+ontology) with a url, saying "dc:title" rather than just saying the property is "title", is something
+of a burden in creating metadata.  The previous examples involve functions that already know what 
+kind of metadata is being provided (a title, a data creator, etc), and so can provide the appropriate
+vocabulary automatically.  The main advantage of linking to an external vocabulary, however, is the
+ability to express concepts that are not already "built in" to the data model.  `RNeXML` provides 
+tools to generate arbitrary metadata annotations given an external vocabulary, such as we illustrate
+here.
+
+From the list of namespaces in `prefixes`, we can follow links a handful of established vocabularies
+and find terms that are not included in `add_basic_meta`.  For instance, in the `prism` prefix,
+
+```{r}
+prefixes["prism"]
+```
+
+we see that there is a term for `modificationDate` for a file.  We create a `meta` element 
+containing this annotation using the `meta` function:
+
+```{r}
+modified <- meta(property = "prism:modificationDate", content = "2013-10-04")
+```
+
+We can add this annotation to our existing `birds` nexml file using the `add_meta` function.
+Because we do not specify a level, it is added to the root node, refering to the nexml file as a whole.
+
+```{r}
+birds <- add_meta(modified, birds) 
+```
+
+The built-in vocabularies are just the tip of the iceburg of established vocabularies. Here 
+we add an annotation from the `skos` namespace which describes the history of where the
+data comes from:
+
+```{r}
+history <- meta(property = "skos:historyNote",
+  content = "Mapped from the bird.orders data in the ape package using RNeXML")
+```
+
+Because `skos` is not in the current namespace list, we add it with a url when
+adding this meta element.  We also specify that this annotation be placed 
+at the level of the `trees` subnode in the NeXML file.
+
+```{r}
+birds <- add_meta(history, 
+                birds, 
+                level = "trees",
+                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
+```
+
+If we forget to add the namespace with add_meta, we can always add it later
+using the `add_namespaces` function,
+
+```{r}
+birds <- add_namespaces(c(skos = "http://www.w3.org/2004/02/skos/core#"), birds)
+```
+
+(The `add_namespaces` function will avoid creating any duplicate definitions)
+For finer control of the level at which a `meta` element is added, we will manipulate 
+the nexml R object directly using S4 subsetting, as shown later.
+
+<!-- Ugh. y'all have much more practice explaining this well and without tangents, 
+so please feel free to edit or re-write.  -->
 
@@
 
+<!--
+- Utilities to add common metadata
+
+  - Add a citation to any papers associated with the phylogeny data.  
+  - Writing in methodological information on how tree has been generated or analyzed (including R code)
+  - Add annotations to any taxon.
+
+Temporal, spatial, taxonomic coverage for dryad/knb?
+-->
+
 ### Reading NeXML metadata
 
-To illustrate metadata access, we begin by reading in a nexml file.
+To illustrate metadata access, we begin by reading in a NeXML file.
 

### 2014-03-17 inst/doc/pubs/manuscript.Rmd
a few writing edits
draft updated, examples all seem to run at this time for me anyway

@@
 
-``{r}
+```{r eval=FALSE}
 plot(phy)
@@
 ```{r }
-nex <- nexml_read("comp_analysis.xml")
+comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
+nex <- nexml_read(comp_analysis)
 get_characters(nex)
@@
 
-```{r}
+```{r message=FALSE, results='hide'}
 birds <- taxize_nexml(birds, "NCBI")
@@
 
-To illustrate metadata access, we begin by reading in a NeXML file.
+A call to the nexml object prints some metadata summarizing the data structure: <!-- need to make this method faster!-->
 
 ```{r }
-S100 <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
-```
-
-A call to the nexml object prints some metadata summarizing the data structure:
-
-```{r }
-S100
+birds
 ```
@@
 ```{r}
-meta <- get_metadata(S100) 
+meta <- get_metadata(birds) 
 ```
@@
 This returns a named list of available metadata. We can see the kinds
-of metadata recorded from the names:
+of metadata recorded from the names (showing the first 4):
 
 ```{r}
-names(meta)
+names(meta)[1:4]
 ```
@@
 meta[["dc:title"]]
-meta[["dc:publisher"]]
 ```
@@
 ```{r}
-prefixes <- get_namespaces(S100)
+prefixes <- get_namespaces(birds)
 prefixes["dc"]
@@
 ```{r}
-get_citation(S100)
+get_citation(birds)
 ```
@@
 ```{r}
-get_taxa(S100)
+get_taxa(birds)
 ```
@@
 ```{r} 
-otu_meta <- get_metadata(S100, level="otu")
-otu_meta[1:2]
+otu_meta <- get_metadata(birds, level="otu")
+otu_meta[1:4]
 ```
@@
 ```{r}
-rdf <- get_rdf(S100)
+rdf <- get_rdf(birds)
 ```
@@
 ```{r eval = FALSE}
-id <- nexml_publish(nexml, visibility = "public", repository="figshare")
+doi <- nexml_publish(birds, visibility = "public", repository="figshare")
 ```

### 2014-03-18 inst/doc/pubs/manuscript.Rmd
edits additions to rnexml manuscript

--- 2014-03-14 inst/doc/pubs/manuscript.Rmd
@@
 
-<!-- Emphasize: 
-1) storing character data and tree data together, convenient for the comparative methods community
-2) Adding metadata, including automated addition of metadata and semantic utilities 
-3) Extending the NeXML standard
-
--->
-
-Though R provides the most extensive and widely used collection of
-comparative phylogenetic methods tools, R users lack a convenient,
-extensible format to exchange and archive comparative methods data.
+Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient, extensible format to exchange and archive comparative methods data.
 
@@
 
+A summar of some phylogenetic formats in R:
+
+| Class | R package | R Object |
+|-------|-----------|-------|
+| phylo | ape | S3 |
+| phylo4 | phylobase | S4 |
+| phylo4d | adephylo | S4 |
+| ouchtree | ouch | S4 |
+| treeshape | apTreeshape | list |
+
 ### NeXML
@@
 
-Mention of data sharing challenges and opportunities.  See [Drew et al 2013]
+Mention of data sharing challenges and opportunities.  See [Drew et al 2013](10.1371/journal.pbio.1001636)
 
@@
 
-* Why NeXML -- because Nexus standard isn't a standard.  Developers extend
-it arbitrarily to meet their needs, making it incompatible.  Cannot
-represent horizontal gene transfer, ambiguous use of symbols, cannot be
-extended reliably, no validator.  NeXML addresses all of these problems
-[Vos _et al._ 2012] with a state-of-the-art data management format.
+* Why NeXML -- because Nexus standard isn't a standard.
+Developers extend it arbitrarily to meet their needs, making it incompatible <!-- with?
+-->.
+For example, Nexus cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, and there is no way to validate that it is proper Nexus.
+NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
 
-* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
-All would benefit by being able to read NeXML, both as a robust standard
-free of the pitfalls of Nexus, and more importantly, as a much richer
-source of data thanks to its extendable annotation model.  (2) Write
-NeXML.  R packages and R users add rich information to the phylogenies
-they work with, but need a platform independent data exchange format to
-share and archive this. In particular, we illustrate how NeXML can
-be used to distribute both phylogenies and character trait data used in 
-many popular phylogenetic comparative methods tools in R.  (3) Annotate 
-and extend nexml. `RNeXML` provides convenient tools for adding metadata
-to NeXML files, including the automated generation of metadata and examples
-of how to extend the NeXML standard to represent new data features without
-breaking validation or backwards compatability of the file type.
+* Why RNeXML: Huge ecosystem of phylogenetics packages.
+(1) Read nexml.
+All would benefit by being able to read NeXML, both as a robust standard free of the pitfalls of Nexus, and more importantly, as a much richer source of data thanks to its extendable annotation model.
+(2) Write nexml.
+R packages and R users add rich information to the phylogenies they work with, but need a platform independent data exchange format to share and archive this.
+(3) Annotate and extend nexml.
+
+* as your own personal treebase
+
+* Adding metadata to capture workflow
 
@@
 
-A nexml file can be thought of as having three components: (1) phylogeny
-data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
-provide a seemless way to convert these to and from their corresponding
-R object representations, such as `ape::phylo` phylogenies, character
-matrices (`data.frames`) and metadata lists (or for advanced users,
-`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
-information about the study in which the phylogeny was generated or
-applied (such as authors, citations to publications, timestamps, and even
-descriptions of methods), but can also be used to annotate individual
-data components.  We illustrate how metadata annotations can be added to
-the operational taxonomic units of the phylogeny and can be used to extend
-the data representation to new uses such as stochastic character maps.
+A nexml file can be thought of as having three components: (1) phylogeny data (2) character-trait data, and (3) metadata.
+`RNeXML` seeks to provide a seemless way to convert these to and from their corresponding R object representations, such as `ape::phylo` phylogenies, character matrices (`data.frames`) and metadata lists (or for advanced users, `XMLInternalDocument` and `rrdf` triplestores).
+Metadata can provide information about the study in which the phylogeny was generated or applied (such as authors, citations to publications, timestamps, and even descriptions of methods), but can also be used to annotate individual data components.
+We illustrate how metadata annotations can be added to the operational taxonomic units of the phylogeny and can be used to extend the data representation to new uses such as stochastic character maps.
 
-The current stable version of the `RNeXML` package can be installed from
-the CRAN library using the standard <!-- Actually not yet -->
+The current stable version of the `RNeXML` package can be installed from the CRAN library <!-- Actually not yet -->
+using the standard installation method
 
@@
 f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
+nex <- nexml_read("trees.xml")
 ```
@@
 
-<!-- have we already said all this above?--> The ability to read in
+<!-- have we already said all this above?--> <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
 phylogenetic trees in the `nexml` format opens up a wide and rapidly
@@
 NEXML standard, every valid NeXML file on TreeBASE can now successfully
-be parsed.
+be parsed to R objects.
 
@@
 the phylogenetic tree, we find this ideal for character data used in
-comparative methods popular in R. Methods to get extract character data
+comparative methods popular in R. Methods to extract character data
 work much like those for the phylogenetic data:
@@
 
-Returns a `data.frame` with columns as characters and rows as
+`get_characters` returns a `data.frame` with columns as characters and rows as
 taxa. Continuous and discrete characters are represented as seperate
 blocks in NeXML files, but will be combined as seperate columns of a
-single data frame by `get_characters()` if they correspond to the same
+single `data.frame` by `get_characters()` if they correspond to the same
 taxa. The `get_characters_list()` function will consistently recieve
@@
 
-This allows us to include multiple trees and character matrices to a single NeXML file.  
-<!-- We can add relevant metadata to each tree, such as the citation information 
-     indicating the paper in which it was first described.  -->
-
 ### Metadata
 
-At the heart of the NeXML standard is the ability to encode arbitrary
-metadata in precisely specified, machine-readable formats that maintain
-backwards compabitibilty with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations", or RDFa [W3C 2013],
-as described in [Vos et al 2011].  These annotations can be used to add 
-information to any particular element of the data, (a tree, node, edge,
-or taxonomic unit; a character matrix or specific trait, or even another 
-metadata annotation or the NeXML data file as a whole). `RNeXML` provides
-a range of tools for accessing metadata embedded in a NeXML file.
+At the heart of the NeXML standard is the ability to encode arbitrary metadata in precisely specified, machine-readable formats that maintain backwards compabitibilty with any existing NeXML parser.
+NeXML uses the "Resource Description Framework in Annotations", or RDFa [W3C 2013], as described in [Vos et al 2011].
+
+### Reading NeXML metadata
+
+Metadata from Treebase NeXML
+- article citation information
+- Summary data: taxa, number of taxa, number of trees, number of tips.  
+- Metadata of nodes, complete RDF extraction
+
+```{r } 
+birds <- nexml_read("test.xml")
+get_taxa(birds)
+get_metadata(birds) 
+```
+
+This allows us to include multiple trees and character matrices to a single NeXML file.
+
+<!-- explain OTU blocks somewhere -->
+
+### Taxonomic identifiers
+
+Add taxonomic identifier metadata to the OTU elements:
+
+```{r}
+nex <- add_trees(bird.orders)
+nex <- nexml_taxize(nex)
+```
+
+Replicate examples from [treebase] metadata section: publication information, tree size, taxanomic coverage
 
@@
 
-Researchers cannot take advantage of NeXML metadata manipulation if
-there is no metadata there.  The burden of adding good metadata remains a
-significant barrier to increased sharing and reuse of scientific research
-[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
-high quality, machine-readable metadata to NeXML files. This occurs on
-several levels:
+Researchers cannot take advantage of NeXML metadata manipulation if there is no metadata.
+The burden of adding good metadata remains a significant barrier to increased sharing and reuse of scientific research [Tenopir et al.
+2011].
+`RNeXML` seeks to reduce the burden of adding high quality, machine-readable metadata to NeXML files.
+This occurs on several levels:
 
-- Automatic annotation.
-
-`RNeXML` automatically adds basic metadata by default such as a timestamp
-of when the tree was produced and an open data license (CC0, compatible
-with deposition on Dryad data repository), which can be configured
-as needed.  To adjust these and other common metadata fields that should 
-be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
-function.  This function does not assume any familiarity with the concepts
-of namespaces or other more advanced metadata tools discussed later.  The
-`add_basic_meta` takes an existing `nexml` object (or will create a new one
-if none is provided, like the other `add_` functions), and any of the following
-fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
-`citation`.  Additional fields may be added to subsequent versions, see
-`?add_basic_metadata` for the most up-to-date information.  Here we create
-an new `nexml` object with the basic metadata for the `bird.orders` data
-provided in the `ape` package:
-
-```{r}
-data(bird.orders)
-nex <- add_trees(bird.orders)
-nex <- add_basic_meta(nex,
-  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
-
-  description =" This data set describes the phylogenetic relationships of the
-     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
-     and Ahlquist inferred this phylogeny from an extensive number of
-     DNA/DNA hybridization experiments. The ``tapestry'' reported by
-     these two authors (more than 1000 species out of the ca. 9000
-     extant bird species) generated a lot of debates.
-
-     The present tree is based on the relationships among orders. The
-     branch lengths were calculated from the values of Delta T50H as
-     found in Sibley and Ahlquist (1990, fig. 353).",
-
-  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
-     classification of birds: a study in molecular evolution. New
-     Haven: Yale University Press.",
-
-  creator = "Sibley, C. G. and Ahlquist, J. E.")
-```
-
-`RNeXML` provides various functions to assist in automatically adding
-other more specific metadata to various elements of the NeXML data.
+- Automatic annotation.  `RNeXML` automatically adds basic metadata by
+default such as a timestamp of when the tree was produced and an open
+data license (CC0, compatible with deposition on Dryad data repository),
+which can be configured as needed.
 
@@
 
-For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
-data are a common source of frustration in reusing phylogenetic trees. By 
-associating the taxonomic labels against a specific, centralized database
-we can catch typographical errors and indicate more precisely which classification
-we are appealing to in the event of conflicting taxonomic definitions.  In this 
-example, we indicate that we intend the bird orders listed in the tree to correspond
-to the NCBI definitions
-
-```{r}
-nex <- taxize_nexml(nex, "NCBI")
-```
-
-This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
-label against the NCBI database.  If a unique match is found, a metadata annotation
-is added to the taxon providing the NCBI identification number to the taxonomic unit.
-If no match is found, the user is warned to check for possible typographic errors in
-the taxonomic labels provided.
-
 Temporal, spatial, taxonomic coverage for dryad/knb?
 
-Much richer metadata annotation is possible. Later we illustrate how metadata annotation
-can be used to extend the base NeXML format to represent new forms of data while maintaining
-compatability with any NeXML parser. The `RNeXML` package can be easily extended to support
-helper functions such as `taxize_nexml` to add additional metadata without imposing a large
-burden on the user.
+- Extending the NeXML standard through metadata annotation.
 
-### Reading NeXML metadata
+No data standard can anticipate all the uses and extensions researchers may one day need.
+Researchers have repeatedly adapted the existing NEXUS file format to accomidate encoding new kinds of information, but in the process break compatibility between different flavors of NEXUS files, with few reliable methods to distinguish between these flavors.
+NeXML provides a solution to this problem through the use of metadata annotation.
 
-To illustrate metadata access, we begin by reading in a nexml file.
+Here we illustrate this process using the example of stochastic character mapping [Huelsenbeck et al 2003].
+A stochastic character map is simply an annotation of the branches on a phylogeny, assigning each section of each branch to a particular "state" (typically of a morphological characteristic).
 
-```{r } 
-S100 <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
-```
+[Bollback 2006] provides a widely used stand-alone software implementation of this method in the software `simmap`, which modified the standard Newick tree format to express this additional information.
+This can break compatibility with other software[^1], and creates a format that cannot be interpreted without additional information describing this convention.
+By contrast, the NeXML extension is not only backwards compatible but contains a precise and machine-readable description of what it is encoding.
 
-A call to the nexml object prints some metadata summarizing the data structure:
+[^1]: By using the commenting mechanism of the Newick format, it is possible that other software that doesn't also use the comment mechanism for some other such purpose would be able to successfully parse the tree.
+However there is no way to guarentee that this is the case or for the data format to describe its use.
 
-```{r }
-S100
-```
-
-We can extract all metadata pertaining to the NeXML document as a whole
-(annotations of the XML root node, `<nexml>`) with the command
-
-```{r}
-meta <- get_metadata(S100) 
-```
-
-This returns a named list of available metadata. We can see the kinds
-of metadata recorded from the names:
-
-```{r}
-names(meta)
-```
-
-and can ask for a particular element using the standard list subsetting
-mechanism (i.e. either the name of an element or it's numeric position),
-
-```{r}
-meta[["dc:title"]]
-meta[["dc:publisher"]]
-```
-
-All metadata terms must belong to an explicit *namespace* or vocabulary
-that allows a computer to interpret the term precisely. The prefix (before
-the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
-this case.  The `get_namespaces` function tells us the definition of
-the vocabulary using a link:
-
-```{r}
-prefixes <- get_namespaces(S100)
-prefixes["dc"]
-```
-
-Following the link we can see that `dc` refers to the "Dublin Core"
-vocabulary, and the term `title` refers to the title of a creative work
-(as opposed to the title of a person, a legal title, the verb title, etc).
-Such explicit definitions allow a computer to reason about these objects,
-as we shall see later.
-
-Common metadata can be accessed with a few dedicated functions:
-
-```{r}
-get_citation(S100)
-```
-
-```{r}
-get_taxa(S100)
-```
-
-Which returns text from the otu element labels, typically used to define
-taxonomic names, rather than text from explicit meta elements.
-
-We can also access metadata at a specific level (or use `level=all`
-to extract all meta elements in a list).  Here we show only the first
-few results:
-
-```{r} 
-otu_meta <- get_metadata(S100, level="otu")
-otu_meta[1:2]
-```
-
-This approach does not preserve the relationship between the metadata element and the 
-`otu` it was annotating, and as such is most useful for quick visual exploration. 
-Metadata values for specific elements or sets of elemetns can always be accessed using
-the S4 subsetting mechanisms (see section below).
-
-## RDF manipulation and SPARQL queries
-
-Consider a better example than this one:
-
-```{r}
-rdf <- get_rdf(S100)
-```
-
-Query the rdf using XPath expressions:
-
-```{r}
-library(XML)
-xpathSApply(rdf, "//dc:title", xmlValue) 
-```
-
-XPath is an expressive syntax for querying XML-structured data such
-as RDF.  Queries using the XPath syntax are possible in R using the
-XML library [Temple Lang 2013, XML].  (Note that XPath is far more
-expressive for querying the metadata once the RDFa has been extracted,
-since attribute values are now interpreted as namespaced XML nodes and
-not just arbitrary text.)
-
-The most powerful way to take advantage of the machine-readable nature
-of the metadata involves the construction of SPARQL queries. This
-approach can exploit the reasoning and logical deduction capacity of
-formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
-necessary to apply such queries to NeXML files from R.  First we load
-the necessary R library and import the RDF-extracted metadata:
-
-```{r}
-library(rrdf)
-saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
-lib <- load.rdf("rdf_meta.xml")
-```
-
-We can then make arbitrary SPARQL queries directly from R.  For instance,
-here is a simple query to extract the value of any `dc:title` element, analogous
-to the XPath expression above:
-
-```{r}
-sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-```
-
-### Extending the NeXML standard through metadata annotation.
-
-No data standard can anticipate all the uses and extensions researchers
-may one day need. Researchers have repeatedly adapted the existing NEXUS
-file format to accomidate encoding new kinds of information, but in the
-process break compatibility between different flavors of NEXUS files, with
-few reliable methods to distinguish between these flavors.  NeXML provides
-a solution to this problem through the use of metadata annotation.
-
-Here we illustrate this process using the example of stochastic character
-mapping [Huelsenbeck et al 2003]. A stochastic character map is simply
-an annotation of the branches on a phylogeny, assigning each section
-of each branch to a particular "state" (typically of a morphological
-characteristic).
-
-[Bollback 2006] provides a widely used stand-alone software implementation
-of this method in the software `simmap`, which modified the standard
-Newick tree format to express this additional information. This can
-break compatibility with other software[^1], and creates a format that
-cannot be interpreted without additional information describing this
-convention.  By contrast, the NeXML extension is not only backwards
-compatible but contains a precise and machine-readable description of
-what it is encoding.
-
-[^1]: By using the commenting mechanism of the Newick format, it is
-possible that other software that doesn't also use the comment mechanism
-for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the
-data format to describe its use.
-
-[Revell 2011] describes the `phytools` package for R, which includes
-utilities for reading, manipulating, and writing `simmap` files.
+[Revell 2011] describes the `phytools` package for R, which includes utilities for reading, manipulating, and writing `simmap` files.
 
@@
 [Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
-
-### Subsetting nexml objects directly
-
-The `RNeXML` package provides many convenient functions to add and extract
-information from `nexml` objects in the R environment without requiring
-the reader to understand the details of the NeXML data structure and
-making it less likely that a user will generate invalid NeXML syntax
-that could not be read by other parsers. The `nexml` object we have been using
-in all of the examples is built on R's S4 mechanism. Advanced users may
-sometimes prefer to interact with the data structure more directly using 
-R's S4 class mechanism and subsetting methods. Many R users are more familiar
-with the S3 class mechanism (such as in the `ape` package phylo objects)
-rather than the S4 class mechanism used in phylogenetics packages such as 
-`ouch` and `phylobase`. The `phylobase` vignette provides an excellent introduction
-to these data structures.  Users already familiar with subsetting lists and other
-S3 objects in R are likely familar with the use of the `$` operator, such as
-`phy$edge`. S4 objects simply use an `@` operator instead (but cannot be subset
-using numeric arguments such as `phy[[1]]` or named arguments such as phy[["edge"]]).
-
-The `nexml` object is an S4 object, as are all of its components (slots).  It's 
-heirachical structure corresponds exactly with the XML tree of a NeXML file, with 
-the single exception that both XML attributes and children are represented as slots.  
-S4 objects have constructor functions to initialize them.  We create a new `nexml` 
-object with the command:
-
-```{r}
-nex <- new("nexml")
-```
-
-We can see a list of slots contained in this object with
-
-```{r}
-slotNames(nex)
-```
-
-Some of these slots have already been populated for us, for instance, the schema version and default namespaces:
-
-```{r}
-nex@version
-nex@namespaces
-```
-
-Recognize that `nex@namespaces` serves the same role as `get_namespaces` function, but provides direct access to the slot data.
-For instance, with this syntax we could also overwrite the existing namespaces with `nex@namespaces <- NULL`.
-Changing the namespace in this way is not advised.
-
-Some slots can contain multiple elements of the same type, such as `trees`, `characters`, and `otus`.  For instance, we see that
-
-```{r}
-class(nex@characters)
-```
-
-is an object of class `ListOfcharacters`, and is currently empty,
-
-```{r}
-length(nex@characters)
-```
-
-In order to assign an object to a slot, it must match the class definition
-of the slot.  We can create a new element of any given class with the
-`new` function,
-
-```{r}
-nex@characters <- new("ListOfcharacters", list(new("characters")))
-```
-
-and now we have a length-1 list of character matrices,
-
-```{r}
-length(nex@characters)
-```
-
-and we access the first character matrix using the list notation, `[[1]]`. Here we check the class is a `characters` object.
-
-```{r}
-class(nex@characters[[1]])
-```
-
-Direct subsetting has two primary use cases: (a) useful in looking up (and possibly editing) a specific value of an element, or (b) when adding metadata annotations to specific elements.
-Consider the example file
-
-```{r}
-f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
-```
-
-We can look up the species label of the first `otu` in the first `otus` block:
-
-```{r}
-nex@otus[[1]]@otu[[1]]@label
-```
-
-We can add metadata to this particular OTU using this subsetting format
-
-```{r}
-nex@otus[[1]]@otu[[1]]@meta <- 
-  c(meta("skos:note", 
-          "This species was incorrectly identified"),
-         nex@otus[[1]]@otu[[1]]@meta)
-```
-
-Here we use the `c` operator to append this element to any existing meta annotations to this otu.
-
-### Publishing NeXML files from R
-
-Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity [e.g. @Whitlock2011].  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier [Tenopir et al. 2011, Stodden 2013],
-and many phylogenetic trees upon which studies are based are inaccessible or lost 
-to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
-by providing functions to immediately archive NeXML files in scientific repositories.
-
-```{r eval = FALSE}
-id <- nexml_publish(nexml, visibility = "public", repository="figshare")
-```
-
-_Show downloading the nexml file directly from figshare, explain value
-of global paths instead of personal directory paths appearing in scripts, etc_
-
-Robust data archiving need not involve immediate release of the data. The `figshare` 
-repository supports the option of `visibility = "private"`, allowing secure 
-publication that provides a way to backup and share data with collaborators 
-prior to public release.
-
-<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->
 
@@
 There are many promising directions for further development of the
-`RNeXML` package.  The project's Milestones page provides a list of
-steps currently planned or under consideration.  In particular, these
-milestones include increased support for interactive mode of entering data
-(wizards), and support for integration of semantics defined through the
-Web Ontology language (OWL), as discussed in the Challenges section.
+`RNeXML` package.  The project's [Milestones page][milestones] 
+provides a list of steps currently planned or under consideration.  In 
+particular, these milestones include increased support for interactive 
+mode of entering data (wizards), and support for integration of semantics 
+defined through the Web Ontology language (OWL), as discussed in the 
+Challenges section.
 
@@
 [Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
-[Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636
+[milestones]: https://github.com/ropensci/RNeXML/issues/milestones
 

### 2014-03-18 inst/doc/pubs/manuscript.Rmd

--- 2014-03-17 inst/doc/pubs/manuscript.Rmd
@@
 
-* Why NeXML -- because Nexus standard isn't a standard.  Developers extend
-it arbitrarily to meet their needs, making it incompatible.  Cannot
-represent horizontal gene transfer, ambiguous use of symbols, cannot be
-extended reliably, no validator.  NeXML addresses all of these problems
-[Vos _et al._ 2012] with a state-of-the-art data management format.
+* Why NeXML -- because Nexus standard isn't a standard.
+Developers extend it arbitrarily to meet their needs, making it incompatible <!-- with?
+-->.
+For example, Nexus cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, and there is no way to validate that it is proper Nexus.
+NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
 
@@
 The current stable version of the `RNeXML` package can be installed from
-the CRAN library using the standard <!-- Actually not yet -->
+the CRAN library using the standard installation method <!-- Actually not yet -->
 
@@
 
-<!-- have we already said all this above?--> The ability to read in
+<!-- have we already said all this above?-->  <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
 phylogenetic trees in the `nexml` format opens up a wide and rapidly
@@
 There are many promising directions for further development of the
-`RNeXML` package.  The project's Milestones page provides a list of
+`RNeXML` package.  The project's [Milestones page][milestones] page provides a list of
 steps currently planned or under consideration.  In particular, these
@@
 [Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636 
+[milestones]: https://github.com/ropensci/RNeXML/issues/milestones
 

### 2014-03-21 inst/doc/pubs/manuscript.Rmd
writing

@@
 ---
-title: RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R
+layout: preprint
+#layout: review, 11pt
+title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
 author: 
@@
 
+```{r compile-settings, include=FALSE}
+library(knitr)
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile') )
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
+```
+
 Though R provides the most extensive and widely used collection of
@@
 
+<!--
 A summar of some phylogenetic formats in R:
 
-| Class | R package | R Object |
-|-------|-----------|-------|
-| phylo | ape | S3 |
-| phylo4 | phylobase | S4 |
-| phylo4d | adephylo | S4 |
-| ouchtree | ouch | S4 |
-| treeshape | apTreeshape | list |
+| Class     | R package   | R Object 
+|-----------|-------------|-------
+| phylo     | ape         | S3 
+| phylo4    | phylobase   | S4
+| phylo4d   | adephylo    | S4
+| ouchtree  | ouch        | S4 
+| treeshape | apTreeshape | list 
+-->
 
@@
 
+In this example, we illustrate how the additional information required
+to define a stochastic character mapping (a `simmap` mapping) in NeXML.
+
 [^1]: By using the commenting mechanism of the Newick format, it is
@@
 [Revell 2011] describes the `phytools` package for R, which includes
-utilities for reading, manipulating, and writing `simmap` files.
+utilities for reading, manipulating, and writing `simmap` files in R.
+In this example, we also show how to define `RNeXML` functions that 
+map the R representation used by Revell (an extension of the `ape` class)
+into the NeXML extension we have defined by using `RNeXML` functions..
 
-NeXML annotation of an edge that switches from state `s2` to state `s1`:
+Since a stochastic character map simply assigns different states to
+parts of a branch (or edge) on the phylogenetic tree, we can create
+a NeXML representation by annotating the `edge` elements with appropriate
+`meta` elements.  These elements need to describe the character state
+being assigned and the duration (in terms of branch-length) that the edge 
+spends in that state (Stochastic character maps are specific to time-calibrated
+or ultrametric trees).
 
-```xml
-<edge about="#e34" id="e34" source="n16" target="n35" length="0.205223">
-        <meta xsi:type="nex:LiteralMeta" id="m262" property="simmap:reconstructions" datatype="xsd:string">
-          <meta xsi:type="nex:LiteralMeta" id="m260" property="simmap:reconstruction" datatype="xsd:string">
-            <meta xsi:type="nex:LiteralMeta" id="m261" property="simmap:char" datatype="xsd:string" content="cr1"/>
-            <meta xsi:type="nex:LiteralMeta" id="m252" property="simmap:stateChange" datatype="xsd:string">
-              <meta xsi:type="nex:LiteralMeta" id="m253" property="simmap:order" datatype="xsd:decimal" content="1"/>
-              <meta xsi:type="nex:LiteralMeta" id="m254" property="simmap:length" datatype="xsd:decimal" content="0.202947550453246"/>
-              <meta xsi:type="nex:LiteralMeta" id="m255" property="simmap:state" datatype="xsd:string" content="s2"/>
-            </meta>
-            <meta xsi:type="nex:LiteralMeta" id="m256" property="simmap:stateChange" datatype="xsd:string">
-              <meta xsi:type="nex:LiteralMeta" id="m257" property="simmap:order" datatype="xsd:decimal" content="2"/>
-              <meta xsi:type="nex:LiteralMeta" id="m258" property="simmap:length" datatype="xsd:decimal" content="0.00227544954675424"/>
-              <meta xsi:type="nex:LiteralMeta" id="m259" property="simmap:state" datatype="xsd:string" content="s1"/>
-            </meta>
-          </meta>
-        </meta>
-      </edge>
+NeXML already defines the `characters` element to handle descrete character traits (`nex:char`)
+and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
+this purpose, referring to both the character trait and the states by the ids assigned to them
+in that element.  (NeXML's convention of referring to everything by id permits a single canonical
+definition of each term, making it clear where additional annotation belongs).  For each edge, we 
+need to indicate:
+
+<!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?
+     -->
+
+- That our annotation contains a stochastic character mapping reconstruction
+- Since many reconstructions are possible for a single edge, we give each reconstruction an id
+- We indicate for which character trait we are defining the reconstruction 
+- We then indicate which states the character assumes on that edge. 
+  For each state realized on the edge, that involves stating: 
+    + the state assignment
+    + the duration (length of time) for which the edge spends in the given state
+    + the order in which the state changes happen (Though we could just assume 
+      state transitions are listed chronologically, NeXML suggests making all 
+      data explicit, rather than relying on the structure of the data file to
+      convey information).
+
+Thus the the annotation for an edge that switches from state `s2` to state 
+`s1` of character `cr1` would be constructed like this:
+
+```{r}
+ m <- meta("simmap:reconstructions", children = c(
+        meta("simmap:reconstruction", children = c(
+
+          meta("simmap:char", "cr1"),
+          meta("simmap:stateChange", children = c(
+            meta("simmap:order", 1),
+            meta("simmap:length", "0.2030"),
+            meta("simmap:state", "s2"))),
+
+          meta("simmap:char", "cr1"),
+          meta("simmap:stateChange", children = c(
+            meta("simmap:order", 2),
+            meta("simmap:length", "0.0022"),
+            meta("simmap:state", "s1")))
+          ))))
 ```
+
+Of course writing out such a definition manually becomes tedious quickly. Because
+these are just R commands, we can easily define a function that can loop over an
+assignment like this for each edge, extracting the appropriate order, length and
+state from an existing R object such as that provided in the `phytools` package.  
+Likewise, it is straight forward to define a function that reads this data using
+the `RNeXML` utilities and converts it back to the `phytools` package. The full
+implementation of this mapping can be seen in the `simmap_to_nexml()` and the
+`nexml_to_simmap()` functions provided in the `RNeXML` package.
+
+As the code indicates, the key step is simply to define the data in meta elements. In 
+so doing, we have defined a custom namespace, `simmap`, to hold our variables.  This
+allows us to provide a URL with more detailed descriptions of what each of these 
+elements mean:
+
+```{r}
+nex <- add_namespaces(c(simmap = "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
+```
+
+At that URL we post a simple description of each term.
 

### 2014-03-24 inst/doc/pubs/manuscript.Rmd
updated manuscript
update simmap example in manuscript

@@
 ```xml
-<meta id="m2" property="dc:title" datatype="xsd:string" content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
+<meta id="m2" property="dc:title" datatype="xsd:string"
+      content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
 ```
@@
 Newick tree format to express this additional information. This can
-break compatibility with other software[^1], and creates a format that
+break compatibility with other software [^1], and creates a format that
 cannot be interpreted without additional information describing this
@@
 to define a stochastic character mapping (a `simmap` mapping) in NeXML.
-
-[^1]: By using the commenting mechanism of the Newick format, it is
-possible that other software that doesn't also use the comment mechanism
-for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the
-data format to describe its use.
 
@@
 
-At that URL we post a simple description of each term.
+At that URL we have posted a simple description of each term.
+
+<!-- How could this be improved? Should we use something more formal? 
+We want this example to be something other users could reasonably
+do themselves without expertise in RDFa or XML, but also to be a good
+model case that doesn't cut corners.  -->
+
+Using this convention we can generate NeXML files containing simmap
+data, read those files into R, and convert them back into the `phytools`
+package format. These simple functions serve as further illustration of
+how `RNeXML` can be used to extend the NeXML standard.  We illustrate
+their use briefly here, starting with  loading a nexml object containing
+a simmap reconstruction into R:
+
+```{r}
+data(simmap_ex)
+```
+
+The `get_trees` function can be used to return an `ape::phylo` tree as
+usual.  `RNeXML` automatically detects the simmap reconstruction data
+and returns includes this in a `maps` element of the `phylo` object,
+for use with other `phytools` functions.
+
+```{r}
+phy <- nexml_to_simmap(simmap_ex)
+```
+
+We can then use various functions from `phytools` designed for simmap
+objects, such as the plotting function:
+
+```{r}
+library(phytools)
+plotSimmap(phy)
+```
+
+Likewise, we can convert the object back in the nexml format and write
+it out to file to be read by other users.
+
+```{r}
+nex <- simmap_to_nexml(phy) 
+nexml_write(nex, "simmap.xml")
+```
+
+Though other NeXML parsers (for instance, for perl or python) have
+not been written explicitly to express simmap data, those parsers will
+nonetheless be able to successfully parse this file and expose the simmap
+data to the user.
 
@@
 [Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
+
+[^1]: By using the commenting mechanism of the Newick format, it is
+possible that other software that doesn't also use the comment mechanism
+for some other such purpose would be able to successfully parse the tree.
+However there is no way to guarentee that this is the case or for the
+data format to describe its use.
 
@@
 Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity [e.g. @Whitlock2011].  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier [Tenopir et al. 2011, Stodden 2013],
+biodiversity (e.g. [Whitlock2011]).  The burden of preparing and submitting properly
+annotated data to archives continues to be a significant barrier ([Tenopir et al. 2011], [Stodden 2010]),
 and many phylogenetic trees upon which studies are based are inaccessible or lost 

### 2014-03-24 R/get_trees.R

@@ -100,7 +100,8 @@ setAs("nexml", "phylo", function(from){
 #' @param tree an nexml tree element 
 #' @param otus a character string of taxonomic labels, named by the otu ids.  
 #' e.g. (from get_otu_maps for the otus set matching the relevant trees node. 
-#' @return phylo object 
+#' @return phylo object.  If a "reconstructions" annotation is found on the 
+#' edges, return simmap maps slot as well.  
 toPhylo <- function(tree, otus){
 
   ## Extract the nodes list
@@ -167,6 +168,12 @@ toPhylo <- function(tree, otus){
   if(!is.null(edge.length))
     phy$edge.length = edge.length # optional fields
   class(phy) = "phylo"
+
+
+
+  ## Check for simmap  
+
+
   phy
 }
 

### 2014-03-24 R/simmap.R

@@ -113,7 +113,7 @@ nexml_to_simmap <- function(nexml){
   characters <- get_characters(nexml)
 
   ## loop over trees blocks
-  lapply(nexml@trees, function(trees){
+  out <- lapply(nexml@trees, function(trees){
     phys <- lapply(trees@tree, 
                    tree_to_simmap, 
                    get_otu_maps(nexml)[[trees@otus]],
@@ -125,6 +125,15 @@ nexml_to_simmap <- function(nexml){
     class(phys) <- "multiPhylo"
     phys 
   })
+
+
+  if(length(out) == 1){
+    if(length(out[[1]]) > 1){
+      flatten_multiphylo(out)
+    } else {
+      out[[1]][[1]]
+    }
+  }
 
 }
 

### 2014-03-25 inst/doc/pubs/manuscript.Rmd
updated manuscript.Rmd with @rvosa's edits, regenerate manuscript.md

@@
   - name: Rutger Vos
-    affiliation: NCB
+    affiliation: NBC
 address: 
@@
       Department of Biology, Simon Fraser University, Burnaby, Canada
-  - code: NCB
+  - code: NBC
     address: | 
-      NCB Naturalis, Leiden, the Netherlands
+      NBC Naturalis, Leiden, the Netherlands
 abstract: | 
@@
       improvements in reliability, extensibility, and richness over
-      earlier standards such as Nexus.
+      earlier standards such as NEXUS.
 
@@
 
-- Most R users rely on the nexus inherits the limitations of nexus file format
+- Most R users rely on the NEXUS inherits the limitations of NEXUS file format
 
@@
 
-* Why NeXML -- because Nexus standard isn't a standard.
-Developers extend it arbitrarily to meet their needs, making it incompatible <!-- with?
--->.
-For example, Nexus cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, and there is no way to validate that it is proper Nexus.
-NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.
+* Why NeXML -- because NEXUS standard isn't a standard.  Developers extend
+it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
+For example, NEXUS cannot represent horizontal gene transfer, ambiguous
+use of symbols, cannot be extended reliably, and there is no way to
+validate that it is proper NEXUS.  NeXML addresses all of these problems
+[Vos _et al._ 2012] with a state-of-the-art data management format.
 
@@
 All would benefit by being able to read NeXML, both as a robust standard
-free of the pitfalls of Nexus, and more importantly, as a much richer
+free of the pitfalls of NEXUS, and more importantly, as a much richer
 source of data thanks to its extendable annotation model.  (2) Write
@@
 Previously, this was possible using the `treebase` package [@treebase]
-for R, which relied on the nexus parser provided in the `ape` package.
-Due to inconsistencies present in the Nexus standard [Vos _et al._
-2012], a significant number <!-- State explicitly --> of TreeBASE nexus
+for R, which relied on the NEXUS parser provided in the `ape` package.
+Due to inconsistencies present in the NEXUS standard [Vos _et al._
+2012], a significant number <!-- State explicitly --> of TreeBASE NEXUS
 files could not be read in this way.  Using the more precisely defined
@@
 
-Like nexus files, NeXML files can also contain character data.  While
+Like NEXUS files, NeXML files can also contain character data.  While
 most researchers will be familiar with the use of character matrices in
-nexus files as a container for molecular sequence data used to infer
+NEXUS files as a container for molecular sequence data used to infer
 the phylogenetic tree, we find this ideal for character data used in
@@
 If no match is found, the user is warned to check for possible typographic errors in
-the taxonomic labels provided.
+the taxonomic labels provided.  <!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->
 
@@
 provides us with a unique way to refer to this meta element in this document.  The `datatype`
-tells us that the content is a text string (as opposed to a numerical or boolean value).  
+tells us that the content is a text string (as opposed to, say, a numerical or boolean value).  
 The `property` is the attribute that most interests us here; it tells us that the content 
@@
 
-The built-in vocabularies are just the tip of the iceburg of established vocabularies. Here 
+The built-in vocabularies are just the tip of the iceberg of established vocabularies. Here 
 we add an annotation from the `skos` namespace which describes the history of where the
@@
 do themselves without expertise in RDFa or XML, but also to be a good
-model case that doesn't cut corners.  -->
+model case that doesn't cut corners. @cboettig -->
+
+<!-- I don't see how this could be improved much without forcing people
+to fire op protege (let's not!). As it happens, the Dublin Core URI, 
+for example, points to an HTML page so even in that case automatic 
+reasoners would not be able to consume the vocabulary and draw inferences
+about, e.g. identity of predicates (owl:sameAs and such). @rvosa -->
 
@@
 
-```{r}
+```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
 library(phytools)
@@
 
-Though other NeXML parsers (for instance, for perl or python) have
+Though other NeXML parsers (for instance, for Perl or Python) have
 not been written explicitly to express simmap data, those parsers will

### 2014-04-28 inst/doc/pubs/manuscript.Rmd
Update manuscript.Rmd

@@
     address: | 
-      NBC Naturalis, Leiden, the Netherlands
+      Naturalis Biodiversity Center, Leiden, the Netherlands
 abstract: | 
       1. NeXML promises to be the next generation of phylogenetic
-      informatics data and metadata exchange format, offering substantial
+      informatics data and metadata exchange formats, offering substantial
       improvements in reliability, extensibility, and richness over
@@
 
-      2. In this paper we present the RNeXML package which provides
+      2. In this paper we present the RNeXML package, which provides
       a bridge between this rich and reliable data format and the
@@
 
-* Why NeXML -- because NEXUS standard isn't a standard.  Developers extend
+* Why NeXML -- because the NEXUS standard isn't a standard.  Developers extend
 it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
@@
 
-A nexml file can be thought of as having three components: (1) phylogeny
+A NeXML file can be thought of as having three components: (1) phylogeny
 data (2) character-trait data, and (3) metadata. `RNeXML` seeks to

### 2014-04-28 inst/doc/pubs/manuscript.Rmd
minor edits

--- 2014-03-25 inst/doc/pubs/manuscript.Rmd
@@
 abstract: |
+
       1. NeXML promises to be the next generation of phylogenetic
@@
       improvements in reliability, extensibility, and richness over
-      earlier standards such as NEXUS.
+      earlier standards such as NEXUS. In this paper we present the RNeXML
+      package which provides a bridge between this rich and reliable
+      data format and the extensive library of phylogenetic tools already
+      available in R, with a particular emphasis on comparative methods.
 
-      2. In this paper we present the RNeXML package which provides
-      a bridge between this rich and reliable data format and the
-      extensive library of phylogenetic tools already available in R,
-      with a particular emphasis on comparative methods.
-
-      3. RNeXML allows R users to both read and write NeXML files. Yet
+      2. RNeXML allows R users to both read and write NeXML files. Yet
       RNeXML is much richer than a basic parser: here we illustrate
@@
 
-      4. An essential feature of NeXML is to provide both the flexibility
+      3. An essential feature of NeXML is to provide both the flexibility
       to be continually extended to meet the emerging needs of the phylogenetics
@@
       for earlier versions.  Here we illustrate how RNeXML makes it easy
-      to extend NeXML in this way.
+      to extend NeXML in this manner.
 
-      5. RNeXML is perfectly suited for users to preserve and maintain their
+      4. RNeXML is perfectly suited for users to preserve and maintain their
       personal archives of phylogenetic and comparative trait data in consistent,
@@
 <!-- Emphasize:
+
 1) storing character data and tree data together, convenient for the comparative methods community
@@
 3) Extending the NeXML standard  
+4) Accessing / opening up existing rich NeXML data to the R user
 
@@
 using the R software environment and the XML package for parsing and
-generating XML [Temple Lang 2013, XML], and provides conversions into
+generating XML [Temple Lang 2013, XML].
+
+<!--  Find more consise way to make sure dependencies are cited. 
+      How about citing packages that are only on the suggests list?
+
+and provides conversions into
 the popular `phylo` format provided by the `ape` package [Paradis 2004].
@@
 ([Chamberlain 2013]).
+-->
 

### 2014-04-28 inst/doc/pubs/manuscript.Rmd

--- 2014-04-28 inst/doc/pubs/manuscript.Rmd
@@
       improvements in reliability, extensibility, and richness over
-      earlier standards such as NEXUS.
+      earlier standards such as NEXUS. In this paper we present the RNeXML
+      package which provides a bridge between this rich and reliable
+      data format and the extensive library of phylogenetic tools already
+      available in R, with a particular emphasis on comparative methods.
 
@@
       with a particular emphasis on comparative methods. 
-
-      3. RNeXML allows R users to both read and write NeXML files. Yet
       RNeXML is much richer than a basic parser: here we illustrate

### 2014-05-02 manuscripts/manuscript.Rmd
reorganizing into manuscripts

--- 2014-04-28 inst/doc/pubs/manuscript.Rmd
@@
                cache = 1, comment = NA, verbose = TRUE)
-basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile') )
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
+               cache.path = paste("components/cache/", basename, "/", sep=""))
 
@@
 conflicting specifications).
+
+```{r include=FALSE}
+# remove external NeXML files created by running this vignette
+unlink("simmap.xml")
+unlink("birds.xml")
+unlink("geospiza.xml")
+unlink("rdf_meta.xml")
+```

### 2014-05-02 manuscripts/tutorial.Rmd
reorganizing into manuscripts

--- 2014-03-06 inst/doc/pubs/README.Rmd
@@
 library(knitr)
-opts_knit$set(upload.fun = socialR::flickr.url)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=TRUE)
+opts_knit$set(upload.fun = imgur_upload)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
 ```
@@
 ```
+
+```{r include=FALSE}
+unlink("example.xml")
+unlink("meta_example.xml")
+unlink("test.xml")
+```

### 2014-05-02 manuscripts/manuscript.Rmd
call library(methods) internally

@@
 ```{r compile-settings, include=FALSE}
+library(methods)
 library(knitr)

### 2014-06-16 manuscripts/manuscript.Rmd
edits, more work on intro, fixed some citations

--- 2014-04-28 inst/doc/pubs/manuscript.Rmd
@@
     affiliation: cstar
-    email: cboettig@gmail.com
+    email: cboettig(at)gmail.com
     footnote: Corresponding author
@@
 
-      2. In this paper we present the RNeXML package, which provides
-      a bridge between this rich and reliable data format and the
-      extensive library of phylogenetic tools already available in R,
-      with a particular emphasis on comparative methods. 
-      RNeXML is much richer than a basic parser: here we illustrate
+      2. RNeXML is much richer than a basic parser: here we illustrate
       how the package facilitates adding and manipulating semantic 
@@
 
-      4. RNeXML is perfectly suited for users to preserve and maintain their
+      4. RNeXML is well suited for users to preserve and maintain their
       personal archives of phylogenetic and comparative trait data in consistent,
@@
 
+bibliography: components/references.bib
+csl: components/methods-in-ecology-and-evolution.csl
+documentclass: components/elsarticle
+
+output: 
+  pdf_document:
+    template: components/elsarticle.latex
+    keep_tex: false
+    fig_caption: true
+
 ---
 
-<!-- Emphasize:
-
-1) storing character data and tree data together, convenient for the comparative methods community
-2) Adding metadata, including automated addition of metadata and semantic utilities 
-3) Extending the NeXML standard  
-4) Accessing / opening up existing rich NeXML data to the R user
-
--->
+<!------------------------------------------------
+
+Map/Outline:
+------------
+
+- Introduction
+  - Exciting times in phylogenetics!  
+  - Challenges with existing formats 
+  - Overview of NeXML
+  - Map
+
+- Tutorial
+  - Reading Phylogenetic trees
+  - Reading Character data
+  - Writing phylogenetic trees & character data
+
+  - Metadata
+    - Adding basic metadata
+    - Adding taxonomic identifers
+    - Adding Generic metadata
+    - Parsing metadata. XML. RDF.
+    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.
+
+  - Publishing NeXML
+
+- Implementation, Quality Control
+- Software reuse and support
+
+------------------------------------------------>
 
 ```{r compile-settings, include=FALSE}
-library(knitr)
+library("methods")
+library("knitr")
 opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
                cache = 1, comment = NA, verbose = TRUE)
-basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile') )
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
-
-```
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
+               cache.path = paste("components/cache/", basename, "/", sep=""))
+
+library("knitcitations")
+cleanbib()
+cite_options(citation_format = "pandoc", check.entries=FALSE)
+```
+
+# Introduction
+
+<!-- Need to say some of this stuff without repeating too much of existing literature -->
 
@@
 extensible format to exchange and archive comparative methods data.
-
-## Background
-
-### Challenges with existing formats
-See [Vos _et al._ 2012].
-
-### Challenges with R formats
-
-- Most R users rely on the NEXUS inherits the limitations of NEXUS file format
-
-See [phylobase vignette, 2010]
-
-<!--
-A summar of some phylogenetic formats in R:
-
-| Class     | R package   | R Object 
-|-----------|-------------|-------
-| phylo     | ape         | S3 
-| phylo4    | phylobase   | S4
-| phylo4d   | adephylo    | S4
-| ouchtree  | ouch        | S4 
-| treeshape | apTreeshape | list 
--->
-
-### NeXML
-
-See [Vos _et al._ 2012].
-
-Mention of data sharing challenges and opportunities.  See [Drew et al 2013]
-
-Why RNeXML?
-
-* Why NeXML -- because the NEXUS standard isn't a standard.  Developers extend
-it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
-For example, NEXUS cannot represent horizontal gene transfer, ambiguous
-use of symbols, cannot be extended reliably, and there is no way to
-validate that it is proper NEXUS.  NeXML addresses all of these problems
-[Vos _et al._ 2012] with a state-of-the-art data management format.
-
-* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
-All would benefit by being able to read NeXML, both as a robust standard
-free of the pitfalls of NEXUS, and more importantly, as a much richer
-source of data thanks to its extendable annotation model.  (2) Write
-NeXML.  R packages and R users add rich information to the phylogenies
+A common file format for sharing phylogenetic tree data has long played
+an essential role in the advancement of phylogenetics research.
+The rapid increase in phylogenetics data and research methods that make
+use of such data has introduced many new challenges and opportunities 
+to sharing this information both between software environments and 
+between research groups `r citet("Drew PLOS Biology Tree of Life 2013")`.
+
+<!-- FIXME -->
+
+ADD STUFF ABOUT:
+
+- Rapid growth in phylogenetics data (+citations)
+- Rapid growth of research methods and software that use phylogenetic data. Comparative methods in evolution. Ecological phylogenetics. Biodiversity informatics. (+citations)
+- More on data sharing challenges and opportunities a la @Drew_2013
+
+The ever expanding needs of the phylogenetics community are rapidly
+out-pacing the capacity of current standards, which is resulting in a
+radiation of different data representations and a loss of interoperability
+at the very time when the growth of available data and mehtods has made
+that interoprability most valuable.
+
+For example, the currently popular NEXUS file format cannot consistently
+represent horizontal gene transfer or ambiguity in reading a character
+(such as a DNA sequence base pair).  This has led developers to
+repeatedly add small modifications to the standard to meet the needs
+of their own software, which results in mutually incompatible dialects
+that largely defeat the advantages of a shared standard file format.
+NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
+by drawing on state-of-the-art data management technology.
+
+<!-- What do we need to say, if anything, about the current R formats?  about phylobase?  -->
+
+The R lanugage `r citep(citation())` has become an important enviroment for 
+phylogenetics research, with over 30 packages available including several
+with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
+though stand-alone programs remain an important part of this ecosystem
+for tasks such as generating phylogenetic trees from sequence data.  
+Within this ecosystem a variety of different data formats for
+phylogenetic data have been developed and modified. Most but not all 
+are based on the simple and popular `phylo` format provided by the
+ape package `r citep(citation("ape"))`.
+
+Though certain functions allow R users to switch between these formats
+within R, these formats frequently contain more information than their
+NEXUS file counterparts, and users lack a suitable alternative for
+archiving and sharing phylogenetic data outside of the R ecosystem. 
+Consequently, information added to phylogenies by these R programs cannot 
+easily be shared, archived, and made available to other software;
+but inherits the limiations of the NEXUS format.
+
+`RNeXML` opens up a rich world of both currently available and rapidly
+growing data sets in the NeXML format to the R user and the large ecosystem
+of R packages currently avaliable for handling phylogenetic data  
+R packages and R users frequently add rich information to the phylogenies
 they work with, but need a platform independent data exchange format to
-share and archive this. In particular, we illustrate how NeXML can
+share and archive this. We illustrate how NeXML can
 be used to distribute both phylogenies and character trait data used in 
-many popular phylogenetic comparative methods tools in R.  (3) Annotate 
-and extend nexml. `RNeXML` provides convenient tools for adding metadata
+many popular phylogenetic comparative methods tools in R.  
+`RNeXML` also provides convenient tools for adding metadata
 to NeXML files, including the automated generation of metadata and examples
@@
 
-## Introduction to RNeXML
+<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
+and how RNeXML provides a solid platform from which to do so.  
+-->
+
+<!-- Map paragraph -->
+
+In this paper, we introduce the reader to the concepts and use of
+RNeXML and related technologies. We begin by discussing methods to read
+phylogenetic tree data and character trait data in the NeXML format into
+R, and also how to write phylogenetic data and character trait data into
+NeXML. The heart of this section focuses on tools for reading and writing
+the kind of semantically rich metadata that most sets NeXML apart from
+other phylogenetic data formats. Starting simple, we use examples of
+increasing complexity to illustrate concepts and tools from biodiversity
+informatics that allow users to express and intepret complicated data
+structures automatically.  We end with an example that uses metadata to
+extend the NeXML standard to describe stochastic character mapping, one
+of the kinds of data annotation that is currently available in R formats
+but lacks a consistent, extensible export format. We illustrate how NeXML
+files can be formally published to a data respository directly from R.
+Finally, we provide an brief description of how the software has been
+implemented, what quality control measures are in place, as well as
+information about software reuse and getting support.
+
+This paper has been written as a dynamic document 
+`r citep(citation("knitr")[[2]])`.  The reader may paste the code blocks
+shown into their R terminal to follow along with the exmples shown.
+
+# Introduction to RNeXML
 
@@
 
+## Getting Started
+
 The current stable version of the `RNeXML` package can be installed from
-the CRAN library using the standard installation method <!-- Actually not yet -->
+the CRAN library[^1].
+
+<!-- actually not yet, but before submission -->
 
 ```r
-install.packages("RNeXML")
-```
-
+install.packages("RNeXML", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
+```
+
+<!-- 
 The latest stable development version can be installed from Github using the `devtools` package:
@@
 library(devtools)
-install_github("RNeXML", "ropensci")
-```
-
-## Working with phylogenetic trees
-
-#### Read methods
+install_github("RNeXML", "ropensci", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
+```
+-->
+
+## Reading phylogenetic trees
 
@@
 
-We begin by reading in an example NeXML file provided with the package.
+We begin by reading in an example NeXML file provided with the package.[^2]
 
@@
 
-(For convenience `nexml_read` is also aliased as `read.nexml`, for
-users familiar with the somewhat analogous `read.nexus` function).
 The resulting `nex` object is an R object representation of the NeXML.  
+This is a native R object for representing phylogenetic trees (as well as 
+character traits, and metadata), analgous to the `phylo` object from the 
+popular `ape` R package `r citep(citation("ape"))`.
+
+While R functions and packages could be written to use this `nexml` object
+directly, most users will prefer to
+
 From this object we can extract any phylogenies it contains in the
@@
 phylogenetics community (including coercion into other representations,
-such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny
+such as `ouch` `r citep(citation("ouch"))` and `phylobase` `r citep(citation("phylobase"))` phylogeny
 formats). For instance, here we use the standard plotting function from
@@
 ```{r eval=FALSE}
-plot(phy)
+plot(phy[[1]])
 ```
@@
 
-Previously, this was possible using the `treebase` package [@treebase]
+Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`.
 for R, which relied on the NEXUS parser provided in the `ape` package.
-Due to inconsistencies present in the NEXUS standard [Vos _et al._
-2012], a significant number <!-- State explicitly --> of TreeBASE NEXUS
+Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
+a significant number <!-- State explicitly --> of TreeBASE NEXUS
 files could not be read in this way.  Using the more precisely defined
@@
 
-#### Character data
+## Reading character data
 
@@
 
-#### Write methods
+## Writing phylogenetic trees
 
@@
 ```{r}
-library(geiger)
+library("geiger")
 data(geospiza)
@@
 ```
+
+These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger"))`.
 
@@
 This allows us to include multiple trees and character matrices to a single NeXML file.
-<!-- We can add relevant metadata to each tree, such as the citation information 
-     indicating the paper in which it was first described.  -->
-
-### Metadata
+
+# Metadata
 
@@
 backwards compabitibilty with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations", or RDFa [W3C 2013],
-as described in [Vos et al 2011].  These annotations can be used to add 
+"Resource Description Framework in Annotations", or RDFa `r citep("http://www.w3.org/TR/xhtml-rdfa-primer")`,
+as described in @Vos_2012.  These annotations can be used to add 
 information to any particular element of the data, (a tree, node, edge,
@@
 
-### Writing NeXML metadata
+## Writing NeXML metadata
 
@@
 significant barrier to increased sharing and reuse of scientific research
-[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
+`r citep("Tenopir Data Sharing by Scientists: Practices and Perceptions")`.
+`RNeXML` seeks to reduce the burden of adding
 high quality, machine-readable metadata to NeXML files. This occurs on
@@
 
-- Automatic annotation.
+- Automatic annotation
+- Adding common metadata fields (`add_basic_metadata`)
+- Taxonomic identifiers (`taxize_nexml`)
+- Generic metadata (`add_meta`)
+- Extending the NeXML standard using RDFa
 
@@
 with deposition on Dryad data repository), which can be configured
-as needed.  To adjust these and other common metadata fields that should 
+as needed.
+
+To adjust these and other common metadata fields that should 
 be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
@@
 
-or to papers by simply using the object's DOI:
+or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`):
 
@@
 library("knitcitations")
-geiger_nex <- add_basic_meta(geiger_nex, citation=cite("10.2307/2408428"))
+geiger_nex <- add_basic_meta(geiger_nex, citation = bib_metadata("10.2307/2408428"))
 ```
@@
 (Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)
+
+## Taxonomic identifiers
 
@@
 other more specific metadata to various elements of the NeXML data.
-
-For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
-data are a common source of frustration in reusing phylogenetic trees. By 
-associating the taxonomic labels against a specific, centralized database
-we can catch typographical errors and indicate more precisely which classification
-we are appealing to in the event of conflicting taxonomic definitions.  In this 
-example, we indicate that we intend the bird orders listed in the tree to correspond
-to the NCBI definitions
+One common use case is that of taxonomic identifers.  Misspellings or
+inconsistent use of taxonomic names in phylogenetic data are a common
+source of frustration in reusing phylogenetic trees. By associating the
+taxonomic labels against a specific, centralized database we can catch
+typographical errors and indicate more precisely which classification
+we are appealing to in the event of conflicting taxonomic definitions.
+In this example, we indicate that we intend the bird orders listed in
+the tree to correspond to the NCBI definitions
 
@@
 
-This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
-label against the NCBI database.  If a unique match is found, a metadata annotation
-is added to the taxon providing the NCBI identification number to the taxonomic unit.
-If no match is found, the user is warned to check for possible typographic errors in
-the taxonomic labels provided.  <!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->
-
-### Custom metadata extensions
-
-So far our examples have relied on functions that have a built-in understanding of the 
-kind of metadata being provided, and can thus generate the appropriate metadata with 
-little further guidence.  In order to be machine-readable and understandable, 
-NeXML requires all metadata annotations follow the RDFa format which explicitly declares 
-the meaning of the annotation in a way a computer can understand.  For instance, the 
-title we gave to the `bird.orders` phylogeny appears in the NeXML as:
+This function uses the `taxize` R library `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
+If a unique match is found, a metadata annotation is added to the taxon
+providing the NCBI identification number to the taxonomic unit.  If no
+match is found, the user is warned to check for possible typographic
+errors in the taxonomic labels provided.
+
+<!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->
+
+## Custom metadata extensions
+
+So far our examples have relied on functions that have a built-in
+understanding of the kind of metadata being provided, and can thus
+generate the appropriate metadata with little further guidence.
+In order to be machine-readable and understandable, NeXML requires
+all metadata annotations follow the RDFa format (see @Vos_2012) which
+explicitly declares the meaning of the annotation in a way a computer
+can understand.  For instance, the title we gave to the `bird.orders`
+phylogeny appears in the NeXML as:
 
@@
 
-We see the title appears as the `content` of this `meta` element.  The `id` attribute
-provides us with a unique way to refer to this meta element in this document.  The `datatype`
-tells us that the content is a text string (as opposed to, say, a numerical or boolean value).  
-The `property` is the attribute that most interests us here; it tells us that the content 
-is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
-the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:
+We see the title appears as the `content` of this `meta` element.
+The `id` attribute provides us with a unique way to refer to this meta
+element in this document.  The `datatype` tells us that the content
+is a text string (as opposed to, say, a numerical or boolean value).
+The `property` is the attribute that most interests us here; it tells us
+that the content is a title.  The prefix `dc` indicates the vocabulary
+(called a namespace in XML) to which the word `title` belongs.  The prefix
+is defined elsewhere in the NeXML file, as we can see with:
 
@@
 
-This URL links us to a webpage providing both a human and machine-readable version of precisely
-what the term `title` means, showing us that it refers to `title` in the sense of a creative work, 
-rather than the title of a person, a legal title, the verb title, etc.  This precision can allow
-us to perform precise computer reasoning on metadata, without relying on heuristics or assumptions,
-as we illustrate later.  For the moment though, the need to refer to an explicit controlled vocabulary (or
-ontology) with a url, saying "dc:title" rather than just saying the property is "title", is something
-of a burden in creating metadata.  The previous examples involve functions that already know what 
-kind of metadata is being provided (a title, a data creator, etc), and so can provide the appropriate
-vocabulary automatically.  The main advantage of linking to an external vocabulary, however, is the
-ability to express concepts that are not already "built in" to the data model.  `RNeXML` provides 
-tools to generate arbitrary metadata annotations given an external vocabulary, such as we illustrate
-here.
-
-From the list of namespaces in `prefixes`, we can follow links a handful of established vocabularies
-and find terms that are not included in `add_basic_meta`.  For instance, in the `prism` prefix,
+This URL links us to a webpage providing both a human and machine-readable
+version of precisely what the term `title` means, showing us that it
+refers to `title` in the sense of a creative work, rather than the title
+of a person, a legal title, the verb title, etc.  This precision can allow
+us to perform precise computer reasoning on metadata, without relying on
+heuristics or assumptions, as we illustrate later.  For the moment though,
+the need to refer to an explicit controlled vocabulary (or ontology)
+with a url, saying "dc:title" rather than just saying the property is
+"title", is something of a burden in creating metadata.  The previous
+examples involve functions that already know what kind of metadata is
+being provided (a title, a data creator, etc), and so can provide the
+appropriate vocabulary automatically.  The main advantage of linking to
+an external vocabulary, however, is the ability to express concepts that
+are not already "built in" to the data model.  `RNeXML` provides tools
+to generate arbitrary metadata annotations given an external vocabulary,
+such as we illustrate here.
+
+From the list of namespaces in `prefixes`, we can follow links a handful
+of established vocabularies and find terms that are not included in
+`add_basic_meta`.  For instance, in the `prism` prefix,
 
@@
 
-we see that there is a term for `modificationDate` for a file.  We create a `meta` element 
-containing this annotation using the `meta` function:
+we see that there is a term for `modificationDate` for a file.  We create
+a `meta` element containing this annotation using the `meta` function:
 
@@
 
-We can add this annotation to our existing `birds` nexml file using the `add_meta` function.
-Because we do not specify a level, it is added to the root node, refering to the nexml file as a whole.
+We can add this annotation to our existing `birds` nexml file using the
+`add_meta` function.  Because we do not specify a level, it is added to
+the root node, refering to the nexml file as a whole.
 
@@
 
-The built-in vocabularies are just the tip of the iceberg of established vocabularies. Here 
-we add an annotation from the `skos` namespace which describes the history of where the
-data comes from:
+The built-in vocabularies are just the tip of the iceberg of established
+vocabularies. Here we add an annotation from the `skos` namespace which
+describes the history of where the data comes from:
 
@@
 
-Because `skos` is not in the current namespace list, we add it with a url when
-adding this meta element.  We also specify that this annotation be placed 
-at the level of the `trees` subnode in the NeXML file.
+Because `skos` is not in the current namespace list, we add it with a
+url when adding this meta element.  We also specify that this annotation
+be placed at the level of the `trees` subnode in the NeXML file.
 
@@
 
-If we forget to add the namespace with add_meta, we can always add it later
-using the `add_namespaces` function,
+If we forget to add the namespace with add_meta, we can always add it
+later using the `add_namespaces` function,
 
@@
 
-(The `add_namespaces` function will avoid creating any duplicate definitions)
-For finer control of the level at which a `meta` element is added, we will manipulate 
-the nexml R object directly using S4 subsetting, as shown later.
-
-<!-- Ugh. y'all have much more practice explaining this well and without tangents, 
+Note that the `add_namespaces` function will automatically avoid creating
+any duplicate definitions.
+
+For finer control of the level at which a `meta` element is added,
+we will manipulate the nexml R object directly using S4 subsetting,
+as shown later.
+
+<!--FIXME Ugh. y'all have much more practice explaining this well and without tangents, 
 so please feel free to edit or re-write.  -->
 
-Much richer metadata annotation is possible. Later we illustrate how metadata annotation
-can be used to extend the base NeXML format to represent new forms of data while maintaining
-compatability with any NeXML parser. The `RNeXML` package can be easily extended to support
-helper functions such as `taxize_nexml` to add additional metadata without imposing a large
-burden on the user.
-
-<!--
-- Utilities to add common metadata
-
-  - Add a citation to any papers associated with the phylogeny data.  
-  - Writing in methodological information on how tree has been generated or analyzed (including R code)
-  - Add annotations to any taxon.
-
-Temporal, spatial, taxonomic coverage for dryad/knb?
--->
-
-### Reading NeXML metadata
-
-A call to the nexml object prints some metadata summarizing the data structure: <!-- need to make this method faster!-->
+Much richer metadata annotation is possible. Later we illustrate how
+metadata annotation can be used to extend the base NeXML format to
+represent new forms of data while maintaining compatability with any
+NeXML parser. The `RNeXML` package can be easily extended to support
+helper functions such as `taxize_nexml` to add additional metadata
+without imposing a large burden on the user.
+
+## Reading NeXML metadata
+
+A call to the nexml object prints some metadata summarizing the data structure:
 
@@
 ```{r}
-library(XML)
+library("XML")
 xpathSApply(rdf, "//dc:title", xmlValue) 
@@
 as RDF.  Queries using the XPath syntax are possible in R using the
-XML library [Temple Lang 2013, XML].  (Note that XPath is far more
+XML library `r citep(citation("XML"))`.  (Note that XPath is far more
 expressive for querying the metadata once the RDFa has been extracted,
@@
 formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
+queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the sytnax
 necessary to apply such queries to NeXML files from R.  First we load
-the necessary R library and import the RDF-extracted metadata:
-
-```{r}
-library(rrdf)
+the necessary R library `r citep(citation("rrdf"))` and import the RDF-extracted metadata:
+
+```{r}
+library("rrdf")
 saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
@@
 
-### Extending the NeXML standard through metadata annotation.
+## Extending the NeXML standard through metadata annotation.
 
@@
 Here we illustrate this process using the example of stochastic character
-mapping [Huelsenbeck et al 2003]. A stochastic character map is simply
+mapping `r citep("Huelsenbeck stochastic character mapping")`. A stochastic character map is simply
 an annotation of the branches on a phylogeny, assigning each section
@@
 
-[Bollback 2006] provides a widely used stand-alone software implementation
+`r citet("Bollback simmap 2006")` provides a widely used stand-alone software implementation
 of this method in the software `simmap`, which modified the standard
 Newick tree format to express this additional information. This can
-break compatibility with other software [^1], and creates a format that
+break compatibility with other software [^3], and creates a format that
 cannot be interpreted without additional information describing this
@@
 
-[Revell 2011] describes the `phytools` package for R, which includes
+`r citet("Revell phytools 2011")` describes the `phytools` package for R, which includes
 utilities for reading, manipulating, and writing `simmap` files in R.
@@
 map the R representation used by Revell (an extension of the `ape` class)
-into the NeXML extension we have defined by using `RNeXML` functions..
+into the NeXML extension we have defined by using `RNeXML` functions.
 
@@
 We can then use various functions from `phytools` designed for simmap
-objects, such as the plotting function:
+objects `r citep(citation("phytools"))`, such as the plotting function:
 
 ```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
-library(phytools)
+library("phytools")
 plotSimmap(phy)
@@
 
+<!--
 [Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
@@
 [Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
-
-[^1]: By using the commenting mechanism of the Newick format, it is
-possible that other software that doesn't also use the comment mechanism
-for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the
-data format to describe its use.
-
-### Subsetting nexml objects directly
+-->
+
+<!-- Consider moving to the supplement? -->
+
+## Understanding the `nexml` S4 object
 
@@
 
-### Publishing NeXML files from R
+## Publishing NeXML files from R
 
 Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity (e.g. [Whitlock2011]).  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier ([Tenopir et al. 2011], [Stodden 2010]),
+biodiversity (e.g. `r citet("Whitlock data archiving Evolution 2010")`).  The burden of preparing and submitting properly
+annotated data to archives continues to be a significant barrier (@Tenopir_2011, `r citet("10.2139/ssrn.1550193")`),
 and many phylogenetic trees upon which studies are based are inaccessible or lost 
-to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
+to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
 by providing functions to immediately archive NeXML files in scientific repositories.
@@
 
-### Implementation and architecture
+# Implementation, architecture and quality control
 
@@
 using the R software environment and the XML package for parsing and
-generating XML [Temple Lang 2013, XML].
+generating XML `r citep(citation("XML"))`,  provides conversions into
+the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
+`RNeXML` can automatically check taxonomic names and add taxonomic
+identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
+Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
+and internet interfaces by the `r citep(citation("httr"))` package. The
+packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .
 
 <!--  Find more consise way to make sure dependencies are cited. 
-      How about citing packages that are only on the suggests list?
-
-and provides conversions into
-the popular `phylo` format provided by the `ape` package [Paradis 2004].
-`RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package
-([Chamberlain 2013]).
+      How about citing packages that are only on the suggests list? --> 
 --> 
-
 <!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
@@
 
-### Quality control
-
-The `RNeXML` package performs severak levels of validation on NeXML files
+The `RNeXML` package performs several levels of validation on NeXML files
 written from or read into R: validating the XML, and validating against
@@
 mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult [Vos et al. 2011].
+interoperability difficult [@Vos_2012].
 
@@
 the `RNeXML` package includes a complete test suite using the `testthat`
-package [Wickham 2011]. Package development on Github includes continuous 
+package `r citep(citation("testthat"))`. Package development on Github includes continuous 
 integration through the `travis` platform, in which the full test suite
@@
 
-### Software reuse and support
+# Software reuse and support
 
@@
 to submit bug reports or feature requests in the issues log.  Contact
-ropensci-discuss@googlegroups.com or the package maintainer for help.
+`ropensci-discuss@googlegroups.com` or the package maintainer for help,
+or email the phylogenetics R users group list at `r-sig-phylo@r-project.org`.
 
 There are many promising directions for further development of the
-`RNeXML` package.  The project's [Milestones page][milestones] page provides a list of
+`RNeXML` package.  The project's [Milestones page][https://github.com/ropensci/RNeXML/issues/milestones] page provides a list of
 steps currently planned or under consideration.  In particular, these
@@
 
-### Acknowledgements
+## Acknowledgements
 
 This project was supported in part by a grant from the Alfred P Sloan
-Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part
-of the rOpenSci project, [http://ropensci.org].
-
-[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
-[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
-[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
-[@ouch]: http://cran.r-project.org/web/packages/ouch/
-[@phylobase]: http://cran.r-project.org/web/packages/phylobase
-[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
-[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
-[Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636 
-[milestones]: https://github.com/ropensci/RNeXML/issues/milestones
+Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... <!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->
+
+`RNeXML` is part of the rOpenSci project, [http://ropensci.org].
+
+```{r cleanup, include=FALSE, cache=FALSE}
+write.bibtex(file="components/references.bib")
+unlink("simmap.xml")
+unlink("birds.xml")
+unlink("geospiza.xml")
+unlink("rdf_meta.xml")
+```
+
+<!-- Footnotes -->
+
+[^1]: By providing the optional `dependencies` argument, R will install
+additional packages used in some of the advanced examples shown later
+but which are not required for basic functionality.
+
+[^2]: For convenience `nexml_read` is also aliased as `read.nexml`,
+for users familiar with the somewhat analogous `read.nexus` function.
+
+[^3]: By using the commenting mechanism of the Newick format, it is
+possible that other software that doesn't also use the comment mechanism
+for some other such purpose would be able to successfully parse the tree.
+However there is no way to guarentee that this is the case or for the
+data format to describe its use.
+
+<!-- MISC / SCRATCH
+
+What to say about existing formats?  
+A summary of some phylogenetic formats in R:
+
+| Class     | R package   | R Object 
+|-----------|-------------|-------
+| phylo     | ape         | S3 
+| phylo4    | phylobase   | S4
+| phylo4d   | adephylo    | S4
+| ouchtree  | ouch        | S4 
+| treeshape | apTreeshape | list
+
+See [phylobase vignette, 2010]
 
@@
 conflicting specifications).
+
+-->

### 2014-06-17 manuscripts/manuscript.Rmd
remove extraneous citations

@@
 We can then leverage the rich suite of methods available from the R
-phylogenetics community (including coercion into other representations,
-such as `ouch` `r citep(citation("ouch"))` and `phylobase` `r citep(citation("phylobase"))` phylogeny
-formats). For instance, here we use the standard plotting function from
+phylogenetics community. For instance, here we use the standard plotting function from
 the `ape` package to plot the resulting `ape::phylo` object returned by

### 2014-06-17 manuscripts/manuscript.Rmd
fix more citations

@@
 
-These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger"))`.
+These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger")[[4]])`.
 
@@
 backwards compabitibilty with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations", or RDFa `r citep("http://www.w3.org/TR/xhtml-rdfa-primer")`,
+"Resource Description Framework in Annotations", or RDFa,
 as described in @Vos_2012.  These annotations can be used to add 
@@
 
-`r citet("Revell phytools 2011")` describes the `phytools` package for R, which includes
+`r citet(citation("phytools"))` describes the `phytools` package for R, which includes
 utilities for reading, manipulating, and writing `simmap` files in R.

### 2014-06-23 manuscripts/manuscript.Rmd
added citation to Stolzfus et al (#58)

@@
 to sharing this information both between software environments and 
-between research groups `r citet("Drew PLOS Biology Tree of Life 2013")`.
+between research groups `r citep(c("Drew PLOS Biology Tree of Life 2013", "10.1186/1756-0500-5-574"))`.
 

### 2014-06-24 manuscripts/manuscript.Rmd
removed fluff about "the next generation of phylogenetic informatics data"

--- 2014-06-17 manuscripts/manuscript.Rmd
@@
 abstract: | 
-      1. NeXML promises to be the next generation of phylogenetic
-      informatics data and metadata exchange formats, offering substantial
-      improvements in reliability, extensibility, and richness over
-      earlier standards such as NEXUS. In this paper we present the RNeXML
-      package which provides a bridge between this rich and reliable
+      1. NeXML is a new phylogenetic informatics data and metadata exchange 
+      format that offers substantial improvements in reliability, extensibility, 
+      and richness over earlier standards such as NEXUS. In this paper we present 
+      the RNeXML package which provides a bridge between this rich and reliable
       data format and the extensive library of phylogenetic tools already

### 2014-06-24 manuscripts/manuscript.Rmd
clarified meaning of "semantic annotations"

@@
       how the package facilitates adding and manipulating semantic 
-      annotations.
+      annotations, that is annotations whose meaning can be defined externally
+      by way of semantic web technologies.
 

### 2014-06-24 manuscripts/manuscript.Rmd
added the use case of interacting with NeXML-based web services

@@
       personal archives of phylogenetic and comparative trait data in consistent,
-      query-able, and cross-platform format, or leverage the rapidly growing 
-      public community archives of rich NeXML files.
+      query-able, and cross-platform format, to leverage the rapidly growing 
+      public community archives of rich NeXML files, and to interact with web
+      services that deliver their results (with their provenance) as NeXML.
 

### 2014-06-24 manuscripts/manuscript.Rmd
added another thing to add to the intro

@@
 - More on data sharing challenges and opportunities a la @Drew_2013
+- More on the hoops that users need to jump through if trees, data
+and other things all need to come from separate files and need to be
+joined on taxon labels.
 

### 2014-06-24 manuscripts/manuscript.Rmd
changed in an attempt to draw the distinction between community conventions for flat text, and standards that have been described in technical recommendations under the auspices of the W3C (i.e. "web standards")

@@
 of their own software, which results in mutually incompatible dialects
-that largely defeat the advantages of a shared standard file format.
+that largely defeat the advantages of a shared  file format.
 NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
-by drawing on state-of-the-art data management technology.
+by drawing on robust web standards in data representation.
 

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix

@@
 easily be shared, archived, and made available to other software;
-but inherits the limiations of the NEXUS format.
+but inherits the limitations of the NEXUS format.
 

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix

@@
 growing data sets in the NeXML format to the R user and the large ecosystem
-of R packages currently avaliable for handling phylogenetic data  
+of R packages currently available for handling phylogenetic data  
 R packages and R users frequently add rich information to the phylogenies

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix

@@
 of how to extend the NeXML standard to represent new data features without
-breaking validation or backwards compatability of the file type.
+breaking validation or backwards compatibility of the file type.
 

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix

@@
 increasing complexity to illustrate concepts and tools from biodiversity
-informatics that allow users to express and intepret complicated data
+informatics that allow users to express and interpret complicated data
 structures automatically.  We end with an example that uses metadata to

### 2014-06-24 manuscripts/manuscript.Rmd
prefer to use "document" instead of "file", as NeXML data can be obtained from a handle that reads from a remote location, i.e. the data never reside as a file in the system
although this is denormalized into just trees, data and metadata, NeXML documents have four components, not three.
typo fix, re-flow of paragraph

@@
 
-A NeXML file can be thought of as having three components: (1) phylogeny
-data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
-provide a seemless way to convert these to and from their corresponding
-R object representations, such as `ape::phylo` phylogenies, character
+A NeXML document can be thought of as having multiple components: (1) phylogeny
+data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
+`RNeXML` seeks to provide a seemless way to convert these to and from their 
+corresponding R object representations, such as `ape::phylo` phylogenies, character
 matrices (`data.frames`) and metadata lists (or for advanced users,
@@
 
-The resulting `nex` object is an R object representation of the NeXML.  
-This is a native R object for representing phylogenetic trees (as well as 
-character traits, and metadata), analgous to the `phylo` object from the 
+The resulting `nex` object is an R object representation of the NeXML. This is 
+a native R object for representing phylogenetic trees (as well as 
+character traits, and metadata), analogous to the `phylo` object from the 
 popular `ape` R package `r citep(citation("ape"))`.

### 2014-06-24 manuscripts/manuscript.Rmd
completed sentence, re-flowed paragraph

@@
 character traits, and metadata), analogous to the `phylo` object from the 
-popular `ape` R package `r citep(citation("ape"))`.
-
-While R functions and packages could be written to use this `nexml` object
-directly, most users will prefer to
-
-From this object we can extract any phylogenies it contains in the
-`ape::phylo` format:
+popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
+could be written to use this `nexml` object directly, most users will prefer to use
+already existing methods provided by packages they know. From `nex` object we can 
+therefore extract any phylogenies it contains in the `ape::phylo` format:
 

### 2014-06-24 manuscripts/manuscript.Rmd
s/file/document/
added clarification about multiple blocks
typo fix

@@
 
-NeXML can not only contain multiple phylogenetic trees in a single file,
-but can also group sets of trees into separate blocks.  The `get_trees`
-function attempts to return the most sensible R object for NeXML: if
-only a single tree is available, it will return an `ape::phylo` object;
-if multiple trees are available in a single `trees` block, it will return
+NeXML can not only contain multiple phylogenetic trees in a single document,
+but can also group sets of trees into separate blocks (e.g. as resulting from
+different workflows). The `get_trees` function attempts to return the most sensible 
+R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
+object; if multiple trees are available in a single `trees` block, it will return
 an `ape::multiPhylo` object (a list of `phylo` objects).  If multiple
@@
 (such as looping over many NeXML files) when having a variable return
-type is undisirable.  To always get back a list of `multiPhylo` objects,
+type is undesirable.  To always get back a list of `multiPhylo` objects,
 even when only one tree is available, use `get_trees_list()`.

### 2014-06-24 manuscripts/manuscript.Rmd
corrected capitalization

@@
 <!-- have we already said all this above?-->  <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
-phylogenetic trees in the `nexml` format opens up a wide and rapidly
+phylogenetic trees in the NeXML format opens up a wide and rapidly
 growing array of resources to the R user.  For instance, a user can access

### 2014-06-24 manuscripts/manuscript.Rmd
corrected capitalization
s/file/document/
recorded my doubts about having aliased method names

@@
 
+<!-- I still question the wisdom of having multiple names for the same
+method. At first this might seem user-friendly, but it will just be confusing
+and create a document and source code maintenance nightmare. Why not just pick
+read.nexml and be done with it? -->
 We begin by reading in an example NeXML file provided with the package.[^2]
@@
 files could not be read in this way.  Using the more precisely defined
-NEXML standard, every valid NeXML file on TreeBASE can now successfully
+NeXML standard, every valid NeXML document on TreeBASE can now successfully
 be parsed to R objects.

### 2014-06-24 manuscripts/manuscript.Rmd
s/file/document/, also added parenthetical statement about using morphological data to infer trees

@@
 
-Like NEXUS files, NeXML files can also contain character data.  While
+Like NEXUS files, NeXML documents can also contain character data.  While
 most researchers will be familiar with the use of character matrices in
-NEXUS files as a container for molecular sequence data used to infer
-the phylogenetic tree, we find this ideal for character data used in
+NEXUS files as a container for molecular sequence (or morphological) data 
+used to infer the phylogenetic tree, we find this ideal for character data used in
 comparative methods popular in R. Methods to get extract character data

### 2014-06-24 manuscripts/manuscript.Rmd
removed redundant 'get'
typo fix

@@
 used to infer the phylogenetic tree, we find this ideal for character data used in
-comparative methods popular in R. Methods to get extract character data
+comparative methods popular in R. Methods to extract character data
 work much like those for the phylogenetic data:
@@
 single data frame by `get_characters()` if they correspond to the same
-taxa. The `get_characters_list()` function will consistently recieve
+taxa. The `get_characters_list()` function will consistently receive
 a list of data frames corresponding to each character matrix block of

### 2014-06-24 manuscripts/manuscript.Rmd
one word
added comment about caper's comparative.data method

@@
 
-Generating NeXML files from R is likewise straight forward.  Here we
+Generating NeXML files from R is likewise straightforward.  Here we
 write a phylogeny from the `ape::phylo` format out to a NeXML file:
@@
 
+<!-- in caper there is this method comparative.data to join data
+and trees on taxon labels. it would be nice if this merged data
+structure could also be passed to nexml_write (or should that be
+write.nexml?) -->
 The first argument to `nexml_write` can be a `phylo` phylogeny,

### 2014-06-24 manuscripts/manuscript.Rmd
added comment about the need, perhaps, for an add_taxa method
typo fix

@@
 
+<!-- how about add_taxa? -->
 NeXML files can also be built up bit by bit using the `add_characters`,
@@
 metadata in precisely specified, machine-readable formats that maintain
-backwards compabitibilty with any existing NeXML parser. NeXML uses the
+backwards compatibility with any existing NeXML parser. NeXML uses the
 "Resource Description Framework in Annotations", or RDFa,

### 2014-06-24 manuscripts/manuscript.Rmd
it might be better to cite the Technical Recommendation about RDFa from the WWW consortium, to indicate that this is not something we pulled out of thin air

@@
 "Resource Description Framework in Annotations", or RDFa,
-as described in @Vos_2012.  These annotations can be used to add 
+as described in @Vos_2012. <!-- better to cite the W3C's RDFa TR -->
+These annotations can be used to add 
 information to any particular element of the data, (a tree, node, edge,

### 2014-06-24 manuscripts/manuscript.Rmd
note the possibility of using the user name (env var $USER) as an annotation

@@
 
+<!-- in the Perl library I also embed the value of the environment variable
+$USER as a document-level annotation -->
 `RNeXML` automatically adds basic metadata by default such as a timestamp

### 2014-06-24 manuscripts/manuscript.Rmd
removed redundant 'also'
expanded a bit more on the problems with names-based data integration

@@
 
-`RNeXML` also provides various functions to assist in automatically adding
+`RNeXML` provides various functions to assist in automatically adding
 other more specific metadata to various elements of the NeXML data.
@@
 inconsistent use of taxonomic names in phylogenetic data are a common
-source of frustration in reusing phylogenetic trees. By associating the
+source of frustration in reusing phylogenetic trees, especially when 
+integrating data from different sources: trees might come from one place
+that uses one taxonomy, character data and occurrence data might come from
+other places that use different taxonomies. Hence, not just misspellings, 
+but also synonyms, homonyms and incompatible taxon concepts might plague
+the end user. By associating the
 taxonomic labels against a specific, centralized database we can catch
-typographical errors and indicate more precisely which classification
+some of these inconsistencies and indicate more precisely which classification
 we are appealing to in the event of conflicting taxonomic definitions.

### 2014-06-24 manuscripts/manuscript.Rmd
clarified that 'content' is the name of an attribute, not the text data that is subtended by the focal DOM node
clarified that CURIE prefixes are not the same as XML namespace prefixes

@@
 
-We see the title appears as the `content` of this `meta` element.
+We see the title appears as the value of the `content` attribute of this `meta` element.
 The `id` attribute provides us with a unique way to refer to this meta
@@
 that the content is a title.  The prefix `dc` indicates the vocabulary
-(called a namespace in XML) to which the word `title` belongs.  The prefix
+(here identified as an XML namespace) to which the word `title` belongs.  The prefix
 is defined elsewhere in the NeXML file, as we can see with:

### 2014-06-24 manuscripts/manuscript.Rmd
formulated a bit more carefully: the prefix identifies a namespace, which here has a URL location (e.g. it's not an URN), which we can visit so that we may learn about the modificationDate predicate
request for clarification: can we use undefined prefixes?

@@
 of established vocabularies and find terms that are not included in
-`add_basic_meta`.  For instance, in the `prism` prefix,
+`add_basic_meta`.  For instance, by resolving the location identified
+by the `prism` prefix,
 
@@
 
-we see that there is a term for `modificationDate` for a file.  We create
+we will find that there is a term for `modificationDate` for a file.  We create
 a `meta` element containing this annotation using the `meta` function:
@@
 
+<!-- does this mean we can use CURIEs whose prefix has not yet been 
+bound to a namespace? that would be bad. there should probably be an
+error message when that happens -->
 If we forget to add the namespace with add_meta, we can always add it

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix

@@
 metadata annotation can be used to extend the base NeXML format to
-represent new forms of data while maintaining compatability with any
+represent new forms of data while maintaining compatibility with any
 NeXML parser. The `RNeXML` package can be easily extended to support

### 2014-06-24 manuscripts/manuscript.Rmd
typo fix
didn't we already cover dc:title?

@@
 and can ask for a particular element using the standard list subsetting
-mechanism (i.e. either the name of an element or it's numeric position),
+mechanism (i.e. either the name of an element or its numeric position),
 
@@
 
+<!-- I feel like we already covered what dc:title is -->
 Following the link we can see that `dc` refers to the "Dublin Core"

### 2014-06-24 manuscripts/manuscript.Rmd
note formatting error: the citations run off the page of the PDF
maybe not go into XPath applied to RDF?

@@
 
+<!-- the citations run off the page -->
 Common metadata can be accessed with a few dedicated functions:
@@
 
+<!-- I think it is unwise to go here: RDF is not structured data -->
 Query the rdf using XPath expressions:

### 2014-06-26 manuscripts/manuscript.Rmd
typo fix
added SPARQL comment

@@
 formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the sytnax
+queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the syntax
 necessary to apply such queries to NeXML files from R.  First we load
@@
 
+<!-- it would be nice if we had a more persuasive example here. directly getting the value
+of some predicate can be done the syntax-oriented way by treating the data as XML. what I
+think we should be going for here is an example of a slightly more complex query that would
+be painful to perform by scanning the XML document. I'm thinking of some sort of join of
+multiple predicate values. Nothing great comes to mind yet, though... ->
 We can then make arbitrary SPARQL queries directly from R.  For instance,

### 2014-06-26 manuscripts/manuscript.Rmd
typo fix

@@
 may one day need. Researchers have repeatedly adapted the existing NEXUS
-file format to accomidate encoding new kinds of information, but in the
+file format to accommodate encoding new kinds of information, but in the
 process break compatibility between different flavors of NEXUS files, with

### 2014-06-26 manuscripts/manuscript.Rmd
typo fix
typo fixes

@@
 
-NeXML already defines the `characters` element to handle descrete character traits (`nex:char`)
+NeXML already defines the `characters` element to handle discrete character traits (`nex:char`)
 and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
@@
 
-The `nexml` object is an S4 object, as are all of its components (slots).  It's 
-heirachical structure corresponds exactly with the XML tree of a NeXML file, with 
+The `nexml` object is an S4 object, as are all of its components (slots).  Its 
+hierarchical structure corresponds exactly with the XML tree of a NeXML file, with 
 the single exception that both XML attributes and children are represented as slots.  

### 2014-06-26 manuscripts/manuscript.Rmd
included httr package name before citation

@@
 Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
-and internet interfaces by the `r citep(citation("httr"))` package. The
+and internet interfaces by the `httr` package `r citep(citation("httr"))`. The
 packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .

### 2014-06-26 manuscripts/manuscript.Rmd
re-flow paragraph
request for clarification
removed duplicate comment delimiter

@@
 though stand-alone programs remain an important part of this ecosystem
-for tasks such as generating phylogenetic trees from sequence data.  
-Within this ecosystem a variety of different data formats for
+for tasks such as generating phylogenetic trees from sequence data.  Within 
+this ecosystem a variety of different data formats for
 phylogenetic data have been developed and modified. Most but not all 
@@
 any duplicate definitions. 
-
 For finer control of the level at which a `meta` element is added,
@@
 so please feel free to edit or re-write.  -->
+<!-- I am not quite sure what points we are trying to make here -->
 
@@
       How about citing packages that are only on the suggests list? --> 
---> 
 <!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 

### 2014-07-08 manuscripts/manuscript.Rmd
create supplement.Rmd - moves implementation details and S4 subsetting to supplement - adds a render call to the Makefile to make supplement pdf - removes XPath query on RDF

@@
 ---
-layout: preprint
-#layout: review, 11pt
+#layout: preprint
+layout: review, 11pt
 title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
@@
 
-<!-- I think it is unwise to go here: RDF is not structured data -->
-Query the rdf using XPath expressions:
-
-```{r}
-library("XML")
-xpathSApply(rdf, "//dc:title", xmlValue) 
-```
-
-XPath is an expressive syntax for querying XML-structured data such
-as RDF.  Queries using the XPath syntax are possible in R using the
-XML library `r citep(citation("XML"))`.  (Note that XPath is far more
-expressive for querying the metadata once the RDFa has been extracted,
-since attribute values are now interpreted as namespaced XML nodes and
-not just arbitrary text.)
-
 The most powerful way to take advantage of the machine-readable nature
@@
 
-<!-- it would be nice if we had a more persuasive example here. directly getting the value
+<!-- FIXME it would be nice if we had a more persuasive example here. directly getting the value
 of some predicate can be done the syntax-oriented way by treating the data as XML. what I
@@
 multiple predicate values. Nothing great comes to mind yet, though... ->
+
 We can then make arbitrary SPARQL queries directly from R.  For instance,
@@
 <!-- I don't see how this could be improved much without forcing people
-to fire op protege (let's not!). As it happens, the Dublin Core URI, 
+to fire up protege (let's not!). As it happens, the Dublin Core URI, 
 for example, points to an HTML page so even in that case automatic 
@@
 
-<!--
-[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
-[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
-[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
--->
-
-<!-- Consider moving to the supplement? -->
-
-## Understanding the `nexml` S4 object
-
-The `RNeXML` package provides many convenient functions to add and extract
-information from `nexml` objects in the R environment without requiring
-the reader to understand the details of the NeXML data structure and
-making it less likely that a user will generate invalid NeXML syntax
-that could not be read by other parsers. The `nexml` object we have been using
-in all of the examples is built on R's S4 mechanism. Advanced users may
-sometimes prefer to interact with the data structure more directly using 
-R's S4 class mechanism and subsetting methods. Many R users are more familiar
-with the S3 class mechanism (such as in the `ape` package phylo objects)
-rather than the S4 class mechanism used in phylogenetics packages such as 
-`ouch` and `phylobase`. The `phylobase` vignette provides an excellent introduction
-to these data structures.  Users already familiar with subsetting lists and other
-S3 objects in R are likely familar with the use of the `$` operator, such as
-`phy$edge`. S4 objects simply use an `@` operator instead (but cannot be subset
-using numeric arguments such as `phy[[1]]` or named arguments such as phy[["edge"]]).
-
-The `nexml` object is an S4 object, as are all of its components (slots).  Its 
-hierarchical structure corresponds exactly with the XML tree of a NeXML file, with 
-the single exception that both XML attributes and children are represented as slots.  
-S4 objects have constructor functions to initialize them.  We create a new `nexml` 
-object with the command:
-
-```{r}
-nex <- new("nexml")
-```
-
-We can see a list of slots contained in this object with
-
-```{r}
-slotNames(nex)
-```
-
-Some of these slots have already been populated for us, for instance, the schema version and default namespaces:
-
-```{r}
-nex@version
-nex@namespaces
-```
-
-Recognize that `nex@namespaces` serves the same role as `get_namespaces` function, but provides direct access to the slot data.
-For instance, with this syntax we could also overwrite the existing namespaces with `nex@namespaces <- NULL`.
-Changing the namespace in this way is not advised.
-
-Some slots can contain multiple elements of the same type, such as `trees`, `characters`, and `otus`.  For instance, we see that
-
-```{r}
-class(nex@characters)
-```
-
-is an object of class `ListOfcharacters`, and is currently empty,
-
-```{r}
-length(nex@characters)
-```
-
-In order to assign an object to a slot, it must match the class definition
-of the slot.  We can create a new element of any given class with the
-`new` function,
-
-```{r}
-nex@characters <- new("ListOfcharacters", list(new("characters")))
-```
-
-and now we have a length-1 list of character matrices,
-
-```{r}
-length(nex@characters)
-```
-
-and we access the first character matrix using the list notation, `[[1]]`. Here we check the class is a `characters` object.
-
-```{r}
-class(nex@characters[[1]])
-```
-
-Direct subsetting has two primary use cases: (a) useful in looking up (and possibly editing) a specific value of an element, or (b) when adding metadata annotations to specific elements.
-Consider the example file
-
-```{r}
-f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
-```
-
-We can look up the species label of the first `otu` in the first `otus` block:
-
-```{r}
-nex@otus[[1]]@otu[[1]]@label
-```
-
-We can add metadata to this particular OTU using this subsetting format
-
-```{r}
-nex@otus[[1]]@otu[[1]]@meta <- 
-  c(meta("skos:note", 
-          "This species was incorrectly identified"),
-         nex@otus[[1]]@otu[[1]]@meta)
-```
-
-Here we use the `c` operator to append this element to any existing meta annotations to this otu.
-
 ## Publishing NeXML files from R
@@
 <!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->
-
-# Implementation, architecture and quality control
-
-The `RNeXML` package is built for the NeXML Schema, [version 2009],
-using the R software environment and the XML package for parsing and
-generating XML `r citep(citation("XML"))`,  provides conversions into
-the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
-`RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
-Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
-and internet interfaces by the `httr` package `r citep(citation("httr"))`. The
-packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .
-
-<!--  Find more consise way to make sure dependencies are cited. 
-      How about citing packages that are only on the suggests list? --> 
-<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
-<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->
-
-The `RNeXML` package performs several levels of validation on NeXML files
-written from or read into R: validating the XML, and validating against
-the current NeXML schema, and performing additional checks through the 
-NeXML online validation tool for elements that cannot be enforced by
-schema validation alone.  Validation serves to improve consistency in
-data encoding which facilitates the use of NeXML across a wide variety
-of platforms and software.  A lack of a consistent standard and validation
-mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult [@Vos_2012].
-
-Several mechanisms are in place to evaluate the quality of the `RNeXML`
-package functions themselves.  In addition to the examples from the
-documentation and the automated package checks provided for the R system,
-the `RNeXML` package includes a complete test suite using the `testthat`
-package `r citep(citation("testthat"))`. Package development on Github includes continuous 
-integration through the `travis` platform, in which the full test suite
-is run every time the code is changed.  This not only gives users confidence
-that the development copy can be successfully installed, but should assist
-other developers seeking to contribute to the package to test that their proposed 
-changes have not negatively impacted any of the existing code base.
 
@@
 data format to describe its use.
-
-<!-- MISC / SCRATCH
-
-What to say about existing formats?  
-A summary of some phylogenetic formats in R:
-
-| Class     | R package   | R Object 
-|-----------|-------------|-------
-| phylo     | ape         | S3 
-| phylo4    | phylobase   | S4
-| phylo4d   | adephylo    | S4
-| ouchtree  | ouch        | S4 
-| treeshape | apTreeshape | list
-
-See [phylobase vignette, 2010]
-
-phylobase::phylo4 was an attempt to solve the problems associated with
-the popular ape::phylo format (unpredictable errors, loose and multiple
-conflicting specifications).
-
--->

### 2014-07-08 manuscripts/supplement.Rmd
create supplement.Rmd - moves implementation details and S4 subsetting to supplement - adds a render call to the Makefile to make supplement pdf - removes XPath query on RDF

--- 2014-04-28 inst/doc/pubs/manuscript.Rmd
@@
 ---
-layout: preprint
-#layout: review, 11pt
-title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
+layout: review, 11pt
+title: "Supplementary material for -- RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
 author: 
@@
     address: | 
-      NBC Naturalis, Leiden, the Netherlands
-abstract: |
+      Naturalis Biodiversity Center, Leiden, the Netherlands
 
-      1. NeXML promises to be the next generation of phylogenetic
-      informatics data and metadata exchange format, offering substantial
-      improvements in reliability, extensibility, and richness over
-      earlier standards such as NEXUS. In this paper we present the RNeXML
-      package which provides a bridge between this rich and reliable
-      data format and the extensive library of phylogenetic tools already
-      available in R, with a particular emphasis on comparative methods.
+bibliography: components/references.bib
+csl: components/methods-in-ecology-and-evolution.csl
+documentclass: components/elsarticle
 
-      2. RNeXML allows R users to both read and write NeXML files. Yet
-      RNeXML is much richer than a basic parser: here we illustrate
-      how the package facilitates adding and manipulating semantic 
-      annotations.
-
-      3. An essential feature of NeXML is to provide both the flexibility
-      to be continually extended to meet the emerging needs of the phylogenetics
-      community while remaining backwards-compatible with software developed
-      for earlier versions.  Here we illustrate how RNeXML makes it easy
-      to extend NeXML in this manner.
-
-      4. RNeXML is perfectly suited for users to preserve and maintain their
-      personal archives of phylogenetic and comparative trait data in consistent,
-      query-able, and cross-platform format, or leverage the rapidly growing 
-      public community archives of rich NeXML files.
+output: 
+  pdf_document:
+    template: components/elsarticle.latex
+    keep_tex: false
+    fig_caption: true
 
@@
 
-<!-- Emphasize:
-
-1) storing character data and tree data together, convenient for the comparative methods community
-2) Adding metadata, including automated addition of metadata and semantic utilities 
-3) Extending the NeXML standard  
-4) Accessing / opening up existing rich NeXML data to the R user
-
--->
-
-```{r compile-settings, include=FALSE}
-library(knitr)
-opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
-               cache = 1, comment = NA, verbose = TRUE)
-basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile') )
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
-
-```
-
-Though R provides the most extensive and widely used collection of
-comparative phylogenetic methods tools, R users lack a convenient,
-extensible format to exchange and archive comparative methods data.
-
-## Background
-
-### Challenges with existing formats
-See [Vos _et al._ 2012].
-
-### Challenges with R formats
-
-- Most R users rely on the NEXUS inherits the limitations of NEXUS file format
-
-See [phylobase vignette, 2010]
-
-<!--
-A summar of some phylogenetic formats in R:
-
-| Class     | R package   | R Object 
-|-----------|-------------|-------
-| phylo     | ape         | S3 
-| phylo4    | phylobase   | S4
-| phylo4d   | adephylo    | S4
-| ouchtree  | ouch        | S4 
-| treeshape | apTreeshape | list 
--->
-
-### NeXML
-
-See [Vos _et al._ 2012].
-
-Mention of data sharing challenges and opportunities.  See [Drew et al 2013]
-
-Why RNeXML?
-
-* Why NeXML -- because NEXUS standard isn't a standard.  Developers extend
-it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
-For example, NEXUS cannot represent horizontal gene transfer, ambiguous
-use of symbols, cannot be extended reliably, and there is no way to
-validate that it is proper NEXUS.  NeXML addresses all of these problems
-[Vos _et al._ 2012] with a state-of-the-art data management format.
-
-* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
-All would benefit by being able to read NeXML, both as a robust standard
-free of the pitfalls of NEXUS, and more importantly, as a much richer
-source of data thanks to its extendable annotation model.  (2) Write
-NeXML.  R packages and R users add rich information to the phylogenies
-they work with, but need a platform independent data exchange format to
-share and archive this. In particular, we illustrate how NeXML can
-be used to distribute both phylogenies and character trait data used in 
-many popular phylogenetic comparative methods tools in R.  (3) Annotate 
-and extend nexml. `RNeXML` provides convenient tools for adding metadata
-to NeXML files, including the automated generation of metadata and examples
-of how to extend the NeXML standard to represent new data features without
-breaking validation or backwards compatability of the file type.
-
-## Introduction to RNeXML
-
-A nexml file can be thought of as having three components: (1) phylogeny
-data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
-provide a seemless way to convert these to and from their corresponding
-R object representations, such as `ape::phylo` phylogenies, character
-matrices (`data.frames`) and metadata lists (or for advanced users,
-`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
-information about the study in which the phylogeny was generated or
-applied (such as authors, citations to publications, timestamps, and even
-descriptions of methods), but can also be used to annotate individual
-data components.  We illustrate how metadata annotations can be added to
-the operational taxonomic units of the phylogeny and can be used to extend
-the data representation to new uses such as stochastic character maps.
-
-The current stable version of the `RNeXML` package can be installed from
-the CRAN library using the standard installation method <!-- Actually not yet -->
-
-```r
-install.packages("RNeXML")
-```
-
-The latest stable development version can be installed from Github using the `devtools` package:
-
-```r
-library(devtools)
-install_github("RNeXML", "ropensci")
-```
-
-## Working with phylogenetic trees
-
-#### Read methods
-
-```{r echo=FALSE}
-library(RNeXML)
-```
-
-We begin by reading in an example NeXML file provided with the package.
-
-```{r}
-f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
-```
-
-(For convenience `nexml_read` is also aliased as `read.nexml`, for
-users familiar with the somewhat analogous `read.nexus` function).
-The resulting `nex` object is an R object representation of the NeXML.
-From this object we can extract any phylogenies it contains in the
-`ape::phylo` format:
-
-```{r}
-phy <- get_trees(nex)
-```
-
-We can then leverage the rich suite of methods available from the R
-phylogenetics community (including coercion into other representations,
-such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny
-formats). For instance, here we use the standard plotting function from
-the `ape` package to plot the resulting `ape::phylo` object returned by
-`get_trees`.
-
-```{r eval=FALSE}
-plot(phy)
-```
-
-NeXML can not only contain multiple phylogenetic trees in a single file,
-but can also group sets of trees into separate blocks.  The `get_trees`
-function attempts to return the most sensible R object for NeXML: if
-only a single tree is available, it will return an `ape::phylo` object;
-if multiple trees are available in a single `trees` block, it will return
-an `ape::multiPhylo` object (a list of `phylo` objects).  If multiple
-blocks are available, it will return a list of `multiPhylo` blocks.
-While most users will find this behavior convenient, there are times
-(such as looping over many NeXML files) when having a variable return
-type is undisirable.  To always get back a list of `multiPhylo` objects,
-even when only one tree is available, use `get_trees_list()`.
-
-<!-- have we already said all this above?-->  <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
-phylogenetic trees in the `nexml` format opens up a wide and rapidly
-growing array of resources to the R user.  For instance, a user can access
-all the phylogenies available in TreeBASE through the `nexml` format.
-`RNeXML` can read directly from a URL:
-
-```{r eval=FALSE}
-nex = nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
-```
-
-Previously, this was possible using the `treebase` package [@treebase]
-for R, which relied on the NEXUS parser provided in the `ape` package.
-Due to inconsistencies present in the NEXUS standard [Vos _et al._
-2012], a significant number <!-- State explicitly --> of TreeBASE NEXUS
-files could not be read in this way.  Using the more precisely defined
-NEXML standard, every valid NeXML file on TreeBASE can now successfully
-be parsed to R objects.
-
-#### Character data
-
-Like NEXUS files, NeXML files can also contain character data.  While
-most researchers will be familiar with the use of character matrices in
-NEXUS files as a container for molecular sequence data used to infer
-the phylogenetic tree, we find this ideal for character data used in
-comparative methods popular in R. Methods to get extract character data
-work much like those for the phylogenetic data:
-
-```{r }
-comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
-nex <- nexml_read(comp_analysis)
-get_characters(nex)
-```
-
-Returns a `data.frame` with columns as characters and rows as
-taxa. Continuous and discrete characters are represented as seperate
-blocks in NeXML files, but will be combined as seperate columns of a
-single data frame by `get_characters()` if they correspond to the same
-taxa. The `get_characters_list()` function will consistently recieve
-a list of data frames corresponding to each character matrix block of
-the NeXML.
-
-#### Write methods
-
-Generating NeXML files from R is likewise straight forward.  Here we
-write a phylogeny from the `ape::phylo` format out to a NeXML file:
-
-```{r}
-data(bird.orders)
-nexml_write(bird.orders, file = "birds.xml")
-```
-
-The first argument to `nexml_write` can be a `phylo` phylogeny,
-`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
-can be coerced into the `nexml` class.  As we have seen, NeXML files
-can contain multiple phylogenetic trees, as well as character matrices.
-We can add both trees  and character data using `nexml_write`,
-
-```{r}
-library(geiger)
-data(geospiza)
-nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
-```
-
-NeXML files can also be built up bit by bit using the `add_characters`,
-`add_trees`, and `add_meta`.  Each of these functions can add data to an
-existing `nexml` object or create a new nexml object if none is provided.
-Here we first create a `nexml` object containing the phylogeny data,
-and then add the character data to it:
-
-```{r}
-geiger_nex <- add_trees(geospiza$phy)
-geiger_nex <- add_characters(geospiza$dat, geiger_nex)
-```
-
-The data need not share the same taxa.  We can append additional
-phylogenies and characters corresponding to different taxa:
-
-```{r}
-data(primates)
-geiger_nex <- add_trees(primates$phy, geiger_nex)
-geiger_nex <- add_characters(primates$dat, geiger_nex)
-```
-
-This allows us to include multiple trees and character matrices to a single NeXML file.  
-<!-- We can add relevant metadata to each tree, such as the citation information 
-     indicating the paper in which it was first described.  -->
-
-### Metadata
-
-At the heart of the NeXML standard is the ability to encode arbitrary
-metadata in precisely specified, machine-readable formats that maintain
-backwards compabitibilty with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations", or RDFa [W3C 2013],
-as described in [Vos et al 2011].  These annotations can be used to add 
-information to any particular element of the data, (a tree, node, edge,
-or taxonomic unit; a character matrix or specific trait, or even another 
-metadata annotation or the NeXML data file as a whole). `RNeXML` provides
-a range of tools for accessing metadata embedded in a NeXML file.
-
-### Writing NeXML metadata
-
-Researchers cannot take advantage of NeXML metadata manipulation if
-there is no metadata there.  The burden of adding good metadata remains a
-significant barrier to increased sharing and reuse of scientific research
-[Tenopir et al. 2011].  `RNeXML` seeks to reduce the burden of adding
-high quality, machine-readable metadata to NeXML files. This occurs on
-several levels:
-
-- Automatic annotation.
-
-`RNeXML` automatically adds basic metadata by default such as a timestamp
-of when the tree was produced and an open data license (CC0, compatible
-with deposition on Dryad data repository), which can be configured
-as needed.  To adjust these and other common metadata fields that should 
-be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
-function.  This function does not assume any familiarity with the concepts
-of namespaces or other more advanced metadata tools discussed later.  The
-`add_basic_meta` takes an existing `nexml` object (or will create a new one
-if none is provided, like the other `add_` functions), and any of the following
-fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
-`citation`.  Additional fields may be added to subsequent versions, see
-`?add_basic_metadata` for the most up-to-date information.  Here we create
-an new `nexml` object with the basic metadata for the `bird.orders` data
-provided in the `ape` package:
-
-```{r}
-data(bird.orders)
-birds <- add_trees(bird.orders)
-birds <- add_basic_meta(birds,
-  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
-
-  description =" This data set describes the phylogenetic relationships of the
-     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
-     and Ahlquist inferred this phylogeny from an extensive number of
-     DNA/DNA hybridization experiments. The ``tapestry'' reported by
-     these two authors (more than 1000 species out of the ca. 9000
-     extant bird species) generated a lot of debates.
-
-     The present tree is based on the relationships among orders. The
-     branch lengths were calculated from the values of Delta T50H as
-     found in Sibley and Ahlquist (1990, fig. 353).",
-
-  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
-     classification of birds: a study in molecular evolution. New
-     Haven: Yale University Press.",
-
-  creator = "Sibley, C. G. and Ahlquist, J. E.")
-```
-
-Citations can also be added using R's `bibentry` type, which allows
-a user to add citations to R packages,
-
-```{r}
-birds <- add_basic_meta(birds, citation = citation("ape"))
-```
-
-or to papers by simply using the object's DOI:
-
-```{r}
-library("knitcitations")
-geiger_nex <- add_basic_meta(geiger_nex, citation=cite("10.2307/2408428"))
-```
-
-(Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)
-
-`RNeXML` also provides various functions to assist in automatically adding
-other more specific metadata to various elements of the NeXML data.
-
-For instance, misspellings or inconsistent use of taxonomic names in phylogenetic
-data are a common source of frustration in reusing phylogenetic trees. By 
-associating the taxonomic labels against a specific, centralized database
-we can catch typographical errors and indicate more precisely which classification
-we are appealing to in the event of conflicting taxonomic definitions.  In this 
-example, we indicate that we intend the bird orders listed in the tree to correspond
-to the NCBI definitions
-
-```{r message=FALSE, results='hide'}
-birds <- taxize_nexml(birds, "NCBI")
-```
-
-This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
-label against the NCBI database.  If a unique match is found, a metadata annotation
-is added to the taxon providing the NCBI identification number to the taxonomic unit.
-If no match is found, the user is warned to check for possible typographic errors in
-the taxonomic labels provided.  <!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->
-
-### Custom metadata extensions
-
-So far our examples have relied on functions that have a built-in understanding of the 
-kind of metadata being provided, and can thus generate the appropriate metadata with 
-little further guidence.  In order to be machine-readable and understandable, 
-NeXML requires all metadata annotations follow the RDFa format which explicitly declares 
-the meaning of the annotation in a way a computer can understand.  For instance, the 
-title we gave to the `bird.orders` phylogeny appears in the NeXML as:
-
-```xml
-<meta id="m2" property="dc:title" datatype="xsd:string"
-      content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
-```
-
-We see the title appears as the `content` of this `meta` element.  The `id` attribute
-provides us with a unique way to refer to this meta element in this document.  The `datatype`
-tells us that the content is a text string (as opposed to, say, a numerical or boolean value).  
-The `property` is the attribute that most interests us here; it tells us that the content 
-is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
-the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:
-
-```{r}
-prefixes <- get_namespaces(birds)
-prefixes["dc"]
-```
-
-This URL links us to a webpage providing both a human and machine-readable version of precisely
-what the term `title` means, showing us that it refers to `title` in the sense of a creative work, 
-rather than the title of a person, a legal title, the verb title, etc.  This precision can allow
-us to perform precise computer reasoning on metadata, without relying on heuristics or assumptions,
-as we illustrate later.  For the moment though, the need to refer to an explicit controlled vocabulary (or
-ontology) with a url, saying "dc:title" rather than just saying the property is "title", is something
-of a burden in creating metadata.  The previous examples involve functions that already know what 
-kind of metadata is being provided (a title, a data creator, etc), and so can provide the appropriate
-vocabulary automatically.  The main advantage of linking to an external vocabulary, however, is the
-ability to express concepts that are not already "built in" to the data model.  `RNeXML` provides 
-tools to generate arbitrary metadata annotations given an external vocabulary, such as we illustrate
-here.
-
-From the list of namespaces in `prefixes`, we can follow links a handful of established vocabularies
-and find terms that are not included in `add_basic_meta`.  For instance, in the `prism` prefix,
-
-```{r}
-prefixes["prism"]
-```
-
-we see that there is a term for `modificationDate` for a file.  We create a `meta` element 
-containing this annotation using the `meta` function:
-
-```{r}
-modified <- meta(property = "prism:modificationDate", content = "2013-10-04")
-```
-
-We can add this annotation to our existing `birds` nexml file using the `add_meta` function.
-Because we do not specify a level, it is added to the root node, refering to the nexml file as a whole.
-
-```{r}
-birds <- add_meta(modified, birds) 
-```
-
-The built-in vocabularies are just the tip of the iceberg of established vocabularies. Here 
-we add an annotation from the `skos` namespace which describes the history of where the
-data comes from:
-
-```{r}
-history <- meta(property = "skos:historyNote",
-  content = "Mapped from the bird.orders data in the ape package using RNeXML")
-```
-
-Because `skos` is not in the current namespace list, we add it with a url when
-adding this meta element.  We also specify that this annotation be placed 
-at the level of the `trees` subnode in the NeXML file.
-
-```{r}
-birds <- add_meta(history, 
-                birds, 
-                level = "trees",
-                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
-```
-
-If we forget to add the namespace with add_meta, we can always add it later
-using the `add_namespaces` function,
-
-```{r}
-birds <- add_namespaces(c(skos = "http://www.w3.org/2004/02/skos/core#"), birds)
-```
-
-(The `add_namespaces` function will avoid creating any duplicate definitions)
-For finer control of the level at which a `meta` element is added, we will manipulate 
-the nexml R object directly using S4 subsetting, as shown later.
-
-<!-- Ugh. y'all have much more practice explaining this well and without tangents, 
-so please feel free to edit or re-write.  -->
-
-Much richer metadata annotation is possible. Later we illustrate how metadata annotation
-can be used to extend the base NeXML format to represent new forms of data while maintaining
-compatability with any NeXML parser. The `RNeXML` package can be easily extended to support
-helper functions such as `taxize_nexml` to add additional metadata without imposing a large
-burden on the user.
-
-<!--
-- Utilities to add common metadata
-
-  - Add a citation to any papers associated with the phylogeny data.  
-  - Writing in methodological information on how tree has been generated or analyzed (including R code)
-  - Add annotations to any taxon.
-
-Temporal, spatial, taxonomic coverage for dryad/knb?
--->
-
-### Reading NeXML metadata
-
-A call to the nexml object prints some metadata summarizing the data structure: <!-- need to make this method faster!-->
-
-```{r }
-birds
-```
-
-We can extract all metadata pertaining to the NeXML document as a whole
-(annotations of the XML root node, `<nexml>`) with the command
-
-```{r}
-meta <- get_metadata(birds) 
-```
-
-This returns a named list of available metadata. We can see the kinds
-of metadata recorded from the names (showing the first 4):
-
-```{r}
-names(meta)[1:4]
-```
-
-and can ask for a particular element using the standard list subsetting
-mechanism (i.e. either the name of an element or it's numeric position),
-
-```{r}
-meta[["dc:title"]]
-```
-
-All metadata terms must belong to an explicit *namespace* or vocabulary
-that allows a computer to interpret the term precisely. The prefix (before
-the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
-this case.  The `get_namespaces` function tells us the definition of
-the vocabulary using a link:
-
-```{r}
-prefixes <- get_namespaces(birds)
-prefixes["dc"]
-```
-
-Following the link we can see that `dc` refers to the "Dublin Core"
-vocabulary, and the term `title` refers to the title of a creative work
-(as opposed to the title of a person, a legal title, the verb title, etc).
-Such explicit definitions allow a computer to reason about these objects,
-as we shall see later.
-
-Common metadata can be accessed with a few dedicated functions:
-
-```{r}
-get_citation(birds)
-```
-
-```{r}
-get_taxa(birds)
-```
-
-Which returns text from the otu element labels, typically used to define
-taxonomic names, rather than text from explicit meta elements.
-
-We can also access metadata at a specific level (or use `level=all`
-to extract all meta elements in a list).  Here we show only the first
-few results:
-
-```{r} 
-otu_meta <- get_metadata(birds, level="otu")
-otu_meta[1:4]
-```
-
-This approach does not preserve the relationship between the metadata element and the 
-`otu` it was annotating, and as such is most useful for quick visual exploration. 
-Metadata values for specific elements or sets of elemetns can always be accessed using
-the S4 subsetting mechanisms (see section below).
-
-## RDF manipulation and SPARQL queries
-
-Consider a better example than this one:
-
-```{r}
-rdf <- get_rdf(birds)
-```
-
-Query the rdf using XPath expressions:
-
-```{r}
-library(XML)
-xpathSApply(rdf, "//dc:title", xmlValue) 
-```
-
-XPath is an expressive syntax for querying XML-structured data such
-as RDF.  Queries using the XPath syntax are possible in R using the
-XML library [Temple Lang 2013, XML].  (Note that XPath is far more
-expressive for querying the metadata once the RDFa has been extracted,
-since attribute values are now interpreted as namespaced XML nodes and
-not just arbitrary text.)
-
-The most powerful way to take advantage of the machine-readable nature
-of the metadata involves the construction of SPARQL queries. This
-approach can exploit the reasoning and logical deduction capacity of
-formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see CITE SOMETHING.  Meanwhile, we just introduce the sytnax
-necessary to apply such queries to NeXML files from R.  First we load
-the necessary R library and import the RDF-extracted metadata:
-
-```{r}
-library(rrdf)
-saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
-lib <- load.rdf("rdf_meta.xml")
-```
-
-We can then make arbitrary SPARQL queries directly from R.  For instance,
-here is a simple query to extract the value of any `dc:title` element, analogous
-to the XPath expression above:
-
-```{r}
-sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-```
-
-### Extending the NeXML standard through metadata annotation.
-
-No data standard can anticipate all the uses and extensions researchers
-may one day need. Researchers have repeatedly adapted the existing NEXUS
-file format to accomidate encoding new kinds of information, but in the
-process break compatibility between different flavors of NEXUS files, with
-few reliable methods to distinguish between these flavors.  NeXML provides
-a solution to this problem through the use of metadata annotation.
-
-Here we illustrate this process using the example of stochastic character
-mapping [Huelsenbeck et al 2003]. A stochastic character map is simply
-an annotation of the branches on a phylogeny, assigning each section
-of each branch to a particular "state" (typically of a morphological
-characteristic).
-
-[Bollback 2006] provides a widely used stand-alone software implementation
-of this method in the software `simmap`, which modified the standard
-Newick tree format to express this additional information. This can
-break compatibility with other software [^1], and creates a format that
-cannot be interpreted without additional information describing this
-convention.  By contrast, the NeXML extension is not only backwards
-compatible but contains a precise and machine-readable description of
-what it is encoding.
-
-In this example, we illustrate how the additional information required
-to define a stochastic character mapping (a `simmap` mapping) in NeXML.
-
-[Revell 2011] describes the `phytools` package for R, which includes
-utilities for reading, manipulating, and writing `simmap` files in R.
-In this example, we also show how to define `RNeXML` functions that 
-map the R representation used by Revell (an extension of the `ape` class)
-into the NeXML extension we have defined by using `RNeXML` functions..
-
-Since a stochastic character map simply assigns different states to
-parts of a branch (or edge) on the phylogenetic tree, we can create
-a NeXML representation by annotating the `edge` elements with appropriate
-`meta` elements.  These elements need to describe the character state
-being assigned and the duration (in terms of branch-length) that the edge 
-spends in that state (Stochastic character maps are specific to time-calibrated
-or ultrametric trees).
-
-NeXML already defines the `characters` element to handle descrete character traits (`nex:char`)
-and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
-this purpose, referring to both the character trait and the states by the ids assigned to them
-in that element.  (NeXML's convention of referring to everything by id permits a single canonical
-definition of each term, making it clear where additional annotation belongs).  For each edge, we 
-need to indicate:
-
-<!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?
-     -->
-
-- That our annotation contains a stochastic character mapping reconstruction
-- Since many reconstructions are possible for a single edge, we give each reconstruction an id
-- We indicate for which character trait we are defining the reconstruction 
-- We then indicate which states the character assumes on that edge. 
-  For each state realized on the edge, that involves stating: 
-    + the state assignment
-    + the duration (length of time) for which the edge spends in the given state
-    + the order in which the state changes happen (Though we could just assume 
-      state transitions are listed chronologically, NeXML suggests making all 
-      data explicit, rather than relying on the structure of the data file to
-      convey information).
-
-Thus the the annotation for an edge that switches from state `s2` to state 
-`s1` of character `cr1` would be constructed like this:
-
-```{r}
- m <- meta("simmap:reconstructions", children = c(
-        meta("simmap:reconstruction", children = c(
-
-          meta("simmap:char", "cr1"),
-          meta("simmap:stateChange", children = c(
-            meta("simmap:order", 1),
-            meta("simmap:length", "0.2030"),
-            meta("simmap:state", "s2"))),
-
-          meta("simmap:char", "cr1"),
-          meta("simmap:stateChange", children = c(
-            meta("simmap:order", 2),
-            meta("simmap:length", "0.0022"),
-            meta("simmap:state", "s1")))
-          ))))
-```
-
-Of course writing out such a definition manually becomes tedious quickly. Because
-these are just R commands, we can easily define a function that can loop over an
-assignment like this for each edge, extracting the appropriate order, length and
-state from an existing R object such as that provided in the `phytools` package.  
-Likewise, it is straight forward to define a function that reads this data using
-the `RNeXML` utilities and converts it back to the `phytools` package. The full
-implementation of this mapping can be seen in the `simmap_to_nexml()` and the
-`nexml_to_simmap()` functions provided in the `RNeXML` package.
-
-As the code indicates, the key step is simply to define the data in meta elements. In 
-so doing, we have defined a custom namespace, `simmap`, to hold our variables.  This
-allows us to provide a URL with more detailed descriptions of what each of these 
-elements mean:
-
-```{r}
-nex <- add_namespaces(c(simmap = "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
-```
-
-At that URL we have posted a simple description of each term.
-
-<!-- How could this be improved? Should we use something more formal? 
-We want this example to be something other users could reasonably
-do themselves without expertise in RDFa or XML, but also to be a good
-model case that doesn't cut corners. @cboettig -->
-
-<!-- I don't see how this could be improved much without forcing people
-to fire op protege (let's not!). As it happens, the Dublin Core URI, 
-for example, points to an HTML page so even in that case automatic 
-reasoners would not be able to consume the vocabulary and draw inferences
-about, e.g. identity of predicates (owl:sameAs and such). @rvosa -->
-
-Using this convention we can generate NeXML files containing simmap
-data, read those files into R, and convert them back into the `phytools`
-package format. These simple functions serve as further illustration of
-how `RNeXML` can be used to extend the NeXML standard.  We illustrate
-their use briefly here, starting with  loading a nexml object containing
-a simmap reconstruction into R:
-
-```{r}
-data(simmap_ex)
-```
-
-The `get_trees` function can be used to return an `ape::phylo` tree as
-usual.  `RNeXML` automatically detects the simmap reconstruction data
-and returns includes this in a `maps` element of the `phylo` object,
-for use with other `phytools` functions.
-
-```{r}
-phy <- nexml_to_simmap(simmap_ex)
-```
-
-We can then use various functions from `phytools` designed for simmap
-objects, such as the plotting function:
-
-```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
-library(phytools)
-plotSimmap(phy)
-```
-
-Likewise, we can convert the object back in the nexml format and write
-it out to file to be read by other users.
-
-```{r}
-nex <- simmap_to_nexml(phy) 
-nexml_write(nex, "simmap.xml")
-```
-
-Though other NeXML parsers (for instance, for Perl or Python) have
-not been written explicitly to express simmap data, those parsers will
-nonetheless be able to successfully parse this file and expose the simmap
-data to the user.
-
-[Revell 2011]: http://doi.org/10.1111/j.2041-210X.2011.00169.x
-[Huelsenbeck et al 2003]: http://doi.org/10.1080/10635150390192780 "Stochastic Mapping of Morphological Characters"
-[Bollback 2006]: http://doi.org/10.1186/1471-2105-7-88 "SIMMAP: stochastic character mapping of discrete traits on phylogenies."
-
-[^1]: By using the commenting mechanism of the Newick format, it is
-possible that other software that doesn't also use the comment mechanism
-for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the
-data format to describe its use.
-
-### Subsetting nexml objects directly
+## Understanding the `nexml` S4 object
 
@@
 
-### Publishing NeXML files from R
-
-Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity (e.g. [Whitlock2011]).  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier ([Tenopir et al. 2011], [Stodden 2010]),
-and many phylogenetic trees upon which studies are based are inaccessible or lost 
-to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
-by providing functions to immediately archive NeXML files in scientific repositories.
-
-```{r eval = FALSE}
-doi <- nexml_publish(birds, visibility = "public", repository="figshare")
-```
-
-_Show downloading the nexml file directly from figshare, explain value
-of global paths instead of personal directory paths appearing in scripts, etc_
-
-Robust data archiving need not involve immediate release of the data. The `figshare` 
-repository supports the option of `visibility = "private"`, allowing secure 
-publication that provides a way to backup and share data with collaborators 
-prior to public release.
-
-<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->
-
-### Implementation and architecture
+# Implementation, architecture and quality control
 
@@
 using the R software environment and the XML package for parsing and
-generating XML [Temple Lang 2013, XML].
+generating XML `r citep(citation("XML"))`,  provides conversions into
+the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
+`RNeXML` can automatically check taxonomic names and add taxonomic
+identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
+Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
+and internet interfaces by the `httr` package `r citep(citation("httr"))`. The
+packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .
 
 <!--  Find more consise way to make sure dependencies are cited. 
-      How about citing packages that are only on the suggests list?
-
-and provides conversions into
-the popular `phylo` format provided by the `ape` package [Paradis 2004].
-`RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package
-([Chamberlain 2013]).
--->
-
+      How about citing packages that are only on the suggests list? --> 
 <!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
@@
 
-### Software reuse and support
+<!-- MISC / SCRATCH
 
-The `RNeXML` package is available through the CRAN repository under
-a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
-The current development
-branch can be found on Github at [https://github.com/ropensci/RNeXML].
-See the NEWS file for changes in recent versions. Users are encouraged
-to submit bug reports or feature requests in the issues log.  Contact
-ropensci-discuss@googlegroups.com or the package maintainer for help.
+What to say about existing formats?  
+A summary of some phylogenetic formats in R:
 
-There are many promising directions for further development of the
-`RNeXML` package.  The project's [Milestones page][milestones] page provides a list of
-steps currently planned or under consideration.  In particular, these
-milestones include increased support for interactive mode of entering data
-(wizards), and support for integration of semantics defined through the
-Web Ontology language (OWL), as discussed in the Challenges section.
+| Class     | R package   | R Object 
+|-----------|-------------|-------
+| phylo     | ape         | S3 
+| phylo4    | phylobase   | S4
+| phylo4d   | adephylo    | S4
+| ouchtree  | ouch        | S4 
+| treeshape | apTreeshape | list
 
-### Acknowledgements
-
-This project was supported in part by a grant from the Alfred P Sloan
-Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... `RNeXML` is part
-of the rOpenSci project, [http://ropensci.org].
-
-[Temple Lang 2013, XML]: http://cran.r-project.org/web/packages/XML/
-[Wickham 2011]: http://vita.had.co.nz/papers/testthat.html
-[Chamberlain 2013]: http://dx.doi.org/10.12688/f1000research.2-191.v2
-[@ouch]: http://cran.r-project.org/web/packages/ouch/
-[@phylobase]: http://cran.r-project.org/web/packages/phylobase
-[W3C 2013]: http://www.w3.org/TR/xhtml-rdfa-primer/ "RDFa 1.1 Primer - Second Edition: Rich Structured Data Markup for Web Documents"
-[Tenopir et al. 2011]: http://doi.org/10.1371/journal.pone.0021101 "Data Sharing by Scientists: Practices and Perceptions"
-[Drew et al 2013]: http://doi.org/10.1371/journal.pbio.1001636 
-[milestones]: https://github.com/ropensci/RNeXML/issues/milestones
+See [phylobase vignette, 2010]
 

### 2014-07-11 manuscripts/manuscript.Rmd
update runs of manuscript

@@
     template: components/elsarticle.latex
-    keep_tex: false
+    keep_tex: true
     fig_caption: true
@@
     - Adding basic metadata
-    - Adding taxonomic identifers
+    - Adding taxonomic identifiers
     - Adding Generic metadata
@@
 
-<!-- Need to say some of this stuff without repeating too much of existing literature -->
-
 Though R provides the most extensive and widely used collection of
@@
 
-<!-- FIXME -->
+<!-- FIXME
 
@@
 
+-->
+
 The ever expanding needs of the phylogenetics community are rapidly
@@
 radiation of different data representations and a loss of interoperability
-at the very time when the growth of available data and mehtods has made
-that interoprability most valuable.
+at the very time when the growth of available data and methods has made
+that interoperability most valuable.
 
@@
 by drawing on robust web standards in data representation.
+The design and expressiveness of NeXML makes it much easier to meet
+the rising bar for best practices in data sharing in phylogenetic 
+research `r citep("10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645")`.
 
@@
 
-The R lanugage `r citep(citation())` has become an important enviroment for 
+The R language `r citep(citation())` has become an important environment for 
 phylogenetics research, with over 30 packages available including several
@@
 but lacks a consistent, extensible export format. We illustrate how NeXML
-files can be formally published to a data respository directly from R.
+files can be formally published to a data repository directly from R.
 Finally, we provide an brief description of how the software has been
@@
 data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
-`RNeXML` seeks to provide a seemless way to convert these to and from their 
+`RNeXML` seeks to provide a seamless way to convert these to and from their 
 corresponding R object representations, such as `ape::phylo` phylogenies, character
@@
 
-<!-- have we already said all this above?-->  <!-- no, seems appropriate to repeat here, -scott --> The ability to read in
-phylogenetic trees in the NeXML format opens up a wide and rapidly
-growing array of resources to the R user.  For instance, a user can access
-all the phylogenies available in TreeBASE through the `nexml` format.
-`RNeXML` can read directly from a URL:
+The ability to read in phylogenetic trees in the NeXML format opens up a
+wide and rapidly growing array of resources to the R user.  For instance,
+a user can access all the phylogenies available in TreeBASE through the
+`nexml` format.  `RNeXML` can read directly from a URL:
 
@@
 
-Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`.
+Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`
 for R, which relied on the NEXUS parser provided in the `ape` package.
@@
 Returns a `data.frame` with columns as characters and rows as
-taxa. Continuous and discrete characters are represented as seperate
-blocks in NeXML files, but will be combined as seperate columns of a
+taxa. Continuous and discrete characters are represented as separate
+blocks in NeXML files, but will be combined as separate columns of a
 single data frame by `get_characters()` if they correspond to the same
@@
 $USER as a document-level annotation -->
-`RNeXML` automatically adds basic metadata by default such as a timestamp
+`RNeXML` automatically adds basic metadata by default such as a time-stamp
 of when the tree was produced and an open data license (CC0, compatible
@@
 other more specific metadata to various elements of the NeXML data.
-One common use case is that of taxonomic identifers.  Misspellings or
+One common use case is that of taxonomic identifiers.  Misspellings or
 inconsistent use of taxonomic names in phylogenetic data are a common
@@
 but also synonyms, homonyms and incompatible taxon concepts might plague
-the end user. By associating the
-taxonomic labels against a specific, centralized database we can catch
-some of these inconsistencies and indicate more precisely which classification
-we are appealing to in the event of conflicting taxonomic definitions.
-In this example, we indicate that we intend the bird orders listed in
-the tree to correspond to the NCBI definitions
+the end user. By associating the taxonomic labels against a specific,
+centralized database we can catch some of these inconsistencies and
+indicate more precisely which classification we are appealing to in the
+event of conflicting taxonomic definitions.  In this example, we indicate
+that we intend the bird orders listed in the tree to correspond to the
+NCBI definitions
 
@@
 understanding of the kind of metadata being provided, and can thus
-generate the appropriate metadata with little further guidence.
+generate the appropriate metadata with little further guidance.
 In order to be machine-readable and understandable, NeXML requires
@@
 
-This URL links us to a webpage providing both a human and machine-readable
+This URL links us to a web page providing both a human and machine-readable
 version of precisely what the term `title` means, showing us that it
@@
 
-We can add this annotation to our existing `birds` nexml file using the
+We can add this annotation to our existing `birds` NeXML file using the
 `add_meta` function.  Because we do not specify a level, it is added to
-the root node, refering to the nexml file as a whole.
+the root node, referring to the NeXML file as a whole.
 
@@
 url when adding this meta element.  We also specify that this annotation
-be placed at the level of the `trees` subnode in the NeXML file.
+be placed at the level of the `trees` sub-node in the NeXML file.
 
@@
 
-<!-- does this mean we can use CURIEs whose prefix has not yet been 
-bound to a namespace? that would be bad. there should probably be an
-error message when that happens -->
-If we forget to add the namespace with add_meta, we can always add it
-later using the `add_namespaces` function,
-
-```{r}
-birds <- add_namespaces(c(skos = "http://www.w3.org/2004/02/skos/core#"), birds)
-```
-
-Note that the `add_namespaces` function will automatically avoid creating
-any duplicate definitions. 
 For finer control of the level at which a `meta` element is added,
-we will manipulate the nexml R object directly using S4 subsetting,
-as shown later.
-
-<!--FIXME Ugh. y'all have much more practice explaining this well and without tangents, 
-so please feel free to edit or re-write.  -->
-<!-- I am not quite sure what points we are trying to make here -->
+we will manipulate the `nexml` R object directly using S4 sub-setting,
+as shown in the supplement.
 
@@
 
-A call to the nexml object prints some metadata summarizing the data structure:
+A call to the `nexml` object prints some metadata summarizing the data structure:
 
@@
 
-and can ask for a particular element using the standard list subsetting
+and can ask for a particular element using the standard list sub-setting
 mechanism (i.e. either the name of an element or its numeric position),
@@
 ```
-
-<!-- I feel like we already covered what dc:title is -->
-Following the link we can see that `dc` refers to the "Dublin Core"
-vocabulary, and the term `title` refers to the title of a creative work
-(as opposed to the title of a person, a legal title, the verb title, etc).
-Such explicit definitions allow a computer to reason about these objects,
-as we shall see later.
 
@@
 `otu` it was annotating, and as such is most useful for quick visual exploration. 
-Metadata values for specific elements or sets of elemetns can always be accessed using
-the S4 subsetting mechanisms (see section below).
+Metadata values for specific elements or sets of elements can always be accessed using
+the S4 sub-setting mechanisms (see section below).
 
 ## RDF manipulation and SPARQL queries
-
-Consider a better example than this one:
-
-```{r}
-rdf <- get_rdf(birds)
-```
 
@@
 formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see `r citet("Hartig Introduction to SPARQL")`.  Meanwhile, we just introduce the syntax
-necessary to apply such queries to NeXML files from R.  First we load
-the necessary R library `r citep(citation("rrdf"))` and import the RDF-extracted metadata:
+queries, see `r citet("http://www.w3.org/TR/rdf-sparql-query/")` and
+`r citet("Hartig Introduction to SPARQL")`.
 
-```{r}
-library("rrdf")
-saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
-lib <- load.rdf("rdf_meta.xml")
-```
-
-<!-- FIXME it would be nice if we had a more persuasive example here. directly getting the value
-of some predicate can be done the syntax-oriented way by treating the data as XML. what I
-think we should be going for here is an example of a slightly more complex query that would
-be painful to perform by scanning the XML document. I'm thinking of some sort of join of
-multiple predicate values. Nothing great comes to mind yet, though... ->
-
-We can then make arbitrary SPARQL queries directly from R.  For instance,
-here is a simple query to extract the value of any `dc:title` element, analogous
-to the XPath expression above:
-
-```{r}
-sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-```
+The real expressiveness of SPARQL is better illustrated through more
+complex examples such as the one shown in the supplementary materials
+(which can also be run from R using the command: `demo("sparql", "RNeXML")`).
+This demonstration constructs a cladogram based on the taxonomic ranks
+provided in the NeXML. SPARQL allows us to make explicit use of ontological
+concepts such as manipulating sub-classes (i.e. that "Genus" is a subclass 
+of "Family"), or joining multiple predicates (i.e. retrieve the NCBI identifier
+of the OTU that has taxon rank of "Order").
 
@@
 
-Thus the the annotation for an edge that switches from state `s2` to state 
+Thus the annotation for an edge that switches from state `s2` to state 
 `s1` of character `cr1` would be constructed like this:
@@
 
-<!-- How could this be improved? Should we use something more formal? 
-We want this example to be something other users could reasonably
-do themselves without expertise in RDFa or XML, but also to be a good
-model case that doesn't cut corners. @cboettig -->
-
-<!-- I don't see how this could be improved much without forcing people
-to fire up protege (let's not!). As it happens, the Dublin Core URI, 
-for example, points to an HTML page so even in that case automatic 
-reasoners would not be able to consume the vocabulary and draw inferences
-about, e.g. identity of predicates (owl:sameAs and such). @rvosa -->
-
 Using this convention we can generate NeXML files containing simmap
@@
 how `RNeXML` can be used to extend the NeXML standard.  We illustrate
-their use briefly here, starting with  loading a nexml object containing
+their use briefly here, starting with  loading a `nexml` object containing
 a simmap reconstruction into R:
@@
 
-Likewise, we can convert the object back in the nexml format and write
+Likewise, we can convert the object back in the NeXML format and write
 it out to file to be read by other users.
@@
 
-_Show downloading the nexml file directly from figshare, explain value
-of global paths instead of personal directory paths appearing in scripts, etc_
-
 Robust data archiving need not involve immediate release of the data. The `figshare` 
@@
 prior to public release.
-
-<!-- Wouldn't it be nice to publish to Dryad or TreeBASE in this way?  -->
 
@@
 for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarentee that this is the case or for the
+However there is no way to guarantee that this is the case or for the
 data format to describe its use.

### 2014-07-11 manuscripts/supplement.Rmd
update runs of manuscript

@@
 ---
+
+```{r supplement-compile-settings, include=FALSE}
+library("methods")
+library("knitr")
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- 'supplement' 
+opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
+               cache.path = paste("components/cache/", basename, "/", sep=""))
+
+library("knitcitations")
+cleanbib()
+cite_options(citation_format = "pandoc", check.entries=FALSE)
+```
+
+## SPARQL Queries
+
+```{r}
+library(rrdf)
+library(phytools)
+library(RNeXML)
+```
+
+```{r}
+nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
+```
+
+Extract the RDF graph from the nexml
+
+```{r}
+rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
+graph <- load.rdf("rdf_meta.xml")
+```
+
+fetch the NCBI URI for the taxon that has rank 'Order', i.e. the root of the primates. The dot operator `.` between clauses implies a join, in this case
+
+```{r}
+root <- sparql.rdf(graph, 
+    "SELECT ?uri WHERE { 
+        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
+        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
+    }"               
+)
+```
+
+Define a function to get the name
+
+```{r}
+get_name <- function(id) {
+    max <- length(nexml@otus[[1]]@otu)
+    for(i in 1:max) {
+        if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
+            label <- nexml@otus[[1]]@otu[[i]]@label
+            label <- gsub(" ","_",label)
+            return(label)
+        }
+    }
+}
+```
+
+ define a recursive function to build newick
+
+```{r}
+recurse <- function(node){
+
+    # fetch the taxonomic rank and id string
+    rank_query <- paste0(
+        "SELECT ?rank ?id WHERE {
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
+          }")
+    result <- sparql.rdf(graph, rank_query)
+
+    # get the local ID, strip URI part
+    id <- result[2]
+    id <- gsub("^.+#", "", id, perl = TRUE)
+
+    # if rank is terminal, return the name
+    if (result[1] == "http://rs.tdwg.org/ontology/voc/TaxonRank#Species") {
+        return(get_name(id))
+    }
+
+    # recurse deeper
+    else {
+        child_query <- paste0(
+            "SELECT ?uri WHERE {
+                ?id <http://www.w3.org/2000/01/rdf-schema#subClassOf> <",node,"> .
+                ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
+            }")
+        children <- sparql.rdf(graph, child_query)
+
+        # the newick can be made to contain interior node labels by inserting get_name(id), before the sep="" argument
+        return(paste("(", 
+                     paste(sapply(children, recurse), 
+                           sep = ",", collapse = "," ), 
+                     ")",  sep = "", collapse = ""))
+    }
+}
+```
+
+build the tree and visualize it
+
+```{r}
+newick <- paste(recurse(root), ";", sep = "", collapse = "")
+tree <- read.newick(text = newick)
+collapsed <- collapse.singles(tree)
+plot(collapsed, type = "cladogram")
+```
 

### 2014-07-11 manuscripts/supplement.Rmd
tweaking mansucript

@@
 
-bibliography: components/references.bib
+bibliography: components/sup-references.bib
 csl: components/methods-in-ecology-and-evolution.csl
@@
 mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult [@Vos_2012].
+interoperability difficult `r citep("Vos NeXML 2012")`.
 
@@
 changes have not negatively impacted any of the existing code base.
+
+```{r cleanup, include=FALSE, cache=FALSE}
+write.bibtex(file="components/sup-references.bib")
+unlink("simmap.xml")
+unlink("birds.xml")
+unlink("geospiza.xml")
+unlink("rdf_meta.xml")
+```
 

### 2014-07-11 manuscripts/manuscript.tex
commit additional component outputs

\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic
Data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[NESCent]{Hilmar Lapp}

    \author[SimonFraser]{Scott Chamberlain}

    \author[NBC]{Rutger Vos}

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
    \address[SimonFraser]{Department of Biology, Simon Fraser University, Burnaby, Canada}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    NeXML is a new phylogenetic informatics data and metadata exchange
    format that offers substantial improvements in reliability,
    extensibility, and richness over earlier standards such as NEXUS. In
    this paper we present the RNeXML package which provides a bridge
    between this rich and reliable data format and the extensive library
    of phylogenetic tools already available in R, with a particular
    emphasis on comparative methods.
  \item
    RNeXML is much richer than a basic parser: here we illustrate how the
    package facilitates adding and manipulating semantic annotations, that
    is annotations whose meaning can be defined externally by way of
    semantic web technologies.
  \item
    An essential feature of NeXML is to provide both the flexibility to be
    continually extended to meet the emerging needs of the phylogenetics
    community while remaining backwards-compatible with software developed
    for earlier versions. Here we illustrate how RNeXML makes it easy to
    extend NeXML in this manner.
  \item
    RNeXML is well suited for users to preserve and maintain their
    personal archives of phylogenetic and comparative trait data in
    consistent, query-able, and cross-platform format, to leverage the
    rapidly growing public community archives of rich NeXML files, and to
    interact with web services that deliver their results (with their
    provenance) as NeXML.
  \end{enumerate}
  \end{abstract}

 \end{frontmatter}

\section{Introduction}\label{introduction}

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data. A
common file format for sharing phylogenetic tree data has long played an
essential role in the advancement of phylogenetics research. The rapid
increase in phylogenetics data and research methods that make use of
such data has introduced many new challenges and opportunities to
sharing this information both between software environments and between
research groups (Stoltzfus \emph{et al.} 2012; Drew \emph{et al.} 2013).

The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of
interoperability at the very time when the growth of available data and
methods has made that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair). This has led developers to
repeatedly add small modifications to the standard to meet the needs of
their own software, which results in mutually incompatible dialects that
largely defeat the advantages of a shared file format. NeXML was
developed to addresses each of these problems (Vos \emph{et al.}
2012)\\by drawing on robust web standards in data representation. The
design and expressiveness of NeXML makes it much easier to meet the
rising bar for best practices in data sharing in phylogenetic research
(Cranston \emph{et al.} 2014).

The R language (R Core Team 2014) has become an important environment
for phylogenetics research, with over 30 packages available including
several with over 1000 citations (`CRAN task view' 2014), though
stand-alone programs remain an important part of this ecosystem for
tasks such as generating phylogenetic trees from sequence data. Within
this ecosystem a variety of different data formats for phylogenetic data
have been developed and modified. Most but not all are based on the
simple and popular \texttt{phylo} format provided by the ape package
(Paradis \emph{et al.} 2004).

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts, and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem.
Consequently, information added to phylogenies by these R programs
cannot easily be shared, archived, and made available to other software;
but inherits the limitations of the NEXUS format.

\texttt{RNeXML} opens up a rich world of both currently available and
rapidly growing data sets in the NeXML format to the R user and the
large ecosystem of R packages currently available for handling
phylogenetic data\\R packages and R users frequently add rich
information to the phylogenies they work with, but need a platform
independent data exchange format to share and archive this. We
illustrate how NeXML can be used to distribute both phylogenies and
character trait data used in many popular phylogenetic comparative
methods tools in R.\\\texttt{RNeXML} also provides convenient tools for
adding metadata to NeXML files, including the automated generation of
metadata and examples of how to extend the NeXML standard to represent
new data features without breaking validation or backwards compatibility
of the file type.

In this paper, we introduce the reader to the concepts and use of RNeXML
and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and
writing the kind of semantically rich metadata that most sets NeXML
apart from other phylogenetic data formats. Starting simple, we use
examples of increasing complexity to illustrate concepts and tools from
biodiversity informatics that allow users to express and interpret
complicated data structures automatically. We end with an example that
uses metadata to extend the NeXML standard to describe stochastic
character mapping, one of the kinds of data annotation that is currently
available in R formats but lacks a consistent, extensible export format.
We illustrate how NeXML files can be formally published to a data
repository directly from R. Finally, we provide an brief description of
how the software has been implemented, what quality control measures are
in place, as well as information about software reuse and getting
support.

This paper has been written as a dynamic document (Xie 2013). The reader
may paste the code blocks shown into their R terminal to follow along
with the exmples shown.

\section{Introduction to RNeXML}\label{introduction-to-rnexml}

A NeXML document can be thought of as having multiple components: (1)
phylogeny data (2) character-trait data, (3) operational taxonomic
units, and (4) metadata. \texttt{RNeXML} seeks to provide a seamless way
to convert these to and from their corresponding R object
representations, such as \texttt{ape::phylo} phylogenies, character
matrices (\texttt{data.frames}) and metadata lists (or for advanced
users, \texttt{XMLInternalDocument} and \texttt{rrdf} triplestores).
Metadata can provide information about the study in which the phylogeny
was generated or applied (such as authors, citations to publications,
timestamps, and even descriptions of methods), but can also be used to
annotate individual data components. We illustrate how metadata
annotations can be added to the operational taxonomic units of the
phylogeny and can be used to extend the data representation to new uses
such as stochastic character maps.

\subsection{Getting Started}\label{getting-started}

The current stable version of the \texttt{RNeXML} package can be
installed from the CRAN library\footnote{By providing the optional
  \texttt{dependencies} argument, R will install additional packages
  used in some of the advanced examples shown later but which are not
  required for basic functionality.}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{, }\DataTypeTok{dependencies=}\KeywordTok{c}\NormalTok{(}\StringTok{"DEPENDS"}\NormalTok{, }\StringTok{"IMPORTS"}\NormalTok{, }\StringTok{"SUGGESTS"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\subsection{Reading phylogenetic
trees}\label{reading-phylogenetic-trees}

We begin by reading in an example NeXML file provided with the
package.\footnote{For convenience \texttt{nexml\_read} is also aliased
  as \texttt{read.nexml}, for users familiar with the somewhat analogous
  \texttt{read.nexus} function.}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(f)}
\end{Highlighting}
\end{Shaded}

The resulting \texttt{nex} object is an R object representation of the
NeXML. This is a native R object for representing phylogenetic trees (as
well as character traits, and metadata), analogous to the \texttt{phylo}
object from the popular \texttt{ape} R package (Paradis \emph{et al.}
2004). While R functions and packages could be written to use this
\texttt{nexml} object directly, most users will prefer to use already
existing methods provided by packages they know. From \texttt{nex}
object we can therefore extract any phylogenies it contains in the
\texttt{ape::phylo} format:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{phy <-}\StringTok{ }\KeywordTok{get_trees}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

We can then leverage the rich suite of methods available from the R
phylogenetics community. For instance, here we use the standard plotting
function from the \texttt{ape} package to plot the resulting
\texttt{ape::phylo} object returned by \texttt{get\_trees}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(phy[[}\DecValTok{1}\NormalTok{]])}
\end{Highlighting}
\end{Shaded}

NeXML can not only contain multiple phylogenetic trees in a single
document, but can also group sets of trees into separate blocks (e.g.~as
resulting from different workflows). The \texttt{get\_trees} function
attempts to return the most sensible R object for NeXML: if only a
single tree is available, it will return an \texttt{ape::phylo} object;
if multiple trees are available in a single \texttt{trees} block, it
will return an \texttt{ape::multiPhylo} object (a list of \texttt{phylo}
objects). If multiple blocks are available, it will return a list of
\texttt{multiPhylo} blocks. While most users will find this behavior
convenient, there are times (such as looping over many NeXML files) when
having a variable return type is undesirable. To always get back a list
of \texttt{multiPhylo} objects, even when only one tree is available,
use \texttt{get\_trees\_list()}.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user. For instance,
a user can access all the phylogenies available in TreeBASE through the
\texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex =}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Previously, this was possible using the \texttt{treebase} package
(Boettiger \& {Temple Lang} 2012) for R, which relied on the NEXUS
parser provided in the \texttt{ape} package. Due to inconsistencies
present in the NEXUS standard (Vos \emph{et al.} 2012), a significant
number of TreeBASE NEXUS files could not be read in this way. Using the
more precisely defined NeXML standard, every valid NeXML document on
TreeBASE can now successfully be parsed to R objects.

\subsection{Reading character data}\label{reading-character-data}

Like NEXUS files, NeXML documents can also contain character data. While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological)
data used to infer the phylogenetic tree, we find this ideal for
character data used in comparative methods popular in R. Methods to
extract character data work much like those for the phylogenetic data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_analysis <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"comp_analysis.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(comp_analysis)}
\KeywordTok{get_characters}\NormalTok{(nex)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
         log snout-vent length reef-dwelling
taxon_8                 -3.278             0
taxon_9                  2.096             1
taxon_10                 3.137             0
taxon_1                  4.753             1
taxon_2                 -2.762             0
taxon_3                  2.105             0
taxon_4                 -4.950             0
taxon_5                  1.271             1
taxon_6                  6.259             1
taxon_7                  0.910             1
\end{verbatim}

Returns a \texttt{data.frame} with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by \texttt{get\_characters()} if they correspond to
the same taxa. The \texttt{get\_characters\_list()} function will
consistently receive a list of data frames corresponding to each
character matrix block of the NeXML.

\subsection{Writing phylogenetic
trees}\label{writing-phylogenetic-trees}

Generating NeXML files from R is likewise straightforward. Here we write
a phylogeny from the \texttt{ape::phylo} format out to a NeXML file:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data}\NormalTok{(bird.orders)}
\KeywordTok{nexml_write}\NormalTok{(bird.orders, }\DataTypeTok{file =} \StringTok{"birds.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "birds.xml"
\end{verbatim}

The first argument to \texttt{nexml\_write} can be a \texttt{phylo}
phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
object -- anything that can be coerced into the \texttt{nexml} class. As
we have seen, NeXML files can contain multiple phylogenetic trees, as
well as character matrices. We can add both trees and character data
using \texttt{nexml\_write},

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"geiger"}\NormalTok{)}
\KeywordTok{data}\NormalTok{(geospiza)}
\KeywordTok{nexml_write}\NormalTok{(}\DataTypeTok{trees =} \NormalTok{geospiza$phy, }\DataTypeTok{characters =} \NormalTok{geospiza$dat, }\DataTypeTok{file=}\StringTok{"geospiza.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "geospiza.xml"
\end{verbatim}

These examples use data from the popular phylogenetics software
\texttt{geiger} (Harmon \emph{et al.} 2008).

NeXML files can also be built up bit by bit using the
\texttt{add\_characters}, \texttt{add\_trees}, and \texttt{add\_meta}.
Each of these functions can add data to an existing \texttt{nexml}
object or create a new nexml object if none is provided. Here we first
create a \texttt{nexml} object containing the phylogeny data, and then
add the character data to it:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(geospiza$phy)}
\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_characters}\NormalTok{(geospiza$dat, geiger_nex)}
\end{Highlighting}
\end{Shaded}

The data need not share the same taxa. We can append additional
phylogenies and characters corresponding to different taxa:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data}\NormalTok{(primates)}
\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(primates$phy, geiger_nex)}
\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_characters}\NormalTok{(primates$dat, geiger_nex)}
\end{Highlighting}
\end{Shaded}

This allows us to include multiple trees and character matrices to a
single NeXML file.

\section{Metadata}\label{metadata}

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
``Resource Description Framework in Annotations'', or RDFa, as described
in Vos et al. (2012). These annotations can be used to add information
to any particular element of the data, (a tree, node, edge, or taxonomic
unit; a character matrix or specific trait, or even another metadata
annotation or the NeXML data file as a whole). \texttt{RNeXML} provides
a range of tools for accessing metadata embedded in a NeXML file.

\subsection{Writing NeXML metadata}\label{writing-nexml-metadata}

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there. The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific
research (Tenopir \emph{et al.} 2011). \texttt{RNeXML} seeks to reduce
the burden of adding high quality, machine-readable metadata to NeXML
files. This occurs on several levels:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Automatic annotation
\item
  Adding common metadata fields (\texttt{add\_basic\_metadata})
\item
  Taxonomic identifiers (\texttt{taxize\_nexml})
\item
  Generic metadata (\texttt{add\_meta})
\item
  Extending the NeXML standard using RDFa
\end{itemize}

\texttt{RNeXML} automatically adds basic metadata by default such as a
time-stamp of when the tree was produced and an open data license (CC0,
compatible with deposition on Dryad data repository), which can be
configured as needed.

To adjust these and other common metadata fields that should be part of
any data deposition, \texttt{RNeXML} provides the
\texttt{add\_basic\_meta()} function. This function does not assume any
familiarity with the concepts of namespaces or other more advanced
metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
existing \texttt{nexml} object (or will create a new one if none is
provided, like the other \texttt{add\_} functions), and any of the
following fields: \texttt{title}, \texttt{description},
\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
\texttt{citation}. Additional fields may be added to subsequent
versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
information. Here we create an new \texttt{nexml} object with the basic
metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
package:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data}\NormalTok{(bird.orders)}
\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(bird.orders)}
\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(birds,}
  \DataTypeTok{title =} \StringTok{"Phylogeny of the Orders of Birds From Sibley and Ahlquist"}\NormalTok{,}

  \DataTypeTok{description =}\StringTok{" This data set describes the phylogenetic relationships of the}
\StringTok{     orders of birds as reported by Sibley and Ahlquist (1990). Sibley}
\StringTok{     and Ahlquist inferred this phylogeny from an extensive number of}
\StringTok{     DNA/DNA hybridization experiments. The ``tapestry'' reported by}
\StringTok{     these two authors (more than 1000 species out of the ca. 9000}
\StringTok{     extant bird species) generated a lot of debates.}

\StringTok{     The present tree is based on the relationships among orders. The}
\StringTok{     branch lengths were calculated from the values of Delta T50H as}
\StringTok{     found in Sibley and Ahlquist (1990, fig. 353)."}\NormalTok{,}

  \DataTypeTok{citation =} \StringTok{"Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and}
\StringTok{     classification of birds: a study in molecular evolution. New}
\StringTok{     Haven: Yale University Press."}\NormalTok{,}

  \DataTypeTok{creator =} \StringTok{"Sibley, C. G. and Ahlquist, J. E."}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Citations can also be added using R's \texttt{bibentry} type, which
allows a user to add citations to R packages,

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(birds, }\DataTypeTok{citation =} \KeywordTok{citation}\NormalTok{(}\StringTok{"ape"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

or to papers by simply using the object's DOI (using Boettiger (2014)):

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"knitcitations"}\NormalTok{)}
\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(geiger_nex, }\DataTypeTok{citation =} \KeywordTok{bib_metadata}\NormalTok{(}\StringTok{"10.2307/2408428"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

(Which is the paper on which the \texttt{geospiza} phylogeny from the
geiger dataset is based)

\subsection{Taxonomic identifiers}\label{taxonomic-identifiers}

\texttt{RNeXML} provides various functions to assist in automatically
adding other more specific metadata to various elements of the NeXML
data. One common use case is that of taxonomic identifiers. Misspellings
or inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come
from other places that use different taxonomies. Hence, not just
misspellings, but also synonyms, homonyms and incompatible taxon
concepts might plague the end user. By associating the taxonomic labels
against a specific, centralized database we can catch some of these
inconsistencies and indicate more precisely which classification we are
appealing to in the event of conflicting taxonomic definitions. In this
example, we indicate that we intend the bird orders listed in the tree
to correspond to the NCBI definitions

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{birds <-}\StringTok{ }\KeywordTok{taxize_nexml}\NormalTok{(birds, }\StringTok{"NCBI"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

This function uses the \texttt{taxize} R library (Chamberlain \& Sz{ö}cs
2013) to check each taxon label against the NCBI database. If a unique
match is found, a metadata annotation is added to the taxon providing
the NCBI identification number to the taxonomic unit. If no match is
found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.

\subsection{Custom metadata
extensions}\label{custom-metadata-extensions}

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance. In order
to be machine-readable and understandable, NeXML requires all metadata
annotations follow the RDFa format (see Vos et al. (2012)) which
explicitly declares the meaning of the annotation in a way a computer
can understand. For instance, the title we gave to the
\texttt{bird.orders} phylogeny appears in the NeXML as:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{<meta}\OtherTok{ id=}\StringTok{"m2"}\OtherTok{ property=}\StringTok{"dc:title"}\OtherTok{ datatype=}\StringTok{"xsd:string"}
\OtherTok{      content=}\StringTok{"Phylogeny of the Orders of Birds From Sibley and Ahlquist"}\KeywordTok{/>} 
\end{Highlighting}
\end{Shaded}

We see the title appears as the value of the \texttt{content} attribute
of this \texttt{meta} element. The \texttt{id} attribute provides us
with a unique way to refer to this meta element in this document. The
\texttt{datatype} tells us that the content is a text string (as opposed
to, say, a numerical or boolean value). The \texttt{property} is the
attribute that most interests us here; it tells us that the content is a
title. The prefix \texttt{dc} indicates the vocabulary (here identified
as an XML namespace) to which the word \texttt{title} belongs. The
prefix is defined elsewhere in the NeXML file, as we can see with:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{prefixes <-}\StringTok{ }\KeywordTok{get_namespaces}\NormalTok{(birds)}
\NormalTok{prefixes[}\StringTok{"dc"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
                                dc 
"http://purl.org/dc/elements/1.1/" 
\end{verbatim}

This URL links us to a web page providing both a human and
machine-readable version of precisely what the term \texttt{title}
means, showing us that it refers to \texttt{title} in the sense of a
creative work, rather than the title of a person, a legal title, the
verb title, etc. This precision can allow us to perform precise computer
reasoning on metadata, without relying on heuristics or assumptions, as
we illustrate later. For the moment though, the need to refer to an
explicit controlled vocabulary (or ontology) with a url, saying
``dc:title'' rather than just saying the property is ``title'', is
something of a burden in creating metadata. The previous examples
involve functions that already know what kind of metadata is being
provided (a title, a data creator, etc), and so can provide the
appropriate vocabulary automatically. The main advantage of linking to
an external vocabulary, however, is the ability to express concepts that
are not already ``built in'' to the data model. \texttt{RNeXML} provides
tools to generate arbitrary metadata annotations given an external
vocabulary, such as we illustrate here.

From the list of namespaces in \texttt{prefixes}, we can follow links a
handful of established vocabularies and find terms that are not included
in \texttt{add\_basic\_meta}. For instance, by resolving the location
identified by the \texttt{prism} prefix,

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{prefixes[}\StringTok{"prism"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
                                           prism 
"http://prismstandard.org/namespaces/1.2/basic/" 
\end{verbatim}

we will find that there is a term for \texttt{modificationDate} for a
file. We create a \texttt{meta} element containing this annotation using
the \texttt{meta} function:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{modified <-}\StringTok{ }\KeywordTok{meta}\NormalTok{(}\DataTypeTok{property =} \StringTok{"prism:modificationDate"}\NormalTok{, }\DataTypeTok{content =} \StringTok{"2013-10-04"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can add this annotation to our existing \texttt{birds} NeXML file
using the \texttt{add\_meta} function. Because we do not specify a
level, it is added to the root node, referring to the NeXML file as a
whole.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_meta}\NormalTok{(modified, birds) }
\end{Highlighting}
\end{Shaded}

The built-in vocabularies are just the tip of the iceberg of established
vocabularies. Here we add an annotation from the \texttt{skos} namespace
which describes the history of where the data comes from:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{history <-}\StringTok{ }\KeywordTok{meta}\NormalTok{(}\DataTypeTok{property =} \StringTok{"skos:historyNote"}\NormalTok{,}
  \DataTypeTok{content =} \StringTok{"Mapped from the bird.orders data in the ape package using RNeXML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Because \texttt{skos} is not in the current namespace list, we add it
with a url when adding this meta element. We also specify that this
annotation be placed at the level of the \texttt{trees} sub-node in the
NeXML file.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_meta}\NormalTok{(history, }
                \NormalTok{birds, }
                \DataTypeTok{level =} \StringTok{"trees"}\NormalTok{,}
                \DataTypeTok{namespaces =} \KeywordTok{c}\NormalTok{(}\DataTypeTok{skos =} \StringTok{"http://www.w3.org/2004/02/skos/core#"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

For finer control of the level at which a \texttt{meta} element is
added, we will manipulate the \texttt{nexml} R object directly using S4
sub-setting, as shown in the supplement.

Much richer metadata annotation is possible. Later we illustrate how
metadata annotation can be used to extend the base NeXML format to
represent new forms of data while maintaining compatibility with any
NeXML parser. The \texttt{RNeXML} package can be easily extended to
support helper functions such as \texttt{taxize\_nexml} to add
additional metadata without imposing a large burden on the user.

\subsection{Reading NeXML metadata}\label{reading-nexml-metadata}

A call to the \texttt{nexml} object prints some metadata summarizing the
data structure:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{birds}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
A nexml object representing:
     1 phylogenetic tree blocks, where: 
     block 1 contains 1 phylogenetic trees 
     44 meta elements 
     0 character matrices 
     23 taxonomic units 
 Taxa:   Struthioniformes, Tinamiformes, Craciformes, Galliformes, Anseriformes, Turniciformes ...

 NeXML generated by RNeXML using schema version: 0.9 
 size: 370.7 Kb 
\end{verbatim}

We can extract all metadata pertaining to the NeXML document as a whole
(annotations of the XML root node,
\texttt{\textless{}nexml\textgreater{}}) with the command

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds) }
\end{Highlighting}
\end{Shaded}

This returns a named list of available metadata. We can see the kinds of
metadata recorded from the names (showing the first 4):

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{names}\NormalTok{(meta)[}\DecValTok{1}\NormalTok{:}\DecValTok{4}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "dc:title"                      "dc:creator"                   
[3] "dc:description"                "dcterms:bibliographicCitation"
\end{verbatim}

and can ask for a particular element using the standard list sub-setting
mechanism (i.e.~either the name of an element or its numeric position),

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{meta[[}\StringTok{"dc:title"}\NormalTok{]]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "Phylogeny of the Orders of Birds From Sibley and Ahlquist"
\end{verbatim}

All metadata terms must belong to an explicit \emph{namespace} or
vocabulary that allows a computer to interpret the term precisely. The
prefix (before the \texttt{:}) indicates to which vocabulary the term
belongs, e.g. \texttt{dc} in this case. The \texttt{get\_namespaces}
function tells us the definition of the vocabulary using a link:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{prefixes <-}\StringTok{ }\KeywordTok{get_namespaces}\NormalTok{(birds)}
\NormalTok{prefixes[}\StringTok{"dc"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
                                dc 
"http://purl.org/dc/elements/1.1/" 
\end{verbatim}

Common metadata can be accessed with a few dedicated functions:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{get_citation}\NormalTok{(birds)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
     classification of birds: a study in molecular evolution. New
     Haven: Yale University Press. Paradis E, Claude J and Strimmer K (2004). "APE: analyses of
phylogenetics and evolution in R language." _Bioinformatics_,
*20*, pp. 289-290.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{get_taxa}\NormalTok{(birds)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
 [1] "Struthioniformes" "Tinamiformes"     "Craciformes"     
 [4] "Galliformes"      "Anseriformes"     "Turniciformes"   
 [7] "Piciformes"       "Galbuliformes"    "Bucerotiformes"  
[10] "Upupiformes"      "Trogoniformes"    "Coraciiformes"   
[13] "Coliiformes"      "Cuculiformes"     "Psittaciformes"  
[16] "Apodiformes"      "Trochiliformes"   "Musophagiformes" 
[19] "Strigiformes"     "Columbiformes"    "Gruiformes"      
[22] "Ciconiiformes"    "Passeriformes"   
\end{verbatim}

Which returns text from the otu element labels, typically used to define
taxonomic names, rather than text from explicit meta elements.

We can also access metadata at a specific level (or use
\texttt{level=all} to extract all meta elements in a list). Here we show
only the first few results:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{otu_meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds, }\DataTypeTok{level=}\StringTok{"otu"}\NormalTok{)}
\NormalTok{otu_meta[}\DecValTok{1}\NormalTok{:}\DecValTok{4}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
$`tc:toTaxon`
[1] "http://ncbi.nlm.nih.gov/taxonomy/8798"

$`tc:toTaxon`
[1] "http://ncbi.nlm.nih.gov/taxonomy/8802"

$`tc:toTaxon`
[1] "http://ncbi.nlm.nih.gov/taxonomy/8976"

$`tc:toTaxon`
[1] "http://ncbi.nlm.nih.gov/taxonomy/8976"
\end{verbatim}

This approach does not preserve the relationship between the metadata
element and the \texttt{otu} it was annotating, and as such is most
useful for quick visual exploration. Metadata values for specific
elements or sets of elements can always be accessed using the S4
sub-setting mechanisms (see section below).

\subsection{RDF manipulation and SPARQL
queries}\label{rdf-manipulation-and-sparql-queries}

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see Prud'hommeaux (2014) and Hartig (2012).

The real expressiveness of SPARQL is better illustrated through more
complex examples such as the one shown in the supplementary materials
(which can also be run from R using the command:
\texttt{demo("sparql", "RNeXML")}). This demonstration constructs a
cladogram based on the taxonomic ranks provided in the NeXML. SPARQL
allows us to make explicit use of ontological concepts such as
manipulating sub-classes (i.e.~that ``Genus'' is a subclass of
``Family''), or joining multiple predicates (i.e.~retrieve the NCBI
identifier of the OTU that has taxon rank of ``Order'').

\subsection{Extending the NeXML standard through metadata
annotation.}\label{extending-the-nexml-standard-through-metadata-annotation.}

No data standard can anticipate all the uses and extensions researchers
may one day need. Researchers have repeatedly adapted the existing NEXUS
file format to accommodate encoding new kinds of information, but in the
process break compatibility between different flavors of NEXUS files,
with few reliable methods to distinguish between these flavors. NeXML
provides a solution to this problem through the use of metadata
annotation.

Here we illustrate this process using the example of stochastic
character mapping (Huelsenbeck \emph{et al.} 2003). A stochastic
character map is simply an annotation of the branches on a phylogeny,
assigning each section of each branch to a particular ``state''
(typically of a morphological characteristic).

Bollback (2006) provides a widely used stand-alone software
implementation of this method in the software \texttt{simmap}, which
modified the standard Newick tree format to express this additional
information. This can break compatibility with other software,\footnote{By
  using the commenting mechanism of the Newick format, it is possible
  that other software that doesn't also use the comment mechanism for
  some other such purpose would be able to successfully parse the tree.
  However there is no way to guarantee that this is the case or for the
  data format to describe its use.} and creates a format that cannot be
interpreted without additional information describing this convention.
By contrast, the NeXML extension is not only backwards compatible but
contains a precise and machine-readable description of what it is
encoding.

In this example, we illustrate how the additional information required
to define a stochastic character mapping (a \texttt{simmap} mapping) in
NeXML.

Revell (2012) describes the \texttt{phytools} package for R, which
includes utilities for reading, manipulating, and writing
\texttt{simmap} files in R. In this example, we also show how to define
\texttt{RNeXML} functions that map the R representation used by Revell
(an extension of the \texttt{ape} class) into the NeXML extension we
have defined by using \texttt{RNeXML} functions.

Since a stochastic character map simply assigns different states to
parts of a branch (or edge) on the phylogenetic tree, we can create a
NeXML representation by annotating the \texttt{edge} elements with
appropriate \texttt{meta} elements. These elements need to describe the
character state being assigned and the duration (in terms of
branch-length) that the edge spends in that state (Stochastic character
maps are specific to time-calibrated or ultrametric trees).

NeXML already defines the \texttt{characters} element to handle discrete
character traits (\texttt{nex:char}) and the states they can assume
(\texttt{nex:state}). We will thus reuse the \texttt{characters} element
for this purpose, referring to both the character trait and the states
by the ids assigned to them in that element. (NeXML's convention of
referring to everything by id permits a single canonical definition of
each term, making it clear where additional annotation belongs). For
each edge, we need to indicate:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  That our annotation contains a stochastic character mapping
  reconstruction
\item
  Since many reconstructions are possible for a single edge, we give
  each reconstruction an id
\item
  We indicate for which character trait we are defining the
  reconstruction
\item
  We then indicate which states the character assumes on that edge. For
  each state realized on the edge, that involves stating:

  \begin{itemize}
  \itemsep1pt\parskip0pt\parsep0pt
  \item
    the state assignment
  \item
    the duration (length of time) for which the edge spends in the given
    state
  \item
    the order in which the state changes happen (Though we could just
    assume state transitions are listed chronologically, NeXML suggests
    making all data explicit, rather than relying on the structure of
    the data file to convey information).
  \end{itemize}
\end{itemize}

Thus the annotation for an edge that switches from state \texttt{s2} to
state \texttt{s1} of character \texttt{cr1} would be constructed like
this:

\begin{Shaded}
\begin{Highlighting}[]
 \NormalTok{m <-}\StringTok{ }\KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:reconstructions"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
        \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:reconstruction"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}

          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:char"}\NormalTok{, }\StringTok{"cr1"}\NormalTok{),}
          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:stateChange"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:order"}\NormalTok{, }\DecValTok{1}\NormalTok{),}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:length"}\NormalTok{, }\StringTok{"0.2030"}\NormalTok{),}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:state"}\NormalTok{, }\StringTok{"s2"}\NormalTok{))),}

          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:char"}\NormalTok{, }\StringTok{"cr1"}\NormalTok{),}
          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:stateChange"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:order"}\NormalTok{, }\DecValTok{2}\NormalTok{),}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:length"}\NormalTok{, }\StringTok{"0.0022"}\NormalTok{),}
            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:state"}\NormalTok{, }\StringTok{"s1"}\NormalTok{)))}
          \NormalTok{))))}
\end{Highlighting}
\end{Shaded}

Of course writing out such a definition manually becomes tedious
quickly. Because these are just R commands, we can easily define a
function that can loop over an assignment like this for each edge,
extracting the appropriate order, length and state from an existing R
object such as that provided in the \texttt{phytools}
package.\\Likewise, it is straight forward to define a function that
reads this data using the \texttt{RNeXML} utilities and converts it back
to the \texttt{phytools} package. The full implementation of this
mapping can be seen in the \texttt{simmap\_to\_nexml()} and the
\texttt{nexml\_to\_simmap()} functions provided in the \texttt{RNeXML}
package.

As the code indicates, the key step is simply to define the data in meta
elements. In so doing, we have defined a custom namespace,
\texttt{simmap}, to hold our variables. This allows us to provide a URL
with more detailed descriptions of what each of these elements mean:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{add_namespaces}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DataTypeTok{simmap =} \StringTok{"https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

At that URL we have posted a simple description of each term.

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the
\texttt{phytools} package format. These simple functions serve as
further illustration of how \texttt{RNeXML} can be used to extend the
NeXML standard. We illustrate their use briefly here, starting with
loading a \texttt{nexml} object containing a simmap reconstruction into
R:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data}\NormalTok{(simmap_ex)}
\end{Highlighting}
\end{Shaded}

The \texttt{get\_trees} function can be used to return an
\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
the simmap reconstruction data and returns includes this in a
\texttt{maps} element of the \texttt{phylo} object, for use with other
\texttt{phytools} functions.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{phy <-}\StringTok{ }\KeywordTok{nexml_to_simmap}\NormalTok{(simmap_ex)}
\end{Highlighting}
\end{Shaded}

We can then use various functions from \texttt{phytools} designed for
simmap objects (Revell 2012), such as the plotting function:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"phytools"}\NormalTok{)}
\KeywordTok{plotSimmap}\NormalTok{(phy)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
no colors provided. using the following legend:
       A        B        C 
 "black"    "red" "green3" 
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-unnamed-chunk-31.pdf}
\caption{plot of chunk unnamed-chunk-31}
\end{figure}

Likewise, we can convert the object back in the NeXML format and write
it out to file to be read by other users.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nex <-}\StringTok{ }\KeywordTok{simmap_to_nexml}\NormalTok{(phy) }
\KeywordTok{nexml_write}\NormalTok{(nex, }\StringTok{"simmap.xml"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] "simmap.xml"
\end{verbatim}

Though other NeXML parsers (for instance, for Perl or Python) have not
been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the
simmap data to the user.

\subsection{Publishing NeXML files from
R}\label{publishing-nexml-files-from-r}

Data archiving is increasingly required by journals in evolutionary
biology and biodiversity (e.g. Rausher et al. (2010)). The burden of
preparing and submitting properly annotated data to archives continues
to be a significant barrier (Tenopir et al. (2011), Stodden (2014)), and
many phylogenetic trees upon which studies are based are inaccessible or
lost to the research community (Drew \emph{et al.} 2013).
\texttt{RNeXML} seeks to lower these barriers by providing functions to
immediately archive NeXML files in scientific repositories.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{doi <-}\StringTok{ }\KeywordTok{nexml_publish}\NormalTok{(birds, }\DataTypeTok{visibility =} \StringTok{"public"}\NormalTok{, }\DataTypeTok{repository=}\StringTok{"figshare"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Robust data archiving need not involve immediate release of the data.
The \texttt{figshare} repository supports the option of
\texttt{visibility = "private"}, allowing secure publication that
provides a way to backup and share data with collaborators prior to
public release.

\section{Software reuse and support}\label{software-reuse-and-support}

The \texttt{RNeXML} package is available through the CRAN repository
under a CC0 license. At the time of writing, the current version is
1.0-2. The current development branch can be found on Github at
{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
changes in recent versions. Users are encouraged to submit bug reports
or feature requests in the issues log. Contact
\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
help, or email the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org}.

There are many promising directions for further development of the
\texttt{RNeXML} package. The project's {[}Milestones
page{]}{[}\url{https://github.com/ropensci/RNeXML/issues/milestones}{]}
page provides a list of steps currently planned or under consideration.
In particular, these milestones include increased support for
interactive mode of entering data (wizards), and support for integration
of semantics defined through the Web Ontology language (OWL), as
discussed in the Challenges section.

\subsection{Acknowledgements}\label{acknowledgements}

This project was supported in part by a grant from the Alfred P Sloan
Foundation (CB and SC), NSF Grant DBI-1306697 (CB), \ldots{}

\texttt{RNeXML} is part of the rOpenSci project,
{[}\url{http://ropensci.org}{]}.

Boettiger, C. (2014). \emph{knitcitations: Citations for knitr markdown
files}. Retrieved from \url{https://github.com/cboettig/knitcitations}

Boettiger, C. \& {Temple Lang}, D. (2012). Treebase: an r package for
discovery, access and manipulation of online phylogenies (L. Harmon,
Ed.). \emph{Methods Ecol Evol}, \textbf{3}, 1060--1066. Retrieved from
\url{http://dx.doi.org/10.1111/j.2041-210x.2012.00247.x}

Bollback, J. (2006).\emph{BMC Bioinformatics}, \textbf{7}, 88. Retrieved
from \url{http://dx.doi.org/10.1186/1471-2105-7-88}

Chamberlain, S.A. \& Sz{ö}cs, E. (2013). taxize: taxonomic search and
retrieval in r. \emph{F1000Research}. Retrieved from
\url{http://dx.doi.org/10.12688/f1000research.2-191.v2}

CRAN task view: Phylogenetics, especially comparative methods. (2014).
Retrieved from
\url{http://cran.r-project.org/web/views/Phylogenetics.html}

Cranston, K., Harmon, L.J., O'Leary, M.A. \& Lisle, C. (2014). Best
practices for data sharing in phylogenetic research. \emph{PLoS Curr}.
Retrieved from
\url{http://dx.doi.org/10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645}

Drew, B.T., Gazis, R., Cabezas, P., Swithers, K.S., Deng, J., Rodriguez,
R., Katz, L.A., Crandall, K.A., Hibbett, D.S. \& Soltis, D.E. (2013).
Lost branches on the tree of life. \emph{PLoS Biol}, \textbf{11},
e1001636. Retrieved from
\url{http://dx.doi.org/10.1371/journal.pbio.1001636}

Harmon, L., Weir, J., Brock, C., Glor, R. \& Challenger, W. (2008).
GEIGER: investigating evolutionary radiations. \emph{Bioinformatics},
\textbf{24}, 129--131.

Hartig, O. (2012). An introduction to sPARQL and queries over linked
data. \emph{Web engineering} pp. 506--507. Springer Science + Business
Media. Retrieved from
\url{http://dx.doi.org/10.1007/978-3-642-31753-8_56}

Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
mapping of morphological characters. \emph{Systematic Biology},
\textbf{52}, 131--158. Retrieved from
\url{http://dx.doi.org/10.1080/10635150390192780}

Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: analyses of
phylogenetics and evolution in R language. \emph{Bioinformatics},
\textbf{20}, 289--290.

Prud'hommeaux, E. (2014). SPARQL query language for rDF. \emph{W3C}.
Retrieved from \url{http://www.w3.org/TR/rdf-sparql-query/}

R Core Team. (2014). \emph{R: A language and environment for statistical
computing}. R Foundation for Statistical Computing, Vienna, Austria.
Retrieved from \url{http://www.R-project.org/}

Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
M.C. (2010). DATA aRCHIVING. \emph{Evolution}, \textbf{64}, 603--604.
Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}

Revell, L.J. (2012). phytools: An r package for phylogenetic comparative
biology (and other things). \emph{Methods in Ecology and Evolution},
\textbf{3}, 217--223.

Stodden, V. (2014). The scientific method in practice: Reproducibility
in the computational sciences. \emph{SSRN Journal}. Retrieved from
\url{http://dx.doi.org/10.2139/ssrn.1550193}

Stoltzfus, A., O'Meara, B., Whitacre, J., Mounce, R., Gillespie, E.L.,
Kumar, S., Rosauer, D.F. \& Vos, R.A. (2012). Sharing and re-use of
phylogenetic trees (and associated data) to facilitate synthesis.
\emph{BMC Research Notes}, \textbf{5}, 574. Retrieved from
\url{http://dx.doi.org/10.1186/1756-0500-5-574}

Tenopir, C., Allard, S., Douglass, K., Aydinoglu, A.U., Wu, L., Read,
E., Manoff, M. \& Frame, M. (2011). Data sharing by scientists:
Practices and perceptions (C. Neylon, Ed.). \emph{PLoS ONE}, \textbf{6},
e21101. Retrieved from
\url{http://dx.doi.org/10.1371/journal.pone.0021101}

Vos, R.A., Balhoff, J.P., Caravas, J.A., Holder, M.T., Lapp, H.,
Maddison, W.P., Midford, P.E., Priyam, A., Sukumaran, J., Xia, X. \&
Stoltzfus, A. (2012). NeXML: Rich, extensible, and verifiable
representation of comparative data and metadata. \emph{Systematic
Biology}, \textbf{61}, 675--689. Retrieved from
\url{http://dx.doi.org/10.1093/sysbio/sys025}

Xie, Y. (2013). \emph{Dynamic documents with R and knitr}. Chapman;
Hall/CRC, Boca Raton, Florida. Retrieved from
\url{http://yihui.name/knitr/}

\end{document}

### 2014-07-17 manuscripts/supplement.Rmd
update supplement with more text describing the SPARQL example #73, doesn't run because of error in newick parsing now

@@
 
-## SPARQL Queries
-
-```{r}
-library(rrdf)
-library(phytools)
-library(RNeXML)
-```
-
-```{r}
-nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
-```
-
-Extract the RDF graph from the nexml
-
-```{r}
-rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
-saveXML(rdf, "rdf_meta.xml") # rrdf requires a file name, so we must write the XML out first
-graph <- load.rdf("rdf_meta.xml")
-```
-
-fetch the NCBI URI for the taxon that has rank 'Order', i.e. the root of the primates. The dot operator `.` between clauses implies a join, in this case
-
-```{r}
-root <- sparql.rdf(graph, 
-    "SELECT ?uri WHERE { 
-        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
-        ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
-    }"               
-)
-```
-
-Define a function to get the name
-
-```{r}
-get_name <- function(id) {
-    max <- length(nexml@otus[[1]]@otu)
-    for(i in 1:max) {
-        if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
-            label <- nexml@otus[[1]]@otu[[i]]@label
-            label <- gsub(" ","_",label)
-            return(label)
-        }
-    }
-}
-```
-
- define a recursive function to build newick
-
-```{r}
-recurse <- function(node){
-
-    # fetch the taxonomic rank and id string
-    rank_query <- paste0(
-        "SELECT ?rank ?id WHERE {
-            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
-            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
-          }")
-    result <- sparql.rdf(graph, rank_query)
-
-    # get the local ID, strip URI part
-    id <- result[2]
-    id <- gsub("^.+#", "", id, perl = TRUE)
-
-    # if rank is terminal, return the name
-    if (result[1] == "http://rs.tdwg.org/ontology/voc/TaxonRank#Species") {
-        return(get_name(id))
-    }
-
-    # recurse deeper
-    else {
-        child_query <- paste0(
-            "SELECT ?uri WHERE {
-                ?id <http://www.w3.org/2000/01/rdf-schema#subClassOf> <",node,"> .
-                ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
-            }")
-        children <- sparql.rdf(graph, child_query)
-
-        # the newick can be made to contain interior node labels by inserting get_name(id), before the sep="" argument
-        return(paste("(", 
-                     paste(sapply(children, recurse), 
-                           sep = ",", collapse = "," ), 
-                     ")",  sep = "", collapse = ""))
-    }
-}
-```
-
-build the tree and visualize it
-
-```{r}
-newick <- paste(recurse(root), ";", sep = "", collapse = "")
-tree <- read.newick(text = newick)
-collapsed <- collapse.singles(tree)
-plot(collapsed, type = "cladogram")
+```{r include=FALSE}
+library("RNeXML")
 ```
@@
 
-Recognize that `nex@namespaces` serves the same role as `get_namespaces` function, but provides direct access to the slot data.
-For instance, with this syntax we could also overwrite the existing namespaces with `nex@namespaces <- NULL`.
-Changing the namespace in this way is not advised.
+Recognize that `nex@namespaces` serves the same role as `get_namespaces`
+function, but provides direct access to the slot data.  For instance,
+with this syntax we could also overwrite the existing namespaces with
+`nex@namespaces <- NULL`.  Changing the namespace in this way is not
+advised.
 
-Some slots can contain multiple elements of the same type, such as `trees`, `characters`, and `otus`.  For instance, we see that
+Some slots can contain multiple elements of the same type, such as
+`trees`, `characters`, and `otus`.  For instance, we see that
 
@@
 
-and we access the first character matrix using the list notation, `[[1]]`. Here we check the class is a `characters` object.
+and we access the first character matrix using the list notation,
+`[[1]]`. Here we check the class is a `characters` object.
 
@@
 
-Direct subsetting has two primary use cases: (a) useful in looking up (and possibly editing) a specific value of an element, or (b) when adding metadata annotations to specific elements.
-Consider the example file
+Direct subsetting has two primary use cases: (a) useful in looking up
+(and possibly editing) a specific value of an element, or (b) when adding
+metadata annotations to specific elements. Consider the example file
 
@@
 
+## SPARQL Queries
+
+Rich, semantically meaningful metadata lies at the heart of the NeXML
+standard.  R provides a rich environment to unlock this information.
+While our previous examples have relied on the user knowing exactly
+what metadata they intend to extract (title, publication date, citation
+information, and so forth), _semantic_ metadata has meaning that a
+computer can make use of, allowing us to make much more conceptually
+rich queries than those simple examples.  The SPARQL query language
+is a powerful way to make use of such semantic information in making 
+complex queries.
+
+While users should consult a formal introduction to SPARQL for further
+background, here we illustrate how SPARQL can be used in combination
+with R functions in ways that would be much more tedious to assemble
+with only traditional/non-semantic queries.  The SPARQL query language
+is provided for the R environment through the `rrdf` package `r citep(citation("rrdf"))`,
+so we start by loading that package.  We will also make use of functions
+from `phytools` and `RNeXML`.
+
+```{r}
+library(rrdf)
+library(phytools)
+library(RNeXML)
+```
+
+We read in an example file that contains semantic metadata annotations
+describing the taxonomic units (OTUs) used in the tree.
+
+```{r}
+nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
+```
+
+In particular, this example declares the taxon rank, NCBI identifier and parent taxon
+for each OTU, such as:
+
+```xml
+<otu about="#ou541" id="ou541" label="Alouatta guariba">
+      <meta href="http://ncbi.nlm.nih.gov/taxonomy/182256" 
+            id="ma20" 
+            rel="concept:toTaxon" 
+            xsi:type="nex:ResourceMeta"/>
+      <meta href="http://rs.tdwg.org/ontology/voc/TaxonRank#Species" 
+            id="ma21" 
+            rel="concept:rank" 
+            xsi:type="nex:ResourceMeta"/>
+      <meta href="http://ncbi.nlm.nih.gov/taxonomy/9499" 
+            id="ma22" 
+            rel="rdfs:subClassOf" 
+            xsi:type="nex:ResourceMeta"/>
+    </otu>
+
+```
+
+In this example, we will construct a cladogram by using this information
+to identify the taxonomic rank of each OTU, and its shared parent
+taxonomic rank.  (If this example looks complex, try writing down the
+steps to do this without the aid of the SPARQL queries).  These examples
+show the manipulation of semantic triples, Unique Resource Identifiers
+(URIs) and use of the SPARQL "Join" operator.
+
+Note that this example can be run using `demo("sparql", "RNeXML")` to see the 
+code displayed in the R terminal and to avoid character errors that
+can occur in having to copy and paste from PDF files.
+
+We begin by extracting the RDF graph from the NeXML,
+
+```{r}
+rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+tmp <- tempfile()  # rrdf requires a file name, so we must write the XML out first
+saveXML(rdf, tmp) 
+graph <- load.rdf(tmp)
+```
+
+We then fetch the NCBI URI for the taxon that has rank 'Order', i.e. the
+root of the primates phylogeny. The dot operator `.` between clauses
+implies a join, in this case
+
+```{r}
+root <- sparql.rdf(graph, 
+"SELECT ?uri WHERE { 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
+}")
+```
+
+This makes use of the SPARQL query language provided by the `rrdf`
+package.  We will also define some helper functions that use SPARQL
+queries.  Here we define a function to get the name
+
+```{r}
+get_name <- function(id) {
+  max <- length(nexml@otus[[1]]@otu)
+  for(i in 1:max) {
+    if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
+      label <- nexml@otus[[1]]@otu[[i]]@label
+      label <- gsub(" ","_",label)
+      return(label)
+    }
+  }
+}
+```
+
+Next, we define a recursive function to build a newick tree from the taxonomic rank information.
+
+```{r}
+recurse <- function(node){
+
+    # fetch the taxonomic rank and id string
+    rank_query <- paste0(
+        "SELECT ?rank ?id WHERE {
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
+          }")
+    result <- sparql.rdf(graph, rank_query)
+
+    # get the local ID, strip URI part
+    id <- result[2]
+    id <- gsub("^.+#", "", id, perl = TRUE)
+
+    # if rank is terminal, return the name
+    if (result[1] == "http://rs.tdwg.org/ontology/voc/TaxonRank#Species") {
+        return(get_name(id))
+    }
+
+    # recurse deeper
+    else {
+        child_query <- paste0(
+            "SELECT ?uri WHERE {
+                ?id <http://www.w3.org/2000/01/rdf-schema#subClassOf> <",node,"> .
+                ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
+            }")
+        children <- sparql.rdf(graph, child_query)
+
+        return(paste("(", 
+                     paste(sapply(children, recurse), 
+                           sep = ",", collapse = "," ), 
+                     ")",  
+                     get_name(id), # label interior nodes
+                     sep = "", collapse = ""))
+    }
+}
+
+```
+
+With these functions in place, it is straight forward to build the tree
+from the semantic RDFa data and then visualize it
+
+<!-- hoping Rutger can add something here so that read.newick works when we have interior tip labels. -->
+
+```{r}
+newick <- paste(recurse(root), ";", sep = "", collapse = "")
+tree <- read.newick(text = newick)
+collapsed <- collapse.singles(tree)
+plot(collapsed, type = "cladogram")
+```
+
 # Implementation, architecture and quality control
@@
 packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .
-
-<!--  Find more consise way to make sure dependencies are cited. 
-      How about citing packages that are only on the suggests list? --> 
-<!-- Possibly cite all dependencies and suggests from the DESCRIPTION --> 
-<!-- Reference any installation and configuration instructions necessary (e.g. API credentials for figshare) -->
 
@@
 unlink("geospiza.xml")
-unlink("rdf_meta.xml")
+unlink(tmp)
 ```

### 2014-07-18 manuscripts/manuscript.Rmd
tweaking mansucript
Avoid running tests that may fail due to missing dependencies on certain CRAN architectures (e.g. Solaris, see NEWS)

@@
   - name: Scott Chamberlain
-    affiliation: SimonFraser
+    affiliation: ropensci
   - name: Rutger Vos
@@
       National Evolutionary Synthesis Center, Durham, NC, USA
-  - code: SimonFraser
+  - code: ropensci
     address: |
-      Department of Biology, Simon Fraser University, Burnaby, Canada
+      University of California, Berkeley, CA, USA
   - code: NBC
@@
 
-```{r}
+```{r get_citation}
 get_citation(birds)
@@
 
-```{r}
+```{r get_taxa}
 get_taxa(birds)
@@
 This project was supported in part by a grant from the Alfred P Sloan
-Foundation (CB and SC), NSF Grant DBI-1306697 (CB), ... <!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->
+Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB), ... <!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->
 

### 2014-07-18 tests/testthat/test_global_ids.R

@@ -2,19 +2,20 @@ context("Set global (uuid) identifiers")
 
 test_that("We can generate valid EML with uuid ids on all elements", {
 
-  options(uuid = TRUE)
+  if(require("uuid")){
+    options(uuid = TRUE)
 
-  data(geospiza)
-  add_trees(geospiza$phy)
-  nexml <- add_characters(geospiza$dat)
-  write.nexml(nexml, file = "geospiza.xml")
+    data(geospiza)
+    add_trees(geospiza$phy)
+    nexml <- add_characters(geospiza$dat)
+    write.nexml(nexml, file = "geospiza.xml")
 
-  results <- xmlSchemaValidate("http://www.nexml.org/2009/nexml.xsd", "geospiza.xml")
-  expect_equal(results$status, 0)
-  expect_equal(length(results$errors), 0)
+    results <- xmlSchemaValidate("http://www.nexml.org/2009/nexml.xsd", "geospiza.xml")
+    expect_equal(results$status, 0)
+    expect_equal(length(results$errors), 0)
 
-  
-  expect_true(nexml_validate("geospiza.xml"))
-  unlink("geospiza.xml")
-
+    
+    expect_true(nexml_validate("geospiza.xml"))
+    unlink("geospiza.xml")
+  }
 })

### 2014-07-21 manuscripts/manuscript.Rmd
update title as per #76

@@
 layout: review, 11pt
-title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
+title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
 author: 

### 2014-07-21 manuscripts/supplement.Rmd
update title as per #76

@@
 layout: review, 11pt
-title: "Supplementary material for -- RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
+title: "Supplementary material"
 author: 

### 2014-07-25 manuscripts/manuscript.Rmd

@@
 files can be formally published to a data repository directly from R.
-Finally, we provide an brief description of how the software has been
+Finally, we provide a brief description of how the software has been
 implemented, what quality control measures are in place, as well as
@@
 
+<!-- I think it would be useful to provide an overview of what is contained in this -->
+<!-- example file here. The reader will know better what to expect (e.g., the -->
+<!-- object contains 2 trees.) -->
+
 ```{r}
@@
 The resulting `nex` object is an R object representation of the NeXML. This is 
-a native R object for representing phylogenetic trees (as well as 
+a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
 character traits, and metadata), analogous to the `phylo` object from the 
@@
 could be written to use this `nexml` object directly, most users will prefer to use
-already existing methods provided by packages they know. From `nex` object we can 
+already existing methods provided by packages they know. From a `nexml` object, we can 
 therefore extract any phylogenies it contains in the `ape::phylo` format:
@@
 phylogenetics community. For instance, here we use the standard plotting function from
-the `ape` package to plot the resulting `ape::phylo` object returned by
+the `ape` package to plot the first tree in the resulting `ape::phylo` object returned by
 `get_trees`.
+
+<!-- I am not sure it really matters but `phy` is actually a `multiPhylo` -->
+<!-- object -->
 
@@
 wide and rapidly growing array of resources to the R user.  For instance,
-a user can access all the phylogenies available in TreeBASE through the
+a user can access all the phylogenies available in TreeBASE <!-- first mention -->
+<!-- of TreeBase, include URL? or introduce it in intro? --> through the
 `nexml` format.  `RNeXML` can read directly from a URL:
@@
 ```{r eval=FALSE}
-nex = nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
 ```
@@
 ```
+<!-- I think it's better to be explicit rather than referring to the chunk above -->
 
-Returns a `data.frame` with columns as characters and rows as
+`get_characters` returns a `data.frame` with columns as characters and rows as
 taxa. Continuous and discrete characters are represented as separate
@@
 `?add_basic_metadata` for the most up-to-date information.  Here we create
-an new `nexml` object with the basic metadata for the `bird.orders` data
+a new `nexml` object with the basic metadata for the `bird.orders` data
 provided in the `ape` package:
@@
 
-  description =" This data set describes the phylogenetic relationships of the
+  description = "This data set describes the phylogenetic relationships of the
      orders of birds as reported by Sibley and Ahlquist (1990). Sibley
@@
 
-or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`):
+or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`).
+For instance to add the citation information of the paper that generated the
+`geospiza` phylogeny included in the `geiger` package:
 
@@
 ```
-
-(Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)
 
@@
 
-This function uses the `taxize` R library `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
+This function uses the R package `taxize` `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
 If a unique match is found, a metadata annotation is added to the taxon
@@
 all metadata annotations follow the RDFa format (see @Vos_2012) which
-explicitly declares the meaning of the annotation in a way a computer
-can understand.  For instance, the title we gave to the `bird.orders`
-phylogeny appears in the NeXML as:
+explicitly declares the meaning of the annotation <!-- in a way a computer
+can understand -- this is redundant with the beginning of the sentence -->.  For instance, the title we gave to the `bird.orders`
+phylogeny appears in the NeXML file as:
 
@@
 state from an existing R object such as that provided in the `phytools` package.  
-Likewise, it is straight forward to define a function that reads this data using
+Likewise, it is straightforward to define a function that reads this data using
 the `RNeXML` utilities and converts it back to the `phytools` package. The full

### 2014-07-25 manuscripts/manuscript.Rmd
add Francois to acknowledgements

--- 2014-07-21 manuscripts/manuscript.Rmd
@@
 There are many promising directions for further development of the
-`RNeXML` package.  The project's [Milestones page][https://github.com/ropensci/RNeXML/issues/milestones] page provides a list of
+`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
 steps currently planned or under consideration.  In particular, these
@@
 ## Acknowledgements
+<!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->.
 
 This project was supported in part by a grant from the Alfred P Sloan
-Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB), ... <!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->
-
+Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB), ... 
+We thank F. Michonneau for helpful comments on an earlier version
+of this manuscript.
 `RNeXML` is part of the rOpenSci project, [http://ropensci.org].

### 2014-07-31 manuscripts/supplement.Rmd
supplement updated using newer phytools to plot sparql cladogram

@@
 collapsed <- collapse.singles(tree)
-plot(collapsed, type = "cladogram")
+plot(collapsed, 
+     type='cladogram', 
+     show.tip.label=FALSE, 
+     show.node.label=TRUE, 
+     cex=0.75, 
+     edge.color='grey60', 
+     label.offset=-9)
 ```

### 2014-07-31 manuscripts/manuscript.Rmd
edits to address #81 and #82

@@
 phylogenetic data have been developed and modified. Most but not all 
-are based on the simple and popular `phylo` format provided by the
+are based on the simple and popular `ape::phylo` format provided by the
 ape package `r citep(citation("ape"))`.
@@
 
-<!-- actually not yet, but before submission -->
-
 ```r
@@
 ```
-
-<!-- 
-The latest stable development version can be installed from Github using the `devtools` package:
-
-```r
-library(devtools)
-install_github("RNeXML", "ropensci", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
-```
--->
 
@@
 read.nexml and be done with it? -->
-We begin by reading in an example NeXML file provided with the package.[^2]
+<!-- For whatever reason, R has native support for aliasing function names;
+     so I don't see this creating a maintenance/documentation issue.  If I 
+     had to pick, I'd go with read_nexml though; a lot of the R developer 
+     community strongly opposes the `.` in function names, since it falsely
+     mimics the S3 class structure.  - CB -->
 
-<!-- I think it would be useful to provide an overview of what is contained in this -->
-<!-- example file here. The reader will know better what to expect (e.g., the -->
-<!-- object contains 2 trees.) -->
+We begin by loading the `RNeXML` package and reading in an example
+NeXML file provided with the package, containing two small phylogenetic
+trees.[^2]
 
 ```{r}
+library("RNeXML")
 f <- system.file("examples", "trees.xml", package="RNeXML")
@@
 a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
-character traits, and metadata), analogous to the `phylo` object from the 
+character traits, and metadata), analogous to the `ape::phylo` object from the 
 popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
@@
 already existing methods provided by packages they know. From a `nexml` object, we can 
-therefore extract any phylogenies it contains in the `ape::phylo` format:
+therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
+case we will get back a `multiPhylo` object, since the file contains two phylogenies.
 
@@
 
-<!-- I am not sure it really matters but `phy` is actually a `multiPhylo` -->
-<!-- object -->
-
 ```{r eval=FALSE}
@@
 object; if multiple trees are available in a single `trees` block, it will return
-an `ape::multiPhylo` object (a list of `phylo` objects).  If multiple
+an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
 blocks are available, it will return a list of `multiPhylo` blocks.
@@
 comparative methods popular in R. Methods to extract character data
-work much like those for the phylogenetic data:
+work much like those for the phylogenetic data. Here we read in an example
+NeXML file that contains both a phylogeny and character data. 
+`get_characters` returns a `data.frame` with columns as characters and rows as
+taxa:
 
@@
 ```
-<!-- I think it's better to be explicit rather than referring to the chunk above -->
 
-`get_characters` returns a `data.frame` with columns as characters and rows as
-taxa. Continuous and discrete characters are represented as separate
+As this example shows, continuous and discrete characters are represented as separate
 blocks in NeXML files, but will be combined as separate columns of a
@@
 write.nexml?) -->
-The first argument to `nexml_write` can be a `phylo` phylogeny,
+<!-- Great observation.  I think we should add a coercion method
+from the caper structure to the nexml S4 structure.  Then nexml_write
+can automatically handle this. added as issue now  -->
+
+The first argument to `nexml_write` can be a `ape::phylo` phylogeny,
 `multiPhylo` list of phylogenies, or a `nexml` object -- anything that
@@
 <!-- how about add_taxa? -->
+<!-- A taxa block is added when trees or characters are added.  Yeah,
+     I guess it would make sense to be able to add taxa directly,
+     (e.g. a NeXML file containing only RDFa metadata about taxa, 
+     independent of characters or trees), but I worry about confusing
+     people into thinking they must add taxa blocks separately, and
+     about when we should merge or not merge taxa blocks of matching
+     species. See https://github.com/ropensci/RNeXML/issues/43 
+     Added new issue for this, https://github.com/ropensci/RNeXML/issues/85 -->
+
 NeXML files can also be built up bit by bit using the `add_characters`,
@@
 $USER as a document-level annotation -->
+<!-- Good idea-->
+
 `RNeXML` automatically adds basic metadata by default such as a time-stamp
@@
 that we intend the bird orders listed in the tree to correspond to the
-NCBI definitions
+NCBI definitions.
+
+The `taxize_nexml` function uses the R package `taxize` 
+`r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
+If a unique match is found, a metadata annotation is added to the taxon
+providing the NCBI identification number to the taxonomic unit.  If no
+match is found, the user is warned to check for possible typographic
+errors in the taxonomic labels provided.
 
@@
 ```
-
-This function uses the R package `taxize` `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
-If a unique match is found, a metadata annotation is added to the taxon
-providing the NCBI identification number to the taxonomic unit.  If no
-match is found, the user is warned to check for possible typographic
-errors in the taxonomic labels provided.
 
@@
 all metadata annotations follow the RDFa format (see @Vos_2012) which
-explicitly declares the meaning of the annotation <!-- in a way a computer
-can understand -- this is redundant with the beginning of the sentence -->.  For instance, the title we gave to the `bird.orders`
+explicitly declares the meaning of the annotation. For instance, the title we gave to the `bird.orders`
 phylogeny appears in the NeXML file as:
@@
 
-Using this convention we can generate NeXML files containing simmap
+Using this convention we can generate NeXML files containing `simmap`
 data, read those files into R, and convert them back into the `phytools`
@@
 their use briefly here, starting with  loading a `nexml` object containing
-a simmap reconstruction into R:
+a `simmap` reconstruction into R:
 
@@
 The `get_trees` function can be used to return an `ape::phylo` tree as
-usual.  `RNeXML` automatically detects the simmap reconstruction data
-and returns includes this in a `maps` element of the `phylo` object,
+usual.  `RNeXML` automatically detects the `simmap` reconstruction data
+and returns includes this in a `maps` element of the `ape::phylo` object,
 for use with other `phytools` functions.
@@
 
-We can then use various functions from `phytools` designed for simmap
+We can then use various functions from `phytools` designed for `simmap`
 objects `r citep(citation("phytools"))`, such as the plotting function:
@@
 Though other NeXML parsers (for instance, for Perl or Python) have
-not been written explicitly to express simmap data, those parsers will
-nonetheless be able to successfully parse this file and expose the simmap
+not been written explicitly to express `simmap` data, those parsers will
+nonetheless be able to successfully parse this file and expose the `simmap`
 data to the user.

### 2014-07-31 manuscripts/manuscript.Rmd
Improve explanation of XML vocab #79 and namespaces #80 lines 565-600

@@
 all metadata annotations follow the RDFa format (see @Vos_2012) which
-explicitly declares the meaning of the annotation. For instance, the title we gave to the `bird.orders`
-phylogeny appears in the NeXML file as:
+explicitly declares the meaning of the annotation.
+
+Some basic definitions about the XML syntax used to express NeXML
+will be helpful for this discussion, though a more comprehensive 
+introduction can be found in @Vos_2012.  For instance, consider how
+the title we gave to the `bird.orders` phylogeny appears in the NeXML file
+using the XML format:
 
@@
 
+The format begins with an XML "tag" and the name "meta", indicated by the opening
+angle bracket; `<meta`.  A closing tag `\>` indicates the end of this
+XML "element" .  This tag contains several "attributes", named "id",
+"property", etc., that an R user can think of as names in a `list`. What
+follows the equals sign after the attribute name is called the attribute "value",
+(e.g. the `id` attribute has value "m2").
+
+The name of any element (like "meta") or attribute (like "id") in XML must be associated
+with a particular dictionary called a "namespace" and specified using a "prefix"
+of an abbreviation followed by a colon appearing before the name.  If no prefix
+is given, as is the case for all attributes here: `id`, `property`, `datatype`,
+and `content`, then a default prefix specified at the beginning of the document applies,
+in this case, the NeXML namespace.  This allows users to combine different XML
+that have different definitions of terms like `id` or `property`.
+
+The example element above also uses namespaces on some of the attribute values themselves.
+This is a consequence of RDFa annotation, and not strictly part of XML, though
+it works much the same way.  While we hope the following examples will make these
+concepts clearer, users do not need to understand XML and namespaces to use most
+of the basic annotation features provided by `RNeXML`.
+
+So much for the structure; we can now briefly explain the meaning of this element. 
 We see the title appears as the value of the `content` attribute of this `meta` element.

### 2014-07-31 manuscripts/manuscript.Rmd
mention phylobase among further directions #83

@@
 milestones include increased support for interactive mode of entering data
-(wizards), and support for integration of semantics defined through the
+(wizards), conversions to and from other existing R phylogenetics formats
+such those found in the `phylobase` package `r citep(citation("phylobase"))`,
+and support for integration of semantics defined through the
 Web Ontology language (OWL), as discussed in the Challenges section.

### 2014-07-31 manuscripts/manuscript.tex
bump version and add notes post CRAN-release
mention phylobase among further directions #83

@@
             pdfauthor={},
-            pdftitle={RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R},
+            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
             colorlinks=true,
@@
 
-  \title{RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic
-Data in R}
+  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
+character, and trait data in R}
     \author[cstar]{Carl Boettiger\corref{c1}}
@@
 
-    \author[SimonFraser]{Scott Chamberlain}
+    \author[ropensci]{Scott Chamberlain}
 
@@
     \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
-    \address[SimonFraser]{Department of Biology, Simon Fraser University, Burnaby, Canada}    
+    \address[ropensci]{University of California, Berkeley, CA, USA}    
     \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}
@@
 have been developed and modified. Most but not all are based on the
-simple and popular \texttt{phylo} format provided by the ape package
-(Paradis \emph{et al.} 2004).
+simple and popular \texttt{ape::phylo} format provided by the ape
+package (Paradis \emph{et al.} 2004).
 
@@
 We illustrate how NeXML files can be formally published to a data
-repository directly from R. Finally, we provide an brief description of
+repository directly from R. Finally, we provide a brief description of
 how the software has been implemented, what quality control measures are
@@
 
-We begin by reading in an example NeXML file provided with the
-package.\footnote{For convenience \texttt{nexml\_read} is also aliased
-  as \texttt{read.nexml}, for users familiar with the somewhat analogous
-  \texttt{read.nexus} function.}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+We begin by loading the \texttt{RNeXML} package and reading in an
+example NeXML file provided with the package, containing two small
+phylogenetic trees.\footnote{For convenience \texttt{nexml\_read} is
+  also aliased as \texttt{read.nexml}, for users familiar with the
+  somewhat analogous \texttt{read.nexus} function.}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{)}
 \NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
@@
 The resulting \texttt{nex} object is an R object representation of the
-NeXML. This is a native R object for representing phylogenetic trees (as
-well as character traits, and metadata), analogous to the \texttt{phylo}
-object from the popular \texttt{ape} R package (Paradis \emph{et al.}
-2004). While R functions and packages could be written to use this
-\texttt{nexml} object directly, most users will prefer to use already
-existing methods provided by packages they know. From \texttt{nex}
-object we can therefore extract any phylogenies it contains in the
-\texttt{ape::phylo} format:
+NeXML. This is a native R object, of the \texttt{nexml} class, for
+representing phylogenetic trees (as well as character traits, and
+metadata), analogous to the \texttt{ape::phylo} object from the popular
+\texttt{ape} R package (Paradis \emph{et al.} 2004). While R functions
+and packages could be written to use this \texttt{nexml} object
+directly, most users will prefer to use already existing methods
+provided by packages they know. From a \texttt{nexml} object, we can
+therefore extract any phylogenies it contains in the \texttt{ape::phylo}
+format. In this case we will get back a \texttt{multiPhylo} object,
+since the file contains two phylogenies.
 
@@
 phylogenetics community. For instance, here we use the standard plotting
-function from the \texttt{ape} package to plot the resulting
-\texttt{ape::phylo} object returned by \texttt{get\_trees}.
+function from the \texttt{ape} package to plot the first tree in the
+resulting \texttt{ape::phylo} object returned by \texttt{get\_trees}.
 
@@
 if multiple trees are available in a single \texttt{trees} block, it
-will return an \texttt{ape::multiPhylo} object (a list of \texttt{phylo}
-objects). If multiple blocks are available, it will return a list of
-\texttt{multiPhylo} blocks. While most users will find this behavior
-convenient, there are times (such as looping over many NeXML files) when
-having a variable return type is undesirable. To always get back a list
-of \texttt{multiPhylo} objects, even when only one tree is available,
-use \texttt{get\_trees\_list()}.
+will return an \texttt{ape::multiPhylo} object (a list of
+\texttt{ape::phylo} objects). If multiple blocks are available, it will
+return a list of \texttt{multiPhylo} blocks. While most users will find
+this behavior convenient, there are times (such as looping over many
+NeXML files) when having a variable return type is undesirable. To
+always get back a list of \texttt{multiPhylo} objects, even when only
+one tree is available, use \texttt{get\_trees\_list()}.
 
@@
 \begin{Highlighting}[]
-\NormalTok{nex =}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
+\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
 \end{Highlighting}
@@
 character data used in comparative methods popular in R. Methods to
-extract character data work much like those for the phylogenetic data:
+extract character data work much like those for the phylogenetic data.
+Here we read in an example NeXML file that contains both a phylogeny and
+character data. \texttt{get\_characters} returns a \texttt{data.frame}
+with columns as characters and rows as taxa:
 
@@
 
-Returns a \texttt{data.frame} with columns as characters and rows as
-taxa. Continuous and discrete characters are represented as separate
-blocks in NeXML files, but will be combined as separate columns of a
-single data frame by \texttt{get\_characters()} if they correspond to
-the same taxa. The \texttt{get\_characters\_list()} function will
-consistently receive a list of data frames corresponding to each
-character matrix block of the NeXML.
+As this example shows, continuous and discrete characters are
+represented as separate blocks in NeXML files, but will be combined as
+separate columns of a single data frame by \texttt{get\_characters()} if
+they correspond to the same taxa. The \texttt{get\_characters\_list()}
+function will consistently receive a list of data frames corresponding
+to each character matrix block of the NeXML.
 
@@
 
-The first argument to \texttt{nexml\_write} can be a \texttt{phylo}
+The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
 phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
@@
 versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
-information. Here we create an new \texttt{nexml} object with the basic
+information. Here we create a new \texttt{nexml} object with the basic
 metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
@@
 
-  \DataTypeTok{description =}\StringTok{" This data set describes the phylogenetic relationships of the}
+  \DataTypeTok{description =} \StringTok{"This data set describes the phylogenetic relationships of the}
 \StringTok{     orders of birds as reported by Sibley and Ahlquist (1990). Sibley}
@@
 
-or to papers by simply using the object's DOI (using Boettiger (2014)):
+or to papers by simply using the object's DOI (using Boettiger (2014)).
+For instance to add the citation information of the paper that generated
+the \texttt{geospiza} phylogeny included in the \texttt{geiger} package:
 
@@
 \end{Shaded}
-
-(Which is the paper on which the \texttt{geospiza} phylogeny from the
-geiger dataset is based)
 
@@
 example, we indicate that we intend the bird orders listed in the tree
-to correspond to the NCBI definitions
+to correspond to the NCBI definitions.
+
+The \texttt{taxize\_nexml} function uses the R package \texttt{taxize}
+(Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
+database. If a unique match is found, a metadata annotation is added to
+the taxon providing the NCBI identification number to the taxonomic
+unit. If no match is found, the user is warned to check for possible
+typographic errors in the taxonomic labels provided.
 
@@
 \end{Shaded}
-
-This function uses the \texttt{taxize} R library (Chamberlain \& Sz{ö}cs
-2013) to check each taxon label against the NCBI database. If a unique
-match is found, a metadata annotation is added to the taxon providing
-the NCBI identification number to the taxonomic unit. If no match is
-found, the user is warned to check for possible typographic errors in
-the taxonomic labels provided.
 
@@
 annotations follow the RDFa format (see Vos et al. (2012)) which
-explicitly declares the meaning of the annotation in a way a computer
-can understand. For instance, the title we gave to the
-\texttt{bird.orders} phylogeny appears in the NeXML as:
+explicitly declares the meaning of the annotation.
+
+Some basic definitions about the XML syntax used to express NeXML will
+be helpful for this discussion, though a more comprehensive introduction
+can be found in Vos et al. (2012). For instance, consider how the title
+we gave to the \texttt{bird.orders} phylogeny appears in the NeXML file
+using the XML format:
 
@@
 
-We see the title appears as the value of the \texttt{content} attribute
-of this \texttt{meta} element. The \texttt{id} attribute provides us
-with a unique way to refer to this meta element in this document. The
-\texttt{datatype} tells us that the content is a text string (as opposed
-to, say, a numerical or boolean value). The \texttt{property} is the
-attribute that most interests us here; it tells us that the content is a
-title. The prefix \texttt{dc} indicates the vocabulary (here identified
-as an XML namespace) to which the word \texttt{title} belongs. The
-prefix is defined elsewhere in the NeXML file, as we can see with:
+The format begins with an XML ``tag'' and the name ``meta'', indicated
+by the opening angle bracket; \texttt{\textless{}meta}. A closing tag
+\texttt{\textbackslash{}\textgreater{}} indicates the end of this XML
+``element'' . This tag contains several ``attributes'', named ``id'',
+``property'', etc., that an R user can think of as names in a
+\texttt{list}. What follows the equals sign after the attribute name is
+called the attribute ``value'', (e.g.~the \texttt{id} attribute has
+value ``m2'').
+
+The name of any element (like ``meta'') or attribute (like ``id'') in
+XML must be associated with a particular dictionary called a
+``namespace'' and specified using a ``prefix'' of an abbreviation
+followed by a colon appearing before the name. If no prefix is given, as
+is the case for all attributes here: \texttt{id}, \texttt{property},
+\texttt{datatype}, and \texttt{content}, then a default prefix specified
+at the beginning of the document applies, in this case, the NeXML
+namespace. This allows users to combine different XML that have
+different definitions of terms like \texttt{id} or \texttt{property}.
+
+The example element above also uses namespaces on some of the attribute
+values themselves. This is a consequence of RDFa annotation, and not
+strictly part of XML, though it works much the same way. While we hope
+the following examples will make these concepts clearer, users do not
+need to understand XML and namespaces to use most of the basic
+annotation features provided by \texttt{RNeXML}.
+
+So much for the structure; we can now briefly explain the meaning of
+this element. We see the title appears as the value of the
+\texttt{content} attribute of this \texttt{meta} element. The
+\texttt{id} attribute provides us with a unique way to refer to this
+meta element in this document. The \texttt{datatype} tells us that the
+content is a text string (as opposed to, say, a numerical or boolean
+value). The \texttt{property} is the attribute that most interests us
+here; it tells us that the content is a title. The prefix \texttt{dc}
+indicates the vocabulary (here identified as an XML namespace) to which
+the word \texttt{title} belongs. The prefix is defined elsewhere in the
+NeXML file, as we can see with:
 
@@
 object such as that provided in the \texttt{phytools}
-package.\\Likewise, it is straight forward to define a function that
+package.\\Likewise, it is straightforward to define a function that
 reads this data using the \texttt{RNeXML} utilities and converts it back
@@
 
-Using this convention we can generate NeXML files containing simmap
-data, read those files into R, and convert them back into the
-\texttt{phytools} package format. These simple functions serve as
-further illustration of how \texttt{RNeXML} can be used to extend the
+Using this convention we can generate NeXML files containing
+\texttt{simmap} data, read those files into R, and convert them back
+into the \texttt{phytools} package format. These simple functions serve
+as further illustration of how \texttt{RNeXML} can be used to extend the
 NeXML standard. We illustrate their use briefly here, starting with
-loading a \texttt{nexml} object containing a simmap reconstruction into
-R:
+loading a \texttt{nexml} object containing a \texttt{simmap}
+reconstruction into R:
 
@@
 \texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
-the simmap reconstruction data and returns includes this in a
-\texttt{maps} element of the \texttt{phylo} object, for use with other
-\texttt{phytools} functions.
+the \texttt{simmap} reconstruction data and returns includes this in a
+\texttt{maps} element of the \texttt{ape::phylo} object, for use with
+other \texttt{phytools} functions.
 
@@
 We can then use various functions from \texttt{phytools} designed for
-simmap objects (Revell 2012), such as the plotting function:
+\texttt{simmap} objects (Revell 2012), such as the plotting function:
 
@@
 Though other NeXML parsers (for instance, for Perl or Python) have not
-been written explicitly to express simmap data, those parsers will
-nonetheless be able to successfully parse this file and expose the
-simmap data to the user.
+been written explicitly to express \texttt{simmap} data, those parsers
+will nonetheless be able to successfully parse this file and expose the
+\texttt{simmap} data to the user.
 
@@
 under a CC0 license. At the time of writing, the current version is
-1.0-2. The current development branch can be found on Github at
+1.1.1. The current development branch can be found on Github at
 {[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
@@
 There are many promising directions for further development of the
-\texttt{RNeXML} package. The project's {[}Milestones
-page{]}{[}\url{https://github.com/ropensci/RNeXML/issues/milestones}{]}
-page provides a list of steps currently planned or under consideration.
-In particular, these milestones include increased support for
-interactive mode of entering data (wizards), and support for integration
-of semantics defined through the Web Ontology language (OWL), as
-discussed in the Challenges section.
+\texttt{RNeXML} package. The project's
+\href{https://github.com/ropensci/RNeXML/issues/milestones}{Milestones
+page} page provides a list of steps currently planned or under
+consideration. In particular, these milestones include increased support
+for interactive mode of entering data (wizards), conversions to and from
+other existing R phylogenetics formats such those found in the
+\texttt{phylobase} package (al. 2014), and support for integration of
+semantics defined through the Web Ontology language (OWL), as discussed
+in the Challenges section.
 
@@
 
+.
+
 This project was supported in part by a grant from the Alfred P Sloan
-Foundation (CB and SC), NSF Grant DBI-1306697 (CB), \ldots{}
-
-\texttt{RNeXML} is part of the rOpenSci project,
-{[}\url{http://ropensci.org}{]}.
+Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB),
+\ldots{} We thank F. Michonneau for helpful comments on an earlier
+version of this manuscript. \texttt{RNeXML} is part of the rOpenSci
+project, {[}\url{http://ropensci.org}{]}.
+
+al., R.H. et. (2014). \emph{phylobase: Base package for phylogenetic
+structures and comparative data}. Retrieved from
+\url{http://CRAN.R-project.org/package=phylobase}
 

### 2014-07-31 manuscripts/manuscript.Rmd
fix figure caption

@@
 
-```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
+```{r Figure1, fig.cap="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML."}
 library("phytools")

### 2014-08-07 manuscripts/manuscript.tex
fix figure caption
suggest phylobase (for citation at least), fix install.packages line

@@
 \begin{Highlighting}[]
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{, }\DataTypeTok{dependencies=}\KeywordTok{c}\NormalTok{(}\StringTok{"DEPENDS"}\NormalTok{, }\StringTok{"IMPORTS"}\NormalTok{, }\StringTok{"SUGGESTS"}\NormalTok{))}
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{, }\DataTypeTok{dependencies=}\OtherTok{TRUE}\NormalTok{)}
 \end{Highlighting}
@@
 \centering
-\includegraphics{components/figure/manuscript-unnamed-chunk-31.pdf}
-\caption{plot of chunk unnamed-chunk-31}
+\includegraphics{components/figure/manuscript-Figure1.pdf}
+\caption{Stochastic character mapping on a phylogeny, as generated by
+the phytools package after parsing the simmap-extended NeXML.}
 \end{figure}
@@
 under a CC0 license. At the time of writing, the current version is
-1.1.1. The current development branch can be found on Github at
+1.1.3. The current development branch can be found on Github at
 {[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for

### 2014-08-18 manuscripts/manuscript.Rmd
suggest phylobase (for citation at least), fix install.packages line
drop dynamic references due to CrossRef isntabilities

@@
 
-library("knitcitations")
-cleanbib()
-cite_options(citation_format = "pandoc", check.entries=FALSE)
 ```
@@
 to sharing this information both between software environments and 
-between research groups `r citep(c("Drew PLOS Biology Tree of Life 2013", "10.1186/1756-0500-5-574"))`.
+between research groups [@Drew_2013; @Stoltzfus_2012].
 
@@
 that largely defeat the advantages of a shared  file format.
-NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
+NeXML was developed to addresses each of these problems [@Vos_2012]  
 by drawing on robust web standards in data representation.
@@
 the rising bar for best practices in data sharing in phylogenetic 
-research `r citep("10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645")`.
+research [@Cranston_2014].
 
@@
 
-The R language `r citep(citation())` has become an important environment for 
+The R language [@R] has become an important environment for 
 phylogenetics research, with over 30 packages available including several
-with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
+with over 1000 citations [@taskview],
 though stand-alone programs remain an important part of this ecosystem
@@
 are based on the simple and popular `ape::phylo` format provided by the
-ape package `r citep(citation("ape"))`.
+ape package [@Paradis_2004].
 
@@
 This paper has been written as a dynamic document 
-`r citep(citation("knitr")[[2]])`.  The reader may paste the code blocks
+[@Xiei_2013].  The reader may paste the code blocks
 shown into their R terminal to follow along with the exmples shown.
@@
 ```r
-install.packages("RNeXML", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
+install.packages("RNeXML", dependencies=TRUE)
 ```
@@
 character traits, and metadata), analogous to the `ape::phylo` object from the 
-popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
+popular `ape` R package [@Paradis_2004]. While R functions and packages 
 could be written to use this `nexml` object directly, most users will prefer to use
@@
 
-Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`
+Previously, this was possible using the `treebase` package [@Boettiger_2012"]
 for R, which relied on the NEXUS parser provided in the `ape` package.
-Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
+Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
 a significant number <!-- State explicitly --> of TreeBASE NEXUS
@@
 
-These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger")[[4]])`.
+These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].
 
@@
 "Resource Description Framework in Annotations", or RDFa,
-as described in @Vos_2012. <!-- better to cite the W3C's RDFa TR -->
+as described in @Vos_2012 or @W3C_2014. 
 These annotations can be used to add 
@@
 significant barrier to increased sharing and reuse of scientific research
-`r citep("Tenopir Data Sharing by Scientists: Practices and Perceptions")`.
+[@Tenopir_2011].
 `RNeXML` seeks to reduce the burden of adding
@@
 
-or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`).
+or to papers by simply using the object's DOI (using @Boettiger_2014).
 For instance to add the citation information of the paper that generated the
@@
 The `taxize_nexml` function uses the R package `taxize` 
-`r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
+[@Chamberlain_2013] to check each taxon label against the NCBI database.
 If a unique match is found, a metadata annotation is added to the taxon
@@
 formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see `r citet("http://www.w3.org/TR/rdf-sparql-query/")` and
-`r citet("Hartig Introduction to SPARQL")`.
+queries, see @W3C_2014 or @Hartig_2012).
 
@@
 Here we illustrate this process using the example of stochastic character
-mapping `r citep("Huelsenbeck stochastic character mapping")`. A stochastic character map is simply
+mapping [@Huelsenbecki_2003]. A stochastic character map is simply
 an annotation of the branches on a phylogeny, assigning each section
@@
 
-`r citet("Bollback simmap 2006")` provides a widely used stand-alone software implementation
+@Bollback_2006 provides a widely used stand-alone software implementation
 of this method in the software `simmap`, which modified the standard
@@
 
-`r citet(citation("phytools"))` describes the `phytools` package for R, which includes
+@Revell_2012 describes the `phytools` package for R, which includes
 utilities for reading, manipulating, and writing `simmap` files in R.
@@
 We can then use various functions from `phytools` designed for `simmap`
-objects `r citep(citation("phytools"))`, such as the plotting function:
+objects [@Revell_2012], such as the plotting function:
 
@@
 Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity (e.g. `r citet("Whitlock data archiving Evolution 2010")`).  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier (@Tenopir_2011, `r citet("10.2139/ssrn.1550193")`),
+biodiversity (e.g. @Whitlock_2010).  The burden of preparing and submitting properly
+annotated data to archives continues to be a significant barrier [@Tenopir_2011; @Stodden_2014],
 and many phylogenetic trees upon which studies are based are inaccessible or lost 
@@
 (wizards), conversions to and from other existing R phylogenetics formats
-such those found in the `phylobase` package `r citep(citation("phylobase"))`,
+such those found in the `phylobase` package [@phylobase],
 and support for integration of semantics defined through the

### 2014-08-18 manuscripts/manuscript.Rmd
fix supplement to static reference list

@@
 ```{r cleanup, include=FALSE, cache=FALSE}
-write.bibtex(file="components/references.bib")
 unlink("simmap.xml")

### 2014-08-18 manuscripts/manuscript.tex
fix supplement to static reference list

@@
 
-This paper has been written as a dynamic document (Xie 2013). The reader
-may paste the code blocks shown into their R terminal to follow along
-with the exmples shown.
+This paper has been written as a dynamic document ({\textbf{???}}). The
+reader may paste the code blocks shown into their R terminal to follow
+along with the exmples shown.
 
@@
 Previously, this was possible using the \texttt{treebase} package
-(Boettiger \& {Temple Lang} 2012) for R, which relied on the NEXUS
-parser provided in the \texttt{ape} package. Due to inconsistencies
-present in the NEXUS standard (Vos \emph{et al.} 2012), a significant
-number of TreeBASE NEXUS files could not be read in this way. Using the
-more precisely defined NeXML standard, every valid NeXML document on
-TreeBASE can now successfully be parsed to R objects.
+{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
+NEXUS parser provided in the \texttt{ape} package. Due to
+inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
+a significant number of TreeBASE NEXUS files could not be read in this
+way. Using the more precisely defined NeXML standard, every valid NeXML
+document on TreeBASE can now successfully be parsed to R objects.
 
@@
 ``Resource Description Framework in Annotations'', or RDFa, as described
-in Vos et al. (2012). These annotations can be used to add information
-to any particular element of the data, (a tree, node, edge, or taxonomic
-unit; a character matrix or specific trait, or even another metadata
-annotation or the NeXML data file as a whole). \texttt{RNeXML} provides
-a range of tools for accessing metadata embedded in a NeXML file.
+in Vos et al. (2012) or Prud'hommeaux (2014). These annotations can be
+used to add information to any particular element of the data, (a tree,
+node, edge, or taxonomic unit; a character matrix or specific trait, or
+even another metadata annotation or the NeXML data file as a whole).
+\texttt{RNeXML} provides a range of tools for accessing metadata
+embedded in a NeXML file.
 
@@
 formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see Prud'hommeaux (2014) and Hartig (2012).
+queries, see Prud'hommeaux (2014) or Hartig (2012)).
 
@@
 Here we illustrate this process using the example of stochastic
-character mapping (Huelsenbeck \emph{et al.} 2003). A stochastic
-character map is simply an annotation of the branches on a phylogeny,
-assigning each section of each branch to a particular ``state''
-(typically of a morphological characteristic).
+character mapping ({\textbf{???}}). A stochastic character map is simply
+an annotation of the branches on a phylogeny, assigning each section of
+each branch to a particular ``state'' (typically of a morphological
+characteristic).
 
@@
 Data archiving is increasingly required by journals in evolutionary
-biology and biodiversity (e.g. Rausher et al. (2010)). The burden of
+biology and biodiversity (e.g. ({\textbf{???}})). The burden of
 preparing and submitting properly annotated data to archives continues
-to be a significant barrier (Tenopir et al. (2011), Stodden (2014)), and
-many phylogenetic trees upon which studies are based are inaccessible or
-lost to the research community (Drew \emph{et al.} 2013).
-\texttt{RNeXML} seeks to lower these barriers by providing functions to
-immediately archive NeXML files in scientific repositories.
+to be a significant barrier (Tenopir \emph{et al.} 2011; Stodden 2014),
+and many phylogenetic trees upon which studies are based are
+inaccessible or lost to the research community (Drew \emph{et al.}
+2013). \texttt{RNeXML} seeks to lower these barriers by providing
+functions to immediately archive NeXML files in scientific repositories.
 
@@
 
-Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
-mapping of morphological characters. \emph{Systematic Biology},
-\textbf{52}, 131--158. Retrieved from
-\url{http://dx.doi.org/10.1080/10635150390192780}
-
 Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: analyses of
@@
 Retrieved from \url{http://www.R-project.org/}
-
-Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
-M.C. (2010). DATA aRCHIVING. \emph{Evolution}, \textbf{64}, 603--604.
-Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}
 
@@
 
-Xie, Y. (2013). \emph{Dynamic documents with R and knitr}. Chapman;
-Hall/CRC, Boca Raton, Florida. Retrieved from
-\url{http://yihui.name/knitr/}
-
 \end{document}

### 2014-08-18 manuscripts/supplement.Rmd
fix supplement to static reference list

@@
 with only traditional/non-semantic queries.  The SPARQL query language
-is provided for the R environment through the `rrdf` package `r citep(citation("rrdf"))`,
+is provided for the R environment through the `rrdf` package [@Willighagen_2014],
 so we start by loading that package.  We will also make use of functions
@@
 rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
-tmp <- tempfile()  # rrdf requires a file name, so we must write the XML out first
+tmp <- tempfile()  # so we must write the XML out first
 saveXML(rdf, tmp) 
@@
 using the R software environment and the XML package for parsing and
-generating XML `r citep(citation("XML"))`,  provides conversions into
-the popular `phylo` format provided by the `ape` package `r citep(citation("ape"))`
+generating XML [@Temple_Lang_2013],  provides conversions into
+the popular `phylo` format provided by the `ape` package [@Paradis_2004],  
 `RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package `r citep(citation("taxize"))`.
-Unique identifiers are provided with the `uuid` package `r citep(citation("uuid"))`,
-and internet interfaces by the `httr` package `r citep(citation("httr"))`. The
-packages `plyr` and `reshape2``r citep(c(citation("plyr"), citation("reshape2")))` are also imported for internal methods .
+identifiers when generating NeXML using methods from the `taxize` package [@Chamberlain_2014].
+Unique identifiers are provided with the `uuid` package [@Urbank_2014],
+and internet interfaces by the `httr` package [@Wickham_2014]. The
+packages `plyr` and `reshape2` [@Wickham_2011; Wickham_2007] are also imported for internal methods.
 
@@
 mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult `r citep("Vos NeXML 2012")`.
+interoperability difficult [@Vos_2012].
 
@@
 the `RNeXML` package includes a complete test suite using the `testthat`
-package `r citep(citation("testthat"))`. Package development on Github includes continuous 
+package [@Wickham_2011a]. Package development on Github includes continuous 
 integration through the `travis` platform, in which the full test suite
@@
 ```{r cleanup, include=FALSE, cache=FALSE}
-write.bibtex(file="components/sup-references.bib")
 unlink("simmap.xml")

### 2014-08-18 manuscripts/manuscript.Rmd
fix reference typos

@@
 This paper has been written as a dynamic document 
-[@Xiei_2013].  The reader may paste the code blocks
+[@Xie_2013].  The reader may paste the code blocks
 shown into their R terminal to follow along with the exmples shown.
@@
 Here we illustrate this process using the example of stochastic character
-mapping [@Huelsenbecki_2003]. A stochastic character map is simply
+mapping [@Huelsenbeck_2003]. A stochastic character map is simply
 an annotation of the branches on a phylogeny, assigning each section
@@
 Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity (e.g. @Whitlock_2010).  The burden of preparing and submitting properly
+biodiversity (e.g. @Rausher_2010).  The burden of preparing and submitting properly
 annotated data to archives continues to be a significant barrier [@Tenopir_2011; @Stodden_2014],

### 2014-08-18 manuscripts/manuscript.tex
fix reference typos

@@
 
-This paper has been written as a dynamic document ({\textbf{???}}). The
-reader may paste the code blocks shown into their R terminal to follow
-along with the exmples shown.
+This paper has been written as a dynamic document (Xie 2013). The reader
+may paste the code blocks shown into their R terminal to follow along
+with the exmples shown.
 
@@
 Here we illustrate this process using the example of stochastic
-character mapping ({\textbf{???}}). A stochastic character map is simply
-an annotation of the branches on a phylogeny, assigning each section of
-each branch to a particular ``state'' (typically of a morphological
-characteristic).
+character mapping (Huelsenbeck \emph{et al.} 2003). A stochastic
+character map is simply an annotation of the branches on a phylogeny,
+assigning each section of each branch to a particular ``state''
+(typically of a morphological characteristic).
 
@@
 Data archiving is increasingly required by journals in evolutionary
-biology and biodiversity (e.g. ({\textbf{???}})). The burden of
+biology and biodiversity (e.g. Rausher et al. (2010)). The burden of
 preparing and submitting properly annotated data to archives continues
@@
 
+Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
+mapping of morphological characters. \emph{Systematic Biology},
+\textbf{52}, 131--158. Retrieved from
+\url{http://dx.doi.org/10.1080/10635150390192780}
+
 Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: analyses of
@@
 Retrieved from \url{http://www.R-project.org/}
+
+Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
+M.C. (2010). Data archiving. \emph{Evolution}, \textbf{64}, 603--604.
+Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}
 
@@
 
+Xie, Y. (2013). \emph{Dynamic documents with R and knitr}. Chapman;
+Hall/CRC, Boca Raton, Florida. Retrieved from
+\url{http://yihui.name/knitr/}
+
 \end{document}

### 2014-08-21 manuscripts/manuscript.Rmd
added a few comments for @cboettig to abstract

@@
       format that offers substantial improvements in reliability, extensibility, 
+      <!-- refer to NeXML as a format and NEXUS as a standard, correct? -->
       and richness over earlier standards such as NEXUS. In this paper we present 
@@
       for earlier versions.  Here we illustrate how RNeXML makes it easy
+      <!-- earlier versions of what? -->
       to extend NeXML in this manner.

### 2014-08-21 manuscripts/manuscript.Rmd
added a few comments to intro for @cboettig, inserted a single period

@@
 with over 1000 citations [@taskview],
+<!-- hmm, taskview citation renders as "'CRAN task view' 2014" intended? -->
 though stand-alone programs remain an important part of this ecosystem
@@
 growing data sets in the NeXML format to the R user and the large ecosystem
-of R packages currently available for handling phylogenetic data  
+of R packages currently available for handling phylogenetic data.  
+<!-- Carl, I assume a period goes here, I inserted one -->
 R packages and R users frequently add rich information to the phylogenies

### 2014-08-21 manuscripts/manuscript.Rmd
removed misplaced single quote after boettiger paper reference
inserted () parens in references to function calls in paragraphs in the manusript, #58

@@
 but can also group sets of trees into separate blocks (e.g. as resulting from
-different workflows). The `get_trees` function attempts to return the most sensible 
+different workflows). The `get_trees()` function attempts to return the most sensible 
 R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
@@
 
-Previously, this was possible using the `treebase` package [@Boettiger_2012"]
+Previously, this was possible using the `treebase` package [@Boettiger_2012]
 for R, which relied on the NEXUS parser provided in the `ape` package.
@@
 NeXML file that contains both a phylogeny and character data. 
-`get_characters` returns a `data.frame` with columns as characters and rows as
+`get_characters()` returns a `data.frame` with columns as characters and rows as
 taxa:
@@
 
-The first argument to `nexml_write` can be a `ape::phylo` phylogeny,
+The first argument to `nexml_write()` can be a `ape::phylo` phylogeny,
 `multiPhylo` list of phylogenies, or a `nexml` object -- anything that
@@
 can contain multiple phylogenetic trees, as well as character matrices.
-We can add both trees  and character data using `nexml_write`,
+We can add both trees and character data using `nexml_write()`,
 
@@
 
-NeXML files can also be built up bit by bit using the `add_characters`,
-`add_trees`, and `add_meta`.  Each of these functions can add data to an
+NeXML files can also be built up bit by bit using `add_characters()`,
+`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
 existing `nexml` object or create a new nexml object if none is provided.
@@
 of namespaces or other more advanced metadata tools discussed later.  The
-`add_basic_meta` takes an existing `nexml` object (or will create a new one
+`add_basic_meta()` function takes an existing `nexml` object (or will create a new one
 if none is provided, like the other `add_` functions), and any of the following
@@
 
-The `taxize_nexml` function uses the R package `taxize` 
+The `taxize_nexml()` function uses the R package `taxize` 
 [@Chamberlain_2013] to check each taxon label against the NCBI database.
@@
 of established vocabularies and find terms that are not included in
-`add_basic_meta`.  For instance, by resolving the location identified
+`add_basic_meta()`.  For instance, by resolving the location identified
 by the `prism` prefix,
@@
 We can add this annotation to our existing `birds` NeXML file using the
-`add_meta` function.  Because we do not specify a level, it is added to
+`add_meta()` function.  Because we do not specify a level, it is added to
 the root node, referring to the NeXML file as a whole.
@@
 
-The `get_trees` function can be used to return an `ape::phylo` tree as
+The `get_trees()` function can be used to return an `ape::phylo` tree as
 usual.  `RNeXML` automatically detects the `simmap` reconstruction data

### 2014-08-21 manuscripts/manuscript.Rmd
added parens after link to render properly

@@
 
-From the list of namespaces in `prefixes`, we can follow links a handful
+From the list of namespaces in `prefixes`, we can follow links to a handful
 of established vocabularies and find terms that are not included in
@@
 The current development
-branch can be found on Github at [https://github.com/ropensci/RNeXML].
+branch can be found on Github at [https://github.com/ropensci/RNeXML]().
 See the NEWS file for changes in recent versions. Users are encouraged

### 2014-08-21 manuscripts/manuscript.Rmd
added comment about phylobase reference

@@
 such those found in the `phylobase` package [@phylobase],
+<!-- hmm, phylobase ref renders for me as 'al. 2014' -->
 and support for integration of semantics defined through the

### 2014-08-21 manuscripts/manuscript.Rmd
added link to treebase where first mentioned

@@
 wide and rapidly growing array of resources to the R user.  For instance,
-a user can access all the phylogenies available in TreeBASE <!-- first mention -->
+a user can access all the phylogenies available in 
+[TreeBASE](http://treebase.org/treebase-web/home.html) <!-- first mention -->
 <!-- of TreeBase, include URL? or introduce it in intro? --> through the

### 2014-08-25 manuscripts/manuscript.Rmd
update get_characters / get_characters_list Adds option to use proper column instead of rownames for the taxa, as suggested in #89.  Closes #89

--- 2014-08-21 manuscripts/manuscript.Rmd
@@
       format that offers substantial improvements in reliability, extensibility, 
-      <!-- refer to NeXML as a format and NEXUS as a standard, correct? -->
-      and richness over earlier standards such as NEXUS. In this paper we present 
+      and richness over earlier formats such as NEXUS. In this paper we present 
       the RNeXML package which provides a bridge between this rich and reliable
@@
 If a unique match is found, a metadata annotation is added to the taxon
-providing the NCBI identification number to the taxonomic unit.  If no
-match is found, the user is warned to check for possible typographic
-errors in the taxonomic labels provided.
+providing the NCBI identification number to the taxonomic unit.
 
@@
 ```
-
-<!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->
+If no match is found, the user is warned to check for possible typographic
+errors in the taxonomic labels provided. If multiple matches are found,
+the user will be prompted to choose between them.
 

### 2014-08-25 R/get_characters.R

@@ -4,13 +4,16 @@
 #' Extract the character matrix
 #'
 #' @param nexml nexml object (e.g. from read.nexml)
+#' @param rownames_as_col option to return character matrix rownames
+#'  (with taxon ids) as it's own column in the data.frame. Default is FALSE
+#'  for compatibility with geiger and similar packages.
 #' @return the list of taxa
 #' @examples
 #' comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
 #' nex <- nexml_read(comp_analysis)
 #' get_characters_list(nex)
 #' @export
-get_characters_list <- function(nexml){
+get_characters_list <- function(nexml, rownames_as_col=FALSE){
 # extract mapping between otus and taxon ids 
   maps <- get_otu_maps(nexml)
 # loop over all character matrices 
@@ -34,6 +37,10 @@ get_characters_list <- function(nexml){
     } else {
       dat <- NULL 
     }
+    if(rownames_as_col){
+      dat <- cbind(taxa = rownames(dat), dat)
+      rownames(dat) <- NULL
+    }
     dat
   })
   # name the character matrices by their labels,
@@ -48,6 +55,9 @@ get_characters_list <- function(nexml){
 #' a list of data.frame's (e.g., as output from \code{\link{get_characters_list}})
 #' @param suffixes Add list element names as suffixes to output data.frame column 
 #' names. 
+#' @param rownames_as_col option to return character matrix rownames
+#'  (with taxon ids) as it's own column in the data.frame. Default is FALSE
+#'  for compatibility with geiger and similar packages.
 #' @export
 #' @examples
 #' \dontrun{
@@ -66,7 +76,7 @@ get_characters_list <- function(nexml){
 #' nex <- read.nexml(f)
 #' get_characters(nex)
 #' }
-get_characters <- function(input, suffixes=FALSE){
+get_characters <- function(input, suffixes=FALSE, rownames_as_col=FALSE){
   
   if(inherits(input, "nexml")){
     list_chars <- get_characters_list(input)
@@ -89,7 +99,14 @@ get_characters <- function(input, suffixes=FALSE){
       row.names(tt) <- tt[,"Row.names"]
       tt[,!names(tt) %in% "Row.names"]
     }
-    return( Reduce(mrecurse, list_chars) )
+
+    out <- Reduce(mrecurse, list_chars)
+
+    if(rownames_as_col){
[…]

### 2014-08-29 manuscripts/manuscript.tex
try simplified Dockerfile

@@
     format that offers substantial improvements in reliability,
-    extensibility, and richness over earlier standards such as NEXUS. In
+    extensibility, and richness over earlier formats such as NEXUS. In
     this paper we present the RNeXML package which provides a bridge
@@
 for phylogenetics research, with over 30 packages available including
-several with over 1000 citations (`CRAN task view' 2014), though
-stand-alone programs remain an important part of this ecosystem for
-tasks such as generating phylogenetic trees from sequence data. Within
-this ecosystem a variety of different data formats for phylogenetic data
-have been developed and modified. Most but not all are based on the
-simple and popular \texttt{ape::phylo} format provided by the ape
-package (Paradis \emph{et al.} 2004).
+several with over 1000 citations (O'Meara 2014), though stand-alone
+programs remain an important part of this ecosystem for tasks such as
+generating phylogenetic trees from sequence data. Within this ecosystem
+a variety of different data formats for phylogenetic data have been
+developed and modified. Most but not all are based on the simple and
+popular \texttt{ape::phylo} format provided by the ape package (Paradis
+\emph{et al.} 2004).
 
@@
 large ecosystem of R packages currently available for handling
-phylogenetic data\\R packages and R users frequently add rich
+phylogenetic data.\\ R packages and R users frequently add rich
 information to the phylogenies they work with, but need a platform
@@
 document, but can also group sets of trees into separate blocks (e.g.~as
-resulting from different workflows). The \texttt{get\_trees} function
+resulting from different workflows). The \texttt{get\_trees()} function
 attempts to return the most sensible R object for NeXML: if only a
@@
 wide and rapidly growing array of resources to the R user. For instance,
-a user can access all the phylogenies available in TreeBASE through the
+a user can access all the phylogenies available in
+\href{http://treebase.org/treebase-web/home.html}{TreeBASE} through the
 \texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:
@@
 Previously, this was possible using the \texttt{treebase} package
-{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
-NEXUS parser provided in the \texttt{ape} package. Due to
-inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
-a significant number of TreeBASE NEXUS files could not be read in this
-way. Using the more precisely defined NeXML standard, every valid NeXML
-document on TreeBASE can now successfully be parsed to R objects.
+(Boettiger \& {Temple Lang} 2012) for R, which relied on the NEXUS
+parser provided in the \texttt{ape} package. Due to inconsistencies
+present in the NEXUS standard (Vos \emph{et al.} 2012), a significant
+number of TreeBASE NEXUS files could not be read in this way. Using the
+more precisely defined NeXML standard, every valid NeXML document on
+TreeBASE can now successfully be parsed to R objects.
 
@@
 Here we read in an example NeXML file that contains both a phylogeny and
-character data. \texttt{get\_characters} returns a \texttt{data.frame}
+character data. \texttt{get\_characters()} returns a \texttt{data.frame}
 with columns as characters and rows as taxa:
@@
 
-The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
-phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
-object -- anything that can be coerced into the \texttt{nexml} class. As
-we have seen, NeXML files can contain multiple phylogenetic trees, as
-well as character matrices. We can add both trees and character data
-using \texttt{nexml\_write},
+The first argument to \texttt{nexml\_write()} can be a
+\texttt{ape::phylo} phylogeny, \texttt{multiPhylo} list of phylogenies,
+or a \texttt{nexml} object -- anything that can be coerced into the
+\texttt{nexml} class. As we have seen, NeXML files can contain multiple
+phylogenetic trees, as well as character matrices. We can add both trees
+and character data using \texttt{nexml\_write()},
 
@@
 
-NeXML files can also be built up bit by bit using the
-\texttt{add\_characters}, \texttt{add\_trees}, and \texttt{add\_meta}.
-Each of these functions can add data to an existing \texttt{nexml}
-object or create a new nexml object if none is provided. Here we first
-create a \texttt{nexml} object containing the phylogeny data, and then
-add the character data to it:
+NeXML files can also be built up bit by bit using
+\texttt{add\_characters()}, \texttt{add\_trees()}, and
+\texttt{add\_meta()}. Each of these functions can add data to an
+existing \texttt{nexml} object or create a new nexml object if none is
+provided. Here we first create a \texttt{nexml} object containing the
+phylogeny data, and then add the character data to it:
 
@@
 familiarity with the concepts of namespaces or other more advanced
-metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
-existing \texttt{nexml} object (or will create a new one if none is
-provided, like the other \texttt{add\_} functions), and any of the
-following fields: \texttt{title}, \texttt{description},
+metadata tools discussed later. The \texttt{add\_basic\_meta()} function
+takes an existing \texttt{nexml} object (or will create a new one if
+none is provided, like the other \texttt{add\_} functions), and any of
+the following fields: \texttt{title}, \texttt{description},
 \texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
@@
 
-The \texttt{taxize\_nexml} function uses the R package \texttt{taxize}
+The \texttt{taxize\_nexml()} function uses the R package \texttt{taxize}
 (Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
@@
 the taxon providing the NCBI identification number to the taxonomic
-unit. If no match is found, the user is warned to check for possible
-typographic errors in the taxonomic labels provided.
+unit.
 
@@
 \end{Shaded}
+
+If no match is found, the user is warned to check for possible
+typographic errors in the taxonomic labels provided. If multiple matches
+are found, the user will be prompted to choose between them.
 
@@
 
-From the list of namespaces in \texttt{prefixes}, we can follow links a
-handful of established vocabularies and find terms that are not included
-in \texttt{add\_basic\_meta}. For instance, by resolving the location
-identified by the \texttt{prism} prefix,
+From the list of namespaces in \texttt{prefixes}, we can follow links to
+a handful of established vocabularies and find terms that are not
+included in \texttt{add\_basic\_meta()}. For instance, by resolving the
+location identified by the \texttt{prism} prefix,
 
@@
 We can add this annotation to our existing \texttt{birds} NeXML file
-using the \texttt{add\_meta} function. Because we do not specify a
+using the \texttt{add\_meta()} function. Because we do not specify a
 level, it is added to the root node, referring to the NeXML file as a
@@
 
-The \texttt{get\_trees} function can be used to return an
+The \texttt{get\_trees()} function can be used to return an
 \texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
@@
 1.1.3. The current development branch can be found on Github at
-{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
+\href{}{\url{https://github.com/ropensci/RNeXML}}. See the NEWS file for
 changes in recent versions. Users are encouraged to submit bug reports
@@
 other existing R phylogenetics formats such those found in the
-\texttt{phylobase} package (al. 2014), and support for integration of
-semantics defined through the Web Ontology language (OWL), as discussed
-in the Challenges section.
+\texttt{phylobase} package (NESCENT R Hackathon Team 2014), and support
+for integration of semantics defined through the Web Ontology language
+(OWL), as discussed in the Challenges section.
 
@@
 
-al., R.H. et. (2014). \emph{phylobase: Base package for phylogenetic
-structures and comparative data}. Retrieved from
-\url{http://CRAN.R-project.org/package=phylobase}
-
 Boettiger, C. (2014). \emph{knitcitations: Citations for knitr markdown
@@
 \url{http://dx.doi.org/10.12688/f1000research.2-191.v2}
-
-CRAN task view: Phylogenetics, especially comparative methods. (2014).
-Retrieved from
-\url{http://cran.r-project.org/web/views/Phylogenetics.html}
 
@@
 \url{http://dx.doi.org/10.1080/10635150390192780}
+
+NESCENT R Hackathon Team. (2014). \emph{phylobase: Base package for
+phylogenetic structures and comparative data}. Retrieved from
+\url{http://CRAN.R-project.org/package=phylobase}
+
+O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
+comparative methods. Retrieved from
+\url{http://cran.r-project.org/web/views/Phylogenetics.html}
 

### 2014-10-06 manuscripts/manuscript.Rmd
Edits for the abstract.
Mostly smoothing flow and enhancing clarity. No substantive changes.

@@
 abstract: | 
-      1. NeXML is a new phylogenetic informatics data and metadata exchange 
-      format that offers substantial improvements in reliability, extensibility, 
-      and richness over earlier formats such as NEXUS. In this paper we present 
-      the RNeXML package which provides a bridge between this rich and reliable
-      data format and the extensive library of phylogenetic tools already
-      available in R, with a particular emphasis on comparative methods.
+      1. NeXML is a new exchange format for phylogenetic data and
+      metadata that offers substantial improvements in reliability,
+      extensibility, and richness over earlier formats such as
+      NEXUS. In this paper we present the RNeXML package which
+      provides a bridge between this data format and the extensive
+      library of phylogenetic tools already available in the popular
+      data analysis environment and programming language R, with a
+      particular emphasis on comparative methods.
 
-      2. RNeXML is much richer than a basic parser: here we illustrate
-      how the package facilitates adding and manipulating semantic 
-      annotations, that is annotations whose meaning can be defined externally
-      by way of semantic web technologies.
+      2. The ability to parse and write NeXML from within R is only a
+      small part of the functionality offered by the package.  We
+      illustrate how RNeXML facilitates adding and manipulating data
+      annotations whose meaning can be defined in a precise and
+      machine-interpretable way by means of formal ontologies and other
+      semantic web technologies.
 
-      3. An essential feature of NeXML is to provide both the flexibility
-      to be continually extended to meet the emerging needs of the phylogenetics
-      community while remaining backwards-compatible with software developed
-      for earlier versions.  Here we illustrate how RNeXML makes it easy
-      <!-- earlier versions of what? -->
-      to extend NeXML in this manner.
+      3. An essential feature of NeXML is to provide both the
+      flexibility to be continually extended to meet the emerging
+      needs of the phylogenetics community without breaking software
+      unaware of such extensions. We illustrate how RNeXML makes
+      it easy to extend NeXML in this manner.
 
-      4. RNeXML is well suited for users to preserve and maintain their
-      personal archives of phylogenetic and comparative trait data in consistent,
-      query-able, and cross-platform format, to leverage the rapidly growing 
-      public community archives of rich NeXML files, and to interact with web
-      services that deliver their results (with their provenance) as NeXML.
+      4. RNeXML is well suited for users to preserve and maintain
+      their personal archives of phylogenetic and comparative trait
+      data in consistent, query-able, and cross-platform format, to
+      leverage the rapidly growing public community archives of rich
+      NeXML files, and to interact with web services that return data
+      and their metadata packaged together as NeXML.
 

### 2014-10-06 R/classes.R
Update CDAO URI (#92) and rebuild README, manuscript, etc

@@ -571,7 +571,7 @@ nexml_namespaces <-
   c("nex"   = "http://www.nexml.org/2009",
     "xsi"   = "http://www.w3.org/2001/XMLSchema-instance",
     "xml"   = "http://www.w3.org/XML/1998/namespace",
-    "cdao"  = "http://www.evolutionaryontology.org/cdao/1.0/cdao.owl#",
+    "cdao"  = "http://purl.obolibrary.org/obo/cdao.owl",
     "xsd"   = "http://www.w3.org/2001/XMLSchema#",
     "dc"    = "http://purl.org/dc/elements/1.1/",
   "dcterms" = "http://purl.org/dc/terms/",

### 2014-11-04 manuscripts/manuscript.Rmd

@@
 within R, these formats frequently contain more information than their
-NEXUS file counterparts, and users lack a suitable alternative for
+NEXUS file counterparts,
+<!-- such as what? we haven't talked about metadata yet. HL -->
+and users lack a suitable alternative for
 archiving and sharing phylogenetic data outside of the R ecosystem. 
@@
 easily be shared, archived, and made available to other software;
+<!-- such as what? should walk through an example here. HL -->
 but inherits the limitations of the NEXUS format.
+<!-- how can information inherit the limitations of a serialization -->
+<!-- format if it's added to data read from that format? HL -->
 
@@
 -->
+<!-- I think this is in fact more important than the paragraph above. I.e., -->
+<!-- whether exchanging basic comparative data needs another format is -->
+<!-- debatable, but that metadata that can hardly be expressed in -->
+<!-- current standards does is a clear point. HL -->
 
@@
 extend the NeXML standard to describe stochastic character mapping, one
-of the kinds of data annotation that is currently available in R formats
+of the kinds of data annotation that is currently available in R
+formats
+<!-- R packages, not formats, right? HL -->
 but lacks a consistent, extensible export format. We illustrate how NeXML
@@
 
-A NeXML document can be thought of as having multiple components: (1) phylogeny
-data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
+A NeXML document can be thought of as having the following components: (1) phylogeny
+data, (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
 `RNeXML` seeks to provide a seamless way to convert these to and from their 
-corresponding R object representations, such as `ape::phylo` phylogenies, character
-matrices (`data.frames`) and metadata lists (or for advanced users,
-`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
-information about the study in which the phylogeny was generated or
+corresponding R object representations, in particular `ape::phylo`,
+`data.frame`, and lists for phylogenies, character
+matrices, and metadata, respectively.
+<!-- I feel the following is a little early - it will likely be over -->
+<!-- most most readers' heads. HL
+(or for advanced users,
+`XMLInternalDocument` and `rrdf` triplestores).
+-->
+Metadata can provide bibliographic and provenance
+information about the study as part of which the phylogeny was generated or
 applied (such as authors, citations to publications, timestamps, and even
-descriptions of methods), but can also be used to annotate individual
-data components.  We illustrate how metadata annotations can be added to
+descriptions of methods). They can also be used to attach annotations to individual
+data elements.  We illustrate how metadata annotations can be added to
 the operational taxonomic units of the phylogeny and can be used to extend
 the data representation to new uses such as stochastic character maps.
+<!-- sounds awkward - need to smoothen the language. HL -->
 
@@
 ```
-
-The resulting `nex` object is an R object representation of the NeXML. This is 
-a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
-character traits, and metadata), analogous to the `ape::phylo` object from the 
-popular `ape` R package [@Paradis_2004]. While R functions and packages 
-could be written to use this `nexml` object directly, most users will prefer to use
-already existing methods provided by packages they know. From a `nexml` object, we can 
+<!-- The system.file() command is not pretty. Can we not just read in -->
+<!-- the file directly? HL -->
+The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
+file. <!-- This seems redundant to me to what has already been -->
+<!-- said. HL
+Objects of type `nexml` represent phylogenetic trees as well as 
+character traits, and metadata, analogous to the `ape::phylo` object from the 
+popular `ape` R package [@Paradis_2004]. --> Although new custom R functions and packages 
+could be written for using `nexml` objects, for most users it will be
+more convienent to use methods of the many already existing R packages for 
+phylogenetics. From a `nexml` object, we can 
 therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
@@
 blocks are available, it will return a list of `multiPhylo` blocks.
+<!-- having a function's return type being more or less unpredictable -->
+<!-- sounds like a bad design pattern. Is there at least precedence for -->
+<!-- this practice in other R packages? HL --> 
 While most users will find this behavior convenient, there are times
@@
 a user can access all the phylogenies available in 
-[TreeBASE](http://treebase.org/treebase-web/home.html) <!-- first mention -->
+[TreeBASE](http://treebase.org/) <!-- first mention -->
 <!-- of TreeBase, include URL? or introduce it in intro? --> through the
@@
 ```
-
-Previously, this was possible using the `treebase` package [@Boettiger_2012]
-for R, which relied on the NEXUS parser provided in the `ape` package.
-Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
+<!-- the example makes me wonder how nexml_read() distinguishes -->
+<!-- between a URL and a file name as its argument. HL -->
+Reading phylogenies from TreeBASE into R was already possible using
+the R package `treebase` [@Boettiger_2012]. However, because it relied
+on the NEXUS parser provided in the `ape` package, and because of
+incompatible flavors of the NEXUS standard [@Vos_2012], 
 a significant number <!-- State explicitly --> of TreeBASE NEXUS
-files could not be read in this way.  Using the more precisely defined
-NeXML standard, every valid NeXML document on TreeBASE can now successfully
-be parsed to R objects.
+files could not be read in this way.  Owing to the syntactically strict
+definition of the
+NeXML standard, the RNeXML package can successfully read every valid
+NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
+<!-- it can't read? the sentence as is suggests so, and so is a little -->
+<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
+<!-- can't read are now invalid NeXML files. Which is probably not -->
+<!-- what you mean to suggest. HL -->
 
@@
 
-Like NEXUS files, NeXML documents can also contain character data.  While
-most researchers will be familiar with the use of character matrices in
-NEXUS files as a container for molecular sequence (or morphological) data 
-used to infer the phylogenetic tree, we find this ideal for character data used in
-comparative methods popular in R. Methods to extract character data
-work much like those for the phylogenetic data. Here we read in an example
-NeXML file that contains both a phylogeny and character data. 
-`get_characters()` returns a `data.frame` with columns as characters and rows as
+Like NEXUS files, NeXML documents can in addition to trees also
+contain character data in the form of matrices.  Character data can be
+of molecular (molecular sequence alignments), discrete (most
+morphological character data), or continuous type (many trait data). 
+Often character data, especially molecular and morphological, is used
+for phylogenetic tree reconstruction, but they are also a common input
+for the many comparative phylogenetic analysis methods available in R. Methods to extract character data
+work much like those for extracting the phylogenetic trees. Here we read in an example
+NeXML file that contains both a phylogeny and character data. The
+function
+`get_characters()` can then be used to obtain a `data.frame` with columns as characters and rows as
 taxa:
@@
 ```
-
-As this example shows, continuous and discrete characters are represented as separate
+<!-- same as above - why can this not be one call? HL -->
+Continuous and discrete character data are represented as separate
 blocks in NeXML files, but will be combined as separate columns of a
 single data frame by `get_characters()` if they correspond to the same
-taxa. The `get_characters_list()` function will consistently receive
-a list of data frames corresponding to each character matrix block of
-the NeXML.
+taxa. If this behavior is undesired, the function
+`get_characters_list()` can be used to obtain
+a list of data frames, where list elements correspond to the character matrix blocks of
+the NeXML file.
 
@@
 
-The first argument to `nexml_write()` can be a `ape::phylo` phylogeny,
-`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
-can be coerced into the `nexml` class.  As we have seen, NeXML files
-can contain multiple phylogenetic trees, as well as character matrices.
-We can add both trees and character data using `nexml_write()`,
+The first argument to `nexml_write()` is a `nexml` object, or any
+object that can be coerced to it, such as a `ape::phylo` phylogeny,
+or a `multiPhylo` list of phylogenies. The `nexml_write()` function
+also allows to add character data:
 
@@
 
-NeXML files can also be built up bit by bit using `add_characters()`,
+Instead of packaging trees and characters at the time of writing a
+NeXML file, the document can also be built up step by step using the
+functions `add_characters()`,
 `add_trees()`, and `add_meta()`.  Each of these functions can add data to an
-existing `nexml` object or create a new nexml object if none is provided.
+existing `nexml` object, or create a new nexml object if none is provided.
 Here we first create a `nexml` object containing the phylogeny data,
@@
 
-This allows us to include multiple trees and character matrices to a single NeXML file.
+This allows us to include multiple trees and character matrices to a
+single NeXML file.
+<!-- perhaps give a motivating use case for why one might -->
+<!-- want to do this? HL -->
 
@@
 backwards compatibility with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations", or RDFa,
+"Resource Description Framework in Annotations" (RDFa),
 as described in @Vos_2012 or @W3C_2014. 
@@
 
-Researchers cannot take advantage of NeXML metadata manipulation if
-there is no metadata there.  The burden of adding good metadata remains a
+The burden typically involved with documenting and sharing good
+metadata alongside the data remains a
 significant barrier to increased sharing and reuse of scientific research
-[@Tenopir_2011].
-`RNeXML` seeks to reduce the burden of adding
-high quality, machine-readable metadata to NeXML files. This occurs on
+[@Tenopir_2011]. To promote sharing of well-documented phylogenetic data,
+`RNeXML` aims to provide a simple interface to adding well-defined machine-readable metadata to NeXML files. This occurs on
 several levels:
@@
 
-`RNeXML` automatically adds basic metadata by default such as a time-stamp
-of when the tree was produced and an open data license (CC0, compatible
-with deposition on Dryad data repository), which can be configured
-as needed.
+`RNeXML` automatically adds basic metadata by default
+<!-- what does "by default" mean here? It doesn't seem like calling -->
+<!-- any RNeXML function will add such metadata if it's not there -->
+<!-- already? HL -->
+such as a time-stamp
+of when a tree was produced, and an open data license. The latter by
+default is the Creative Commons Zero (CC0) public domain waiver, which
+is compatible
+with deposition in the Dryad data repository, but can be configured
+as a different license if needed.
 
-To adjust these and other common metadata fields that should 
-be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
-function.  This function does not assume any familiarity with the concepts
-of namespaces or other more advanced metadata tools discussed later.  The
-`add_basic_meta()` function takes an existing `nexml` object (or will create a new one
-if none is provided, like the other `add_` functions), and any of the following
-fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
-`citation`.  Additional fields may be added to subsequent versions, see
-`?add_basic_metadata` for the most up-to-date information.  Here we create
-a new `nexml` object with the basic metadata for the `bird.orders` data
-provided in the `ape` package:
+To adjust these and other metadata fields commonly recommended to be
+included with publicly archived data, `RNeXML` provides the `add_basic_meta()`
+function.  This function hides the complexities of anchoring metadata
+attribute names in common vocabularies from the user.  The
+`add_basic_meta()` function takes as input an existing `nexml` object
+(like the other `add_` functions, if none is provided it will create one), and at the time
+of this writing any of the following
+parameters: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
+`citation`.  Other metadata elements and corresponding parameters may
+be added in the future.
 
+Here we create a `nexml` object for the `bird.orders` data
+provided in the `ape` package, and then add basic metadata:
+<!-- I notice that other add_* methods take the nexml object as the -->
+<!-- last argument, not the first as seems to be the case here. This -->
+<!-- is inconsistent and should be fixed. HL -->
 ```{r}
@@
 
-Citations can also be added using R's `bibentry` type, which allows
-a user to add citations to R packages,
+Instead of a literal string, citations can also be provided in R's
+`bibentry` type, which is the one in which R package citations are obtained:
 
@@
 
-or to papers by simply using the object's DOI (using @Boettiger_2014).
-For instance to add the citation information of the paper that generated the
+Alternatively, a citation to a published paper with a Digital
+Object Identifier (DOI) can be provided in the form of the DOI,
+which the package `knitcitations` [@Boettiger_2014] can turn into a
+formatted citation using. As an example, to add the citation information of the paper that generated the
 `geospiza` phylogeny included in the `geiger` package:
@@
 inconsistent use of taxonomic names in phylogenetic data are a common
-source of frustration in reusing phylogenetic trees, especially when
+source of frustration when reusing phylogenetic trees, especially when
 integrating data from different sources: trees might come from one place

### 2014-11-04 manuscripts/manuscript.Rmd
merge, edit and reply to @hlapp's edits

@@
 
-<!-- FIXME
-
-ADD STUFF ABOUT:
-
-- Rapid growth in phylogenetics data (+citations)
-- Rapid growth of research methods and software that use phylogenetic data. Comparative methods in evolution. Ecological phylogenetics. Biodiversity informatics. (+citations)
-- More on data sharing challenges and opportunities a la @Drew_2013
-- More on the hoops that users need to jump through if trees, data
-and other things all need to come from separate files and need to be
-joined on taxon labels.
-
--->
-
 The ever expanding needs of the phylogenetics community are rapidly
@@
 
-<!-- What do we need to say, if anything, about the current R formats?  about phylobase?  -->
-
 The R language [@R] has become an important environment for 
@@
 with over 1000 citations [@taskview],
-<!-- hmm, taskview citation renders as "'CRAN task view' 2014" intended? -->
 though stand-alone programs remain an important part of this ecosystem
@@
 <!-- format if it's added to data read from that format? HL -->
+<!-- CB: Sorry, was refering to data _not_ read from NEXUS format,
+but rather data created in an R format/data-structure.  Information
+that is in that R data structure (e.g. a simmap) is lost when using
+write.nexus(simmap) -->
 
@@
 of R packages currently available for handling phylogenetic data. 
-<!-- Carl, I assume a period goes here, I inserted one -->
 R packages and R users frequently add rich information to the phylogenies
@@
 many popular phylogenetic comparative methods tools in R.
+
 `RNeXML` also provides convenient tools for adding metadata
@@
 extend the NeXML standard to describe stochastic character mapping, one
-of the kinds of data annotation that is currently available in R
-formats
-<!-- R packages, not formats, right? HL -->
+of the kinds of data annotation that is currently available in data structures
+provided by existing R packages
 but lacks a consistent, extensible export format. We illustrate how NeXML
@@
 
-<!-- I still question the wisdom of having multiple names for the same
-method. At first this might seem user-friendly, but it will just be confusing
-and create a document and source code maintenance nightmare. Why not just pick
-read.nexml and be done with it? -->
-<!-- For whatever reason, R has native support for aliasing function names;
-     so I don't see this creating a maintenance/documentation issue.  If I 
-     had to pick, I'd go with read_nexml though; a lot of the R developer 
-     community strongly opposes the `.` in function names, since it falsely
-     mimics the S3 class structure.  - CB -->
-
 We begin by loading the `RNeXML` package and reading in an example
@@
 library("RNeXML")
-f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
-```
-<!-- The system.file() command is not pretty. Can we not just read in -->
-<!-- the file directly? HL -->
+nex <- nexml_read("components/trees.xml")
+```
 The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
-file. <!-- This seems redundant to me to what has already been -->
-<!-- said. HL
+file. <!-- This seems redundant to me to what has already been 
+said. HL -->
+<!-- CB: Tried to revise what follows to emphasize that we have a new data structure / object class here,
+that has unique merits -->
 Objects of type `nexml` represent phylogenetic trees as well as 
-character traits, and metadata, analogous to the `ape::phylo` object from the 
-popular `ape` R package [@Paradis_2004]. --> Although new custom R functions and packages 
+character traits, and semantic metadata. As such, this new R object format differs
+from existing formats such as `ape::phylo` format [@Paradis2004],
+which describes trees alone, or any of the `phylobase` classes of object that contain 
+trees and traits but more limited metadata [@phylobase].
+
+Although new custom R functions and packages 
 could be written for using `nexml` objects, for most users it will be
@@
 <!-- this practice in other R packages? HL -->
+
+<!-- Yes, sadly it is.  Really this is a class definition issue that multiPhylo isn't a
+proper generalization of a phylo object, but rather a different class entirely. 
+Users who don't like this function can use the get_trees_list() to always be predictable.  -->
+
 While most users will find this behavior convenient, there are times
@@
 a user can access all the phylogenies available in 
-[TreeBASE](http://treebase.org/) <!-- first mention -->
-<!-- of TreeBase, include URL? or introduce it in intro? --> through the
+[TreeBASE](http://treebase.org/) 
+ through the
 `nexml` format.  `RNeXML` can read directly from a URL:
@@
 <!-- between a URL and a file name as its argument. HL -->
+<!-- just type `nexml_read` and you'll find out ;-).  We use a simple regex  CB -->
+
 Reading phylogenies from TreeBASE into R was already possible using
@@
 definition of the
-NeXML standard, the RNeXML package can successfully read every valid
+NeXML standard, the RNeXML package can successfully read every 
 NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
@@
 <!-- what you mean to suggest. HL -->
+<!-- Correct, there are indeed a few invalid ones (e.g. some are purely
+erroneous HTML, see issues on supertreebase repo) which it cannot
+parse.  But like you say, no need to confuse the reader.  I've dropped the 
+word 'valid'.  -->
 
@@
 ```{r }
-comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
-nex <- nexml_read(comp_analysis)
+nex <- nexml_read("components/comp_analysis.xml")
 get_characters(nex)
@@
 <!-- same as above - why can this not be one call? HL -->
+<!-- CB: I've dropped the system.file, even though that is sadly the 
+standard way for R to load data that comes with a package but is not in CSV, R, or .rda format.
+
+Indeed `get_characters` could call the file path directly, but there is key
+conceptual reason for not doing this.  Note that all the examples call nexml_read to get a nex object
+before using get_characters, get_trees, etc.  That is because we really want the user to get used
+to manipulating the nexml R data structure, rather than using the external XML file directly.
+
+The functions embody this design principle deep down.  We could have written get_characters etc
+in a way that it just yanked the characters out of the XML and left everything else, rather than 
+converting the whole NeXML file into a serialized R object stored in memory.  But the latter strategy
+gives us a native R object that people can extend, write functions for, and subset and query with 
+the standard R syntax.  Without this, the package isn't really providing a native format, and users 
+and developers are left on their own to either also work directly with the XML or just convert immediately
+into a more limited R format like ape::phylo.  
+-->
+
 Continuous and discrete character data are represented as separate
@@
 
-<!-- in caper there is this method comparative.data to join data
-and trees on taxon labels. it would be nice if this merged data
-structure could also be passed to nexml_write (or should that be
-write.nexml?) -->
-<!-- Great observation.  I think we should add a coercion method
-from the caper structure to the nexml S4 structure.  Then nexml_write
-can automatically handle this. added as issue now  -->
-
 The first argument to `nexml_write()` is a `nexml` object, or any
@@
 These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].
-
-<!-- how about add_taxa? -->
-<!-- A taxa block is added when trees or characters are added.  Yeah,
-     I guess it would make sense to be able to add taxa directly,
-     (e.g. a NeXML file containing only RDFa metadata about taxa, 
-     independent of characters or trees), but I worry about confusing
-     people into thinking they must add taxa blocks separately, and
-     about when we should merge or not merge taxa blocks of matching
-     species. See https://github.com/ropensci/RNeXML/issues/43 
-     Added new issue for this, https://github.com/ropensci/RNeXML/issues/85 -->
-
 Instead of packaging trees and characters at the time of writing a
@@
 This allows us to include multiple trees and character matrices to a
-single NeXML file.
-<!-- perhaps give a motivating use case for why one might -->
-<!-- want to do this? HL -->
+single NeXML file, for instance, to represent samples from posterior
+probability distributions of inferred or simulated phylogenies or
+character states.
 
@@
 
-<!-- in the Perl library I also embed the value of the environment variable
-$USER as a document-level annotation -->
-<!-- Good idea-->
-
 `RNeXML` automatically adds basic metadata by default
@@
 <!-- already? HL -->
+<!-- Indeed it does.  We add the time stamp, default license, and user ID even though
+     it's not "there" in the data.  -->
 such as a time-stamp
@@
 <!-- is inconsistent and should be fixed. HL -->
+
+<!-- Not so, the nexml object is the second argument here too.  If no existing 
+nexml object is given, the add function starts a new nexml object.  If one is
+given, it appends it's first argument onto the existing nexml, as described above.
+
+Notice that bird.orders is an ape::phylo object, which we are adding to a new 
+nexml object.  (R allows for default arguments, the default second argument is 
+new("nexml") -->
+
 ```{r}

### 2014-11-21 manuscripts/manuscript.Rmd
Fix the add_basic_meta argument order

@@
 birds <- add_trees(bird.orders)
-birds <- add_basic_meta(birds,
+birds <- add_basic_meta(
   title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
@@
 
-  creator = "Sibley, C. G. and Ahlquist, J. E.")
+  creator = "Sibley, C. G. and Ahlquist, J. E.",
+  nexml = birds)
 ```
@@
 ```{r}
-birds <- add_basic_meta(birds, citation = citation("ape"))
+birds <- add_basic_meta(citation = citation("ape"), nexml=birds)
 ```
@@
 library("knitcitations")
-geiger_nex <- add_basic_meta(geiger_nex, citation = bib_metadata("10.2307/2408428"))
+geiger_nex <- add_basic_meta(citation = bib_metadata("10.2307/2408428"), nexml=geiger_nex)
 ```

### 2014-11-21 manuscripts/manuscript.Rmd
Resolved a number of comments during conference call.

--- 2014-11-04 manuscripts/manuscript.Rmd
@@
 
-<!-- Map paragraph -->
-
 In this paper, we introduce the reader to the concepts and use of
@@
 matrices, and metadata, respectively.
-<!-- I feel the following is a little early - it will likely be over -->
-<!-- most most readers' heads. HL
-(or for advanced users,
-`XMLInternalDocument` and `rrdf` triplestores).
--->
 Metadata can provide bibliographic and provenance
@@
 The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
-file. <!-- This seems redundant to me to what has already been 
-said. HL -->
-<!-- CB: Tried to revise what follows to emphasize that we have a new data structure / object class here,
-that has unique merits -->
+file. 
 Objects of type `nexml` represent phylogenetic trees as well as 
@@
 incompatible flavors of the NEXUS standard [@Vos_2012], 
-a significant number <!-- State explicitly --> of TreeBASE NEXUS
-files could not be read in this way.  Owing to the syntactically strict
+a sizable number of TreeBASE NEXUS
+files could not be read in this way <!-- CB to rephrase to make a -->
+<!-- broader point and refer back to arguments made in the NeXML -->
+<!-- paper, possibly moving this larger point to the introduction -->.  Owing to the syntactically strict
 definition of the
 NeXML standard, the RNeXML package can successfully read every 
-NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
-<!-- it can't read? the sentence as is suggests so, and so is a little -->
-<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
-<!-- can't read are now invalid NeXML files. Which is probably not -->
-<!-- what you mean to suggest. HL -->
-<!-- Correct, there are indeed a few invalid ones (e.g. some are purely
-erroneous HTML, see issues on supertreebase repo) which it cannot
-parse.  But like you say, no need to confuse the reader.  I've dropped the 
-word 'valid'.  -->
+NeXML document from TreeBASE.
 
@@
 ```
-<!-- same as above - why can this not be one call? HL -->
-<!-- CB: I've dropped the system.file, even though that is sadly the 
-standard way for R to load data that comes with a package but is not in CSV, R, or .rda format.
-
-Indeed `get_characters` could call the file path directly, but there is key
-conceptual reason for not doing this.  Note that all the examples call nexml_read to get a nex object
-before using get_characters, get_trees, etc.  That is because we really want the user to get used
-to manipulating the nexml R data structure, rather than using the external XML file directly.
-
-The functions embody this design principle deep down.  We could have written get_characters etc
-in a way that it just yanked the characters out of the XML and left everything else, rather than 
-converting the whole NeXML file into a serialized R object stored in memory.  But the latter strategy
-gives us a native R object that people can extend, write functions for, and subset and query with 
-the standard R syntax.  Without this, the package isn't really providing a native format, and users 
-and developers are left on their own to either also work directly with the XML or just convert immediately
-into a more limited R format like ape::phylo.  
--->
-
 Continuous and discrete character data are represented as separate
@@
 provided in the `ape` package, and then add basic metadata:
-<!-- I notice that other add_* methods take the nexml object as the -->
-<!-- last argument, not the first as seems to be the case here. This -->
-<!-- is inconsistent and should be fixed. HL -->
-
-<!-- Not so, the nexml object is the second argument here too.  If no existing 
-nexml object is given, the add function starts a new nexml object.  If one is
-given, it appends it's first argument onto the existing nexml, as described above.
-
-Notice that bird.orders is an ape::phylo object, which we are adding to a new 
-nexml object.  (R allows for default arguments, the default second argument is 
-new("nexml") -->
 
@@
 ```
+<!-- the above will get fixed by CB, this is just a reminder and -->
+<!-- placeholder to adjust when that's done. -->
 

### 2014-11-21 manuscripts/manuscript.Rmd
Some edits from morning's chat
- Clarify scripted use vs interactive use in `get_trees`, `get_trees_list`
- Add future directions
- a few minor tweaks

Still need to revise introduction to emphasize semantics.

@@
 [@Xie_2013].  The reader may paste the code blocks
-shown into their R terminal to follow along with the exmples shown.
+shown into their R terminal to follow along with the examples shown.
 
@@
 
-NeXML can not only contain multiple phylogenetic trees in a single document,
+Note that NeXML can not only contain multiple phylogenetic trees in a single document,
 but can also group sets of trees into separate blocks (e.g. as resulting from
-different workflows). The `get_trees()` function attempts to return the most sensible 
-R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
+different workflows).
+The `get_trees()` function is designed for interactive use, and thus attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo`
 object; if multiple trees are available in a single `trees` block, it will return
@@
 blocks are available, it will return a list of `multiPhylo` blocks.
-<!-- having a function's return type being more or less unpredictable -->
-<!-- sounds like a bad design pattern. Is there at least precedence for -->
-<!-- this practice in other R packages? HL -->
 
-<!-- Yes, sadly it is.  Really this is a class definition issue that multiPhylo isn't a
-proper generalization of a phylo object, but rather a different class entirely. 
-Users who don't like this function can use the get_trees_list() to always be predictable.  -->
-
-While most users will find this behavior convenient, there are times
-(such as looping over many NeXML files) when having a variable return
-type is undesirable.  To always get back a list of `multiPhylo` objects,
+For scripted use, it is desirable to have a consistent return type.  To always get back a list of `multiPhylo` objects,
 even when only one tree is available, use `get_trees_list()`.
@@
  through the
-`nexml` format.  `RNeXML` can read directly from a URL:
+`nexml` format.  `RNeXML`'s `nexml_read` can also read a file directly from a URL by automatically recognizing the URL patterns and first downloading the file to a working directory:
 
@@
 ```
-<!-- the example makes me wonder how nexml_read() distinguishes -->
-<!-- between a URL and a file name as its argument. HL -->
-<!-- just type `nexml_read` and you'll find out ;-).  We use a simple regex  CB -->
 
@@
 
+## Future Directions
+
+For convenience, the RNeXML package can parse compressed XML binaries (gzipped XML) directly, facilitating work with very large data files.
+As most R functions, especially those in common phylogenetics packages, require in-memory data structures, RNeXML currently loads this data directly into memory.
+Future work could facilitate manipulation of NeXML data directly from the disk.
+
+Emerging formats such as the JSON based NeXSON format provide alternate representations of semantic, machine-readable phylogenetic data.
+At this time, R users must still rely on other tools that can convert these formats into NeXML.
+Because the NeXML syntax is extensible, semantic, and machine-readable, this can be done without loss, unlike NEXUS formats.
+Nevertheless, a possible future direction would be to provide direct support of such formats through RNeXML or a companion R package.
+
 ## Acknowledgements

### 2014-11-21 manuscripts/manuscript.Rmd
Emphasize semantics in the intro
and lots of other revisions trying to be more clear about the advantages of NeXML over NEXUS.

@@
   - Overview of NeXML
+  - Challenges with R
   - Map
@@
 comparative phylogenetic methods tools, R users lack a convenient,
-extensible format to exchange and archive comparative methods data.
-A common file format for sharing phylogenetic tree data has long played
-an essential role in the advancement of phylogenetics research.
+extensible, and semantically aware format to import, exchange and archive 
+phylogenetic trees, traits, and metadata.
 The rapid increase in phylogenetics data and research methods that make
 use of such data has introduced many new challenges and opportunities 
-to sharing this information both between software environments and 
+for sharing, both between software environments and 
 between research groups [@Drew_2013; @Stoltzfus_2012].
-
 The ever expanding needs of the phylogenetics community are rapidly
@@
 represent horizontal gene transfer or ambiguity in reading a character
-(such as a DNA sequence base pair).  This has led developers to
-repeatedly add small modifications to the standard to meet the needs
-of their own software, which results in mutually incompatible dialects
-that largely defeat the advantages of a shared  file format.
-NeXML was developed to addresses each of these problems [@Vos_2012]  
-by drawing on robust web standards in data representation.
-The design and expressiveness of NeXML makes it much easier to meet
+(such as a DNA sequence base pair).  This has led many developers to
+modify NEXUS in different ways for different needs, with the unfortunate
+result that NEXUS files generated by one program can be incompatible
+with another [@Vos_2012]. Without a formal grammar, software based on 
+NEXUS files may also make inconsistent assumptions about tokens, quoting,
+or element lengths.  @Vos_2012 estimates as many as 15% of the NEXUS
+files in the CIPRES portal contain unrecoverable but hard to diagnose errors.
+
+NeXML was developed to addresses each of these problems and also open doors
+to new opportunities in handling phylogenetic data. First, NeXML is defined
+by a precise grammar that can be programmatically **validated**. This helps
+ensure that any NeXML file can always be read without errors by any 
+NeXML parsing software. Second, NeXML is **extensible**: a user can define
+representations of new information (as we will illustrate) without breaking
+this compatibility. Third and most importantly, NeXML is **semantically rich**.
+The use of formal semantics in NeXML provides a representation of data that
+is not only machine-readable, but can also permits machine reasoning about
+the data provided (see @Parr2011 for an overview on semantic reasoning in 
+evolutionary informatics). Instead of asking the computer to
+return "the data on line #93", we can expect the computer to reason about
+taxonomic concepts (as we illustrate in the supplement).
+
+The R language [@R] has become an important environment for phylogenetics
+research, with over 30 packages available including several with over 
+1000 citations [@taskview]. `RNeXML` makes the large and rapidly growing
+archive of richly annotated, NeXML-based phylogenetics data available to
+the R user and this powerful ecosystem of R tools for phylogenetics.
+Further, R packages and users frequently add their own additional 
+information to the phylogenies they work with: annotations of species,
+stochastic character maps, trait values, model estimates and parameter 
+values. By leveraging the semantic and extensible nature of NeXML, `RNeXML` 
+provides R users a platform independent data exchange format to capture,
+further annotate, exchange, and preserve this information. With `RNeXML`,
+R users and developers can both better contribute to and take advantage 
+of the rapidly growing infrastructure of evolutionary informatics [@Parr2011].  
+By capturing information already available in the R environment, together
+with the appropriate choice of defaults, `RNeXML` can facilitate meeting
 the rising bar for best practices in data sharing in phylogenetic 
-research [@Cranston_2014].
-
-The R language [@R] has become an important environment for 
-phylogenetics research, with over 30 packages available including several
-with over 1000 citations [@taskview],
-though stand-alone programs remain an important part of this ecosystem
-for tasks such as generating phylogenetic trees from sequence data.  Within 
-this ecosystem a variety of different data formats for
-phylogenetic data have been developed and modified. Most but not all 
-are based on the simple and popular `ape::phylo` format provided by the
-ape package [@Paradis_2004].
-
-Though certain functions allow R users to switch between these formats
-within R, these formats frequently contain more information than their
-NEXUS file counterparts,
-<!-- such as what? we haven't talked about metadata yet. HL -->
-and users lack a suitable alternative for
-archiving and sharing phylogenetic data outside of the R ecosystem. 
-Consequently, information added to phylogenies by these R programs cannot 
-easily be shared, archived, and made available to other software;
-<!-- such as what? should walk through an example here. HL -->
-but inherits the limitations of the NEXUS format.
-<!-- how can information inherit the limitations of a serialization -->
-<!-- format if it's added to data read from that format? HL -->
-<!-- CB: Sorry, was refering to data _not_ read from NEXUS format,
-but rather data created in an R format/data-structure.  Information
-that is in that R data structure (e.g. a simmap) is lost when using
-write.nexus(simmap) -->
-
-`RNeXML` opens up a rich world of both currently available and rapidly
-growing data sets in the NeXML format to the R user and the large ecosystem
-of R packages currently available for handling phylogenetic data. 
-R packages and R users frequently add rich information to the phylogenies
-they work with, but need a platform independent data exchange format to
-share and archive this. We illustrate how NeXML can
-be used to distribute both phylogenies and character trait data used in 
-many popular phylogenetic comparative methods tools in R.
-
-`RNeXML` also provides convenient tools for adding metadata
-to NeXML files, including the automated generation of metadata and examples
-of how to extend the NeXML standard to represent new data features without
-breaking validation or backwards compatibility of the file type.
-
-<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
-and how RNeXML provides a solid platform from which to do so.
--->
-<!-- I think this is in fact more important than the paragraph above. I.e., -->
-<!-- whether exchanging basic comparative data needs another format is -->
-<!-- debatable, but that metadata that can hardly be expressed in -->
-<!-- current standards does is a clear point. HL -->
+research [@Cranston_2014] while at the same reducing the burden
+on the user.
 
@@
 informatics that allow users to express and interpret complicated data
-structures automatically.  We end with an example that uses metadata to
-extend the NeXML standard to describe stochastic character mapping, one
-of the kinds of data annotation that is currently available in data structures
+structures automatically.  We end with an example that uses semantic metadata to
+extend the NeXML standard to describe stochastic character mapping: a kind of
+data annotation that is currently available in data structures
 provided by existing R packages

### 2014-11-21 manuscripts/manuscript.Rmd
clarify which function 'automatically adds metadata'

@@
 
-Reading phylogenies from TreeBASE into R was already possible using
-the R package `treebase` [@Boettiger_2012]. However, because it relied
-on the NEXUS parser provided in the `ape` package, and because of
-incompatible flavors of the NEXUS standard [@Vos_2012], 
-a sizable number of TreeBASE NEXUS
-files could not be read in this way <!-- CB to rephrase to make a -->
-<!-- broader point and refer back to arguments made in the NeXML -->
-<!-- paper, possibly moving this larger point to the introduction -->.  Owing to the syntactically strict
-definition of the
-NeXML standard, the RNeXML package can successfully read every 
-NeXML document from TreeBASE.
-
 ## Reading character data
@@
 
-`RNeXML` automatically adds basic metadata by default
-<!-- what does "by default" mean here? It doesn't seem like calling -->
-<!-- any RNeXML function will add such metadata if it's not there -->
-<!-- already? HL -->
-<!-- Indeed it does.  We add the time stamp, default license, and user ID even though
-     it's not "there" in the data.  -->
-such as a time-stamp
-of when a tree was produced, and an open data license. The latter by
-default is the Creative Commons Zero (CC0) public domain waiver, which
-is compatible
-with deposition in the Dryad data repository, but can be configured
-as a different license if needed.
+`RNeXML`'s `nexml_write` function adds certain basic metadata by automatically
+if it has not yet been provided; using default values
+consistent with recommended best practices [@Cranston_2014]. At this time,
+this includes naming the software generating the NeXML, a time-stamp of
+when a tree was produced, and an open data license.
 
-To adjust these and other metadata fields commonly recommended to be
-included with publicly archived data, `RNeXML` provides the `add_basic_meta()`
+To adjust these values and also add other metadata fields commonly recommended for
+includion with publicly archived data, `RNeXML` provides the `add_basic_meta()`
 function.  This function hides the complexities of anchoring metadata

### 2014-11-27 manuscripts/manuscript.Rmd
Edits to and partial rewrite of the Introduction.
Overall, I tried to provide for better flow, less redudancy, and clearer
motivation preceding or at least quickly following all statements about
features and capabilities.

@@
 
-Though R provides the most extensive and widely used collection of
-comparative phylogenetic methods tools, R users lack a convenient,
-extensible, and semantically aware format to import, exchange and archive 
-phylogenetic trees, traits, and metadata.
-The rapid increase in phylogenetics data and research methods that make
-use of such data has introduced many new challenges and opportunities 
-for sharing, both between software environments and 
-between research groups [@Drew_2013; @Stoltzfus_2012].
-The ever expanding needs of the phylogenetics community are rapidly
-out-pacing the capacity of current standards, which is resulting in a
-radiation of different data representations and a loss of interoperability
-at the very time when the growth of available data and methods has made
-that interoperability most valuable.
+The popular statistical and mathematical computing platform R [@R] arguably provides the most extensive and widely used collection of
+comparative phylogenetic methods tools <!-- needs a citation to support it -->.
+Yet, R users lack a convenient and extensible way to import, exchange and archive 
+phylogenetic data, trait data, and their accompanying rich metadata in
+a data exchange standard that comprehensively supports current data
+exchange needs, is predictably machine-readable, and is forward compatible.
+In response to the rapid increase in the variety of data
+relevant to comparative analysis, and the rising data interoperability
+needs of integrative comparative research approaches, such an exhcange standard,
+called NeXML, has recently
+been developed [@Vos_2012], and is beginning to be adopted. One of the
+hurdles to broader adoption of any exchange standard is the
+availability of easy-to-use programmtic interfaces (APIs) to the
+format, in particular for developers of tools that consume or generate
+relevant data, so that supporting the standard for reading input and
+writing output data does not require additional effort, compared to,
+say, using other but inferior exchange formats. Here we present such
+an API in the form of an R package, called RNeXML. RNeXML aims to
+provide easy programmatic access to reading and writing NeXML documents,
+tailored for the kinds of use-cases that will be common for users and
+developers of the wealth of evolutionary, and in particular
+comparative analysis methods within the R ecosystem.
 
-For example, the currently popular NEXUS file format cannot consistently
+Increasingly many comparative research questions have come to exploit
+the opportunities from, and thus rely on the ability to reuse and
+integrate previously generated or published data and
+metadata. However, the ever expanding needs of the evolutionary
+research community are rapidly outpacing the capabilities of the
+currently still most widely used exchange standards, which were all
+developed a decade or more ago. This has resulted in a radiation of
+different data representations and exchange standard "flavors" that
+are no longer interoperable at the very time when the growth of
+available data and methods has made that interoperability most
+valuable. The problem is particularly acute in light of the challenges
+for finding and sharing phylogenetic data
+[@Drew_2013; @Stoltzfus_2012], in particular in a way that is not
+subject to loss of most data and metadata semantics, or that is
+amenable to predictable machine interpretation [@Cranston_2014]. For
+example, the currently popular NEXUS file format [@Maddison_1997] cannot consistently
 represent horizontal gene transfer or ambiguity in reading a character
-(such as a DNA sequence base pair).  This has led many developers to
-modify NEXUS in different ways for different needs, with the unfortunate
+(such as a DNA sequence base pair).  This has led to
+modifications of NEXUS in different ways for different needs, with the unfortunate
 result that NEXUS files generated by one program can be incompatible
@@
 NEXUS files may also make inconsistent assumptions about tokens, quoting,
-or element lengths.  @Vos_2012 estimates as many as 15% of the NEXUS
+or element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
 files in the CIPRES portal contain unrecoverable but hard to diagnose errors.
 
-NeXML was developed to addresses each of these problems and also open doors
-to new opportunities in handling phylogenetic data. First, NeXML is defined
-by a precise grammar that can be programmatically **validated**. This helps
-ensure that any NeXML file can always be read without errors by any 
-NeXML parsing software. Second, NeXML is **extensible**: a user can define
-representations of new information (as we will illustrate) without breaking
-this compatibility. Third and most importantly, NeXML is **semantically rich**.
-The use of formal semantics in NeXML provides a representation of data that
-is not only machine-readable, but can also permits machine reasoning about
-the data provided (see @Parr2011 for an overview on semantic reasoning in 
+A detailed account of how the NeXML standard addresses these and other
+relevant challenges can be found in @Vos_2012. In brief, NeXML was
+designed with the following important properties. First, NeXML is defined
+by a precise grammar that can be programmatically **validated**. This
+ensures that any NeXML file can be expected to be parsed and read
+without errors by any NeXML-compliant software, provided the file
+passes validation. Second, NeXML is **extensible**: a user can define
+representations of new, previously unanticipated information (as we
+will illustrate) without violating its defining grammar. Third and
+most importantly, NeXML is **semantically rich**.
+The use of formal semantics in NeXML makes its data representation
+not only machine-readable in a predictable manner, but also enables machine reasoning about
+the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
 evolutionary informatics). Instead of asking the computer to
-return "the data on line #93", we can expect the computer to reason about
+return "the data on line #93" <!-- this is a pretty obscure example, -->
+<!-- and I'm not even clear what you mean here, so having a hard time -->
+<!-- coming up with an improvement -->, we can expect the computer to reason about
 taxonomic concepts (as we illustrate in the supplement).
 
-The R language [@R] has become an important environment for phylogenetics
-research, with over 30 packages available including several with over 
-1000 citations [@taskview]. `RNeXML` makes the large and rapidly growing
-archive of richly annotated, NeXML-based phylogenetics data available to
-the R user and this powerful ecosystem of R tools for phylogenetics.
-Further, R packages and users frequently add their own additional 
-information to the phylogenies they work with: annotations of species,
+The R language and platform [@R] has become a powerful ecosystem for
+evolutionary analysis, with over 30 packages available alone for
+comparative phylogenetics, several of which have accumulated over 
+1000 citations [@taskview]. `RNeXML` makes the sizable amount of
+richly annotated phylogenetic data already existing in NeXML format
+available to R users and R's ecosystem, in a way, as we show here,
+that is no more involved than reading or writing data from other
+phylogenetic data formats. `RNeXML` also carries immediate benefits
+for its users. For example, comparative analysis R packages and users
+frequently add their own metadata annotations to the phylogenies they work with, including annotations of species,
 stochastic character maps, trait values, model estimates and parameter 
-values. By leveraging the semantic and extensible nature of NeXML, `RNeXML` 
-provides R users a platform independent data exchange format to capture,
-further annotate, exchange, and preserve this information. With `RNeXML`,
-R users and developers can both better contribute to and take advantage 
-of the rapidly growing infrastructure of evolutionary informatics [@Parr2011].  
-By capturing information already available in the R environment, together
-with the appropriate choice of defaults, `RNeXML` can facilitate meeting
-the rising bar for best practices in data sharing in phylogenetic 
-research [@Cranston_2014] while at the same reducing the burden
-on the user.
+values. `RNeXML` affords R the capability to harness machine-readable
+semantics and an extensible metadata schema to capture, preserve, and
+share these and other kinds of information, all through an API instead
+of having to understand in detail the schema underlying the NeXML
+standard. To assist users in meeting the rising bar for best practices
+in data sharing in phylogenetic research [@Cranston_2014], `RNeXML`
+captures metadata information from the R environment to the extent
+possible, and applies reasonable defaults. Beyond these more immediate
+benefits, `RNeXML` plugs the R ecosystem into the growing informatics
+infrastructure of evolutionary informatics [@Parr2011], in which
+interoperable data and machine-readable semantics play important
+enabling roles.
 
-In this paper, we introduce the reader to the concepts and use of
-RNeXML and related technologies. We begin by discussing methods to read
-phylogenetic tree data and character trait data in the NeXML format into
-R, and also how to write phylogenetic data and character trait data into
-NeXML. The heart of this section focuses on tools for reading and writing
-the kind of semantically rich metadata that most sets NeXML apart from
-other phylogenetic data formats. Starting simple, we use examples of
+The rest of this paper is organized as follows. First, we introduce
+the basic use and concepts of RNeXML for reading and writing
+phylogenetic trees and character trait data from and to NeXML. We then
+focus on RNeXML's support for reading and writing the kind of
+semantically rich metadata that most sets NeXML apart from
+other phylogenetic data formats. We include examples of
 increasing complexity to illustrate concepts and tools from biodiversity
 informatics that allow users to express and interpret complicated data
-structures automatically.  We end with an example that uses semantic metadata to
-extend the NeXML standard to describe stochastic character mapping: a kind of
-data annotation that is currently available in data structures
-provided by existing R packages
-but lacks a consistent, extensible export format. We illustrate how NeXML
-files can be formally published to a data repository directly from R.
-Finally, we provide a brief description of how the software has been
+structures automatically.  We end this section with a concrete example illustrating
+how RNeXML facilitates harnessing the forward-compatibility property
+of NeXML, by devising a serialization for stochastic character
+mappings, a data annotation generated by existing R packages but
+lacking a standard exchange format. Next we illustrate how RNeXML
+promotes data archiving and sharing by providing commands for directly
+publishing NeXML files to a data repository. We conclude the paper
+with a brief description of how the software has been
 implemented, what quality control measures are in place, as well as
@@
 
-This paper has been written as a dynamic document 
-[@Xie_2013].  The reader may paste the code blocks
-shown into their R terminal to follow along with the examples shown.
+This paper has been written as a dynamic document in R Markddown
+[@Xie_2013], which directly exercises all of the code blocks shown every time
+the typeset version is regenerated, providing immediate feedback on
+whether the code as shown is complete and continues to work. It also
+means that the reader may paste the code blocks into their own R
+terminal to follow along with the examples.
 

### 2014-11-28 manuscripts/manuscript.Rmd
Smoothes and slightly tightens Intro to RNeXML leader.

@@
 
-A NeXML document can be thought of as having the following components: (1) phylogeny
-data, (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
-`RNeXML` seeks to provide a seamless way to convert these to and from their 
-corresponding R object representations, in particular `ape::phylo`,
-`data.frame`, and lists for phylogenies, character
-matrices, and metadata, respectively.
-Metadata can provide bibliographic and provenance
-information about the study as part of which the phylogeny was generated or
-applied (such as authors, citations to publications, timestamps, and even
-descriptions of methods). They can also be used to attach annotations to individual
-data elements.  We illustrate how metadata annotations can be added to
-the operational taxonomic units of the phylogeny and can be used to extend
-the data representation to new uses such as stochastic character maps.
-<!-- sounds awkward - need to smoothen the language. HL -->
+Conceptually, a NeXML document has the following components: (1) phylogeny
+data, (2) character or trait data in matrix form, (3) operational
+taxonomic units (OTUs), and (4) metadata. When reading and writing NeXML,
+`RNeXML` aims to map these to and from, respectively, their most
+natural representations in R. These are the `phylo` object from the
+`ape` package [@Paradis_2004], R's native `data.frame` list structure for
+phylogenies, character matrices, and metadata, respectively.
+<!-- compared to the preceding enumeration, this is missing where the OTUs go -->
+
+NeXML allows attaching metadata to any data element, and even to
+metadata themselves. At the level of the document as a whole or the
+phylogeny, metadata can provide bibliographic and provenance
+information, for example about the study as part of which the
+phylogeny was generated or applied, and with which methods it was
+generated. In the form of annotations, metadata can also be attached
+to very specific elements of the data, such as individual OTUs, nodes,
+or even edges of the phylogeny.
 

### 2014-11-29 manuscripts/manuscript.Rmd
Edits for Reading Phylogenetics Trees section.

@@
 phylogenies, character matrices, and metadata, respectively.
-<!-- compared to the preceding enumeration, this is missing where the OTUs go -->
+<!-- compared to the preceding enumeration, this is missing where the OTUs go. HL -->
 
@@
 
-The current stable version of the `RNeXML` package can be installed from
-the CRAN library[^1].
+As for other R packages, prior to using `RNeXML`, the package needs to
+be installed and loaded. The current stable version can be installed from
+the CRAN library[^1]:
 
@@
 
-## Reading phylogenetic trees
+This will also install its dependencies. Once installed, R scripts
+using the library only need to load it:
 
@@
 
-We begin by loading the `RNeXML` package and reading in an example
-NeXML file provided with the package, containing two small phylogenetic
-trees.[^2]
+## Reading phylogenetic trees
+
+We begin by reading in an example NeXML file provided with the
+package, containing two small phylogenetic trees.[^2]
 
 ```{r}
-library("RNeXML")
 nex <- nexml_read("components/trees.xml")
+typeof(nex)
 ```
-The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
-file. 
-Objects of type `nexml` represent phylogenetic trees as well as 
-character traits, and semantic metadata. As such, this new R object format differs
-from existing formats such as `ape::phylo` format [@Paradis2004],
-which describes trees alone, or any of the `phylobase` classes of object that contain 
-trees and traits but more limited metadata [@phylobase].
 
-Although new custom R functions and packages 
-could be written for using `nexml` objects, for most users it will be
-more convienent to use methods of the many already existing R packages for 
-phylogenetics. From a `nexml` object, we can 
-therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
-case we will get back a `multiPhylo` object, since the file contains two phylogenies.
+The method `nexml_read()` reads NeXML files, either from a local
+file, or from a remote location via its URL, and returns an object of
+type `nexml`. This object type is a representation of the NeXML
+document contents, and therefore holds phylogenetic trees as well as 
+character or trait matrices, and all metadata. This is similar in
+design to the classes in the `phylobase` package, which contain 
+trees and traits but hold only limited metadata [@phylobase], but
+different from R object types such as `ape::phylo` (the `phylo` class
+in the `ape` package [@Paradis2004]), which represents trees alone.
+
+To interface conveniently with the many already existing R packages for 
+phylogenetics, which conventionally accept phylogenies as `ape::phylo`
+objects, the method `get_trees_list()` returns a `ape::multiPhylo`
+object, which can be treated as a list of `ape::phylo` objects:
 
 ```{r}
-phy <- get_trees(nex)
+phy <- get_trees_list(nex)
+typeof(phy)
 ```
 
-We can then leverage the rich suite of methods available from the R
-phylogenetics community. For instance, here we use the standard plotting function from
-the `ape` package to plot the first tree in the resulting `ape::phylo` object returned by
-`get_trees`.
+As an example for leveraging the wealth of resources in R
+interoperating with the `ape::phylo` type, we use `ape`'s standard
+plotting function to plot the first tree in the list of trees:
 
@@
 
-Note that NeXML can not only contain multiple phylogenetic trees in a single document,
-but can also group sets of trees into separate blocks (e.g. as resulting from
-different workflows).
-The `get_trees()` function is designed for interactive use, and thus attempts to return the most sensible R object for NeXML: if only a single tree is available, it will return an `ape::phylo`
-object; if multiple trees are available in a single `trees` block, it will return
-an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
-blocks are available, it will return a list of `multiPhylo` blocks.
+The `get_trees_list()` method is designed for use in scripts, for
+which a consistent and predictable return type (namely
+`ape::multiPhylo`) is desirable.
+<!-- the open (and previously not covered) issue here is what does -->
+<!-- get_trees_list() return in the case of multiple trees blocks? HL -->
+Following a common design practice
+for R packages, `RNeXML` also contains a method `get_trees()`, which
+is designed for interactive use and returns the R object most
+intuitive given the arrangement of phylogeny data in the source NeXML
+document. NeXML inherits from NEXUS the ability to contain multiple
+phylogenies, and to group sets of phylogenies into "blocks". The
+`get_trees()` method will return an `ape::phylo` object if the NeXML
+document contains a single phylogeny, an `ape::multiPhylo` object if
+it contains multiple phylogenies arranged in a single `trees` block,
+and a list of `ape::multiPhylo` objects if it contains multiple
+`trees` blocks. This design spares an interactive user from having to
+type unnecessary data conversion commands, at the expense of needing
+to know which return type to expect (or verifying the type via
+introspection commands such as `typeof()`).
 
-For scripted use, it is desirable to have a consistent return type.  To always get back a list of `multiPhylo` objects,
-even when only one tree is available, use `get_trees_list()`.
+Because the example file supplied with the package contains two
+phylogenies in a single `trees` block, in this case the return type of both
+phylogeny accessor methods is the same:
 
-The ability to read in phylogenetic trees in the NeXML format opens up a
-wide and rapidly growing array of resources to the R user.  For instance,
-a user can access all the phylogenies available in 
-[TreeBASE](http://treebase.org/) 
- through the
-`nexml` format.  `RNeXML`'s `nexml_read` can also read a file directly from a URL by automatically recognizing the URL patterns and first downloading the file to a working directory:
+```{r}
+phy <- get_trees(nex)
+typeof(phy)
+plot(phy[[1]])
+```
+
+As mentioned above, the `nexml_read()` method allows specifying the
+NeXML document to read not only as the name of a local file, but also
+as the URL of a document on the web. If the location parameter is
+recognized as a URL, the method will automatically download the
+document to the local working directory and read it from there. This
+gives convenient and rapid access to phylogenetic data published in
+NeXML format on the web, which includes the content of the
+phylogenetic data repository TreeBASE [@Piel_2002]. For
+example, the following plots a tree in TreeBASE:
 
 ```{r eval=FALSE}
-nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+tb_nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+tb_phy <- get_trees_list(tb_nex)
+plot(tb_phy[[1]])
 ```

### 2014-11-29 manuscripts/manuscript.Rmd
Edits for Reading Character Data.
Less extensive edits here, mostly to tighten text and improve flow.

@@
 
-Like NEXUS files, NeXML documents can in addition to trees also
-contain character data in the form of matrices.  Character data can be
-of molecular (molecular sequence alignments), discrete (most
-morphological character data), or continuous type (many trait data). 
-Often character data, especially molecular and morphological, is used
-for phylogenetic tree reconstruction, but they are also a common input
-for the many comparative phylogenetic analysis methods available in R. Methods to extract character data
-work much like those for extracting the phylogenetic trees. Here we read in an example
-NeXML file that contains both a phylogeny and character data. The
-function
-`get_characters()` can then be used to obtain a `data.frame` with columns as characters and rows as
-taxa:
+NeXML documents can contain character or trait data in the form of
+matrices. A data matrix can be of molecular (for molecular sequence
+alignments), discrete (for most morphological character data), or
+continuous type (for many trait data). Character data, especially
+molecular and morphological matrices, are often used for phylogenetic
+tree reconstruction, but they are also a common input for the many
+comparative phylogenetic analysis methods available in R. The method
+`get_characters()` obtains character data from a `nexml` object, and
+returns them as a standard `data.frame` R object with columns as
+characters and rows as taxa. The following example, which uses a file
+that comes with `RNeXML`, demonstrates this:
 
@@
 ```
-Continuous and discrete character data are represented as separate
-blocks in NeXML files, but will be combined as separate columns of a
-single data frame by `get_characters()` if they correspond to the same
-taxa. If this behavior is undesired, the function
-`get_characters_list()` can be used to obtain
-a list of data frames, where list elements correspond to the character matrix blocks of
-the NeXML file.
+
+To allow strict validation of data types, the NeXML schema requires
+continuous and discrete character data to be separated into different
+"characters" blocks. Since the `data.frame` data structure in R has no
+such constraints, the `get_characters()` method will combine such
+blocks as separate columns into a single `data.frame` object, provided
+they correspond to the same taxa.
+<!-- does the example data file actually demonstrate this? if not, it -->
+<!-- should. HL -->
+Otherwise, a list of `data.frame`s
+is returned, with list elements corresponding to characters
+blocks. Similar to the distinction between `get_trees()` and
+`get_trees_list()`, there is also a method `get_characters_list()`,
+which always returns a list of `data.frame`s.
 

### 2014-12-01 manuscripts/manuscript.Rmd
Edits the Writing NeXML section.
Most of these are fairly minor. I did separate the explanation for creating
`nexml` objects into its own subsection, for better consistency with the
part about reading NeXML files.

@@
 
-## Writing phylogenetic trees
-
-Generating NeXML files from R is likewise straightforward.  Here we
-write a phylogeny from the `ape::phylo` format out to a NeXML file:
+## Writing phylogenetic trees and character data
+
+Aside from providing an easy-to-use interface to data already in
+NeXML, `RNeXML` also aims to facilitate broader adoption of the
+standard for sharing phylogenetic data, and thus to make it
+straightforward to generate them. The method `nexml_write()`, as the
+name suggests, generates a NeXML file from its input parameters. In
+its simplest invocation, the method writes a tree to a file:
 
@@
 
-The first argument to `nexml_write()` is a `nexml` object, or any
-object that can be coerced to it, such as a `ape::phylo` phylogeny,
-or a `multiPhylo` list of phylogenies. The `nexml_write()` function
-also allows to add character data:
+The first argument to `nexml_write()` is an object of type `nexml`, or any
+object that can be coerced to it, such as in the above example an
+`ape::phylo` phylogeny. Alternatively, passing a `multiPhylo` object
+would write a list of phylogenies to the file.
+
+In addition to trees, the `nexml_write()` method also to specify
+character data as another parameter. The following example uses data
+from the comparative phylogenetics R package `geiger` [@Harmon_2008].
 
@@
 data(geospiza)
-nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
-```
-
-These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].
-Instead of packaging trees and characters at the time of writing a
-NeXML file, the document can also be built up step by step using the
-functions `add_characters()`,
-`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
-existing `nexml` object, or create a new nexml object if none is provided.
-Here we first create a `nexml` object containing the phylogeny data,
-and then add the character data to it:
-
-```{r}
-geiger_nex <- add_trees(geospiza$phy)
-geiger_nex <- add_characters(geospiza$dat, geiger_nex)
-```
-
-The data need not share the same taxa.  We can append additional
-phylogenies and characters corresponding to different taxa:
+nexml_write(trees = geospiza$phy, characters = geospiza$dat,
+            file="geospiza.xml")
+```
+
+## Creating and populating `nexml` objects
+
+Instead of packaging the various components for a NeXML file at the
+time of writing the file, the user can also create and iteratively
+populate an in-memory R object representation of the NeXML document,
+i.e., an object of type `nexml`. The methods to do this are `add_characters()`,
+`add_trees()`, and `add_meta()`, for adding characters, trees, and
+metadata, respectively. Each of these functions will automatically
+create a new nexml object if not supplied with an existing one as the
+last (optional) argument.
+
+For example, here we use `add_trees()` to first create a `nexml`
+object with the phylogeny data, and then add the character data to it:
+
+```{r}
+nexObj <- add_trees(geospiza$phy)
+nexObj <- add_characters(geospiza$dat, nexObj)
+```
+
+The data components being added to a `nexml` object need not share the
+same OTUs.  For example, here we append additional phylogenies and characters
+using different OTUs to the `nexml` object originally populated with
+the `geospiza` phylogeny and character data:
 
@@
 data(primates)
-geiger_nex <- add_trees(primates$phy, geiger_nex)
-geiger_nex <- add_characters(primates$dat, geiger_nex)
-```
-
-This allows us to include multiple trees and character matrices to a
-single NeXML file, for instance, to represent samples from posterior
-probability distributions of inferred or simulated phylogenies or
-character states.
+nexObj <- add_trees(primates$phy, nexObj)
+nexObj <- add_characters(primates$dat, nexObj)
+```
+
+<!-- Might be worth adding some brief explanation here as to what -->
+<!-- happens internally in such a case. For example, automatic -->
+<!-- generation of multiple `otus` blocks? HL --> 
+Other than storage size, there is no limit to the number of
+phylogenies and character matrices that can be included in a
+single NeXML file. This allows, for example, to capture samples from a
+posterior probability distribution of inferred or simulated phylogenies or
+character states in one NeXML file.
 

### 2014-12-01 manuscripts/manuscript.Rmd
A few edits in response to @hlapp's suggestions

@@
 the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
-evolutionary informatics). Instead of asking the computer to
-return "the data on line #93" <!-- this is a pretty obscure example, -->
-<!-- and I'm not even clear what you mean here, so having a hard time -->
-<!-- coming up with an improvement -->, we can expect the computer to reason about
-taxonomic concepts (as we illustrate in the supplement).
+evolutionary informatics). For instance, through the use of semantics we can
+expect the computer to reason about taxonomic concepts
+(as we illustrate in the supplement). 
+<!-- Sorry, just trying to contrast what it means to make a 'semantic' query 
+vs a non-semantic one.  Most R users won't be familiar with SQL queries either anyway.  Let me know if this is better.-->
 
@@
 nex <- nexml_read("components/trees.xml")
-typeof(nex)
 ```
@@
 different from R object types such as `ape::phylo` (the `phylo` class
-in the `ape` package [@Paradis2004]), which represents trees alone.
+in the `ape` package [@Paradis_2004]), which represents trees alone.
 
@@
 The `get_trees_list()` method is designed for use in scripts, for
-which a consistent and predictable return type (namely
-`ape::multiPhylo`) is desirable.
-<!-- the open (and previously not covered) issue here is what does -->
-<!-- get_trees_list() return in the case of multiple trees blocks? HL -->
-Following a common design practice
-for R packages, `RNeXML` also contains a method `get_trees()`, which
+which a consistent and predictable return type (namely a `list` of
+`ape::multiPhylo` objects, reflecting the nested structure of NeXML) is desirable.
+`RNeXML` also contains a method `get_trees()`, which
 is designed for interactive use and returns the R object most
@@
 to know which return type to expect (or verifying the type via
-introspection commands such as `typeof()`).
+introspection commands such as `class()`).
 
@@
 phy <- get_trees(nex)
-typeof(phy)
 plot(phy[[1]])
@@
 blocks as separate columns into a single `data.frame` object, provided
-they correspond to the same taxa.
-<!-- does the example data file actually demonstrate this? if not, it -->
-<!-- should. HL -->
-Otherwise, a list of `data.frame`s
+they correspond to the same taxa. Otherwise, a list of `data.frame`s
 is returned, with list elements corresponding to characters
@@
 data(geospiza)
-nexml_write(trees = geospiza$phy, characters = geospiza$dat,
+nexml_write(trees = geospiza$phy, 
+            characters = geospiza$dat,
             file="geospiza.xml")
@@
 
-<!-- Might be worth adding some brief explanation here as to what -->
-<!-- happens internally in such a case. For example, automatic -->
-<!-- generation of multiple `otus` blocks? HL --> 
+`RNeXML` automatically adds new, separate OTU blocks into the NeXML file
+for each of these new data sets.
 Other than storage size, there is no limit to the number of
@@
 ```
+
 <!-- the above will get fixed by CB, this is just a reminder and -->
 <!-- placeholder to adjust when that's done. -->
+<!-- Um, sorry, remind me what I am supposed to fix here? formatting? -->
 
@@
 which the package `knitcitations` [@Boettiger_2014] can turn into a
-formatted citation using. As an example, to add the citation information of the paper that generated the
-`geospiza` phylogeny included in the `geiger` package:
-
-```{r}
+formatted citation using. As an example, to add the above citation
+using the DOI:
+
+```{r }
 library("knitcitations")
-geiger_nex <- add_basic_meta(citation = bib_metadata("10.2307/2408428"), nexml=geiger_nex)
+birds <- add_basic_meta(citation = bib_metadata("10.1093/bioinformatics/btg412"), nexml=birds)
 ```

### 2014-12-02 manuscripts/manuscript.Rmd
Clarify lack of OTU block in ape::phylo & explain merging character blocks into data frame with get_characters as suggested by @hlapp in #101

@@
 `ape` package [@Paradis_2004], R's native `data.frame` list structure for
-phylogenies, character matrices, and metadata, respectively.
-<!-- compared to the preceding enumeration, this is missing where the OTUs go. HL -->
+phylogenies, character matrices, and metadata, respectively. Note that
+unlike NeXML or NEXUS, the `phylo` object stores taxonomic information
+only as tip labels and not as a separate block of data.
 
@@
 
+Observe that `RNeXML` has automatically merged the different NeXML character
+blocks, (representing continuous and discrete characters respectively) 
+into a single data frame with appropriate column classes. 
 To allow strict validation of data types, the NeXML schema requires
@@
 `get_trees_list()`, there is also a method `get_characters_list()`,
-which always returns a list of `data.frame`s.
+which always returns a list of `data.frame`s, one for each
+character block.
 

### 2014-12-03 manuscripts/manuscript.Rmd
update supertreebase URL, closes #102

@@
 ```{r eval=FALSE}
-tb_nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+tb_nex <- nexml_read("https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml")
 tb_phy <- get_trees_list(tb_nex)

### 2014-12-05 manuscripts/manuscript.Rmd
formatting tweaks for 2.0.0

--- 2014-12-02 manuscripts/manuscript.Rmd
@@
 phy <- get_trees_list(nex)
-typeof(phy)
+class(phy)
 ```
@@
 
-```{r}
+```{r fig.show="hide"}
 phy <- get_trees(nex)
@@
 ```{r eval=FALSE}
-tb_nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
+tb_nex <- nexml_read(
+"https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml")
 tb_phy <- get_trees_list(tb_nex)
-plot(tb_phy[[1]])
 ```

### 2014-12-29 manuscripts/manuscript.Rmd
Rewrite/arrangemnt of introduction.
This is to tighten up the text for submission as an Application Note.

@@
 
-The popular statistical and mathematical computing platform R [@R] arguably provides the most extensive and widely used collection of
-comparative phylogenetic methods tools <!-- needs a citation to support it -->.
-Yet, R users lack a convenient and extensible way to import, exchange and archive 
-phylogenetic data, trait data, and their accompanying rich metadata in
-a data exchange standard that comprehensively supports current data
-exchange needs, is predictably machine-readable, and is forward compatible.
-In response to the rapid increase in the variety of data
-relevant to comparative analysis, and the rising data interoperability
-needs of integrative comparative research approaches, such an exhcange standard,
-called NeXML, has recently
-been developed [@Vos_2012], and is beginning to be adopted. One of the
-hurdles to broader adoption of any exchange standard is the
-availability of easy-to-use programmtic interfaces (APIs) to the
-format, in particular for developers of tools that consume or generate
-relevant data, so that supporting the standard for reading input and
-writing output data does not require additional effort, compared to,
-say, using other but inferior exchange formats. Here we present such
-an API in the form of an R package, called RNeXML. RNeXML aims to
-provide easy programmatic access to reading and writing NeXML documents,
-tailored for the kinds of use-cases that will be common for users and
-developers of the wealth of evolutionary, and in particular
-comparative analysis methods within the R ecosystem.
-
-Increasingly many comparative research questions have come to exploit
-the opportunities from, and thus rely on the ability to reuse and
-integrate previously generated or published data and
-metadata. However, the ever expanding needs of the evolutionary
-research community are rapidly outpacing the capabilities of the
-currently still most widely used exchange standards, which were all
+Users of the popular statistical and mathematical computing platform R
+[@R] enjoy a wealth of readily installable comparative phylogenetic
+methods and tools <!-- needs a citation to support it -->. Exploiting
+the opportunities arising from this wealth for complex and integrative
+comparative research questions relies on the ability to reuse and
+integrate previously generated or published data and metadata. The
+resulting expanding data exchange needs of the evolutionary research
+community are rapidly outpacing the capabilities of the currently
+still most widely used data exchange standards, which were all
 developed a decade or more ago. This has resulted in a radiation of
@@
 available data and methods has made that interoperability most
-valuable. The problem is particularly acute in light of the challenges
-for finding and sharing phylogenetic data
-[@Drew_2013; @Stoltzfus_2012], in particular in a way that is not
-subject to loss of most data and metadata semantics, or that is
-amenable to predictable machine interpretation [@Cranston_2014]. For
-example, the currently popular NEXUS file format [@Maddison_1997] cannot consistently
-represent horizontal gene transfer or ambiguity in reading a character
-(such as a DNA sequence base pair).  This has led to
-modifications of NEXUS in different ways for different needs, with the unfortunate
-result that NEXUS files generated by one program can be incompatible
-with another [@Vos_2012]. Without a formal grammar, software based on 
-NEXUS files may also make inconsistent assumptions about tokens, quoting,
-or element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
-files in the CIPRES portal contain unrecoverable but hard to diagnose errors.
+valuable. In response to the unmet needs for standardized data
+exchange in phylogenetics, a modern XML-based exchange standard,
+called NeXML, has recently been developed [@Vos_2012], and is
+beginning to be adopted. NeXML comprehensively supports current data
+exchange needs, is predictably machine-readable, and is forward
+compatible. To make the capibilities of NeXML available to R users in
+an easy-to-use form, and to lower the hurdles to adoption of the
+standard, we present RNeXML, an R package that aims to provide easy
+programmatic access to reading and writing NeXML documents, tailored
+for the kinds of use-cases that will be common for users and
+developers of the wealth of evolutionary, and in particular
+comparative analysis methods within the R ecosystem.
+
+The exchange problem for phylogenetic data is particularly acute in
+light of the challenges in finding and sharing phylogenetic data
+without the otherwise common loss of most data and metadata semantics
+[@Drew_2013; @Stoltzfus_2012], or with semantics that is predictably
+machine interpretable [@Cranston_2014]. For example, the still popular
+NEXUS file format [@Maddison_1997] cannot consistently represent
+horizontal gene transfer or ambiguity in reading a character (such as
+a DNA sequence base pair).  This has led to modifications of NEXUS in
+different ways for different needs, with the unfortunate result that
+NEXUS files generated by one program can be incompatible with another
+[@Vos_2012]. Without a formal grammar, software based on NEXUS files
+may also make inconsistent assumptions about tokens, quoting, or
+element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
+files in the CIPRES portal contain unrecoverable but hard to diagnose
+errors.
 
@@
 relevant challenges can be found in @Vos_2012. In brief, NeXML was
-designed with the following important properties. First, NeXML is defined
-by a precise grammar that can be programmatically **validated**. This
-ensures that any NeXML file can be expected to be parsed and read
-without errors by any NeXML-compliant software, provided the file
-passes validation. Second, NeXML is **extensible**: a user can define
-representations of new, previously unanticipated information (as we
-will illustrate) without violating its defining grammar. Third and
-most importantly, NeXML is **semantically rich**.
-The use of formal semantics in NeXML makes its data representation
-not only machine-readable in a predictable manner, but also enables machine reasoning about
-the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
-evolutionary informatics). For instance, through the use of semantics we can
-expect the computer to reason about taxonomic concepts
-(as we illustrate in the supplement). 
-<!-- Sorry, just trying to contrast what it means to make a 'semantic' query 
-vs a non-semantic one.  Most R users won't be familiar with SQL queries either anyway.  Let me know if this is better.-->
+designed with the following important properties. First, NeXML is
+defined by a precise grammar that can be programmatically
+**validated**. This ensures that any NeXML file can be expected to be
+parsed and read without errors by any NeXML-compliant software,
+provided the file passes validation. Second, NeXML is **extensible**:
+a user can define representations of new, previously unanticipated
+information (as we will illustrate) without violating its defining
+grammar. Third and most importantly, NeXML is **rich in computable
+semantics**. The formal semantics used in NeXML is machine-readable in
+a predictable manner, and also enables machine reasoning about the
+data and metadata (see @Parr2011), especially when metadata attributes
+and values are anchored in appropriately axiomatized ontologies.
 
-The R language and platform [@R] has become a powerful ecosystem for
-evolutionary analysis, with over 30 packages available alone for
-comparative phylogenetics, several of which have accumulated over 
-1000 citations [@taskview]. `RNeXML` makes the sizable amount of
-richly annotated phylogenetic data already existing in NeXML format
-available to R users and R's ecosystem, in a way, as we show here,
-that is no more involved than reading or writing data from other
-phylogenetic data formats. `RNeXML` also carries immediate benefits
-for its users. For example, comparative analysis R packages and users
-frequently add their own metadata annotations to the phylogenies they work with, including annotations of species,
-stochastic character maps, trait values, model estimates and parameter 
-values. `RNeXML` affords R the capability to harness machine-readable
-semantics and an extensible metadata schema to capture, preserve, and
-share these and other kinds of information, all through an API instead
-of having to understand in detail the schema underlying the NeXML
-standard. To assist users in meeting the rising bar for best practices
-in data sharing in phylogenetic research [@Cranston_2014], `RNeXML`
-captures metadata information from the R environment to the extent
-possible, and applies reasonable defaults. Beyond these more immediate
-benefits, `RNeXML` plugs the R ecosystem into the growing informatics
-infrastructure of evolutionary informatics [@Parr2011], in which
-interoperable data and machine-readable semantics play important
-enabling roles.
+`RNeXML` not only allows R's ecosystem to read and write data in
+the NeXML format through an interface that is no more involved than
+reading or writing data from other phylogenetic data formats, but also
+carries immediate benefits for its users. For example, comparative
+analysis R packages and users frequently add their own metadata
+annotations to the phylogenies they work with, such as annotations
+of species, stochastic character maps, trait values, model estimates
+and parameter values. `RNeXML` affords R the capability to harness
+machine-readable semantics and an extensible metadata schema to
+capture, preserve, and share these and other kinds of information, all
+through an API instead of having to understand in detail the schema
+underlying the NeXML standard. To assist users in meeting the rising
+bar for best practices in data sharing in phylogenetic research
+[@Cranston_2014], `RNeXML` captures metadata information from the R
+environment to the extent possible, and applies reasonable
+defaults.
 
-The rest of this paper is organized as follows. First, we introduce
-the basic use and concepts of RNeXML for reading and writing
-phylogenetic trees and character trait data from and to NeXML. We then
-focus on RNeXML's support for reading and writing the kind of
-semantically rich metadata that most sets NeXML apart from
-other phylogenetic data formats. We include examples of
-increasing complexity to illustrate concepts and tools from biodiversity
-informatics that allow users to express and interpret complicated data
-structures automatically.  We end this section with a concrete example illustrating
-how RNeXML facilitates harnessing the forward-compatibility property
-of NeXML, by devising a serialization for stochastic character
-mappings, a data annotation generated by existing R packages but
-lacking a standard exchange format. Next we illustrate how RNeXML
-promotes data archiving and sharing by providing commands for directly
-publishing NeXML files to a data repository. We conclude the paper
-with a brief description of how the software has been
-implemented, what quality control measures are in place, as well as
-information about software reuse and getting support.
+# The RNeXML package
 
-This paper has been written as a dynamic document in R Markddown
-[@Xie_2013], which directly exercises all of the code blocks shown every time
-the typeset version is regenerated, providing immediate feedback on
-whether the code as shown is complete and continues to work. It also
-means that the reader may paste the code blocks into their own R
-terminal to follow along with the examples.
-
-# Introduction to RNeXML
-
-Conceptually, a NeXML document has the following components: (1) phylogeny
-data, (2) character or trait data in matrix form, (3) operational
-taxonomic units (OTUs), and (4) metadata. When reading and writing NeXML,
-`RNeXML` aims to map these to and from, respectively, their most
-natural representations in R. These are the `phylo` object from the
-`ape` package [@Paradis_2004], R's native `data.frame` list structure for
-phylogenies, character matrices, and metadata, respectively. Note that
-unlike NeXML or NEXUS, the `phylo` object stores taxonomic information
-only as tip labels and not as a separate block of data.
-
-NeXML allows attaching metadata to any data element, and even to
-metadata themselves. At the level of the document as a whole or the
-phylogeny, metadata can provide bibliographic and provenance
-information, for example about the study as part of which the
-phylogeny was generated or applied, and with which methods it was
-generated. In the form of annotations, metadata can also be attached
-to very specific elements of the data, such as individual OTUs, nodes,
-or even edges of the phylogeny.
-
-## Getting Started
-
-As for other R packages, prior to using `RNeXML`, the package needs to
-be installed and loaded. The current stable version can be installed from
-the CRAN library[^1]:
+`RNeXML` is an R package entirely written in R, and the current stable
+version can be installed from the CRAN library[^1]:
 

### 2014-12-29 manuscripts/manuscript.Rmd
Tightening up read/write section.

@@
 
-## Reading phylogenetic trees
+## Reading and writing phylogenies and character data
 
-We begin by reading in an example NeXML file provided with the
-package, containing two small phylogenetic trees.[^2]
+Conceptually, a NeXML document has the following components: (1)
+phylogeny data, (2) character or trait data in matrix form, (3)
+operational taxonomic units (OTUs), and (4) metadata. When reading and
+writing NeXML, `RNeXML` aims to map these to and from, respectively,
+their most natural representations in R. These are the `phylo` and
+`multiPhylo` objects from the `ape` package [@Paradis_2004] for
+phylogenies, and R's native `data.frame` list structure for character
+matrices.
+
+## Reading phylogenies and character data
+
+The method `nexml_read()` reads NeXML files, either from a local
+file, or from a remote location via its URL, and returns an object of
+type `nexml` [^2]:
 
@@
 
-The method `nexml_read()` reads NeXML files, either from a local
-file, or from a remote location via its URL, and returns an object of
-type `nexml`. This object type is a representation of the NeXML
+The `nexml` object type is a representation of the NeXML
 document contents, and therefore holds phylogenetic trees as well as 
@@
 phy <- get_trees_list(nex)
-class(phy)
-```
-
-As an example for leveraging the wealth of resources in R
-interoperating with the `ape::phylo` type, we use `ape`'s standard
-plotting function to plot the first tree in the list of trees:
-
-```{r eval=FALSE}
-plot(phy[[1]]) 
 ```
@@
 which a consistent and predictable return type (namely a `list` of
-`ape::multiPhylo` objects, reflecting the nested structure of NeXML) is desirable.
-`RNeXML` also contains a method `get_trees()`, which
+`ape::multiPhylo` objects, reflecting the nested structure of NeXML)
+is desirable. `RNeXML` also contains a method `get_trees()`, which
 is designed for interactive use and returns the R object most
@@
 
-Because the example file supplied with the package contains two
-phylogenies in a single `trees` block, in this case the return type of both
-phylogeny accessor methods is the same:
-
-```{r fig.show="hide"}
-phy <- get_trees(nex)
-plot(phy[[1]]) 
-```
-
-As mentioned above, the `nexml_read()` method allows specifying the
-NeXML document to read not only as the name of a local file, but also
-as the URL of a document on the web. If the location parameter is
-recognized as a URL, the method will automatically download the
+If the location parameter with which the `nexml_read()` method is
+invoked is recognized as a URL, the method will automatically download the
 document to the local working directory and read it from there. This
 gives convenient and rapid access to phylogenetic data published in
-NeXML format on the web, which includes the content of the
+NeXML format on the web, such as the content of the
 phylogenetic data repository TreeBASE [@Piel_2002]. For
@@
 
-## Reading character data
-
-NeXML documents can contain character or trait data in the form of
-matrices. A data matrix can be of molecular (for molecular sequence
-alignments), discrete (for most morphological character data), or
-continuous type (for many trait data). Character data, especially
-molecular and morphological matrices, are often used for phylogenetic
-tree reconstruction, but they are also a common input for the many
-comparative phylogenetic analysis methods available in R. The method
-`get_characters()` obtains character data from a `nexml` object, and
-returns them as a standard `data.frame` R object with columns as
-characters and rows as taxa. The following example, which uses a file
-that comes with `RNeXML`, demonstrates this:
+The method `get_characters()` obtains character data matrics from a
+`nexml` object, and returns them as a standard `data.frame` R object
+with columns as characters and rows as taxa:
 
@@
 
-Observe that `RNeXML` has automatically merged the different NeXML character
-blocks, (representing continuous and discrete characters respectively) 
-into a single data frame with appropriate column classes. 
-To allow strict validation of data types, the NeXML schema requires
-continuous and discrete character data to be separated into different
-"characters" blocks. Since the `data.frame` data structure in R has no
-such constraints, the `get_characters()` method will combine such
-blocks as separate columns into a single `data.frame` object, provided
-they correspond to the same taxa. Otherwise, a list of `data.frame`s
-is returned, with list elements corresponding to characters
-blocks. Similar to the distinction between `get_trees()` and
-`get_trees_list()`, there is also a method `get_characters_list()`,
-which always returns a list of `data.frame`s, one for each
-character block.
+A NeXML data matrix can be of molecular (for molecular sequence
+alignments), discrete (for most morphological character data), or
+continuous type (for many trait data).  Similarly as for trees, data
+matrices can also be arranged in blocks; in fact, matrices of
+different data types are required to be separated into different
+"blocks" to allow strict validation of data typs. Since the
+`data.frame` data structure in R has no such constraints, the
+`get_characters()` method will combine such blocks as separate columns
+into a single `data.frame` object, provided they correspond to the
+same taxa. Otherwise, a list of `data.frame`s is returned, with list
+elements corresponding to characters blocks. Similar to the methods
+for obtaining trees, there is also a method `get_characters_list()`,
+which always returns a list of `data.frame`s, one for each character
+block.
 
@@
 
-Aside from providing an easy-to-use interface to data already in
-NeXML, `RNeXML` also aims to facilitate broader adoption of the
-standard for sharing phylogenetic data, and thus to make it
-straightforward to generate them. The method `nexml_write()`, as the
-name suggests, generates a NeXML file from its input parameters. In
+The method `nexml_write()` generates a NeXML file from its input parameters. In
 its simplest invocation, the method writes a tree to a file:
@@
 
-In addition to trees, the `nexml_write()` method also to specify
+In addition to trees, the `nexml_write()` method also allows to specify
 character data as another parameter. The following example uses data

### 2014-12-29 manuscripts/manuscript.Rmd
Moves down metadata section from earlier position.

@@
 # Metadata
+
+NeXML allows attaching metadata to any data element, and even to
+metadata themselves. At the level of the document as a whole or the
+phylogeny, metadata can provide bibliographic and provenance
+information, for example about the study as part of which the
+phylogeny was generated or applied, and with which methods it was
+generated. In the form of annotations, metadata can also be attached
+to very specific elements of the data, such as individual OTUs, nodes,
+or even edges of the phylogeny.
 

### 2014-12-30 manuscripts/manuscript.Rmd
Minor corrections to previous edits.

@@
 
-## Reading phylogenies and character data
-
 The method `nexml_read()` reads NeXML files, either from a local
@@
 alignments), discrete (for most morphological character data), or
-continuous type (for many trait data).  Similarly as for trees, data
-matrices can also be arranged in blocks; in fact, matrices of
-different data types are required to be separated into different
-"blocks" to allow strict validation of data typs. Since the
-`data.frame` data structure in R has no such constraints, the
+continuous type (for many trait data).  Data matrices can be, and
+matrices of different data types are required to be separated into
+different "blocks", to allow strict validation of data types. Since
+the `data.frame` data structure in R has no such constraints, the
 `get_characters()` method will combine such blocks as separate columns
@@
 block.
-
-## Writing phylogenetic trees and character data
 

### 2014-12-31 manuscripts/manuscript.Rmd
Reorganizing RNeXML package description outline.

@@
 
-## Reading and writing phylogenies and character data
+## Representation of NeXML documents in R
 
@@
 phylogeny data, (2) character or trait data in matrix form, (3)
-operational taxonomic units (OTUs), and (4) metadata. When reading and
-writing NeXML, `RNeXML` aims to map these to and from, respectively,
-their most natural representations in R. These are the `phylo` and
-`multiPhylo` objects from the `ape` package [@Paradis_2004] for
-phylogenies, and R's native `data.frame` list structure for character
-matrices.
+operational taxonomic units (OTUs), and (4) metadata. `RNeXML` defines
+the `nexml` object type, which is a representation of the NeXML
+document contents. It therefore holds phylogenetic trees as well as
+character or trait matrices, and all metadata. This is similar in
+design to the classes in the `phylobase` package, which contain trees
+and traits but hold only limited metadata [@phylobase], but different
+from R object types such as `ape::phylo` (the `phylo` class in the
+`ape` package [@Paradis_2004]), which represents trees alone.
+
+When reading and writing NeXML documents, `RNeXML` aims to map their
+components to and from, respectively, their most natural
+representations in R. Hence, the types of objects accepted or returned
+by the package's methods are the `phylo` and `multiPhylo` objects from
+the `ape` package [@Paradis_2004] for phylogenies, and R's native
+`data.frame` list structure for character matrices.
+
+## Reading phylogenies and character data
 
@@
 ```
-
-The `nexml` object type is a representation of the NeXML
-document contents, and therefore holds phylogenetic trees as well as 
-character or trait matrices, and all metadata. This is similar in
-design to the classes in the `phylobase` package, which contain 
-trees and traits but hold only limited metadata [@phylobase], but
-different from R object types such as `ape::phylo` (the `phylo` class
-in the `ape` package [@Paradis_2004]), which represents trees alone.
 
@@
 block.
+
+## Writing phylogenies and character data
 

### 2014-12-31 manuscripts/manuscript.Rmd
Populating nexml in-memory objects section.

@@
 Instead of packaging the various components for a NeXML file at the
-time of writing the file, the user can also create and iteratively
-populate an in-memory R object representation of the NeXML document,
-i.e., an object of type `nexml`. The methods to do this are `add_characters()`,
-`add_trees()`, and `add_meta()`, for adding characters, trees, and
-metadata, respectively. Each of these functions will automatically
-create a new nexml object if not supplied with an existing one as the
-last (optional) argument.
+time of writing the file, `RNeXML` also allows users to create and
+iteratively populate an in-memory `nexml` object. The methods to do
+this are `add_characters()`, `add_trees()`, and `add_meta()`, for
+adding characters, trees, and metadata, respectively. Each of these
+functions will automatically create a new nexml object if not supplied
+with an existing one as the last (optional) argument.
 

### 2014-12-31 manuscripts/manuscript.Rmd
First part of metadata section rewrite.

@@
 
-The data components being added to a `nexml` object need not share the
-same OTUs.  For example, here we append additional phylogenies and characters
-using different OTUs to the `nexml` object originally populated with
-the `geospiza` phylogeny and character data:
+The data with which a `nexml` object is populated need not share the
+same OTUs.  `RNeXML` automatically adds new, separate OTU blocks into
+the NeXML file for each data matrix and tree that uses a set of OTUs
+that differs from others.
 
-```{r}
-data(primates)
-nexObj <- add_trees(primates$phy, nexObj)
-nexObj <- add_characters(primates$dat, nexObj)
-```
-
-`RNeXML` automatically adds new, separate OTU blocks into the NeXML file
-for each of these new data sets.
 Other than storage size, there is no limit to the number of
@@
 
-# Metadata
+## Data documentation and annotation with built-in metadata
 
-NeXML allows attaching metadata to any data element, and even to
-metadata themselves. At the level of the document as a whole or the
-phylogeny, metadata can provide bibliographic and provenance
-information, for example about the study as part of which the
-phylogeny was generated or applied, and with which methods it was
-generated. In the form of annotations, metadata can also be attached
-to very specific elements of the data, such as individual OTUs, nodes,
-or even edges of the phylogeny.
+NeXML allows attaching metadata ("_annotations_") to any data element,
+and even to metadata themselves. Whether at the level of the document
+as a whole or the an individual data matrix or phylogeny, metadata can
+provide bibliographic and provenance information, for example about
+the study as part of which the phylogeny was generated or applied,
+which data matrix and which methods were used to generate it. Metadata
+can also be attached to very specific elements of the data, such as
+specific traits, individual OTUs, nodes, or even edges of the
+phylogeny.
 
-At the heart of the NeXML standard is the ability to encode arbitrary
-metadata in precisely specified, machine-readable formats that maintain
-backwards compatibility with any existing NeXML parser. NeXML uses the
-"Resource Description Framework in Annotations" (RDFa),
-as described in @Vos_2012 or @W3C_2014. 
-These annotations can be used to add 
-information to any particular element of the data, (a tree, node, edge,
-or taxonomic unit; a character matrix or specific trait, or even another 
-metadata annotation or the NeXML data file as a whole). `RNeXML` provides
-a range of tools for accessing metadata embedded in a NeXML file.
-
-## Writing NeXML metadata
+NeXML uses the "Resource Description Framework in Annotations" (RDFa),
+as described in @Vos_2012 or @W3C_2014, to encode metadata
+annotations. This standard provides a precisely specified,
+machine-readable format that maintains backwards compatibility with
+compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
+tool to _parse_ annotations is not predicated on _understanding_
+annotations it has not seen before.
 
 The burden typically involved with documenting and sharing good
-metadata alongside the data remains a
-significant barrier to increased sharing and reuse of scientific research
-[@Tenopir_2011]. To promote sharing of well-documented phylogenetic data,
-`RNeXML` aims to provide a simple interface to adding well-defined machine-readable metadata to NeXML files. This occurs on
-several levels:
+metadata alongside the data remains a significant barrier to increased
+sharing and reuse of scientific research [@Tenopir_2011]. To promote
+sharing of well-documented phylogenetic data, `RNeXML` aims to
+alleviate the burden to recording useful and machine-readable metadata
+at several levels.
 
-- Automatic annotation
+- Automatic annotation: 
 - Adding common metadata fields (`add_basic_metadata`)
@@
 
-`RNeXML`'s `nexml_write` function adds certain basic metadata by automatically
-if it has not yet been provided; using default values
-consistent with recommended best practices [@Cranston_2014]. At this time,
-this includes naming the software generating the NeXML, a time-stamp of
-when a tree was produced, and an open data license.
+First, when writing a NeXML file the package adds certain basic
+metadata automatically if it has not yet been provided, using default
+values consistent with recommended best practices
+[@Cranston_2014]. Currently, this includes naming the software
+generating the NeXML, a time-stamp of when a tree was produced, and an
+open data license.
 
-To adjust these values and also add other metadata fields commonly recommended for
-includion with publicly archived data, `RNeXML` provides the `add_basic_meta()`
-function.  This function hides the complexities of anchoring metadata
-attribute names in common vocabularies from the user.  The
-`add_basic_meta()` function takes as input an existing `nexml` object
-(like the other `add_` functions, if none is provided it will create one), and at the time
-of this writing any of the following
-parameters: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
-`citation`.  Other metadata elements and corresponding parameters may
-be added in the future.
+Second, `RNeXML` provides a simple method, called
+`add_basic_metadata()`, to set metadata attributes commonly
+recommended to be included with publicly archived or shared data
+[@Cranston_2014]. The currently accepted parameters include `title`,
+`description`, `creator`, `pubdate`, `rights`, `publisher`, and
+`citation`. Behind the scenes the method automatically anchors these
+attributes in common vocabularies (such as Dublin Core).
 
-Here we create a `nexml` object for the `bird.orders` data
-provided in the `ape` package, and then add basic metadata:
+Third, `RNeXML` includes methods to assist with common cases of adding
+metadata to specific elements of the NeXML data that increase the
+data's utility for reuse. For example, misspellings and inconsistent
+taxonomic names are a frequent source of frustration especially when
+recombining phylogenetic data from different sources. These obstacles
+can be mitigated by annotating the taxonomic names with identifiers,
+which also allows being explicit about the identifier authority being
+used. `RNeXML` includes the `taxize_nexml()` method, which uses the R
+package `taxize` [@Chamberlain_2013] to match each taxon label against
+the NCBI database, and, provided a unique match is found, annotates
+the respective OTU with the matching NCBI identifier,.
 
-```{r}
-data(bird.orders)
-birds <- add_trees(bird.orders)
-birds <- add_basic_meta(
-  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
+## Data annotation with custom metadata extensions
 
-  description = "This data set describes the phylogenetic relationships of the
-     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
-     and Ahlquist inferred this phylogeny from an extensive number of
-     DNA/DNA hybridization experiments. The ``tapestry'' reported by
-     these two authors (more than 1000 species out of the ca. 9000
-     extant bird species) generated a lot of debates.
+The RDFa standard adopted by NeXML (see @Vos_2012) for all metadata
+annotations defines the structure and syntax for such annotations, but
+does not restrict the content. Specifically, every metadata annotation
+consists of a metadata property and a value of the property; the
+bearer of the property is determined by where the annotation appears
+in the NeXML document (technically, the containing element in the XML
+document model). Technically, this structure corresponds directly to
+Resource Description Framework (RDF) subject-predicate-object triples.
+It therefore allows for arbitrary custom extensions to meet ongoing
+scientific needs without requiring modifications (or "extensions") of
+the syntax, provided the metadata annotations are cast as RDF triples.
 
-     The present tree is based on the relationships among orders. The
-     branch lengths were calculated from the values of Delta T50H as
-     found in Sibley and Ahlquist (1990, fig. 353).",
+The `RNeXML` methods described above for built-in metadata annotations
+perform this cast automatically behind the scenes, including mapping
+metadata attributes to terms in requisite common vocabularies (such as
+Dublin Core for "title", "creator", etc.) or ontologies. Giving
+metadata properties (and sometimes values) as vocabulary or ontology
+terms rather than simple text strings is crucial for allowing machines
+to not only parse but also interpret their semantics. To achieve this
+benefit for custom metadata extensions, the user necessarily needs to
+handle certain technical details from which the `RNeXML` API shields
+her otherwise, in particular the vocabularies or ontologies of
+metadata terms in the form of globally unique identifiers (normally
+HTTP URIs). To be consistent with XML terminology for more technical
+users, `RNeXML` calls these _namespaces_, and their abbreviations
+_prefixes_. For example, the namespace for the Dublin Core Metadata
+Terms vocabulary, commonly abbreviated with "dc", is
+"http://purl.org/dc/elements/1.1/". Using its abbreviation, a metadata
+property "dc:title" expands to the identifier
+"http://purl.org/dc/elements/1.1/title", the identifier for the term
+"title" in said vocabulary.
 
-  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
-     classification of birds: a study in molecular evolution. New
-     Haven: Yale University Press.",
-
-  creator = "Sibley, C. G. and Ahlquist, J. E.",
-  nexml = birds)
-```
-
-<!-- the above will get fixed by CB, this is just a reminder and -->
-<!-- placeholder to adjust when that's done. -->
-<!-- Um, sorry, remind me what I am supposed to fix here? formatting? -->
-
-Instead of a literal string, citations can also be provided in R's
-`bibentry` type, which is the one in which R package citations are obtained:
-
-```{r}
-birds <- add_basic_meta(citation = citation("ape"), nexml=birds)
-```
-
-Alternatively, a citation to a published paper with a Digital
-Object Identifier (DOI) can be provided in the form of the DOI,
-which the package `knitcitations` [@Boettiger_2014] can turn into a
-formatted citation using. As an example, to add the above citation
-using the DOI:
-
-```{r }
-library("knitcitations")
-birds <- add_basic_meta(citation = bib_metadata("10.1093/bioinformatics/btg412"), nexml=birds)
-```
-
-## Taxonomic identifiers
-
-`RNeXML` provides various functions to assist in automatically adding
-other more specific metadata to various elements of the NeXML data.
-One common use case is that of taxonomic identifiers.  Misspellings or
-inconsistent use of taxonomic names in phylogenetic data are a common
-source of frustration when reusing phylogenetic trees, especially when
-integrating data from different sources: trees might come from one place
-that uses one taxonomy, character data and occurrence data might come from
-other places that use different taxonomies. Hence, not just misspellings,
-but also synonyms, homonyms and incompatible taxon concepts might plague
-the end user. By associating the taxonomic labels against a specific,
-centralized database we can catch some of these inconsistencies and
-indicate more precisely which classification we are appealing to in the
-event of conflicting taxonomic definitions.  In this example, we indicate
-that we intend the bird orders listed in the tree to correspond to the
-NCBI definitions.
-
-The `taxize_nexml()` function uses the R package `taxize` 
-[@Chamberlain_2013] to check each taxon label against the NCBI database.
-If a unique match is found, a metadata annotation is added to the taxon
-providing the NCBI identification number to the taxonomic unit.
-
-```{r message=FALSE, results='hide'}
-birds <- taxize_nexml(birds, "NCBI")
-```
-If no match is found, the user is warned to check for possible typographic
-errors in the taxonomic labels provided. If multiple matches are found,
-the user will be prompted to choose between them.
-
-## Custom metadata extensions
-
-So far our examples have relied on functions that have a built-in
-understanding of the kind of metadata being provided, and can thus
-generate the appropriate metadata with little further guidance.
-In order to be machine-readable and understandable, NeXML requires
-all metadata annotations follow the RDFa format (see @Vos_2012) which
-explicitly declares the meaning of the annotation.
-
-Some basic definitions about the XML syntax used to express NeXML
-will be helpful for this discussion, though a more comprehensive 
-introduction can be found in @Vos_2012.  For instance, consider how
-the title we gave to the `bird.orders` phylogeny appears in the NeXML file
-using the XML format:
-
-```xml
-<meta id="m2" property="dc:title" datatype="xsd:string"
-      content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
-```
-
-The format begins with an XML "tag" and the name "meta", indicated by the opening
-angle bracket; `<meta`.  A closing tag `\>` indicates the end of this
-XML "element" .  This tag contains several "attributes", named "id",
-"property", etc., that an R user can think of as names in a `list`. What
-follows the equals sign after the attribute name is called the attribute "value",
-(e.g. the `id` attribute has value "m2").
-
-The name of any element (like "meta") or attribute (like "id") in XML must be associated
-with a particular dictionary called a "namespace" and specified using a "prefix"
-of an abbreviation followed by a colon appearing before the name.  If no prefix
-is given, as is the case for all attributes here: `id`, `property`, `datatype`,
-and `content`, then a default prefix specified at the beginning of the document applies,
-in this case, the NeXML namespace.  This allows users to combine different XML
-that have different definitions of terms like `id` or `property`.
-
-The example element above also uses namespaces on some of the attribute values themselves.
-This is a consequence of RDFa annotation, and not strictly part of XML, though
-it works much the same way.  While we hope the following examples will make these
-concepts clearer, users do not need to understand XML and namespaces to use most
-of the basic annotation features provided by `RNeXML`.
-
-So much for the structure; we can now briefly explain the meaning of this element. 
-We see the title appears as the value of the `content` attribute of this `meta` element.
-The `id` attribute provides us with a unique way to refer to this meta
-element in this document.  The `datatype` tells us that the content
-is a text string (as opposed to, say, a numerical or boolean value).
-The `property` is the attribute that most interests us here; it tells us
-that the content is a title.  The prefix `dc` indicates the vocabulary
-(here identified as an XML namespace) to which the word `title` belongs.  The prefix
-is defined elsewhere in the NeXML file, as we can see with:
+`RNeXML` includes methods to obtain and manipulate these values. The
+`get_namespaces()` method accepts a `nexml` object and returns a list
+with element names being the namespace abbreviations and element
+values being the corresponding identifies (typically HTTP URIs):
 

### 2015-01-01 manuscripts/manuscript.Rmd
Removes reminder of previous structure.
Fixes and edits to custom metadata intro.

@@
 tool to _parse_ annotations is not predicated on _understanding_
-annotations it has not seen before.
+the meaning of annotations it has not seen before.
 
@@
 at several levels.
-
-- Automatic annotation: 
-- Adding common metadata fields (`add_basic_metadata`)
-- Taxonomic identifiers (`taxize_nexml`)
-- Generic metadata (`add_meta`)
-- Extending the NeXML standard using RDFa
 
@@
 terms rather than simple text strings is crucial for allowing machines
-to not only parse but also interpret their semantics. To achieve this
-benefit for custom metadata extensions, the user necessarily needs to
-handle certain technical details from which the `RNeXML` API shields
-her otherwise, in particular the vocabularies or ontologies of
-metadata terms in the form of globally unique identifiers (normally
-HTTP URIs). To be consistent with XML terminology for more technical
-users, `RNeXML` calls these _namespaces_, and their abbreviations
-_prefixes_. For example, the namespace for the Dublin Core Metadata
-Terms vocabulary, commonly abbreviated with "dc", is
-"http://purl.org/dc/elements/1.1/". Using its abbreviation, a metadata
-property "dc:title" expands to the identifier
-"http://purl.org/dc/elements/1.1/title", the identifier for the term
-"title" in said vocabulary.
+to not only parse but also interpret and potentially reason over their
+semantics. To achieve this benefit for custom metadata extensions, the
+user necessarily needs to handle certain technical details from which
+the `RNeXML` API shields her otherwise, in particular the vocabularies
+or ontologies of metadata terms in the form of globally unique
+identifiers (normally HTTP URIs). To be consistent with XML
+terminology for more technical users, `RNeXML` calls these
+_namespaces_, and their abbreviations _prefixes_. For example, the
+namespace for the Dublin Core Metadata Terms vocabulary, commonly
+abbreviated with "dc", is "http://purl.org/dc/elements/1.1/". Using
+its abbreviation, a metadata property "dc:title" expands to the
+identifier "http://purl.org/dc/elements/1.1/title", the identifier for
+the term "title" in said vocabulary. When resolved as a HTTP URI, the
+identifier resolves to a human and machine-readable
+(depending on how the identifier is resolved) definition of precisely
+what the term `title` in Dublin Core means. In contrast, just using
+the text string "title" could also mean the title of a person, a legal title, the verb title, etc.  Although URI identifiers of metadata
+vocabularies and terms are not required to resolve, for machines to
+derive the maximum benefit from them, they should resolve to RDF
+statements defining their semantics.
 
-`RNeXML` includes methods to obtain and manipulate these values. The
+`RNeXML` includes methods to obtain and manipulate metadata
+properties, values, identifiers, and namespaces. The
 `get_namespaces()` method accepts a `nexml` object and returns a list
 with element names being the namespace abbreviations and element
-values being the corresponding identifies (typically HTTP URIs):
+values being the corresponding identifiers (typically HTTP URIs) for
+all annotations attached to the object or one of the contained data elements:
 

### 2015-01-01 manuscripts/manuscript.Rmd
Rewrite of rest of custom metadata section.

@@
 
-This URL links us to a web page providing both a human and machine-readable
-version of precisely what the term `title` means, showing us that it
-refers to `title` in the sense of a creative work, rather than the title
-of a person, a legal title, the verb title, etc.  This precision can allow
-us to perform precise computer reasoning on metadata, without relying on
-heuristics or assumptions, as we illustrate later.  For the moment though,
-the need to refer to an explicit controlled vocabulary (or ontology)
-with a url, saying "dc:title" rather than just saying the property is
-"title", is something of a burden in creating metadata.  The previous
-examples involve functions that already know what kind of metadata is
-being provided (a title, a data creator, etc), and so can provide the
-appropriate vocabulary automatically.  The main advantage of linking to
-an external vocabulary, however, is the ability to express concepts that
-are not already "built in" to the data model.  `RNeXML` provides tools
-to generate arbitrary metadata annotations given an external vocabulary,
-such as we illustrate here.
-
-From the list of namespaces in `prefixes`, we can follow links to a handful
-of established vocabularies and find terms that are not included in
-`add_basic_meta()`.  For instance, by resolving the location identified
-by the `prism` prefix,
+The `get_metadata()` method returns, as named list, the metadata
+annotations for a given `nexml` object at a given level, with the
+whole NeXML document being the default:
 
 ```{r}
-prefixes["prism"]
+meta <- get_metadata(birds) 
 ```
 
-we will find that there is a term for `modificationDate` for a file.  We create
-a `meta` element containing this annotation using the `meta` function:
+The `level` parameter can be used to extract the metadata objects for
+the data at specific levels (such as `"otu"` in the following
+example), or by specifying `"all"` to extract all metadata objects:
+
+```{r} 
+otu_meta <- get_metadata(birds, level="otu")
+```
+
+However, because the returned list does not include the data elements
+to which the metadata are attached, using the method in this way is
+only useful for identifying which metadata are present and their
+values. The recommended approach for accessing the metadata attached
+to specific data elements is described below. <!-- or is that in the vignette now? -->
+
+A few select metadata properties for which support is built-in have
+dedicated accessor methods, for example the citation:
+
+```{r get_citation}
+get_citation(birds)
+```
+
+The `meta()` method creates a new metadata object from a poperty name
+and content (value). For example, the following creates a modification
+date metadata object, using the modificationDate property in the PRISM
+vocabulary:
 
@@
 
-We can add this annotation to our existing `birds` NeXML file using the
-`add_meta()` function.  Because we do not specify a level, it is added to
-the root node, referring to the NeXML file as a whole.
+The `add_meta()` function adds metadata objects as annotations to a
+`nexml` object at a specified level. The default level is the NeXML document as a whole:
 
@@
 
-The built-in vocabularies are just the tip of the iceberg of established
-vocabularies. Here we add an annotation from the `skos` namespace which
-describes the history of where the data comes from:
+If the prefix used by the metadata property is not among the built-in
+ones (which can be obtained using the `get_namespaces()` as described
+above), it has to be provided along with its URI as the `namespaces`
+parameter. For example, the following uses the SKOS vocabulary (which
+currently is not built-in) to add a note to the trees in the `nexml`
+object:
 
@@
   content = "Mapped from the bird.orders data in the ape package using RNeXML")
-```
-
-Because `skos` is not in the current namespace list, we add it with a
-url when adding this meta element.  We also specify that this annotation
-be placed at the level of the `trees` sub-node in the NeXML file.
-
-```{r}
 birds <- add_meta(history, 
@@
 
-For finer control of the level at which a `meta` element is added,
-we will manipulate the `nexml` R object directly using S4 sub-setting,
-as shown in the supplement.
-
-Much richer metadata annotation is possible. Later we illustrate how
-metadata annotation can be used to extend the base NeXML format to
-represent new forms of data while maintaining compatibility with any
-NeXML parser. The `RNeXML` package can be easily extended to support
-helper functions such as `taxize_nexml` to add additional metadata
-without imposing a large burden on the user.
-
-## Reading NeXML metadata
-
-A call to the `nexml` object prints some metadata summarizing the data structure:
-
-```{r }
-birds
-```
-
-We can extract all metadata pertaining to the NeXML document as a whole
-(annotations of the XML root node, `<nexml>`) with the command
-
-```{r}
-meta <- get_metadata(birds) 
-```
-
-This returns a named list of available metadata. We can see the kinds
-of metadata recorded from the names (showing the first 4):
-
-```{r}
-names(meta)[1:4]
-```
-
-and can ask for a particular element using the standard list sub-setting
-mechanism (i.e. either the name of an element or its numeric position),
-
-```{r}
-meta[["dc:title"]]
-```
-
-All metadata terms must belong to an explicit *namespace* or vocabulary
-that allows a computer to interpret the term precisely. The prefix (before
-the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
-this case.  The `get_namespaces` function tells us the definition of
-the vocabulary using a link:
-
-```{r}
-prefixes <- get_namespaces(birds)
-prefixes["dc"]
-```
-
-<!-- the citations run off the page -->
-Common metadata can be accessed with a few dedicated functions:
-
-```{r get_citation}
-get_citation(birds)
-```
-
-```{r get_taxa}
-get_taxa(birds)
-```
-
-Which returns text from the otu element labels, typically used to define
-taxonomic names, rather than text from explicit meta elements.
-
-We can also access metadata at a specific level (or use `level=all`
-to extract all meta elements in a list).  Here we show only the first
-few results:
-
-```{r} 
-otu_meta <- get_metadata(birds, level="otu")
-otu_meta[1:4]
-```
-
-This approach does not preserve the relationship between the metadata element and the 
-`otu` it was annotating, and as such is most useful for quick visual exploration. 
-Metadata values for specific elements or sets of elements can always be accessed using
-the S4 sub-setting mechanisms (see section below).
+`RNeXML`, by virtue of subsetting the S4 `nexml` object, also offers
+fine control of the level at which a `meta` element is added, for
+which the package vignette contains examples.
 

### 2015-01-01 manuscripts/manuscript.Rmd
Parked paragraph deleted from earlier text.

@@
 
+Much richer metadata annotation is possible. Later we illustrate how
+metadata annotation can be used to extend the base NeXML format to
+represent new forms of data while maintaining compatibility with any
+NeXML parser. The `RNeXML` package can be easily extended to support
+helper functions such as `taxize_nexml` to add additional metadata
+without imposing a large burden on the user.
+
 No data standard can anticipate all the uses and extensions researchers

### 2015-01-02 manuscripts/manuscript.Rmd
First part of data extension section.

--- 2015-01-01 manuscripts/manuscript.Rmd
@@
 
-## Data annotation with custom metadata extensions
+## Data annotation with custom metadata
 
@@
 
-## RDF manipulation and SPARQL queries
+Because NeXML expresses all metadata using the RDF standard, and
+stores them compliant with RDFa, they can be extracted as an RDF graph
+using off-the-shelf tools, and the RDF graph obtained in this way can
+be queried in SPARQL (see @W3C_2014 or @Hartig_2012), the query
+language for RDF graphs, using general purpose triple stores and
+SPARQL query engines. The `RNeXML` vignette illustrates this with
+examples, and the package also comes with a demonstration that can be
+run from R using the following command: `demo("sparql", "RNeXML")`).
 
-The most powerful way to take advantage of the machine-readable nature
-of the metadata involves the construction of SPARQL queries. This
-approach can exploit the reasoning and logical deduction capacity of
-formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see @W3C_2014 or @Hartig_2012).
+## Using metadata to extend the NeXML standard
 
-The real expressiveness of SPARQL is better illustrated through more
-complex examples such as the one shown in the supplementary materials
-(which can also be run from R using the command: `demo("sparql", "RNeXML")`).
-This demonstration constructs a cladogram based on the taxonomic ranks
-provided in the NeXML. SPARQL allows us to make explicit use of ontological
-concepts such as manipulating sub-classes (i.e. that "Genus" is a subclass 
-of "Family"), or joining multiple predicates (i.e. retrieve the NCBI identifier
-of the OTU that has taxon rank of "Order").
+No data standard can anticipate all the uses and extensions
+researchers may one day need. As the various "flavors" of the NEXUS
+and other data formats show, this often results in extensions to
+widely used exchange formats that almost invariably break
+interoperability between such extensions and tools that were designed
+to read one but not another. NeXML was designed to prevent this by
+means of a highly flexible metadata system, while not sacrificing 
+strict validation of syntax and structure.
 
-## Extending the NeXML standard through metadata annotation.
-
-No data standard can anticipate all the uses and extensions researchers
-may one day need. Researchers have repeatedly adapted the existing NEXUS
-file format to accommodate encoding new kinds of information, but in the
-process break compatibility between different flavors of NEXUS files, with
-few reliable methods to distinguish between these flavors.  NeXML provides
-a solution to this problem through the use of metadata annotation.
-
-Here we illustrate this process using the example of stochastic character
-mapping [@Huelsenbeck_2003]. A stochastic character map is simply
-an annotation of the branches on a phylogeny, assigning each section
-of each branch to a particular "state" (typically of a morphological
-characteristic).
-
-@Bollback_2006 provides a widely used stand-alone software implementation
-of this method in the software `simmap`, which modified the standard
-Newick tree format to express this additional information. This can
-break compatibility with other software [^3], and creates a format that
-cannot be interpreted without additional information describing this
-convention.  By contrast, the NeXML extension is not only backwards
-compatible but contains a precise and machine-readable description of
-what it is encoding.
-
-In this example, we illustrate how the additional information required
-to define a stochastic character mapping (a `simmap` mapping) in NeXML.
+Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
+it, can be used to record and share a type of phylogenetic data not
+taken into account when NeXML was designed. Specifically, we use the
+example of stochastic character maps [@Huelsenbeck_2003], which assign
+sections of each branch in a phylogeny to a particular "state"
+(typically of a morphological characteristic). The current de-facto
+format for sharing stochastic character maps is a (non-interoperable)
+modification of the standard Newick tree format created by a software
+tool widely used for creating such maps, `simmap` [@Bollback_2006].
 

### 2015-01-03 manuscripts/manuscript.Rmd
Parked paragraph from earlier text.
Fixes up intro to RNeXML package.

@@
 
-`RNeXML` is an R package entirely written in R, and the current stable
-version can be installed from the CRAN library[^1]:
+The `RNeXML` package is written entirely in R and available under a
+CC0 public domain waiver. The current development version can be found
+on Github at [https://github.com/ropensci/RNeXML](), and the stable
+version can be installed from the CRAN repository:
 
@@
 ```
+
+`RNeXML` is part of the rOpenSci project. Users of `RNeXML` are
+encouraged to submit bug reports or feature requests in the issues log
+on Github, or to contact `ropensci-discuss@googlegroups.com` or the
+phylogenetics R users group list at `r-sig-phylo@r-project.org` for
+help.
 
@@
 
+By contrast, the NeXML extension is not only backwards
+compatible but contains a precise and machine-readable description of
+what it is encoding.
+
 <!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?

### 2015-01-03 manuscripts/manuscript.Rmd
Fixes up the custom metadata section.

@@
 
+Metadata annotatuons in `NeXML` can be nested within another
+annotation, which the `meta()` method accommodates by accepting a
+parameter `children`, with the list of nested metadata objects (which
+can themselves be nested) as value.
+
 The `add_meta()` function adds metadata objects as annotations to a
@@
 ```
+
+Alternatively, additional namespaces can also be added in batch using
+the `add_namespaces()` method, which makes a list of namespaces known
+to a given `nexml` object (or created a new `nexml` object, similar to
+other `add_*` methods).
 

### 2015-01-03 manuscripts/manuscript.Rmd
Rewrite of first part of the simmap section.

@@
 example of stochastic character maps [@Huelsenbeck_2003], which assign
-sections of each branch in a phylogeny to a particular "state"
-(typically of a morphological characteristic). The current de-facto
-format for sharing stochastic character maps is a (non-interoperable)
-modification of the standard Newick tree format created by a software
-tool widely used for creating such maps, `simmap` [@Bollback_2006].
+certain parts (corresponding to time) of each branch in a
+time-calibrated phylogeny to a particular "state" (typically of a
+morphological characteristic). The current de-facto format for sharing
+stochastic character maps is a (non-interoperable) modification of the
+standard Newick tree format created by a software tool widely used for
+creating such maps, `simmap` [@Bollback_2006].
 
-@Revell_2012 describes the `phytools` package for R, which includes
-utilities for reading, manipulating, and writing `simmap` files in R.
-In this example, we also show how to define `RNeXML` functions that 
-map the R representation used by Revell (an extension of the `ape` class)
-into the NeXML extension we have defined by using `RNeXML` functions.
-
-Since a stochastic character map simply assigns different states to
-parts of a branch (or edge) on the phylogenetic tree, we can create
-a NeXML representation by annotating the `edge` elements with appropriate
-`meta` elements.  These elements need to describe the character state
-being assigned and the duration (in terms of branch-length) that the edge 
-spends in that state (Stochastic character maps are specific to time-calibrated
-or ultrametric trees).
-
-NeXML already defines the `characters` element to handle discrete character traits (`nex:char`)
-and the states they can assume (`nex:state`).  We will thus reuse the `characters` element for
-this purpose, referring to both the character trait and the states by the ids assigned to them
-in that element.  (NeXML's convention of referring to everything by id permits a single canonical
-definition of each term, making it clear where additional annotation belongs).  For each edge, we 
-need to indicate:
-
-By contrast, the NeXML extension is not only backwards
-compatible but contains a precise and machine-readable description of
-what it is encoding.
-
-<!-- Ugh, do we need to explain that everything in NeXML uses an id so that it can be reused in this way?
-     -->
-
-- That our annotation contains a stochastic character mapping reconstruction
-- Since many reconstructions are possible for a single edge, we give each reconstruction an id
-- We indicate for which character trait we are defining the reconstruction 
-- We then indicate which states the character assumes on that edge. 
-  For each state realized on the edge, that involves stating: 
-    + the state assignment
-    + the duration (length of time) for which the edge spends in the given state
-    + the order in which the state changes happen (Though we could just assume 
-      state transitions are listed chronologically, NeXML suggests making all 
-      data explicit, rather than relying on the structure of the data file to
-      convey information).
-
-Thus the annotation for an edge that switches from state `s2` to state 
-`s1` of character `cr1` would be constructed like this:
+To express stochastic character map data as a custom metadata
+extension, we create metadata annotations for the `edge` elements in
+the phylogeny topology that relate parts of the annotated edge (branch)
+to different states. These annotations need to describe the character state
+being assigned, and the duration (in terms of branch length) that the edge 
+spends in that state. The `NeXML` standard already defines `state`
+elements for discrete character traits, which, as most elements in
+NeXML, also have unique identifies (attribute `id`) by which they can
+be referenced in other parts of the document. In the following
+proposal, we call each series of character state assignments for an edge a
+_reconstruction_, and for each state assignment comprising the
+reconstruction, we provide the length, the identifier of the character
+state, and an ordinal property to record the chronological order of
+the state assignment (the identifier values in the proposal, "cr1",
+"s1", and "s2", are -- fictitious -- identifiers that, in contrast to
+global identifiers, are unique only within a given NeXML document):
 

### 2015-01-03 manuscripts/manuscript.Rmd
Rewrite of simmap section, part II.

@@
 
-Of course writing out such a definition manually becomes tedious quickly. Because
-these are just R commands, we can easily define a function that can loop over an
-assignment like this for each edge, extracting the appropriate order, length and
-state from an existing R object such as that provided in the `phytools` package.  
-Likewise, it is straightforward to define a function that reads this data using
-the `RNeXML` utilities and converts it back to the `phytools` package. The full
-implementation of this mapping can be seen in the `simmap_to_nexml()` and the
-`nexml_to_simmap()` functions provided in the `RNeXML` package.
-
-As the code indicates, the key step is simply to define the data in meta elements. In 
-so doing, we have defined a custom namespace, `simmap`, to hold our variables.  This
-allows us to provide a URL with more detailed descriptions of what each of these 
-elements mean:
+The example uses a prefix, `simmap`, to group the newly introduced
+metadata properties in a vocabulary, for which the `add_namespace()`
+method can be used to give a URI as an identifier:
 
@@
 
-At that URL we have posted a simple description of each term.
+As we illustrate here, even if the URI does not resolve to a fully
+machine-readable definition of the terms and their semantics, it can
+still be used to provide at least a human-readable informal definition
+of the terms, and where to find such documentation _is_
+machine-readable in the resulting NeXML document. Aside from the fact
+that the above representation does not modify the NeXML format, and
+should thus be readable -- even if not necessarily interpretable as a
+stochastic character mapping -- by any compliant NeXML parser, the
+direct and machine-readable link to metadata documentation contrasts
+with metadata or format extensions that are undocumented, or which do
+not directly link to documentation.
 
-Using this convention we can generate NeXML files containing `simmap`
-data, read those files into R, and convert them back into the `phytools`
-package format. These simple functions serve as further illustration of
-how `RNeXML` can be used to extend the NeXML standard.  We illustrate
-their use briefly here, starting with  loading a `nexml` object containing
-a `simmap` reconstruction into R:
-
-```{r}
-data(simmap_ex)
-```
-
-The `get_trees()` function can be used to return an `ape::phylo` tree as
-usual.  `RNeXML` automatically detects the `simmap` reconstruction data
-and returns includes this in a `maps` element of the `ape::phylo` object,
-for use with other `phytools` functions.
-
-```{r}
-phy <- nexml_to_simmap(simmap_ex)
-```
-
-We can then use various functions from `phytools` designed for `simmap`
-objects [@Revell_2012], such as the plotting function:
-
-```{r Figure1, fig.cap="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML."}
-library("phytools")
-plotSimmap(phy)
-```
-
-Likewise, we can convert the object back in the NeXML format and write
-it out to file to be read by other users.
-
-```{r}
-nex <- simmap_to_nexml(phy) 
-nexml_write(nex, "simmap.xml")
-```
-
-Though other NeXML parsers (for instance, for Perl or Python) have
-not been written explicitly to express `simmap` data, those parsers will
-nonetheless be able to successfully parse this file and expose the `simmap`
-data to the user.
+Although writing out metadata annotations as in the above proposal may
+seem tedious, by virtue of `RNeXML`'s API they can be expressed as a
+series of R commands, and it is therefore straightforward to define
+functions that encapsulate the details behind a parameter
+signature. To demonstrate this, we implemented two functions in
+`RNeXML`, `simmap_to_nexml()` and `nexml_to_simmap()`, that map
+bidirectionally between NeXML files with `simmap` metadata according
+to the above proposal, and the `simmap` data representation in R
+devised by @Revell_2012 in the R package `phytools`. The latter uses
+an extension of the `ape::phylo` class. As shown in more detail in the
+`RNeXML` package vignette, this allows generating NeXML files
+containing `simmap` data, reading those files into R, converting them
+into the in-memory representation understood by `phytools`, and using
+the various functions from `phytools` designed for `simmap` objects
+[@Revell_2012], such as `plotSimmap()` for plotting.
 

### 2015-01-03 manuscripts/manuscript.Rmd
Rewrite of the publishing to repositories section.

@@
 
-Data archiving is increasingly required by journals in evolutionary biology and 
-biodiversity (e.g. @Rausher_2010).  The burden of preparing and submitting properly
-annotated data to archives continues to be a significant barrier [@Tenopir_2011; @Stodden_2014],
-and many phylogenetic trees upon which studies are based are inaccessible or lost 
-to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
-by providing functions to immediately archive NeXML files in scientific repositories.
+Data archiving is increasingly required by scientific journals,
+including in evolutionary biology, ecology, and biodiversity
+(e.g. @Rausher_2010). The effort involved with preparing and
+submitting properly annotated data to archives remains a notable
+barrier to the broad adoption of data archiving and sharing as a
+normal part of the scholarly publication workflow
+[@Tenopir_2011; @Stodden_2014]. In particular, the majority of
+phylogenetic trees published in the scholarly record are inaccessible
+or lost to the research community [@Drew_2013].
+
+One of `RNeXML`'s aims is to promote the archival of well-documented
+phylogenetic data in scientific data repositories, in the form of
+NeXML files. To this end, the method `nexml_publish()` provides an API
+directly from within R that allows data archival to become a step programmed
+into data management scripts, rather than a separate manually executed
+act. Initially, the method supports the data repository Figshare (http://figshare.com):
 
@@
 
-Robust data archiving need not involve immediate release of the data. The `figshare` 
-repository supports the option of `visibility = "private"`, allowing secure 
-publication that provides a way to backup and share data with collaborators 
-prior to public release.
+Figshare also supports "private" visibility, allowing to securely backup
+data to a repository and to share them with collaborators prior to public release.
 
-# Software reuse and support
-
-The `RNeXML` package is available through the CRAN repository under
-a CC0 license. At the time of writing, the current version is `r packageDescription("RNeXML", fields="Version")`. 
-The current development
-branch can be found on Github at [https://github.com/ropensci/RNeXML]().
-See the NEWS file for changes in recent versions. Users are encouraged
-to submit bug reports or feature requests in the issues log.  Contact
-`ropensci-discuss@googlegroups.com` or the package maintainer for help,
-or email the phylogenetics R users group list at `r-sig-phylo@r-project.org`.
-
-There are many promising directions for further development of the
-`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
-steps currently planned or under consideration.  In particular, these
-milestones include increased support for interactive mode of entering data
-(wizards), conversions to and from other existing R phylogenetics formats
-such those found in the `phylobase` package [@phylobase],
-and support for integration of semantics defined through the
-Web Ontology language (OWL), as discussed in the Challenges section.
-
-## Future Directions
-
-For convenience, the RNeXML package can parse compressed XML binaries (gzipped XML) directly, facilitating work with very large data files.
-As most R functions, especially those in common phylogenetics packages, require in-memory data structures, RNeXML currently loads this data directly into memory.
-Future work could facilitate manipulation of NeXML data directly from the disk.
+# Future Directions
 

### 2015-01-03 manuscripts/manuscript.Rmd
Edits to Future Directions.

@@
 
-Emerging formats such as the JSON based NeXSON format provide alternate representations of semantic, machine-readable phylogenetic data.
-At this time, R users must still rely on other tools that can convert these formats into NeXML.
-Because the NeXML syntax is extensible, semantic, and machine-readable, this can be done without loss, unlike NEXUS formats.
-Nevertheless, a possible future direction would be to provide direct support of such formats through RNeXML or a companion R package.
+The goals for continued development of `RNeXML` revolve primarily
+around three directions. One is better interoperability with other
+existing phylogenetic data representations in R, such those found in
+the `phylobase` package [@phylobase]. Another is better integration of
+the rich metadata semantics found in ontologies defined in the Web
+Ontology Language (OWL), including programmatic access to machine
+reasoning with such metadata. Finally, and similar to many other R
+packages, including those for comparative phylogenetics, in light of
+the increasing size of phylogenomic and other datasets mechanisms may
+eventually be needed that avoid the requirement to load an entire
+file's content into memory.
 

### 2015-01-03 manuscripts/manuscript.Rmd
Fixes and additions to Acknowledgments.

@@
 ## Acknowledgements
-<!-- FIXME Rutger, Hilmar, add whatever you'd like to acknowledge here -->.
 
-This project was supported in part by a grant from the Alfred P Sloan
-Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB), ... 
-We thank F. Michonneau for helpful comments on an earlier version
-of this manuscript.
-`RNeXML` is part of the rOpenSci project, [http://ropensci.org].
+This project was supported in part by the National Evolutionary
+Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
+Natioal Science Foundation (DBI-1306697) and the Alfred P Sloan
+Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
+the Google Summer of Code, and we thank Kseniia Shumelchyk for taking
+the first steps to implement it. We are grateful to F. Michonneau for
+helpful comments on an earlier version of this manuscript.
 

### 2015-01-03 manuscripts/manuscript.Rmd
Removes footnote #3, which is now obsolete.

@@
 for users familiar with the somewhat analogous `read.nexus` function.
-
-[^3]: By using the commenting mechanism of the Newick format, it is
-possible that other software that doesn't also use the comment mechanism
-for some other such purpose would be able to successfully parse the tree.
-However there is no way to guarantee that this is the case or for the
-data format to describe its use.

### 2015-01-05 manuscripts/manuscript.Rmd
Tightening up: removing (now partly obsolete) outline.

@@
 ---
-
-<!------------------------------------------------
-
-Map/Outline:
-------------
-
-- Introduction
-  - Exciting times in phylogenetics!  
-  - Challenges with existing formats 
-  - Overview of NeXML
-  - Challenges with R
-  - Map
-
-- Tutorial
-  - Reading Phylogenetic trees
-  - Reading Character data
-  - Writing phylogenetic trees & character data
-
-  - Metadata
-    - Adding basic metadata
-    - Adding taxonomic identifiers
-    - Adding Generic metadata
-    - Parsing metadata. XML. RDF.
-    - Extending NeXML with metadata: Example defining simmap stochastic character mappings.
-
-  - Publishing NeXML
-
-- Implementation, Quality Control
-- Software reuse and support
-
------------------------------------------------->
 

### 2015-01-05 manuscripts/manuscript.Rmd
Tightening up: Introduction

@@
 integrate previously generated or published data and metadata. The
-resulting expanding data exchange needs of the evolutionary research
-community are rapidly outpacing the capabilities of the currently
-still most widely used data exchange standards, which were all
+expanding data exchange needs of the evolutionary research
+community are rapidly outpacing the capabilities of most current
+and widely used data exchange standards [@Vos_2012], which were all
 developed a decade or more ago. This has resulted in a radiation of
@@
 exchange in phylogenetics, a modern XML-based exchange standard,
-called NeXML, has recently been developed [@Vos_2012], and is
-beginning to be adopted. NeXML comprehensively supports current data
-exchange needs, is predictably machine-readable, and is forward
-compatible. To make the capibilities of NeXML available to R users in
-an easy-to-use form, and to lower the hurdles to adoption of the
-standard, we present RNeXML, an R package that aims to provide easy
-programmatic access to reading and writing NeXML documents, tailored
-for the kinds of use-cases that will be common for users and
-developers of the wealth of evolutionary, and in particular
-comparative analysis methods within the R ecosystem.
+called NeXML, has recently been developed [@Vos_2012]. NeXML
+comprehensively supports current data exchange needs, is predictably
+machine-readable, and is forward compatible.
 
@@
 without the otherwise common loss of most data and metadata semantics
-[@Drew_2013; @Stoltzfus_2012], or with semantics that is predictably
-machine interpretable [@Cranston_2014]. For example, the still popular
+[@Drew_2013; @Stoltzfus_2012; @Cranston_2014]. For example, the still popular
 NEXUS file format [@Maddison_1997] cannot consistently represent
 horizontal gene transfer or ambiguity in reading a character (such as
-a DNA sequence base pair).  This has led to modifications of NEXUS in
+a DNA sequence base pair).  This and other limitations have led to modifications of NEXUS in
 different ways for different needs, with the unfortunate result that

### 2015-01-05 manuscripts/manuscript.Rmd
Tightening up various places in package description.

@@
 The `RNeXML` package is written entirely in R and available under a
-CC0 public domain waiver. The current development version can be found
+Creative Commons Zero public domain waiver. The current development version can be found
 on Github at [https://github.com/ropensci/RNeXML](), and the stable
-version can be installed from the CRAN repository:
-
-```r
-install.packages("RNeXML", dependencies=TRUE)
-```
-
-This will also install its dependencies. Once installed, R scripts
-using the library only need to load it:
+version can be installed from the CRAN repository.
 
 ```{r echo=FALSE}
+install.packages("RNeXML", dependencies=TRUE)
 library(RNeXML)
@@
 Conceptually, a NeXML document has the following components: (1)
-phylogeny data, (2) character or trait data in matrix form, (3)
-operational taxonomic units (OTUs), and (4) metadata. `RNeXML` defines
-the `nexml` object type, which is a representation of the NeXML
-document contents. It therefore holds phylogenetic trees as well as
-character or trait matrices, and all metadata. This is similar in
-design to the classes in the `phylobase` package, which contain trees
-and traits but hold only limited metadata [@phylobase], but different
-from R object types such as `ape::phylo` (the `phylo` class in the
-`ape` package [@Paradis_2004]), which represents trees alone.
+phylogeny topology data, (2) character or trait data in matrix form, (3)
+operational taxonomic units (OTUs), and (4) metadata. To represent the
+contents of a NeXML document (currently in memory), `RNeXML` defines
+the `nexml` object type. This type therefore holds phylogenetic trees
+as well as character or trait matrices, and all metadata, which is
+similar to the phylogenetic data object types defined in the
+`phylobase` package [@phylobase], but contrasts with the more widely
+used ones defined in the
+`ape` package [@Paradis_2004], which represents trees alone.
 
 When reading and writing NeXML documents, `RNeXML` aims to map their
-components to and from, respectively, their most natural
-representations in R. Hence, the types of objects accepted or returned
+components to and from, respectively, their most widely used
+representations in R. As a result, the types of objects accepted or returned
 by the package's methods are the `phylo` and `multiPhylo` objects from
 the `ape` package [@Paradis_2004] for phylogenies, and R's native
-`data.frame` list structure for character matrices.
+`data.frame` list structure for data matrices.
 
@@
 file, or from a remote location via its URL, and returns an object of
-type `nexml` [^2]:
+type `nexml`:
 
@@
 
-To interface conveniently with the many already existing R packages for 
-phylogenetics, which conventionally accept phylogenies as `ape::phylo`
-objects, the method `get_trees_list()` returns a `ape::multiPhylo`
+The method `get_trees_list()` can be used to extract the phylogenies
+as an `ape::multiPhylo`
 object, which can be treated as a list of `ape::phylo` objects:
@@
 
-The `get_trees_list()` method is designed for use in scripts, for
-which a consistent and predictable return type (namely a `list` of
-`ape::multiPhylo` objects, reflecting the nested structure of NeXML)
-is desirable. `RNeXML` also contains a method `get_trees()`, which
-is designed for interactive use and returns the R object most
+The `get_trees_list()` method is designed for use in scripts, providing
+a consistent and predictable return type regardless of the number of
+phylogenies a NeXML document contains. For greater convenience in
+interactive use, the method `get_trees()` returns the R object most
 intuitive given the arrangement of phylogeny data in the source NeXML
-document. NeXML inherits from NEXUS the ability to contain multiple
-phylogenies, and to group sets of phylogenies into "blocks". The
-`get_trees()` method will return an `ape::phylo` object if the NeXML
+document. For example, the method returns an `ape::phylo` object if the NeXML
 document contains a single phylogeny, an `ape::multiPhylo` object if
@@
 and a list of `ape::multiPhylo` objects if it contains multiple
-`trees` blocks. This design spares an interactive user from having to
-type unnecessary data conversion commands, at the expense of needing
-to know which return type to expect (or verifying the type via
-introspection commands such as `class()`).
+`trees` blocks (the capability for which NeXML inherits from NEXUS).
 
@@
 phylogenetic data repository TreeBASE [@Piel_2002]. For
-example, the following plots a tree in TreeBASE:
+example, the following plots a tree in TreeBASE (using ape's plot function):
 
-```{r eval=FALSE}
+```{r eval=FALSE fig.show="hide"}
 tb_nex <- nexml_read(
@@
 tb_phy <- get_trees_list(tb_nex)
+plot(tb_phy[[1]]) 
 ```
 
-The method `get_characters()` obtains character data matrics from a
+The method `get_characters()` obtains character data matrices from a
 `nexml` object, and returns them as a standard `data.frame` R object
@@
 alignments), discrete (for most morphological character data), or
-continuous type (for many trait data).  Data matrices can be, and
-matrices of different data types are required to be separated into
-different "blocks", to allow strict validation of data types. Since
+continuous type (for many trait data). To enable strict validation of data types NeXML allows, and if their
+data types differ requires multiple data matrices to be separated into
+different "blocks". Since
 the `data.frame` data structure in R has no such constraints, the
-`get_characters()` method will combine such blocks as separate columns
+`get_characters()` method combines such blocks as separate columns
 into a single `data.frame` object, provided they correspond to the
@@
 
-The first argument to `nexml_write()` is an object of type `nexml`, or any
+The first argument to `nexml_write()` is either an object of type `nexml`, or any
 object that can be coerced to it, such as in the above example an
@@
 time of writing the file, `RNeXML` also allows users to create and
-iteratively populate an in-memory `nexml` object. The methods to do
+iteratively populate in-memory `nexml` objects. The methods to do
 this are `add_characters()`, `add_trees()`, and `add_meta()`, for
@@
 same OTUs.  `RNeXML` automatically adds new, separate OTU blocks into
-the NeXML file for each data matrix and tree that uses a set of OTUs
-that differs from others.
+the NeXML file for each data matrix and tree that uses a different set of OTUs.
 
@@
 phylogenies and character matrices that can be included in a
-single NeXML file. This allows, for example, to capture samples from a
-posterior probability distribution of inferred or simulated phylogenies or
-character states in one NeXML file.
+single NeXML document. This allows, for example, to capture samples from a
+posterior probability distribution of inferred or simulated phylogenies and
+character states in a single NeXML file.
 
@@
 
-NeXML allows attaching metadata ("_annotations_") to any data element,
+NeXML allows attaching ("_annotating_") metadata to any data element,
 and even to metadata themselves. Whether at the level of the document
-as a whole or the an individual data matrix or phylogeny, metadata can
+as a whole or an individual data matrix or phylogeny, metadata can
 provide bibliographic and provenance information, for example about
@@
 
-NeXML uses the "Resource Description Framework in Annotations" (RDFa),
-as described in @Vos_2012 or @W3C_2014, to encode metadata
-annotations. This standard provides a precisely specified,
-machine-readable format that maintains backwards compatibility with
+As described in @Vos_2012,  to encode metadata annotations NeXML uses
+the "Resource Description Framework in Annotations" (RDFa)
+[@W3C_2014]. This standard provides for a strict machine-readable
+format yet enables future backwards compatibility with
 compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
@@
 
-The burden typically involved with documenting and sharing good
-metadata alongside the data remains a significant barrier to increased
-sharing and reuse of scientific research [@Tenopir_2011]. To promote
-sharing of well-documented phylogenetic data, `RNeXML` aims to
-alleviate the burden to recording useful and machine-readable metadata
-at several levels.
+To lower the barriers to sharing well-documented phylogenetic data,
+`RNeXML` aims to make recording useful and machine-readable metadata
+easier at several levels.
 
 First, when writing a NeXML file the package adds certain basic
-metadata automatically if it has not yet been provided, using default
+metadata automatically if they are absent, using default
 values consistent with recommended best practices
@@
 `add_basic_metadata()`, to set metadata attributes commonly
-recommended to be included with publicly archived or shared data
+recommended for inclusion with data to be publicly archived or shared
 [@Cranston_2014]. The currently accepted parameters include `title`,
@@
 
-Third, `RNeXML` includes methods to assist with common cases of adding
-metadata to specific elements of the NeXML data that increase the
-data's utility for reuse. For example, misspellings and inconsistent
-taxonomic names are a frequent source of frustration especially when
-recombining phylogenetic data from different sources. These obstacles
-can be mitigated by annotating the taxonomic names with identifiers,
-which also allows being explicit about the identifier authority being
-used. `RNeXML` includes the `taxize_nexml()` method, which uses the R
-package `taxize` [@Chamberlain_2013] to match each taxon label against
-the NCBI database, and, provided a unique match is found, annotates
-the respective OTU with the matching NCBI identifier,.
+Third, `RNeXML` integrates with the R package `taxize`
+[@Chamberlain_2013] to mitigate one of the most common obstacles to
+reuse of phylogenetic data, namely the misspellings and inconsistent
+taxonomic naming with which OTU labels are often fraught. The
+`taxize_nexml()` method in `RNeXML` uses `taxize` to match OTU labels
+against the NCBI database, and, where a unique match is found, it
+annotates the respective OTU with the matching NCBI identifier.
 
@@
 
-The RDFa standard adopted by NeXML (see @Vos_2012) for all metadata
-annotations defines the structure and syntax for such annotations, but
-does not restrict the content. Specifically, every metadata annotation
-consists of a metadata property and a value of the property; the
-bearer of the property is determined by where the annotation appears
-in the NeXML document (technically, the containing element in the XML
-document model). Technically, this structure corresponds directly to
-Resource Description Framework (RDF) subject-predicate-object triples.
-It therefore allows for arbitrary custom extensions to meet ongoing
-scientific needs without requiring modifications (or "extensions") of
-the syntax, provided the metadata annotations are cast as RDF triples.
+The RDFa standard adopted by NeXML (see @Vos_2012) allows for
+arbitrary custom extensions to meet ongoing scientific needs without
+requiring modifications (or "extensions") of the syntax. Specifically,
+every metadata annotation consists of a metadata property and a value
+of the property; the bearer of the property is determined by where the
+annotation appears in the NeXML document (technically, the containing
+element in the XML document model). This structure corresponds
+directly to Resource Description Framework (RDF)
+subject-predicate-object triples, and therefore the only requirement
+for custom metadata is that they be cast as RDF triples.
 
-The `RNeXML` methods described above for built-in metadata annotations
+The `RNeXML` interface described above for built-in metadata annotations
 perform this cast automatically behind the scenes, including mapping
@@
 Dublin Core for "title", "creator", etc.) or ontologies. Giving
-metadata properties (and sometimes values) as vocabulary or ontology
+metadata properties (and where applicable, values) as vocabulary or ontology
 terms rather than simple text strings is crucial for allowing machines
@@
 user necessarily needs to handle certain technical details from which
-the `RNeXML` API shields her otherwise, in particular the vocabularies
-or ontologies of metadata terms in the form of globally unique
-identifiers (normally HTTP URIs). To be consistent with XML
-terminology for more technical users, `RNeXML` calls these
+the `RNeXML` API shields her otherwise, in particular the globally unique
+identifiers (normally HTTP URIs) of metadata terms and vocabularies. To be consistent with XML
+terminology, `RNeXML` calls vocabulary URIs
 _namespaces_, and their abbreviations _prefixes_. For example, the
-namespace for the Dublin Core Metadata Terms vocabulary, commonly
-abbreviated with "dc", is "http://purl.org/dc/elements/1.1/". Using
-its abbreviation, a metadata property "dc:title" expands to the
-identifier "http://purl.org/dc/elements/1.1/title", the identifier for
-the term "title" in said vocabulary. When resolved as a HTTP URI, the
-identifier resolves to a human and machine-readable
-(depending on how the identifier is resolved) definition of precisely
+namespace for the Dublin Core Metadata Terms vocabulary is "http://purl.org/dc/elements/1.1/". Using
+its common abbreviation "dc", a metadata property "dc:title" expands to the
+identifier "http://purl.org/dc/elements/1.1/title". This URI resolves
+to a human and machine-readable (depending on access) definition of precisely
 what the term `title` in Dublin Core means. In contrast, just using
-the text string "title" could also mean the title of a person, a legal title, the verb title, etc.  Although URI identifiers of metadata
-vocabularies and terms are not required to resolve, for machines to
-derive the maximum benefit from them, they should resolve to RDF
-statements defining their semantics.
+the text string "title" could also mean the title of a person, a legal
+title, the verb title, etc.  URI identifiers of metadata vocabularies
+and terms are not mandated to resolve, but if machines are to derive
+the maximum benefit from them, they should resolve to a definition of
+their semantics in RDF.
 
@@
 properties, values, identifiers, and namespaces. The
-`get_namespaces()` method accepts a `nexml` object and returns a list
-with element names being the namespace abbreviations and element
-values being the corresponding identifiers (typically HTTP URIs) for
-all annotations attached to the object or one of the contained data elements:
+`get_namespaces()` method accepts a `nexml` object and returns a named list
+of namespace prefixes and their corresponding identifiers known to the object:
 
@@
 
-The `get_metadata()` method returns, as named list, the metadata
+The `get_metadata()` method returns, as a named list, the metadata
 annotations for a given `nexml` object at a given level, with the
-whole NeXML document being the default:
+whole NeXML document being the default level (`"all"` extracts all metadata objects):
 
@@
 meta <- get_metadata(birds) 
-```
-
-The `level` parameter can be used to extract the metadata objects for
-the data at specific levels (such as `"otu"` in the following
-example), or by specifying `"all"` to extract all metadata objects:
-
-```{r} 
 otu_meta <- get_metadata(birds, level="otu")
@@
 
-However, because the returned list does not include the data elements
-to which the metadata are attached, using the method in this way is
-only useful for identifying which metadata are present and their
-values. The recommended approach for accessing the metadata attached
-to specific data elements is described below. <!-- or is that in the vignette now? -->
-
-A few select metadata properties for which support is built-in have
-dedicated accessor methods, for example the citation:
-
-```{r get_citation}
-get_citation(birds)
-```
+The returned list does not include the data elements to which the
+metadata are attached. Therefore, a different approach, documented in
+the package vignette, is recommended for accessing the metadata
+attached to data elements.
 
@@
 and content (value). For example, the following creates a modification
-date metadata object, using the modificationDate property in the PRISM
-vocabulary:
+date metadata object, using a property in the PRISM vocabulary:
 
@@
 The `add_meta()` function adds metadata objects as annotations to a
-`nexml` object at a specified level. The default level is the NeXML document as a whole:
+`nexml` object at a specified level, with the default level being the
+NeXML document as a whole:
 
@@
 If the prefix used by the metadata property is not among the built-in
-ones (which can be obtained using the `get_namespaces()` as described
-above), it has to be provided along with its URI as the `namespaces`
+ones (which can be obtained using `get_namespaces()`), it has to be provided along with its URI as the `namespaces`
 parameter. For example, the following uses the SKOS vocabulary (which
@@
 Alternatively, additional namespaces can also be added in batch using
-the `add_namespaces()` method, which makes a list of namespaces known
-to a given `nexml` object (or created a new `nexml` object, similar to
-other `add_*` methods).
+the `add_namespaces()` method.
 
-`RNeXML`, by virtue of subsetting the S4 `nexml` object, also offers
-fine control of the level at which a `meta` element is added, for
-which the package vignette contains examples.
+By virtue of subsetting the S4 `nexml` object, `RNeXML` also offers
+fine control of where a `meta` element is added, for which the package
+vignette contains examples.
 
 Because NeXML expresses all metadata using the RDF standard, and
-stores them compliant with RDFa, they can be extracted as an RDF graph
-using off-the-shelf tools, and the RDF graph obtained in this way can
-be queried in SPARQL (see @W3C_2014 or @Hartig_2012), the query
-language for RDF graphs, using general purpose triple stores and
-SPARQL query engines. The `RNeXML` vignette illustrates this with
-examples, and the package also comes with a demonstration that can be
-run from R using the following command: `demo("sparql", "RNeXML")`).
+stores them compliant with RDFa, they can be extracted as an RDF
+graph, queried, analyzed, and mashed up with other RDF data, local or
+on the web, using a wealth of off-the-shelf tools for working with RDF
+(see @W3C_2014 or @Hartig_2012). Examples for these possibilities are
+included in the `RNeXML` vignette, and the package also comes with a
+demonstration that can be run from R using the following command:
+`demo("sparql", "RNeXML")`).
 
@@
 
-No data standard can anticipate all the uses and extensions
-researchers may one day need. As the various "flavors" of the NEXUS
-and other data formats show, this often results in extensions to
-widely used exchange formats that almost invariably break
-interoperability between such extensions and tools that were designed
-to read one but not another. NeXML was designed to prevent this by
-means of a highly flexible metadata system, while not sacrificing 
-strict validation of syntax and structure.
+NeXML was designed to prevent the need for future non-interoperable
+"flavors" of the standard in response to new research directions. Its
+solution to this inevitable problem is a highly flexible metadata
+system without sacrificing strict validation of syntax and structure.
 
-Here we illustrate how NeXML's metadata system, and `RNeXML`'s API to
-it, can be used to record and share a type of phylogenetic data not
-taken into account when NeXML was designed. Specifically, we use the
-example of stochastic character maps [@Huelsenbeck_2003], which assign
-certain parts (corresponding to time) of each branch in a
-time-calibrated phylogeny to a particular "state" (typically of a
-morphological characteristic). The current de-facto format for sharing
-stochastic character maps is a (non-interoperable) modification of the
-standard Newick tree format created by a software tool widely used for
-creating such maps, `simmap` [@Bollback_2006].
+Here we illustrate how `RNeXML`'s interface to NeXML's metadata system
+can be used to record and share a type of phylogenetic data not taken
+into account when NeXML was designed, in this case stochastic
+character maps [@Huelsenbeck_2003]. Such data assign certain parts
+(corresponding to time) of each branch in a time-calibrated phylogeny
+to a particular "state" (typically of a morphological
+characteristic). The current de-facto format for sharing stochastic
+character maps, created by `simmap` [@Bollback_2006], a widely used tool
+for creating such maps, is a non-interoperable modification of the standard
+Newick tree format.
 
-To express stochastic character map data as a custom metadata
+To express these data as a custom metadata
 extension, we create metadata annotations for the `edge` elements in
-the phylogeny topology that relate parts of the annotated edge (branch)
-to different states. These annotations need to describe the character state
+the phylogeny topology. The metadata need to describe the character state
 being assigned, and the duration (in terms of branch length) that the edge 
@@
 elements for discrete character traits, which, as most elements in
-NeXML, also have unique identifies (attribute `id`) by which they can
-be referenced in other parts of the document. In the following
-proposal, we call each series of character state assignments for an edge a
-_reconstruction_, and for each state assignment comprising the
+NeXML, also have unique identifiers by which they can
+be referenced. In the following
+proposal, we term each series of character state assignments for an edge a
+_reconstruction_, and for each state assignment comprising a
 reconstruction, we provide the length, the identifier of the character
 state, and an ordinal property to record the chronological order of
-the state assignment (the identifier values in the proposal, "cr1",
-"s1", and "s2", are -- fictitious -- identifiers that, in contrast to
-global identifiers, are unique only within a given NeXML document):
+the assignment (the identifier values in the proposal, "cr1",
+"s1", and "s2", are identifiers local to a  -- fictitious -- NeXML document):
 
@@
 
-As we illustrate here, even if the URI does not resolve to a fully
-machine-readable definition of the terms and their semantics, it can
-still be used to provide at least a human-readable informal definition
-of the terms, and where to find such documentation _is_
-machine-readable in the resulting NeXML document. Aside from the fact
-that the above representation does not modify the NeXML format, and
-should thus be readable -- even if not necessarily interpretable as a
-stochastic character mapping -- by any compliant NeXML parser, the
-direct and machine-readable link to metadata documentation contrasts
-with metadata or format extensions that are undocumented, or which do
-not directly link to documentation.
+Here the URI does not resolve to a fully
+machine-readable definition of the terms and their semantics, but it can
+nonetheless be used to provide at least a human-readable informal definition
+of the terms.
 
-Although writing out metadata annotations as in the above proposal may
-seem tedious, by virtue of `RNeXML`'s API they can be expressed as a
-series of R commands, and it is therefore straightforward to define
-functions that encapsulate the details behind a parameter
-signature. To demonstrate this, we implemented two functions in
-`RNeXML`, `simmap_to_nexml()` and `nexml_to_simmap()`, that map
-bidirectionally between NeXML files with `simmap` metadata according
-to the above proposal, and the `simmap` data representation in R
-devised by @Revell_2012 in the R package `phytools`. The latter uses
-an extension of the `ape::phylo` class. As shown in more detail in the
-`RNeXML` package vignette, this allows generating NeXML files
-containing `simmap` data, reading those files into R, converting them
-into the in-memory representation understood by `phytools`, and using
-the various functions from `phytools` designed for `simmap` objects
-[@Revell_2012], such as `plotSimmap()` for plotting.
+Because the above representation does not modify the NeXML format, it
+should be readable -- even if not necessarily interpretable as a
+stochastic character mapping -- by any compliant NeXML parser. In
+addition, it includes a direct and machine-readable link to metadata
+documentation.
+
+Writing out metadata annotations as in the above proposal may seem
+tedious. However, by virtue of `RNeXML`'s API it is straightforward to
+define functions that encapsulate these details. As a demonstration,
+`RNeXML` comes with functions `simmap_to_nexml()` and
+`nexml_to_simmap()` that map bidirectionally between NeXML files with
+`simmap` metadata according to the above proposal, and the extension
+of the `ape::phylo` class devised by @Revell_2012 in the R package
+`phytools`. The `RNeXML` package vignette illustrates how this allows
+seamlessly generating and reading NeXML files containing `simmap`
+data, and using the various functions from `phytools` designed for
+`simmap` objects [@Revell_2012].
 
@@
 directly from within R that allows data archival to become a step programmed
-into data management scripts, rather than a separate manually executed
-act. Initially, the method supports the data repository Figshare (http://figshare.com):
+into data management scripts. Initially, the method supports the data repository Figshare (http://figshare.com):
 

### 2015-01-05 manuscripts/manuscript.Rmd
Moves last paragraph of intro to conclusions.
Also tightens up future directions.

@@
 
-`RNeXML` not only allows R's ecosystem to read and write data in
-the NeXML format through an interface that is no more involved than
-reading or writing data from other phylogenetic data formats, but also
-carries immediate benefits for its users. For example, comparative
-analysis R packages and users frequently add their own metadata
-annotations to the phylogenies they work with, such as annotations
-of species, stochastic character maps, trait values, model estimates
-and parameter values. `RNeXML` affords R the capability to harness
-machine-readable semantics and an extensible metadata schema to
-capture, preserve, and share these and other kinds of information, all
-through an API instead of having to understand in detail the schema
-underlying the NeXML standard. To assist users in meeting the rising
-bar for best practices in data sharing in phylogenetic research
-[@Cranston_2014], `RNeXML` captures metadata information from the R
-environment to the extent possible, and applies reasonable
-defaults.
+To make the capabilities of NeXML available to R users in
+an easy-to-use form, and to lower the hurdles to adoption of the
+standard, we present RNeXML, an R package that aims to provide easy
+programmatic access to reading and writing NeXML documents, tailored
+for the kinds of use-cases that will be common for users and
+developers of the wealth of evolutionary analysis methods within the R
+ecosystem.
 
@@
 
-# Future Directions
+# Conclusions and future directions
+
+`RNeXML` allows R's ecosystem to read and write data in the NeXML
+format through an interface that is no more involved than reading or
+writing data from other phylogenetic data formats. It also carries
+immediate benefits for its users compared to other formats. For
+example, comparative analysis R packages and users frequently add
+their own metadata annotations to the phylogenies they work with, such
+as annotations of species, stochastic character maps, trait values,
+model estimates and parameter values. `RNeXML` affords R the
+capability to harness machine-readable semantics and an extensible
+metadata schema to capture, preserve, and share these and other kinds
+of information, all through an API instead of having to understand in
+detail the schema underlying the NeXML standard. To assist users in
+meeting the rising bar for best practices in data sharing in
+phylogenetic research [@Cranston_2014], `RNeXML` captures metadata
+information from the R environment to the extent possible, and applies
+reasonable defaults.
 
 The goals for continued development of `RNeXML` revolve primarily
-around three directions. One is better interoperability with other
-existing phylogenetic data representations in R, such those found in
-the `phylobase` package [@phylobase]. Another is better integration of
-the rich metadata semantics found in ontologies defined in the Web
-Ontology Language (OWL), including programmatic access to machine
-reasoning with such metadata. Finally, and similar to many other R
-packages, including those for comparative phylogenetics, in light of
-the increasing size of phylogenomic and other datasets mechanisms may
-eventually be needed that avoid the requirement to load an entire
-file's content into memory.
+around better interoperability with other existing phylogenetic data
+representations in R, such those found in the `phylobase` package
+[@phylobase]; and better integration of the rich metadata semantics
+found in ontologies defined in the Web Ontology Language (OWL),
+including programmatic access to machine reasoning with such metadata.
 

### 2015-01-07 manuscripts/manuscript.Rmd
Rewrite of the abstract.
This is a little longer than before, but the MEE author guidelines state
350 words as limit, and we remain under that.

@@
 abstract: | 
-      1. NeXML is a new exchange format for phylogenetic data and
-      metadata that offers substantial improvements in reliability,
-      extensibility, and richness over earlier formats such as
-      NEXUS. In this paper we present the RNeXML package which
-      provides a bridge between this data format and the extensive
-      library of phylogenetic tools already available in the popular
-      data analysis environment and programming language R, with a
-      particular emphasis on comparative methods.
+      1. NeXML is a powerful and forward-compatible exchange standard
+      recently proposed to better meet the expanding needs for
+      machine-proof phylogenetic data and metadata sharing. Here we
+      present the RNeXML package, which aims to provide to users of
+      the popular R programming language easy programmatic access to
+      reading and writing NeXML documents, including rich metadata, in
+      a way that interfaces seamlessly with the extensive library of
+      phylogenetic tools already available in the R ecosystem.
 
-      2. The ability to parse and write NeXML from within R is only a
-      small part of the functionality offered by the package.  We
-      illustrate how RNeXML facilitates adding and manipulating data
-      annotations whose meaning can be defined in a precise and
-      machine-interpretable way by means of formal ontologies and other
-      semantic web technologies.
+      2. Wherever possible, we designed RNeXML to map NeXML document
+      contents, whose arrangement is influenced by the format's
+      normative XML Schema definition, to their most intuitive or
+      useful representation in R. To make NeXML's powerful facility
+      for recording semantically rich machine-readable metadata
+      accessible to R users, we designed a functional programming
+      interface to it that hides the semantic web standards leveraged
+      by NeXML from R users who are unfamiliar with them.
 
-      3. An essential feature of NeXML is to provide both the
-      flexibility to be continually extended to meet the emerging
-      needs of the phylogenetics community without breaking software
-      unaware of such extensions. We illustrate how RNeXML makes
-      it easy to extend NeXML in this manner.
+      3. RNeXML can read any NeXML document that validates, and it
+      generates valid NeXML documentes from phylogeny and character
+      data from various R representations in use. The metadata
+      programming interface at a basic level aids fulfilling data
+      documentation best practices, and at an advanced level preserves
+      NeXML's nearly limitless extensibility, for which we provide a
+      fully working demonstration. Furthermore, to lower the barriers
+      to sharing well-documented phylogenetic data, RNeXML has started
+      to integrate with taxonomic metadata augmentation services on
+      the web, and with online repositories for data archiving.
 
-      4. RNeXML is well suited for users to preserve and maintain
-      their personal archives of phylogenetic and comparative trait
-      data in consistent, query-able, and cross-platform format, to
-      leverage the rapidly growing public community archives of rich
-      NeXML files, and to interact with web services that return data
-      and their metadata packaged together as NeXML.
+      4. RNeXML allows R's rich ecosystem to read and write data in
+      the NeXML format through an interface that is no more involved
+      than reading or writing data from other, less powerful data
+      formats. It also provides an interface designed to feel familiar
+      to R programmers and to be consistent with recommended practices
+      for R package development, yet that retains the full power for
+      users to add their own custom data and metadata to the
+      phylogenies they work with, without introducing potentially
+      incompatible changes to the exchange standard.
 
@@
 Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
-the Google Summer of Code, and we thank Kseniia Shumelchyk for taking
+the Google Summer of Code(TM), and we thank Kseniia Shumelchyk for taking
 the first steps to implement it. We are grateful to F. Michonneau for
@@
 ```
-
-<!-- Footnotes -->
-
-[^1]: By providing the optional `dependencies` argument, R will install
-additional packages used in some of the advanced examples shown later
-but which are not required for basic functionality.
-
-[^2]: For convenience `nexml_read` is also aliased as `read.nexml`,
-for users familiar with the somewhat analogous `read.nexus` function.

### 2015-01-07 manuscripts/manuscript.Rmd
Fixes affiliation for HL.

@@
     address: | 
-      National Evolutionary Synthesis Center, Durham, NC, USA
+      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
   - code: ropensci

### 2015-01-08 manuscripts/manuscript.Rmd

--- 2015-01-07 manuscripts/manuscript.Rmd
@@
       3. RNeXML can read any NeXML document that validates, and it
-      generates valid NeXML documentes from phylogeny and character
-      data from various R representations in use. The metadata
+      generates valid NeXML documents from phylogeny and character
+      data in various R representations in use. The metadata
       programming interface at a basic level aids fulfilling data

### 2015-01-08 manuscripts/manuscript.Rmd
load birds nexml first

@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, eval=FALSE}
 install.packages("RNeXML", dependencies=TRUE)
@@
 
-```{r eval=FALSE fig.show="hide"}
+```{r eval=FALSE, fig.show="hide"}
 tb_nex <- nexml_read(
@@
 ```{r}
+birds <- nexml_read("birds.xml")
 prefixes <- get_namespaces(birds)

### 2015-01-08 manuscripts/manuscript.Rmd
need to load the library still, silly!
minor formatting

@@
 
+```{r echo=FALSE}
+install.packages("RNeXML", dependencies=TRUE)
+library(RNeXML)
+```
+
 # The RNeXML package
@@
 version can be installed from the CRAN repository.
-
-```{r echo=FALSE, eval=FALSE}
-install.packages("RNeXML", dependencies=TRUE)
-library(RNeXML)
-```
-
 `RNeXML` is part of the rOpenSci project. Users of `RNeXML` are
 encouraged to submit bug reports or feature requests in the issues log
-on Github, or to contact `ropensci-discuss@googlegroups.com` or the
+on Github, or the
 phylogenetics R users group list at `r-sig-phylo@r-project.org` for
@@
 
-```{r}
+```{r, results='hide'}
 data(bird.orders)
@@
 
-```{r}
+```{r, results='hide'}
 library("geiger")

### 2015-01-08 manuscripts/manuscript.Rmd
break supplement into vignettes

@@
     footnote: Corresponding author
-  - name: Hilmar Lapp
-    affiliation: NESCent
   - name: Scott Chamberlain
@@
     affiliation: NBC
+  - name: Hilmar Lapp
+    affiliation: NESCent
 address: 
@@
 
-Users of the popular statistical and mathematical computing platform R
-[@R] enjoy a wealth of readily installable comparative phylogenetic
-methods and tools <!-- needs a citation to support it -->. Exploiting
-the opportunities arising from this wealth for complex and integrative
-comparative research questions relies on the ability to reuse and
-integrate previously generated or published data and metadata. The
-expanding data exchange needs of the evolutionary research
-community are rapidly outpacing the capabilities of most current
-and widely used data exchange standards [@Vos_2012], which were all
-developed a decade or more ago. This has resulted in a radiation of
-different data representations and exchange standard "flavors" that
-are no longer interoperable at the very time when the growth of
-available data and methods has made that interoperability most
-valuable. In response to the unmet needs for standardized data
-exchange in phylogenetics, a modern XML-based exchange standard,
-called NeXML, has recently been developed [@Vos_2012]. NeXML
-comprehensively supports current data exchange needs, is predictably
-machine-readable, and is forward compatible.
+Users of the popular statistical and mathematical computing platform
+R [@R] enjoy a wealth of readily installable comparative phylogenetic
+methods and tools [@taskview]. Exploiting the opportunities arising from
+this wealth for complex and integrative comparative research questions
+relies on the ability to reuse and integrate previously generated or
+published data and metadata. The expanding data exchange needs of the
+evolutionary research community are rapidly outpacing the capabilities of
+most current and widely used data exchange standards [@Vos_2012], which
+were all developed a decade or more ago. This has resulted in a radiation
+of different data representations and exchange standard "flavors" that are
+no longer interoperable at the very time when the growth of available data
+and methods has made that interoperability most valuable. In response to
+the unmet needs for standardized data exchange in phylogenetics, a modern
+XML-based exchange standard, called NeXML, has recently been developed
+[@Vos_2012]. NeXML comprehensively supports current data exchange needs,
+is predictably machine-readable, and is forward compatible.
 
-The exchange problem for phylogenetic data is particularly acute in
-light of the challenges in finding and sharing phylogenetic data
-without the otherwise common loss of most data and metadata semantics
-[@Drew_2013; @Stoltzfus_2012; @Cranston_2014]. For example, the still popular
-NEXUS file format [@Maddison_1997] cannot consistently represent
-horizontal gene transfer or ambiguity in reading a character (such as
-a DNA sequence base pair).  This and other limitations have led to modifications of NEXUS in
-different ways for different needs, with the unfortunate result that
+The exchange problem for phylogenetic data is particularly acute in light
+of the challenges in finding and sharing phylogenetic data without the
+otherwise common loss of most data and metadata semantics [@Drew_2013;
+@Stoltzfus_2012; @Cranston_2014]. For example, the still popular NEXUS
+file format [@Maddison_1997] cannot consistently represent horizontal
+gene transfer or ambiguity in reading a character (such as a DNA sequence
+base pair).  This and other limitations have led to modifications of NEXUS
+in different ways for different needs, with the unfortunate result that
 NEXUS files generated by one program can be incompatible with another
 [@Vos_2012]. Without a formal grammar, software based on NEXUS files
-may also make inconsistent assumptions about tokens, quoting, or
-element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
-files in the CIPRES portal contain unrecoverable but hard to diagnose
-errors.
+may also make inconsistent assumptions about tokens, quoting, or element
+lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS files in
+the CIPRES portal contain unrecoverable but hard to diagnose errors.
 
-A detailed account of how the NeXML standard addresses these and other
-relevant challenges can be found in @Vos_2012. In brief, NeXML was
-designed with the following important properties. First, NeXML is
-defined by a precise grammar that can be programmatically
+A detailed account of how the NeXML standard addresses these and
+other relevant challenges can be found in @Vos_2012. In brief,
+NeXML was designed with the following important properties. First,
+NeXML is defined by a precise grammar that can be programmatically
 **validated**. This ensures that any NeXML file can be expected to be
-parsed and read without errors by any NeXML-compliant software,
-provided the file passes validation. Second, NeXML is **extensible**:
-a user can define representations of new, previously unanticipated
-information (as we will illustrate) without violating its defining
-grammar. Third and most importantly, NeXML is **rich in computable
-semantics**. The formal semantics used in NeXML is machine-readable in
-a predictable manner, and also enables machine reasoning about the
-data and metadata (see @Parr2011), especially when metadata attributes
-and values are anchored in appropriately axiomatized ontologies.
+parsed and read without errors by any NeXML-compliant software, provided
+the file passes validation. Second, NeXML is **extensible**: a user
+can define representations of new, previously unanticipated information
+(as we will illustrate) without violating its defining grammar. Third
+and most importantly, NeXML is **rich in computable semantics**. The
+formal semantics used in NeXML is machine-readable in a predictable
+manner, and also enables machine reasoning about the data and metadata
+(see @Parr2011), especially when metadata attributes and values are
+anchored in appropriately axiomatized ontologies.
 
-To make the capabilities of NeXML available to R users in
-an easy-to-use form, and to lower the hurdles to adoption of the
-standard, we present RNeXML, an R package that aims to provide easy
-programmatic access to reading and writing NeXML documents, tailored
-for the kinds of use-cases that will be common for users and
-developers of the wealth of evolutionary analysis methods within the R
-ecosystem.
+To make the capabilities of NeXML available to R users in an easy-to-use
+form, and to lower the hurdles to adoption of the standard, we present
+RNeXML, an R package that aims to provide easy programmatic access to
+reading and writing NeXML documents, tailored for the kinds of use-cases
+that will be common for users and developers of the wealth of evolutionary
+analysis methods within the R ecosystem.
 
@@
 The `RNeXML` package is written entirely in R and available under a
-Creative Commons Zero public domain waiver. The current development version can be found
-on Github at [https://github.com/ropensci/RNeXML](), and the stable
-version can be installed from the CRAN repository.
-`RNeXML` is part of the rOpenSci project. Users of `RNeXML` are
-encouraged to submit bug reports or feature requests in the issues log
-on Github, or the
-phylogenetics R users group list at `r-sig-phylo@r-project.org` for
-help.
+Creative Commons Zero public domain waiver. The current development
+version can be found on Github at [https://github.com/ropensci/RNeXML](),
+and the stable version can be installed from the CRAN repository.
+`RNeXML` is part of the rOpenSci project. Users of `RNeXML` are encouraged
+to submit bug reports or feature requests in the issues log on Github,
+or the phylogenetics R users group list at `r-sig-phylo@r-project.org`
+for help.
 
@@
 Conceptually, a NeXML document has the following components: (1)
-phylogeny topology data, (2) character or trait data in matrix form, (3)
-operational taxonomic units (OTUs), and (4) metadata. To represent the
-contents of a NeXML document (currently in memory), `RNeXML` defines
-the `nexml` object type. This type therefore holds phylogenetic trees
-as well as character or trait matrices, and all metadata, which is
-similar to the phylogenetic data object types defined in the
-`phylobase` package [@phylobase], but contrasts with the more widely
-used ones defined in the
+phylogeny topology data, (2) character or trait data in matrix form,
+(3) operational taxonomic units (OTUs), and (4) metadata. To represent
+the contents of a NeXML document (currently in memory), `RNeXML` defines
+the `nexml` object type. This type therefore holds phylogenetic trees as
+well as character or trait matrices, and all metadata, which is similar
+to the phylogenetic data object types defined in the `phylobase` package
+[@phylobase], but contrasts with the more widely used ones defined in the
 `ape` package [@Paradis_2004], which represents trees alone.
 
-When reading and writing NeXML documents, `RNeXML` aims to map their
-components to and from, respectively, their most widely used
-representations in R. As a result, the types of objects accepted or returned
-by the package's methods are the `phylo` and `multiPhylo` objects from
-the `ape` package [@Paradis_2004] for phylogenies, and R's native
-`data.frame` list structure for data matrices.
+When reading and writing NeXML documents, `RNeXML` aims to map
+their components to and from, respectively, their most widely used
+representations in R. As a result, the types of objects accepted
+or returned by the package's methods are the `phylo` and `multiPhylo`
+objects from the `ape` package [@Paradis_2004] for phylogenies, and R's
+native `data.frame` list structure for data matrices.
 
@@
 
-The method `nexml_read()` reads NeXML files, either from a local
-file, or from a remote location via its URL, and returns an object of
-type `nexml`:
+The method `nexml_read()` reads NeXML files, either from a local file, or
+from a remote location via its URL, and returns an object of type `nexml`:
 
@@
 The `get_trees_list()` method is designed for use in scripts, providing
-a consistent and predictable return type regardless of the number of
-phylogenies a NeXML document contains. For greater convenience in
+a consistent and predictable return type regardless of the number
+of phylogenies a NeXML document contains. For greater convenience in
 interactive use, the method `get_trees()` returns the R object most
 intuitive given the arrangement of phylogeny data in the source NeXML
-document. For example, the method returns an `ape::phylo` object if the NeXML
-document contains a single phylogeny, an `ape::multiPhylo` object if
-it contains multiple phylogenies arranged in a single `trees` block,
-and a list of `ape::multiPhylo` objects if it contains multiple
+document. For example, the method returns an `ape::phylo` object if
+the NeXML document contains a single phylogeny, an `ape::multiPhylo`
+object if it contains multiple phylogenies arranged in a single `trees`
+block, and a list of `ape::multiPhylo` objects if it contains multiple
 `trees` blocks (the capability for which NeXML inherits from NEXUS).
@@
 If the location parameter with which the `nexml_read()` method is
-invoked is recognized as a URL, the method will automatically download the
-document to the local working directory and read it from there. This
+invoked is recognized as a URL, the method will automatically download
+the document to the local working directory and read it from there. This
 gives convenient and rapid access to phylogenetic data published in
-NeXML format on the web, such as the content of the
-phylogenetic data repository TreeBASE [@Piel_2002]. For
-example, the following plots a tree in TreeBASE (using ape's plot function):
+NeXML format on the web, such as the content of the phylogenetic data
+repository TreeBASE [@Piel_2002]. For example, the following plots a
+tree in TreeBASE (using ape's plot function):
 
@@
 alignments), discrete (for most morphological character data), or
-continuous type (for many trait data). To enable strict validation of data types NeXML allows, and if their
-data types differ requires multiple data matrices to be separated into
-different "blocks". Since
-the `data.frame` data structure in R has no such constraints, the
-`get_characters()` method combines such blocks as separate columns
-into a single `data.frame` object, provided they correspond to the
-same taxa. Otherwise, a list of `data.frame`s is returned, with list
-elements corresponding to characters blocks. Similar to the methods
-for obtaining trees, there is also a method `get_characters_list()`,
-which always returns a list of `data.frame`s, one for each character
-block.
+continuous type (for many trait data). To enable strict validation of data
+types NeXML allows, and if their data types differ requires multiple data
+matrices to be separated into different "blocks". Since the `data.frame`
+data structure in R has no such constraints, the `get_characters()`
+method combines such blocks as separate columns into a single `data.frame`
+object, provided they correspond to the same taxa. Otherwise, a list of
+`data.frame`s is returned, with list elements corresponding to characters
+blocks. Similar to the methods for obtaining trees, there is also a method
+`get_characters_list()`, which always returns a list of `data.frame`s,
+one for each character block.
 
@@
 
-The method `nexml_write()` generates a NeXML file from its input parameters. In
-its simplest invocation, the method writes a tree to a file:
+The method `nexml_write()` generates a NeXML file from its input
+parameters. In its simplest invocation, the method writes a tree to
+a file:
 
@@
 
-The first argument to `nexml_write()` is either an object of type `nexml`, or any
-object that can be coerced to it, such as in the above example an
+The first argument to `nexml_write()` is either an object of type `nexml`,
+or any object that can be coerced to it, such as in the above example an
 `ape::phylo` phylogeny. Alternatively, passing a `multiPhylo` object
@@
 NeXML allows attaching ("_annotating_") metadata to any data element,
-and even to metadata themselves. Whether at the level of the document
-as a whole or an individual data matrix or phylogeny, metadata can
-provide bibliographic and provenance information, for example about
-the study as part of which the phylogeny was generated or applied,
-which data matrix and which methods were used to generate it. Metadata
-can also be attached to very specific elements of the data, such as
-specific traits, individual OTUs, nodes, or even edges of the
-phylogeny.
+and even to metadata themselves. Whether at the level of the document as
+a whole or an individual data matrix or phylogeny, metadata can provide
+bibliographic and provenance information, for example about the study
+as part of which the phylogeny was generated or applied, which data
+matrix and which methods were used to generate it. Metadata can also be
+attached to very specific elements of the data, such as specific traits,
+individual OTUs, nodes, or even edges of the phylogeny.
 
-As described in @Vos_2012,  to encode metadata annotations NeXML uses
-the "Resource Description Framework in Annotations" (RDFa)
-[@W3C_2014]. This standard provides for a strict machine-readable
-format yet enables future backwards compatibility with
-compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
-tool to _parse_ annotations is not predicated on _understanding_
-the meaning of annotations it has not seen before.
+As described in @Vos_2012,  to encode metadata annotations NeXML uses the
+"Resource Description Framework in Annotations" (RDFa) [@W3C_2014]. This
+standard provides for a strict machine-readable format yet enables future
+backwards compatibility with compliant NeXML parsers (and thus `RNeXML`),
+because the capacity of a tool to _parse_ annotations is not predicated
+on _understanding_ the meaning of annotations it has not seen before.
 
@@
 
-First, when writing a NeXML file the package adds certain basic
-metadata automatically if they are absent, using default
-values consistent with recommended best practices
-[@Cranston_2014]. Currently, this includes naming the software
-generating the NeXML, a time-stamp of when a tree was produced, and an
-open data license.
+First, when writing a NeXML file the package adds certain basic metadata
+automatically if they are absent, using default values consistent with
+recommended best practices [@Cranston_2014]. Currently, this includes
+naming the software generating the NeXML, a time-stamp of when a tree
+was produced, and an open data license.
 
-Second, `RNeXML` provides a simple method, called
-`add_basic_metadata()`, to set metadata attributes commonly
-recommended for inclusion with data to be publicly archived or shared
-[@Cranston_2014]. The currently accepted parameters include `title`,
-`description`, `creator`, `pubdate`, `rights`, `publisher`, and
-`citation`. Behind the scenes the method automatically anchors these
-attributes in common vocabularies (such as Dublin Core).
+Second, `RNeXML` provides a simple method, called `add_basic_metadata()`,
+to set metadata attributes commonly recommended for inclusion with data
+to be publicly archived or shared [@Cranston_2014]. The currently accepted
+parameters include `title`, `description`, `creator`, `pubdate`, `rights`,
+`publisher`, and `citation`. Behind the scenes the method automatically
+anchors these attributes in common vocabularies (such as Dublin Core).
 
-Third, `RNeXML` integrates with the R package `taxize`
-[@Chamberlain_2013] to mitigate one of the most common obstacles to
-reuse of phylogenetic data, namely the misspellings and inconsistent
-taxonomic naming with which OTU labels are often fraught. The
-`taxize_nexml()` method in `RNeXML` uses `taxize` to match OTU labels
-against the NCBI database, and, where a unique match is found, it
-annotates the respective OTU with the matching NCBI identifier.
+Third, `RNeXML` integrates with the R package `taxize` [@Chamberlain_2013]
+to mitigate one of the most common obstacles to reuse of phylogenetic
+data, namely the misspellings and inconsistent taxonomic naming with
+which OTU labels are often fraught. The `taxize_nexml()` method in
+`RNeXML` uses `taxize` to match OTU labels against the NCBI database,
+and, where a unique match is found, it annotates the respective OTU with
+the matching NCBI identifier.
 
@@
 annotation appears in the NeXML document (technically, the containing
-element in the XML document model). This structure corresponds
-directly to Resource Description Framework (RDF)
-subject-predicate-object triples, and therefore the only requirement
-for custom metadata is that they be cast as RDF triples.
+element in the XML document model). This structure corresponds directly
+to Resource Description Framework (RDF) subject-predicate-object triples,
+and therefore the only requirement for custom metadata is that they be
+cast as RDF triples.
 
@@
 perform this cast automatically behind the scenes, including mapping
-metadata attributes to terms in requisite common vocabularies (such as
-Dublin Core for "title", "creator", etc.) or ontologies. Giving
-metadata properties (and where applicable, values) as vocabulary or ontology
-terms rather than simple text strings is crucial for allowing machines
-to not only parse but also interpret and potentially reason over their
-semantics. To achieve this benefit for custom metadata extensions, the
-user necessarily needs to handle certain technical details from which
-the `RNeXML` API shields her otherwise, in particular the globally unique
-identifiers (normally HTTP URIs) of metadata terms and vocabularies. To be consistent with XML
-terminology, `RNeXML` calls vocabulary URIs
-_namespaces_, and their abbreviations _prefixes_. For example, the
-namespace for the Dublin Core Metadata Terms vocabulary is "http://purl.org/dc/elements/1.1/". Using
-its common abbreviation "dc", a metadata property "dc:title" expands to the
-identifier "http://purl.org/dc/elements/1.1/title". This URI resolves
-to a human and machine-readable (depending on access) definition of precisely
-what the term `title` in Dublin Core means. In contrast, just using
-the text string "title" could also mean the title of a person, a legal
-title, the verb title, etc.  URI identifiers of metadata vocabularies
-and terms are not mandated to resolve, but if machines are to derive
-the maximum benefit from them, they should resolve to a definition of
-their semantics in RDF.
+metadata attributes to terms in requisite common vocabularies (such
+as Dublin Core for "title", "creator", etc.) or ontologies. Giving
+metadata properties (and where applicable, values) as vocabulary or
+ontology terms rather than simple text strings is crucial for allowing
+machines to not only parse but also interpret and potentially reason
+over their semantics. To achieve this benefit for custom metadata
+extensions, the user necessarily needs to handle certain technical
+details from which the `RNeXML` API shields her otherwise, in particular
+the globally unique identifiers (normally HTTP URIs) of metadata terms
+and vocabularies. To be consistent with XML terminology, `RNeXML` calls
+vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
+example, the namespace for the Dublin Core Metadata Terms vocabulary
+is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
+"dc", a metadata property "dc:title" expands to the identifier
+"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
+machine-readable (depending on access) definition of precisely what the
+term `title` in Dublin Core means. In contrast, just using the text string
+"title" could also mean the title of a person, a legal title, the verb
+title, etc.  URI identifiers of metadata vocabularies and terms are not
+mandated to resolve, but if machines are to derive the maximum benefit
+from them, they should resolve to a definition of their semantics in RDF.
 
-`RNeXML` includes methods to obtain and manipulate metadata
-properties, values, identifiers, and namespaces. The
-`get_namespaces()` method accepts a `nexml` object and returns a named list
-of namespace prefixes and their corresponding identifiers known to the object:
+`RNeXML` includes methods to obtain and manipulate metadata properties,
+values, identifiers, and namespaces. The `get_namespaces()` method
+accepts a `nexml` object and returns a named list of namespace prefixes
+and their corresponding identifiers known to the object:
 
@@
 The `get_metadata()` method returns, as a named list, the metadata
-annotations for a given `nexml` object at a given level, with the
-whole NeXML document being the default level (`"all"` extracts all metadata objects):
+annotations for a given `nexml` object at a given level, with the whole
+NeXML document being the default level (`"all"` extracts all metadata
+objects):
 
@@
 metadata are attached. Therefore, a different approach, documented in
-the package vignette, is recommended for accessing the metadata
-attached to data elements.
+the package vignette, is recommended for accessing the metadata attached
+to data elements.
 
@@
 
-Metadata annotatuons in `NeXML` can be nested within another
+Metadata annotations in `NeXML` can be nested within another
 annotation, which the `meta()` method accommodates by accepting a
@@
 If the prefix used by the metadata property is not among the built-in
-ones (which can be obtained using `get_namespaces()`), it has to be provided along with its URI as the `namespaces`
-parameter. For example, the following uses the SKOS vocabulary (which
-currently is not built-in) to add a note to the trees in the `nexml`
-object:
+ones (which can be obtained using `get_namespaces()`), it has to be
+provided along with its URI as the `namespaces` parameter. For example,
+the following uses the SKOS vocabulary (which currently is not built-in)
+to add a note to the trees in the `nexml` object:
 
@@
 can be used to record and share a type of phylogenetic data not taken
-into account when NeXML was designed, in this case stochastic
-character maps [@Huelsenbeck_2003]. Such data assign certain parts
-(corresponding to time) of each branch in a time-calibrated phylogeny
-to a particular "state" (typically of a morphological
-characteristic). The current de-facto format for sharing stochastic
-character maps, created by `simmap` [@Bollback_2006], a widely used tool
-for creating such maps, is a non-interoperable modification of the standard
-Newick tree format.
+into account when NeXML was designed, in this case stochastic character
+maps [@Huelsenbeck_2003]. Such data assign certain parts (corresponding
+to time) of each branch in a time-calibrated phylogeny to a particular
+"state" (typically of a morphological characteristic). The current
+de-facto format for sharing stochastic character maps, created by
+`simmap` [@Bollback_2006], a widely used tool for creating such maps,
+is a non-interoperable modification of the standard Newick tree format.
 
-To express these data as a custom metadata
-extension, we create metadata annotations for the `edge` elements in
-the phylogeny topology. The metadata need to describe the character state
-being assigned, and the duration (in terms of branch length) that the edge 
-spends in that state. The `NeXML` standard already defines `state`
-elements for discrete character traits, which, as most elements in
-NeXML, also have unique identifiers by which they can
-be referenced. In the following
-proposal, we term each series of character state assignments for an edge a
-_reconstruction_, and for each state assignment comprising a
+To express these data as a custom metadata extension, we create metadata
+annotations for the `edge` elements in the phylogeny topology. The
+metadata need to describe the character state being assigned, and
+the duration (in terms of branch length) that the edge spends in
+that state. The `NeXML` standard already defines `state` elements for
+discrete character traits, which, as most elements in NeXML, also have
+unique identifiers by which they can be referenced. In the following
+proposal, we term each series of character state assignments for an
+edge a _reconstruction_, and for each state assignment comprising a
 reconstruction, we provide the length, the identifier of the character
 state, and an ordinal property to record the chronological order of
-the assignment (the identifier values in the proposal, "cr1",
-"s1", and "s2", are identifiers local to a  -- fictitious -- NeXML document):
+the assignment (the identifier values in the proposal, "cr1", "s1", and
+"s2", are identifiers local to a  -- fictitious -- NeXML document):
 
@@
 
-Here the URI does not resolve to a fully
-machine-readable definition of the terms and their semantics, but it can
-nonetheless be used to provide at least a human-readable informal definition
-of the terms.
+Here the URI does not resolve to a fully machine-readable definition of
+the terms and their semantics, but it can nonetheless be used to provide
+at least a human-readable informal definition of the terms.
 
-Because the above representation does not modify the NeXML format, it
-should be readable -- even if not necessarily interpretable as a
-stochastic character mapping -- by any compliant NeXML parser. In
+Because the above representation does not modify the NeXML format,
+it should be readable -- even if not necessarily interpretable as
+a stochastic character mapping -- by any compliant NeXML parser. In
 addition, it includes a direct and machine-readable link to metadata
@@
 define functions that encapsulate these details. As a demonstration,
-`RNeXML` comes with functions `simmap_to_nexml()` and
-`nexml_to_simmap()` that map bidirectionally between NeXML files with
-`simmap` metadata according to the above proposal, and the extension
-of the `ape::phylo` class devised by @Revell_2012 in the R package
-`phytools`. The `RNeXML` package vignette illustrates how this allows
-seamlessly generating and reading NeXML files containing `simmap`
-data, and using the various functions from `phytools` designed for
-`simmap` objects [@Revell_2012].
+`RNeXML` comes with functions `simmap_to_nexml()` and `nexml_to_simmap()`
+that map bidirectionally between NeXML files with `simmap` metadata
+according to the above proposal, and the extension of the `ape::phylo`
+class devised by @Revell_2012 in the R package `phytools`. The `RNeXML`
+package vignette illustrates how this allows seamlessly generating and
+reading NeXML files containing `simmap` data, and using the various
+functions from `phytools` designed for `simmap` objects [@Revell_2012].
 
@@
 Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
-Natioal Science Foundation (DBI-1306697) and the Alfred P Sloan
+National Science Foundation (DBI-1306697) and the Alfred P Sloan
 Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for

### 2015-01-08 manuscripts/manuscript.Rmd
add section explicitly showing validation

@@
 
+## Validating NeXML
+
+File validation is a central feature of the NeXML format which ensures
+that any properly implemented NeXML parser will always be able to read
+the NeXML file. The function takes the path to any NeXML file and returns 
+`TRUE` to indicate a valid file, or `FALSE` otherwise, along with a
+display of any error messages generated by the validator.
+
+```{r }
+nexml_validate("geospiza.xml")
+```
+
+The `nexml_validate()` function performs this validation
+using the online NeXML validator (when a network connection is available),
+which performs additional checks not expressed in the NeXML schema itself [@Vos_2012].
+If a network connection is not available, the function falls back on the 
+schema validation method from the `XML` package [@XML].
+
 ## Creating and populating `nexml` objects
@@
 metadata are attached. Therefore, a different approach, documented in
-the package vignette, is recommended for accessing the metadata attached
+the metadata vignette, is recommended for accessing the metadata attached
 to data elements.
@@
 (see @W3C_2014 or @Hartig_2012). Examples for these possibilities are
-included in the `RNeXML` vignette, and the package also comes with a
+included in the `RNeXML` SPARQL vignette, and the package also comes with a
 demonstration that can be run from R using the following command:
@@
 class devised by @Revell_2012 in the R package `phytools`. The `RNeXML`
-package vignette illustrates how this allows seamlessly generating and
+simmap vignette illustrates how this allows seamlessly generating and
 reading NeXML files containing `simmap` data, and using the various

### 2015-01-13 manuscripts/manuscript.Rmd
update vignettes

@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, results='hide'}
 install.packages("RNeXML", dependencies=TRUE)
@@
 ```{r cleanup, include=FALSE, cache=FALSE}
-unlink("simmap.xml")
 unlink("birds.xml")
 unlink("geospiza.xml")
-unlink("rdf_meta.xml")
 ```

### 2015-01-14 manuscripts/manuscript.Rmd
tweak text on simmap

@@
 or the phylogenetics R users group list at `r-sig-phylo@r-project.org`
-for help.
+for help. Vignettes with more detailed examples of specific features
+of RNeXML are distributed with the R package and serve as a supplement
+to this manuscript. Each of the vignettes can be found at
+[http://ropensci.github.io/RNeXML/]().
 
@@
 
-The `meta()` method creates a new metadata object from a poperty name
+The `meta()` method creates a new metadata object from a property name
 and content (value). For example, the following creates a modification
@@
 fine control of where a `meta` element is added, for which the package
-vignette contains examples.
+vignette on S4 subsetting of `nexml` contains examples.
 
@@
 discrete character traits, which, as most elements in NeXML, also have
-unique identifiers by which they can be referenced. In the following
+unique identifiers by which they can be referenced.
+
+In the following
 proposal, we term each series of character state assignments for an
@@
 state, and an ordinal property to record the chronological order of
-the assignment (the identifier values in the proposal, "cr1", "s1", and
-"s2", are identifiers local to a  -- fictitious -- NeXML document):
+the assignment. In this manner, a single edge containing one
+state change could be annotated as follows.
+(the identifier values in the proposal, "cr1", "s1", and
+"s2", are identifiers local to a  -- fictitious -- NeXML document,
+based on the example simmap tree illustrated in @Revell_2012):
 

### 2015-01-15 manuscripts/manuscript.Rmd
add page numbers

@@
 layout: review, 11pt
+linenumbers: true
 title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"

### 2015-04-17 manuscripts/manuscript.Rmd
add References heading (for #112); update Dockerfile inline use notes
no need to install package, already done
edits to manuscript based on reviews

@@
 abstract: | 
-      1. NeXML is a powerful and forward-compatible exchange standard
+      1. NeXML is a powerful and extensible exchange standard
       recently proposed to better meet the expanding needs for
-      machine-proof phylogenetic data and metadata sharing. Here we
-      present the RNeXML package, which aims to provide to users of
-      the popular R programming language easy programmatic access to
+      phylogenetic data and metadata sharing. Here we
+      present the RNeXML package, which provides users of
+      the R programming language with easy-to-use tools for 
       reading and writing NeXML documents, including rich metadata, in
@@
       contents, whose arrangement is influenced by the format's
-      normative XML Schema definition, to their most intuitive or
+      XML Schema definition, to their most intuitive or
       useful representation in R. To make NeXML's powerful facility
@@
 NeXML is defined by a precise grammar that can be programmatically
-**validated**. This ensures that any NeXML file can be expected to be
-parsed and read without errors by any NeXML-compliant software, provided
-the file passes validation. Second, NeXML is **extensible**: a user
+**validated**: meaning that we can verify that a file has followed this
+precise grammar and thus be sure that it can be read (parsed) 
+without errors by any software that uses the NeXML grammar (e.g. RNeXML). 
+Second, NeXML is **extensible**: a user
 can define representations of new, previously unanticipated information
 (as we will illustrate) without violating its defining grammar. Third
-and most importantly, NeXML is **rich in computable semantics**. The
-formal semantics used in NeXML is machine-readable in a predictable
-manner, and also enables machine reasoning about the data and metadata
-(see @Parr2011), especially when metadata attributes and values are
-anchored in appropriately axiomatized ontologies.
+and most importantly, NeXML is rich in **computable semantics**: patterns
+that permit a machine to reason about concepts. Note that
+these semantics are based on logical deduction such as: "all frogs are vertebrates,
+so a query for vertebrates should return all species that are listed as frogs"), 
+rather than heuristic machine learning algorithms.  Such inference requires a
+computer to understand the 'semantic' meaning of a word like 'frog', which
+in must be specified in a kind of dictionary known in informatics research
+as an vocabulary or ontology (see @Parr2011 for a more detailed introduction
+in the context of evolution and biodiversity research).
 
@@
 
+```{r echo=FALSE, results='hide', eval=FALSE}
+install.packages("RNeXML", dependencies=TRUE)
+```
+
 ```{r echo=FALSE, results='hide'}
-install.packages("RNeXML", dependencies=TRUE)
 library(RNeXML)
@@
 Conceptually, a NeXML document has the following components: (1)
-phylogeny topology data, (2) character or trait data in matrix form,
+phylogeny topology and branch length data, (2) character or trait data in matrix form,
 (3) operational taxonomic units (OTUs), and (4) metadata. To represent
@@
 character data as another parameter. The following example uses data
-from the comparative phylogenetics R package `geiger` [@Harmon_2008].
+from the comparative phylogenetics R package `geiger` [@Pennell2014].
 
@@
 ```
+
+Note that the NeXML format is well-suited for incomplete data: for instance,
+here it does not assume the character matrix has data for every tip in the tree.
 
@@
 
-The RDFa standard adopted by NeXML (see @Vos_2012) allows for
-arbitrary custom extensions to meet ongoing scientific needs without
-requiring modifications (or "extensions") of the syntax. Specifically,
-every metadata annotation consists of a metadata property and a value
-of the property; the bearer of the property is determined by where the
-annotation appears in the NeXML document (technically, the containing
-element in the XML document model). This structure corresponds directly
-to Resource Description Framework (RDF) subject-predicate-object triples,
-and therefore the only requirement for custom metadata is that they be
-cast as RDF triples.
-
 The `RNeXML` interface described above for built-in metadata annotations
 perform this cast automatically behind the scenes, including mapping
-metadata attributes to terms in requisite common vocabularies (such
-as Dublin Core for "title", "creator", etc.) or ontologies. Giving
+metadata attributes to terms in requisite common vocabularies or ontologies. Giving
 metadata properties (and where applicable, values) as vocabulary or
@@
 machines to not only parse but also interpret and potentially reason
-over their semantics. To achieve this benefit for custom metadata
+over their semantics [@Parr2011]. To achieve this benefit for custom metadata
 extensions, the user necessarily needs to handle certain technical
-details from which the `RNeXML` API shields her otherwise, in particular
+details from which the `RNeXML` interface shields her otherwise, in particular
 the globally unique identifiers (normally HTTP URIs) of metadata terms
@@
 provided along with its URI as the `namespaces` parameter. For example,
-the following uses the SKOS vocabulary (which currently is not built-in)
-to add a note to the trees in the `nexml` object:
+the following uses the "[Simple Knowledge Organization System](http://www.w3.org/TR/skos-reference/)" (SKOS) 
+vocabulary to add a note to the trees in the `nexml` object:
 
@@
 
-Because NeXML expresses all metadata using the RDF standard, and
-stores them compliant with RDFa, they can be extracted as an RDF
-graph, queried, analyzed, and mashed up with other RDF data, local or
-on the web, using a wealth of off-the-shelf tools for working with RDF
-(see @W3C_2014 or @Hartig_2012). Examples for these possibilities are
-included in the `RNeXML` SPARQL vignette, and the package also comes with a
-demonstration that can be run from R using the following command:
-`demo("sparql", "RNeXML")`).
+Because NeXML expresses all metadata using the RDF standard, and stores
+them compliant with RDFa, they can be extracted as an RDF graph, queried,
+analyzed, and mashed up with other RDF data, local or on the web, using
+a wealth of off-the-shelf tools for working with RDF (see @W3C_2014
+or @Hartig_2012). Examples for these possibilities are included in the
+`RNeXML` SPARQL vignette (a recursive acronym for SPARQL Protocol and
+RDF Query Language, see http://www.w3.org/TR/rdf-sparql-query/), and
+the package also comes with a demonstration that can be run from R using
+the following command: `demo("sparql", "RNeXML")`).
 
@@
 ```{r}
-nex <- add_namespaces(c(simmap = "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
+nex <- add_namespaces(c(simmap = 
+  "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
 ```
@@
 ```{r eval = FALSE}
-doi <- nexml_publish(birds, visibility = "public", repository="figshare")
+doi <- nexml_publish(birds, repository="figshare")
 ```
 
-Figshare also supports "private" visibility, allowing to securely backup
-data to a repository and to share them with collaborators prior to public release.
+This method reserves a permanent identifier (DOI) on the figshare
+repository that can later be made public through the figshare web
+interface.  This also acts as a secure backup of the data to a repository
+and a way to share with collaborators prior to public release.
 
@@
 ```
+
+# References

### 2015-04-17 manuscripts/manuscript.Rmd
further edits, slimming down/simplifying language in metadata and extending nexml/simmap section

@@
 naming the software generating the NeXML, a time-stamp of when a tree
-was produced, and an open data license.
+was produced, and an open data license. These are merely default arguments
+to `add_basic_meta()` and can be configured.
 
@@
 
-The `RNeXML` interface described above for built-in metadata annotations
-perform this cast automatically behind the scenes, including mapping
-metadata attributes to terms in requisite common vocabularies or ontologies. Giving
-metadata properties (and where applicable, values) as vocabulary or
-ontology terms rather than simple text strings is crucial for allowing
+The `RNeXML` interface described above for built-in metadata allows
+users to create precise and semantically rich annotations without 
+confronting any of the complexity of namespaces and ontologies.
+Nevertheless, advanced users may desire the explicit control over
+these semantic tools that takes full advantage of the flexibility
+and extensibility of the NeXML specification [@Vos2012; @Parr2011].
+In this section we detail how to accomplish these more complex uses
+in RNeXML.
+
+Using a vocabulary or ontology terms rather than simple text strings to 
+describe data is crucial for allowing
 machines to not only parse but also interpret and potentially reason
-over their semantics [@Parr2011]. To achieve this benefit for custom metadata
+over their semantics. To achieve this benefit for custom metadata
 extensions, the user necessarily needs to handle certain technical
-details from which the `RNeXML` interface shields her otherwise, in particular
+details from which the `RNeXML` interface shields her otherwise: in particular
 the globally unique identifiers (normally HTTP URIs) of metadata terms
@@
 is a non-interoperable modification of the standard Newick tree format.
+This means that computer programs designed to read Newick or NEXUS formats
+may fail when trying to read in a phylogeny that includes `simmap` annotations.
 
-To express these data as a custom metadata extension, we create metadata
-annotations for the `edge` elements in the phylogeny topology. The
-metadata need to describe the character state being assigned, and
-the duration (in terms of branch length) that the edge spends in
-that state. The `NeXML` standard already defines `state` elements for
-discrete character traits, which, as most elements in NeXML, also have
-unique identifiers by which they can be referenced.
+NEXML's extensibility avoids this problem.  In NeXML, it is perfectly
+possible to add very complex annotations while maintaining a format that
+is valid NeXML syntax which can be read by any NeXML parser. To illustrate
+how RNeXML facilitates extending the NeXML standard to accommodate such 
+complex annotations, we have implemented two functions in the package,
+`nexml_to_simmap` and `simmap_to_nexml`.  These functions illustrate
+how simmap data can be represented as `meta` annotations on the branch
+length elements of a NeXML tree, and provide routines to convert between
+this NeXML representation and the extended `ape::phylo` representation
+of a `simmap` tree in R that was introduced by @Revell_2012. We encourage
+readers to consult the example code in `simmap_to_nexml` to see how
+this is implemented.
 
-In the following
-proposal, we term each series of character state assignments for an
-edge a _reconstruction_, and for each state assignment comprising a
-reconstruction, we provide the length, the identifier of the character
-state, and an ordinal property to record the chronological order of
-the assignment. In this manner, a single edge containing one
-state change could be annotated as follows.
-(the identifier values in the proposal, "cr1", "s1", and
-"s2", are identifiers local to a  -- fictitious -- NeXML document,
-based on the example simmap tree illustrated in @Revell_2012):
-
-```{r}
- m <- meta("simmap:reconstructions", children = c(
-        meta("simmap:reconstruction", children = c(
-
-          meta("simmap:char", "cr1"),
-          meta("simmap:stateChange", children = c(
-            meta("simmap:order", 1),
-            meta("simmap:length", "0.2030"),
-            meta("simmap:state", "s2"))),
-
-          meta("simmap:char", "cr1"),
-          meta("simmap:stateChange", children = c(
-            meta("simmap:order", 2),
-            meta("simmap:length", "0.0022"),
-            meta("simmap:state", "s1")))
-          ))))
-```
-
-The example uses a prefix, `simmap`, to group the newly introduced
+Extensions to NeXML must also be defined in the file's namespace 
+in order to valid.  This provides a way to ensure that a URI providing
+documentation of the extension is always included.  Our examples here
+use the prefix, `simmap`, to group the newly introduced
 metadata properties in a vocabulary, for which the `add_namespace()`
@@
 at least a human-readable informal definition of the terms.
-
-Because the above representation does not modify the NeXML format,
-it should be readable -- even if not necessarily interpretable as
-a stochastic character mapping -- by any compliant NeXML parser. In
-addition, it includes a direct and machine-readable link to metadata
-documentation.
-
-Writing out metadata annotations as in the above proposal may seem
-tedious. However, by virtue of `RNeXML`'s API it is straightforward to
-define functions that encapsulate these details. As a demonstration,
-`RNeXML` comes with functions `simmap_to_nexml()` and `nexml_to_simmap()`
-that map bidirectionally between NeXML files with `simmap` metadata
-according to the above proposal, and the extension of the `ape::phylo`
-class devised by @Revell_2012 in the R package `phytools`. The `RNeXML`
-simmap vignette illustrates how this allows seamlessly generating and
-reading NeXML files containing `simmap` data, and using the various
-functions from `phytools` designed for `simmap` objects [@Revell_2012].
 

### 2015-04-21 manuscripts/manuscript.Rmd
Update manuscript.Rmd

@@
 NeXML format on the web, such as the content of the phylogenetic data
-repository TreeBASE [@Piel_2002]. For example, the following plots a
+repository TreeBASE [@Piel_2002; @Piel_2009]. For example, the following plots a
 tree in TreeBASE (using ape's plot function):

### 2015-04-27 manuscripts/manuscript.Rmd
Edits to the manuscript revision.

@@
   - name: Hilmar Lapp
-    affiliation: NESCent
+    affiliation: dukeplus
 address: 
@@
     address: | 
+      Center for Genomic and Computational Biology, Duke University, Durham, NC, USA
+      and 
       National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
@@
 NeXML is defined by a precise grammar that can be programmatically
-**validated**: meaning that we can verify that a file has followed this
-precise grammar and thus be sure that it can be read (parsed) 
-without errors by any software that uses the NeXML grammar (e.g. RNeXML). 
+**validated**; i.e., it can be verified whether a file precisely
+follows this grammar, and therefore whether it can be read (parsed) 
+without errors by software that uses the NeXML grammar (e.g. RNeXML)
+is predictable. 
 Second, NeXML is **extensible**: a user
@@
 (as we will illustrate) without violating its defining grammar. Third
-and most importantly, NeXML is rich in **computable semantics**: patterns
-that permit a machine to reason about concepts. Note that
-these semantics are based on logical deduction such as: "all frogs are vertebrates,
-so a query for vertebrates should return all species that are listed as frogs"), 
-rather than heuristic machine learning algorithms.  Such inference requires a
-computer to understand the 'semantic' meaning of a word like 'frog', which
-in must be specified in a kind of dictionary known in informatics research
-as an vocabulary or ontology (see @Parr2011 for a more detailed introduction
-in the context of evolution and biodiversity research).
+and most importantly, NeXML is rich in **computable semantics**: it is
+designed for expressing metadata such that machines can understand
+their meaning and make inferences from it. For example, OTUs in a tree
+or character matrix for frog species can be linked to concepts in a
+formally defined hierarchy of taxonomic concepts such as the
+Vertebrate Taxonomy Ontology [@Midford_2014], which enables a machine
+to infer that a query for amphibia is to include the frog data in what
+is returned. (For a more broader discussion of the value of such
+capabilities for evolutionary and biodiversity science we refer the
+reader to @Parr2011.)
 
@@
 Nevertheless, advanced users may desire the explicit control over
-these semantic tools that takes full advantage of the flexibility
+these semantic tools to take full advantage of the flexibility
 and extensibility of the NeXML specification [@Vos2012; @Parr2011].
@@
 extensions, the user necessarily needs to handle certain technical
-details from which the `RNeXML` interface shields her otherwise: in particular
+details from which the `RNeXML` interface shields her otherwise, in particular
 the globally unique identifiers (normally HTTP URIs) of metadata terms
@@
 
-NEXML's extensibility avoids this problem.  In NeXML, it is perfectly
-possible to add very complex annotations while maintaining a format that
-is valid NeXML syntax which can be read by any NeXML parser. To illustrate
-how RNeXML facilitates extending the NeXML standard to accommodate such 
-complex annotations, we have implemented two functions in the package,
-`nexml_to_simmap` and `simmap_to_nexml`.  These functions illustrate
+In contrast, by allowing new data types to be added as --- sometimes
+complex --- metadata annotations NeXML can accommodate data extensions
+without compromise to its grammar and thus syntax In NeXML. To illustrate
+how RNeXML facilitates extending the NeXML standard in this way, we
+have implemented two functions in the package, `nexml_to_simmap` and
+`simmap_to_nexml`.  These functions show
 how simmap data can be represented as `meta` annotations on the branch
@@
 of a `simmap` tree in R that was introduced by @Revell_2012. We encourage
-readers to consult the example code in `simmap_to_nexml` to see how
-this is implemented.
+readers interested in this capability to consult the example code in
+`simmap_to_nexml` to see how this is implemented.
 

### 2015-04-27 manuscripts/manuscript.Rmd
Adds reference to VTO.

@@
 formally defined hierarchy of taxonomic concepts such as the
-Vertebrate Taxonomy Ontology [@Midford_2014], which enables a machine
+Vertebrate Taxonomy Ontology [@Midford2013], which enables a machine
 to infer that a query for amphibia is to include the frog data in what

### 2015-04-27 manuscripts/manuscript.Rmd
final revisions

--- 2015-04-21 manuscripts/manuscript.Rmd
@@
 NeXML format on the web, such as the content of the phylogenetic data
-repository TreeBASE [@Piel_2002; @Piel_2009]. For example, the following plots a
+repository TreeBASE [@Piel_2002]. For example, the following plots a
 tree in TreeBASE (using ape's plot function):
@@
 these semantic tools that takes full advantage of the flexibility
-and extensibility of the NeXML specification [@Vos2012; @Parr2011].
+and extensibility of the NeXML specification [@Vos_2012; @Parr2011].
 In this section we detail how to accomplish these more complex uses
@@
 around better interoperability with other existing phylogenetic data
-representations in R, such those found in the `phylobase` package
+representations in R, such as those found in the `phylobase` package
 [@phylobase]; and better integration of the rich metadata semantics

### 2015-04-27 manuscripts/manuscript.Rmd

--- 2015-04-27 manuscripts/manuscript.Rmd
@@
 Nevertheless, advanced users may desire the explicit control over
-these semantic tools to take full advantage of the flexibility
-and extensibility of the NeXML specification [@Vos2012; @Parr2011].
+these semantic tools that takes full advantage of the flexibility
+and extensibility of the NeXML specification [@Vos_2012; @Parr2011].
 In this section we detail how to accomplish these more complex uses

### 2015-04-27 manuscripts/manuscript.Rmd
merged commit

@@
       Santa Cruz, CA 95064, USA
-  - code: NESCent
+  - code: dukeplus
     address: | 

### 2015-04-27 manuscripts/manuscript.Rmd

--- 2015-04-27 manuscripts/manuscript.Rmd
@@
       Santa Cruz, CA 95064, USA
-  - code: NESCent
-    address: | 
-      Center for Genomic and Computational Biology, Duke University, Durham, NC, USA
-      and 
-      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
   - code: ropensci
@@
       Naturalis Biodiversity Center, Leiden, the Netherlands
+  - code: dukeplus
+    address: | 
+      Center for Genomic and Computational Biology, Duke University,
+      and 
+      National Evolutionary Synthesis Center, Durham, NC, USA
 abstract: | 

### 2015-04-28 inst/doc/S4.Rmd
prebuilt vignettes

--- 2014-08-18 manuscripts/supplement.Rmd
@@
 ---
-layout: review, 11pt
-title: "Supplementary material"
+title: "The nexml S4 Object"
 author:
-  - name: Carl Boettiger
-    affiliation: cstar
-    email: cboettig(at)gmail.com
-    footnote: Corresponding author
-  - name: Hilmar Lapp
-    affiliation: NESCent
-  - name: Scott Chamberlain
-    affiliation: SimonFraser
-  - name: Rutger Vos
-    affiliation: NBC
-address: 
-  - code: cstar
-    address: | 
-      Center for Stock Assessment Research, 
-      Department of Applied Math and Statistics, 
-      University of California, Mail Stop SOE-2,
-      Santa Cruz, CA 95064, USA
-  - code: NESCent
-    address: | 
-      National Evolutionary Synthesis Center, Durham, NC, USA
-  - code: SimonFraser
-    address: |
-      Department of Biology, Simon Fraser University, Burnaby, Canada
-  - code: NBC
-    address: | 
-      Naturalis Biodiversity Center, Leiden, the Netherlands
+  - Carl Boettiger
+  - Scott Chamberlain
+  - Rutger Vos
+  - Hilmar Lapp
+output: html_vignette
+---
 
-bibliography: components/sup-references.bib
-csl: components/methods-in-ecology-and-evolution.csl
-documentclass: components/elsarticle
-
-output: 
-  pdf_document:
-    template: components/elsarticle.latex
-    keep_tex: false
-    fig_caption: true
-
----
+<!--
+%\VignetteEngine{knitr::rmarkdown}
+%\VignetteIndexEntry{The nexml S4 Object}
+-->
 
@@
                cache = 1, comment = NA, verbose = TRUE)
-basename <- 'supplement' 
-opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
-               cache.path = paste("components/cache/", basename, "/", sep=""))
+basename <- 'S4' 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
 
@@
 Here we use the `c` operator to append this element to any existing meta annotations to this otu.
-
-## SPARQL Queries
-
-Rich, semantically meaningful metadata lies at the heart of the NeXML
-standard.  R provides a rich environment to unlock this information.
-While our previous examples have relied on the user knowing exactly
-what metadata they intend to extract (title, publication date, citation
-information, and so forth), _semantic_ metadata has meaning that a
-computer can make use of, allowing us to make much more conceptually
-rich queries than those simple examples.  The SPARQL query language
-is a powerful way to make use of such semantic information in making 
-complex queries.
-
-While users should consult a formal introduction to SPARQL for further
-background, here we illustrate how SPARQL can be used in combination
-with R functions in ways that would be much more tedious to assemble
-with only traditional/non-semantic queries.  The SPARQL query language
-is provided for the R environment through the `rrdf` package [@Willighagen_2014],
-so we start by loading that package.  We will also make use of functions
-from `phytools` and `RNeXML`.
-
-```{r}
-library(rrdf)
-library(phytools)
-library(RNeXML)
-```
-
-We read in an example file that contains semantic metadata annotations
-describing the taxonomic units (OTUs) used in the tree.
-
-```{r}
-nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
-```
-
-In particular, this example declares the taxon rank, NCBI identifier and parent taxon
-for each OTU, such as:
-
-```xml
-<otu about="#ou541" id="ou541" label="Alouatta guariba">
-      <meta href="http://ncbi.nlm.nih.gov/taxonomy/182256" 
-            id="ma20" 
-            rel="concept:toTaxon" 
-            xsi:type="nex:ResourceMeta"/>
-      <meta href="http://rs.tdwg.org/ontology/voc/TaxonRank#Species" 
-            id="ma21" 
-            rel="concept:rank" 
-            xsi:type="nex:ResourceMeta"/>
-      <meta href="http://ncbi.nlm.nih.gov/taxonomy/9499" 
-            id="ma22" 
-            rel="rdfs:subClassOf" 
-            xsi:type="nex:ResourceMeta"/>
-    </otu>
-
-```
-
-In this example, we will construct a cladogram by using this information
-to identify the taxonomic rank of each OTU, and its shared parent
-taxonomic rank.  (If this example looks complex, try writing down the
-steps to do this without the aid of the SPARQL queries).  These examples
-show the manipulation of semantic triples, Unique Resource Identifiers
-(URIs) and use of the SPARQL "Join" operator.
-
-Note that this example can be run using `demo("sparql", "RNeXML")` to see the 
-code displayed in the R terminal and to avoid character errors that
-can occur in having to copy and paste from PDF files.
-
-We begin by extracting the RDF graph from the NeXML,
-
-```{r}
-rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
-tmp <- tempfile()  # so we must write the XML out first
-saveXML(rdf, tmp) 
-graph <- load.rdf(tmp)
-```
-
-We then fetch the NCBI URI for the taxon that has rank 'Order', i.e. the
-root of the primates phylogeny. The dot operator `.` between clauses
-implies a join, in this case
-
-```{r}
-root <- sparql.rdf(graph, 
-"SELECT ?uri WHERE { 
-    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
-    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
-}")
-```
-
-This makes use of the SPARQL query language provided by the `rrdf`
-package.  We will also define some helper functions that use SPARQL
-queries.  Here we define a function to get the name
-
-```{r}
-get_name <- function(id) {
-  max <- length(nexml@otus[[1]]@otu)
-  for(i in 1:max) {
-    if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
-      label <- nexml@otus[[1]]@otu[[i]]@label
-      label <- gsub(" ","_",label)
-      return(label)
-    }
-  }
-}
-```
-
-Next, we define a recursive function to build a newick tree from the taxonomic rank information.
-
-```{r}
-recurse <- function(node){
-
-    # fetch the taxonomic rank and id string
-    rank_query <- paste0(
-        "SELECT ?rank ?id WHERE {
-            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
-            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
-          }")
-    result <- sparql.rdf(graph, rank_query)
-
-    # get the local ID, strip URI part
-    id <- result[2]
-    id <- gsub("^.+#", "", id, perl = TRUE)
-
-    # if rank is terminal, return the name
-    if (result[1] == "http://rs.tdwg.org/ontology/voc/TaxonRank#Species") {
-        return(get_name(id))
-    }
-
-    # recurse deeper
-    else {
-        child_query <- paste0(
-            "SELECT ?uri WHERE {
-                ?id <http://www.w3.org/2000/01/rdf-schema#subClassOf> <",node,"> .
-                ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
-            }")
-        children <- sparql.rdf(graph, child_query)
-
-        return(paste("(", 
-                     paste(sapply(children, recurse), 
-                           sep = ",", collapse = "," ), 
-                     ")",  
-                     get_name(id), # label interior nodes
-                     sep = "", collapse = ""))
-    }
-}
-
-```
-
-With these functions in place, it is straight forward to build the tree
-from the semantic RDFa data and then visualize it
-
-<!-- hoping Rutger can add something here so that read.newick works when we have interior tip labels. -->
-
-```{r}
-newick <- paste(recurse(root), ";", sep = "", collapse = "")
-tree <- read.newick(text = newick)
-collapsed <- collapse.singles(tree)
-plot(collapsed, 
-     type='cladogram', 
-     show.tip.label=FALSE, 
-     show.node.label=TRUE, 
-     cex=0.75, 
-     edge.color='grey60', 
-     label.offset=-9)
-```
-
-# Implementation, architecture and quality control
-
-The `RNeXML` package is built for the NeXML Schema, [version 2009],
-using the R software environment and the XML package for parsing and
-generating XML [@Temple_Lang_2013],  provides conversions into
-the popular `phylo` format provided by the `ape` package [@Paradis_2004],  
-`RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package [@Chamberlain_2014].
-Unique identifiers are provided with the `uuid` package [@Urbank_2014],
-and internet interfaces by the `httr` package [@Wickham_2014]. The
-packages `plyr` and `reshape2` [@Wickham_2011; Wickham_2007] are also imported for internal methods.
-
-The `RNeXML` package performs several levels of validation on NeXML files
-written from or read into R: validating the XML, and validating against
-the current NeXML schema, and performing additional checks through the 
-NeXML online validation tool for elements that cannot be enforced by
-schema validation alone.  Validation serves to improve consistency in
-data encoding which facilitates the use of NeXML across a wide variety
-of platforms and software.  A lack of a consistent standard and validation
-mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult [@Vos_2012].
-
-Several mechanisms are in place to evaluate the quality of the `RNeXML`
-package functions themselves.  In addition to the examples from the
-documentation and the automated package checks provided for the R system,
-the `RNeXML` package includes a complete test suite using the `testthat`
-package [@Wickham_2011a]. Package development on Github includes continuous 
-integration through the `travis` platform, in which the full test suite
-is run every time the code is changed.  This not only gives users confidence
-that the development copy can be successfully installed, but should assist
-other developers seeking to contribute to the package to test that their proposed 
-changes have not negatively impacted any of the existing code base.
-
-```{r cleanup, include=FALSE, cache=FALSE}
-unlink("simmap.xml")
-unlink("birds.xml")
-unlink("geospiza.xml")
-unlink(tmp)
-```
-
-<!-- MISC / SCRATCH
-
-What to say about existing formats?  
-A summary of some phylogenetic formats in R:
-
-| Class     | R package   | R Object 
-|-----------|-------------|-------
-| phylo     | ape         | S3 
-| phylo4    | phylobase   | S4
-| phylo4d   | adephylo    | S4
-| ouchtree  | ouch        | S4 
-| treeshape | apTreeshape | list
-
-See [phylobase vignette, 2010]
-
-phylobase::phylo4 was an attempt to solve the problems associated with
-the popular ape::phylo format (unpredictable errors, loose and multiple
-conflicting specifications).
-
--->

### 2015-04-28 inst/doc/sparql.Rmd
prebuilt vignettes

--- 2014-08-18 manuscripts/supplement.Rmd
@@
 ---
-layout: review, 11pt
-title: "Supplementary material"
+title: "SPARQL with RNeXML"
 author:
-  - name: Carl Boettiger
-    affiliation: cstar
-    email: cboettig(at)gmail.com
-    footnote: Corresponding author
-  - name: Hilmar Lapp
-    affiliation: NESCent
-  - name: Scott Chamberlain
-    affiliation: SimonFraser
-  - name: Rutger Vos
-    affiliation: NBC
-address: 
-  - code: cstar
-    address: | 
-      Center for Stock Assessment Research, 
-      Department of Applied Math and Statistics, 
-      University of California, Mail Stop SOE-2,
-      Santa Cruz, CA 95064, USA
-  - code: NESCent
-    address: | 
-      National Evolutionary Synthesis Center, Durham, NC, USA
-  - code: SimonFraser
-    address: |
-      Department of Biology, Simon Fraser University, Burnaby, Canada
-  - code: NBC
-    address: | 
-      Naturalis Biodiversity Center, Leiden, the Netherlands
+  - Carl Boettiger
+  - Scott Chamberlain
+  - Rutger Vos
+  - Hilmar Lapp
+output: html_vignette
+---
 
-bibliography: components/sup-references.bib
-csl: components/methods-in-ecology-and-evolution.csl
-documentclass: components/elsarticle
-
-output: 
-  pdf_document:
-    template: components/elsarticle.latex
-    keep_tex: false
-    fig_caption: true
-
----
+<!--
+%\VignetteEngine{knitr::rmarkdown}
+%\VignetteIndexEntry{SPARQL with RNeXML}
+-->
 
@@
                cache = 1, comment = NA, verbose = TRUE)
-basename <- 'supplement' 
-opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
-               cache.path = paste("components/cache/", basename, "/", sep=""))
+basename <- 'sparql' 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
 
@@
 ```
-
-## Understanding the `nexml` S4 object
-
-The `RNeXML` package provides many convenient functions to add and extract
-information from `nexml` objects in the R environment without requiring
-the reader to understand the details of the NeXML data structure and
-making it less likely that a user will generate invalid NeXML syntax
-that could not be read by other parsers. The `nexml` object we have been using
-in all of the examples is built on R's S4 mechanism. Advanced users may
-sometimes prefer to interact with the data structure more directly using 
-R's S4 class mechanism and subsetting methods. Many R users are more familiar
-with the S3 class mechanism (such as in the `ape` package phylo objects)
-rather than the S4 class mechanism used in phylogenetics packages such as 
-`ouch` and `phylobase`. The `phylobase` vignette provides an excellent introduction
-to these data structures.  Users already familiar with subsetting lists and other
-S3 objects in R are likely familar with the use of the `$` operator, such as
-`phy$edge`. S4 objects simply use an `@` operator instead (but cannot be subset
-using numeric arguments such as `phy[[1]]` or named arguments such as phy[["edge"]]).
-
-The `nexml` object is an S4 object, as are all of its components (slots).  Its 
-hierarchical structure corresponds exactly with the XML tree of a NeXML file, with 
-the single exception that both XML attributes and children are represented as slots.  
-S4 objects have constructor functions to initialize them.  We create a new `nexml` 
-object with the command:
-
-```{r}
-nex <- new("nexml")
-```
-
-We can see a list of slots contained in this object with
-
-```{r}
-slotNames(nex)
-```
-
-Some of these slots have already been populated for us, for instance, the schema version and default namespaces:
-
-```{r}
-nex@version
-nex@namespaces
-```
-
-Recognize that `nex@namespaces` serves the same role as `get_namespaces`
-function, but provides direct access to the slot data.  For instance,
-with this syntax we could also overwrite the existing namespaces with
-`nex@namespaces <- NULL`.  Changing the namespace in this way is not
-advised.
-
-Some slots can contain multiple elements of the same type, such as
-`trees`, `characters`, and `otus`.  For instance, we see that
-
-```{r}
-class(nex@characters)
-```
-
-is an object of class `ListOfcharacters`, and is currently empty,
-
-```{r}
-length(nex@characters)
-```
-
-In order to assign an object to a slot, it must match the class definition
-of the slot.  We can create a new element of any given class with the
-`new` function,
-
-```{r}
-nex@characters <- new("ListOfcharacters", list(new("characters")))
-```
-
-and now we have a length-1 list of character matrices,
-
-```{r}
-length(nex@characters)
-```
-
-and we access the first character matrix using the list notation,
-`[[1]]`. Here we check the class is a `characters` object.
-
-```{r}
-class(nex@characters[[1]])
-```
-
-Direct subsetting has two primary use cases: (a) useful in looking up
-(and possibly editing) a specific value of an element, or (b) when adding
-metadata annotations to specific elements. Consider the example file
-
-```{r}
-f <- system.file("examples", "trees.xml", package="RNeXML")
-nex <- nexml_read(f)
-```
-
-We can look up the species label of the first `otu` in the first `otus` block:
-
-```{r}
-nex@otus[[1]]@otu[[1]]@label
-```
-
-We can add metadata to this particular OTU using this subsetting format
-
-```{r}
-nex@otus[[1]]@otu[[1]]@meta <- 
-  c(meta("skos:note", 
-          "This species was incorrectly identified"),
-         nex@otus[[1]]@otu[[1]]@meta)
-```
-
-Here we use the `c` operator to append this element to any existing meta annotations to this otu.
 
@@
 ```{r}
-library(rrdf)
-library(phytools)
-library(RNeXML)
+library("rrdf")
+library("XML")
+library("phytools")
+library("RNeXML")
 ```
@@
 
-<!-- hoping Rutger can add something here so that read.newick works when we have interior tip labels. -->
-
 ```{r}
@@
 ```
-
-# Implementation, architecture and quality control
-
-The `RNeXML` package is built for the NeXML Schema, [version 2009],
-using the R software environment and the XML package for parsing and
-generating XML [@Temple_Lang_2013],  provides conversions into
-the popular `phylo` format provided by the `ape` package [@Paradis_2004],  
-`RNeXML` can automatically check taxonomic names and add taxonomic
-identifiers when generating NeXML using methods from the `taxize` package [@Chamberlain_2014].
-Unique identifiers are provided with the `uuid` package [@Urbank_2014],
-and internet interfaces by the `httr` package [@Wickham_2014]. The
-packages `plyr` and `reshape2` [@Wickham_2011; Wickham_2007] are also imported for internal methods.
-
-The `RNeXML` package performs several levels of validation on NeXML files
-written from or read into R: validating the XML, and validating against
-the current NeXML schema, and performing additional checks through the 
-NeXML online validation tool for elements that cannot be enforced by
-schema validation alone.  Validation serves to improve consistency in
-data encoding which facilitates the use of NeXML across a wide variety
-of platforms and software.  A lack of a consistent standard and validation
-mechanism has continually plagued the current NEXUS standard and made 
-interoperability difficult [@Vos_2012].
-
-Several mechanisms are in place to evaluate the quality of the `RNeXML`
-package functions themselves.  In addition to the examples from the
-documentation and the automated package checks provided for the R system,
-the `RNeXML` package includes a complete test suite using the `testthat`
-package [@Wickham_2011a]. Package development on Github includes continuous 
-integration through the `travis` platform, in which the full test suite
-is run every time the code is changed.  This not only gives users confidence
-that the development copy can be successfully installed, but should assist
-other developers seeking to contribute to the package to test that their proposed 
-changes have not negatively impacted any of the existing code base.
-
-```{r cleanup, include=FALSE, cache=FALSE}
-unlink("simmap.xml")
-unlink("birds.xml")
-unlink("geospiza.xml")
-unlink(tmp)
-```
-
-<!-- MISC / SCRATCH
-
-What to say about existing formats?  
-A summary of some phylogenetic formats in R:
-
-| Class     | R package   | R Object 
-|-----------|-------------|-------
-| phylo     | ape         | S3 
-| phylo4    | phylobase   | S4
-| phylo4d   | adephylo    | S4
-| ouchtree  | ouch        | S4 
-| treeshape | apTreeshape | list
-
-See [phylobase vignette, 2010]
-
-phylobase::phylo4 was an attempt to solve the problems associated with
-the popular ape::phylo format (unpredictable errors, loose and multiple
-conflicting specifications).
-
--->

### 2015-04-28 inst/doc/README.R

@@ -1,0 +1,98 @@
+## ----compile-settings, include=FALSE-------------------------------------
+## Set the paths for cache and figure
+library(methods)
+library(knitr)
+opts_chunk$set(cache = 1)
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
+
+
+## ----echo=FALSE----------------------------------------------------------
+library(knitr)
+opts_knit$set(upload.fun = imgur_upload)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, comment = NA, verbose = TRUE)
+
+## ----eval=FALSE----------------------------------------------------------
+#  library(devtools)
+#  install_github("RNeXML", "ropensci")
+#  library(RNeXML)
+
+## ----echo=FALSE----------------------------------------------------------
+library(RNeXML)
+
+## ------------------------------------------------------------------------
+f <- system.file("examples", "comp_analysis.xml", package="RNeXML")
+nexml <- nexml_read(f)
+tr <- get_trees(nexml) # or: as(nexml, "phylo")
+plot(tr)
+
+## ------------------------------------------------------------------------
+data(bird.orders)
+nexml_write(bird.orders, "test.xml")
+
+## ------------------------------------------------------------------------
+nexml_validate("test.xml")
+
+## ------------------------------------------------------------------------
+birds <- nexml_read("test.xml")
+get_taxa(birds)
+get_metadata(birds) 
+
+## ------------------------------------------------------------------------
+  nexml_write(bird.orders, file="meta_example.xml",
+              title = "My test title",
+              description = "A description of my test",
+              creator = "Carl Boettiger <cboettig@gmail.com>",
+              publisher = "unpublished data",
+              pubdate = "2012-04-01")
+
+
+## ------------------------------------------------------------------------
+modified <- meta(property = "prism:modificationDate",
+                 content = "2013-10-04")
+
+## ------------------------------------------------------------------------
+RNeXML:::nexml_namespaces
+
+## ------------------------------------------------------------------------
+website <- meta(href = "http://carlboettiger.info", 
+                rel = "foaf:homepage")
+
+## ------------------------------------------------------------------------
+  history <- meta(property = "skos:historyNote", 
[…]

### 2015-04-28 inst/doc/S4.R

@@ -1,0 +1,49 @@
+## ----supplement-compile-settings, include=FALSE--------------------------
+library("methods")
+library("knitr")
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- 'S4' 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
+
+
+## ------------------------------------------------------------------------
+nex <- new("nexml")
+
+## ------------------------------------------------------------------------
+slotNames(nex)
+
+## ------------------------------------------------------------------------
+nex@version
+nex@namespaces
+
+## ------------------------------------------------------------------------
+class(nex@characters)
+
+## ------------------------------------------------------------------------
+length(nex@characters)
+
+## ------------------------------------------------------------------------
+nex@characters <- new("ListOfcharacters", list(new("characters")))
+
+## ------------------------------------------------------------------------
+length(nex@characters)
+
+## ------------------------------------------------------------------------
+class(nex@characters[[1]])
+
+## ------------------------------------------------------------------------
+f <- system.file("examples", "trees.xml", package="RNeXML")
+nex <- nexml_read(f)
+
+## ------------------------------------------------------------------------
+nex@otus[[1]]@otu[[1]]@label
+
+## ------------------------------------------------------------------------
+nex@otus[[1]]@otu[[1]]@meta <- 
+  c(meta("skos:note", 
+          "This species was incorrectly identified"),
+         nex@otus[[1]]@otu[[1]]@meta)
+

### 2015-04-28 inst/doc/metadata.R

@@ -1,0 +1,94 @@
+## ----compile-settings, include=FALSE-------------------------------------
+library("methods")
+library("knitr")
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
+
+## ------------------------------------------------------------------------
+library('RNeXML')
+library('geiger')
+data(bird.orders)
+
+## ------------------------------------------------------------------------
+birds <- add_trees(bird.orders)
+birds <- add_basic_meta(
+  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
+
+  description = "This data set describes the phylogenetic relationships of the
+     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
+     and Ahlquist inferred this phylogeny from an extensive number of
+     DNA/DNA hybridization experiments. The ``tapestry'' reported by
+     these two authors (more than 1000 species out of the ca. 9000
+     extant bird species) generated a lot of debates.
+
+     The present tree is based on the relationships among orders. The
+     branch lengths were calculated from the values of Delta T50H as
+     found in Sibley and Ahlquist (1990, fig. 353).",
+
+  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
+     classification of birds: a study in molecular evolution. New
+     Haven: Yale University Press.",
+
+  creator = "Sibley, C. G. and Ahlquist, J. E.",
+	nexml=birds)
+
+## ------------------------------------------------------------------------
+birds <- add_basic_meta(citation = citation("ape"), nexml = birds)
+
+## ------------------------------------------------------------------------
+library("knitcitations")
+geiger_nex <- add_basic_meta(citation = bib_metadata("10.2307/2408428"))
+
+## ----message=FALSE, results='hide'---------------------------------------
+birds <- taxize_nexml(birds, "NCBI")
+
+## ------------------------------------------------------------------------
+prefixes <- get_namespaces(birds)
+prefixes["dc"]
+
+## ------------------------------------------------------------------------
+modified <- meta(property = "prism:modificationDate", content = "2013-10-04")
+
+## ------------------------------------------------------------------------
+birds <- add_meta(modified, birds) 
+
+## ------------------------------------------------------------------------
[…]

### 2015-04-28 inst/doc/simmap.R

@@ -1,0 +1,45 @@
+## ----compile-settings, include=FALSE-------------------------------------
+library("methods")
+library("knitr")
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+library("RNeXML")
+
+
+## ------------------------------------------------------------------------
+ m <- meta("simmap:reconstructions", children = c(
+        meta("simmap:reconstruction", children = c(
+
+          meta("simmap:char", "cr1"),
+          meta("simmap:stateChange", children = c(
+            meta("simmap:order", 1),
+            meta("simmap:length", "0.2030"),
+            meta("simmap:state", "s2"))),
+          
+          meta("simmap:char", "cr1"),
+          meta("simmap:stateChange", children = c(
+            meta("simmap:order", 2),
+            meta("simmap:length", "0.0022"),
+            meta("simmap:state", "s1")))
+          ))))
+
+## ------------------------------------------------------------------------
+nex <- add_namespaces(c(simmap = "https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"))
+
+## ------------------------------------------------------------------------
+data(simmap_ex)
+
+## ------------------------------------------------------------------------
+phy <- nexml_to_simmap(simmap_ex)
+
+## ----Figure1, fig.cap="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML."----
+library("phytools")
+plotSimmap(phy)
+
+## ------------------------------------------------------------------------
+nex <- simmap_to_nexml(phy) 
+nexml_write(nex, "simmap.xml")
+

### 2015-04-28 inst/doc/sparql.R

@@ -1,0 +1,96 @@
+## ----supplement-compile-settings, include=FALSE--------------------------
+library("methods")
+library("knitr")
+opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
+               cache = 1, comment = NA, verbose = TRUE)
+basename <- 'sparql' 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
+
+
+## ------------------------------------------------------------------------
+library("rrdf")
+library("XML")
+library("phytools")
+library("RNeXML")
+
+## ------------------------------------------------------------------------
+nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
+
+## ------------------------------------------------------------------------
+rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+tmp <- tempfile()  # so we must write the XML out first
+saveXML(rdf, tmp) 
+graph <- load.rdf(tmp)
+
+## ------------------------------------------------------------------------
+root <- sparql.rdf(graph, 
+"SELECT ?uri WHERE { 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
+}")
+
+## ------------------------------------------------------------------------
+get_name <- function(id) {
+  max <- length(nexml@otus[[1]]@otu)
+  for(i in 1:max) {
+    if ( nexml@otus[[1]]@otu[[i]]@id == id ) {
+      label <- nexml@otus[[1]]@otu[[i]]@label
+      label <- gsub(" ","_",label)
+      return(label)
+    }
+  }
+}
+
+## ------------------------------------------------------------------------
+recurse <- function(node){
+  
+    # fetch the taxonomic rank and id string
+    rank_query <- paste0(
+        "SELECT ?rank ?id WHERE {
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> <",node,"> .
+            ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> ?rank
+          }")
+    result <- sparql.rdf(graph, rank_query)
+    
+    # get the local ID, strip URI part
+    id <- result[2]
+    id <- gsub("^.+#", "", id, perl = TRUE)
+    
[…]

### 2015-05-01 inst/doc/S4.R
skip coveralls on circleci

@@ -2,11 +2,8 @@
 library("methods")
 library("knitr")
 opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
-               cache = 1, comment = NA, verbose = TRUE)
+               cache = FALSE, comment = NA, verbose = TRUE)
 basename <- 'S4' 
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
-
 
 
 ## ----include=FALSE-------------------------------------------------------

### 2015-05-01 inst/doc/metadata.R

@@ -2,15 +2,12 @@
 library("methods")
 library("knitr")
 opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
-               cache = 1, comment = NA, verbose = TRUE)
+               cache = FALSE, comment = NA, verbose = TRUE)
 basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
 
 
 ## ------------------------------------------------------------------------
 library('RNeXML')
-library('geiger')
 data(bird.orders)
 
 ## ------------------------------------------------------------------------
@@ -38,10 +35,6 @@ birds <- add_basic_meta(
 
 ## ------------------------------------------------------------------------
 birds <- add_basic_meta(citation = citation("ape"), nexml = birds)
-
-## ------------------------------------------------------------------------
-library("knitcitations")
-geiger_nex <- add_basic_meta(citation = bib_metadata("10.2307/2408428"))
 
 ## ----message=FALSE, results='hide'---------------------------------------
 birds <- taxize_nexml(birds, "NCBI")

### 2015-05-01 inst/doc/simmap.R

@@ -2,10 +2,8 @@
 library("methods")
 library("knitr")
 opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
-               cache = 1, comment = NA, verbose = TRUE)
+               cache = FALSE, comment = NA, verbose = TRUE)
 basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
 library("RNeXML")
 
 

### 2015-05-01 inst/doc/sparql.R

@@ -2,98 +2,96 @@
 library("methods")
 library("knitr")
 opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, 
-               cache = 1, comment = NA, verbose = TRUE)
+               cache = FALSE, comment = NA, verbose = TRUE, eval=require("rrdf"))
 basename <- 'sparql' 
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
 
 
 
 ## ----include=FALSE-------------------------------------------------------
-library("RNeXML")
+#  library("RNeXML")
 
 ## ------------------------------------------------------------------------
-library("rrdf")
-library("XML")
-library("phytools")
-library("RNeXML")
+#  library("rrdf")
+#  library("XML")
+#  library("phytools")
+#  library("RNeXML")
 
 ## ------------------------------------------------------------------------
-nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
+#  nexml <- nexml_read(system.file("examples/primates.xml", package="RNeXML"))
 
 ## ------------------------------------------------------------------------
-rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
-tmp <- tempfile()  # so we must write the XML out first
-saveXML(rdf, tmp) 
-graph <- load.rdf(tmp)
+#  rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+#  tmp <- tempfile()  # so we must write the XML out first
+#  saveXML(rdf, tmp)
+#  graph <- load.rdf(tmp)
 
 ## ------------------------------------------------------------------------
-root <- sparql.rdf(graph, 
-"SELECT ?uri WHERE { 
-    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
-    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
-}")
+#  root <- sparql.rdf(graph,
+#  "SELECT ?uri WHERE {
+#      ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> .
+#      ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri
+#  }")
 
[…]

### 2015-05-25 manuscripts/manuscript.tex
Update CDAO URI (#92) and rebuild README, manuscript, etc
tex intermediate files for records

--- 2014-08-18 manuscripts/manuscript.tex
@@
 \usepackage{lineno} % add 
+  \linenumbers % turns line numbering on 
 \bibliographystyle{elsarticle-harv}
@@
 \newcommand{\NormalTok}[1]{{#1}}
-\usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
-\makeatother
-\let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 \ifxetex
@@
    \cortext[c1]{Corresponding author}
-    \author[NESCent]{Hilmar Lapp}
-
     \author[ropensci]{Scott Chamberlain}
@@
     \author[NBC]{Rutger Vos}
+
+    \author[dukeplus]{Hilmar Lapp}
 
@@
 95064, USA}    
-    \address[NESCent]{National Evolutionary Synthesis Center, Durham, NC, USA}    
     \address[ropensci]{University of California, Berkeley, CA, USA}    
     \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}    
+    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, and
+National Evolutionary Synthesis Center, Durham, NC, USA}
 
@@
   \item
-    NeXML is a new phylogenetic informatics data and metadata exchange
-    format that offers substantial improvements in reliability,
-    extensibility, and richness over earlier standards such as NEXUS. In
-    this paper we present the RNeXML package which provides a bridge
-    between this rich and reliable data format and the extensive library
-    of phylogenetic tools already available in R, with a particular
-    emphasis on comparative methods.
+    NeXML is a powerful and extensible exchange standard recently proposed
+    to better meet the expanding needs for phylogenetic data and metadata
+    sharing. Here we present the RNeXML package, which provides users of
+    the R programming language with easy-to-use tools for reading and
+    writing NeXML documents, including rich metadata, in a way that
+    interfaces seamlessly with the extensive library of phylogenetic tools
+    already available in the R ecosystem.
   \item
-    RNeXML is much richer than a basic parser: here we illustrate how the
-    package facilitates adding and manipulating semantic annotations, that
-    is annotations whose meaning can be defined externally by way of
-    semantic web technologies.
+    Wherever possible, we designed RNeXML to map NeXML document contents,
+    whose arrangement is influenced by the format's XML Schema definition,
+    to their most intuitive or useful representation in R. To make NeXML's
+    powerful facility for recording semantically rich machine-readable
+    metadata accessible to R users, we designed a functional programming
+    interface to it that hides the semantic web standards leveraged by
+    NeXML from R users who are unfamiliar with them.
   \item
-    An essential feature of NeXML is to provide both the flexibility to be
-    continually extended to meet the emerging needs of the phylogenetics
-    community while remaining backwards-compatible with software developed
-    for earlier versions. Here we illustrate how RNeXML makes it easy to
-    extend NeXML in this manner.
+    RNeXML can read any NeXML document that validates, and it generates
+    valid NeXML documents from phylogeny and character data in various R
+    representations in use. The metadata programming interface at a basic
+    level aids fulfilling data documentation best practices, and at an
+    advanced level preserves NeXML's nearly limitless extensibility, for
+    which we provide a fully working demonstration. Furthermore, to lower
+    the barriers to sharing well-documented phylogenetic data, RNeXML has
+    started to integrate with taxonomic metadata augmentation services on
+    the web, and with online repositories for data archiving.
   \item
-    RNeXML is well suited for users to preserve and maintain their
-    personal archives of phylogenetic and comparative trait data in
-    consistent, query-able, and cross-platform format, to leverage the
-    rapidly growing public community archives of rich NeXML files, and to
-    interact with web services that deliver their results (with their
-    provenance) as NeXML.
+    RNeXML allows R's rich ecosystem to read and write data in the NeXML
+    format through an interface that is no more involved than reading or
+    writing data from other, less powerful data formats. It also provides
+    an interface designed to feel familiar to R programmers and to be
+    consistent with recommended practices for R package development, yet
+    that retains the full power for users to add their own custom data and
+    metadata to the phylogenies they work with, without introducing
+    potentially incompatible changes to the exchange standard.
   \end{enumerate}
@@
 
-Though R provides the most extensive and widely used collection of
-comparative phylogenetic methods tools, R users lack a convenient,
-extensible format to exchange and archive comparative methods data. A
-common file format for sharing phylogenetic tree data has long played an
-essential role in the advancement of phylogenetics research. The rapid
-increase in phylogenetics data and research methods that make use of
-such data has introduced many new challenges and opportunities to
-sharing this information both between software environments and between
-research groups (Stoltzfus \emph{et al.} 2012; Drew \emph{et al.} 2013).
-
-The ever expanding needs of the phylogenetics community are rapidly
-out-pacing the capacity of current standards, which is resulting in a
-radiation of different data representations and a loss of
-interoperability at the very time when the growth of available data and
-methods has made that interoperability most valuable.
-
-For example, the currently popular NEXUS file format cannot consistently
-represent horizontal gene transfer or ambiguity in reading a character
-(such as a DNA sequence base pair). This has led developers to
-repeatedly add small modifications to the standard to meet the needs of
-their own software, which results in mutually incompatible dialects that
-largely defeat the advantages of a shared file format. NeXML was
-developed to addresses each of these problems (Vos \emph{et al.}
-2012)\\by drawing on robust web standards in data representation. The
-design and expressiveness of NeXML makes it much easier to meet the
-rising bar for best practices in data sharing in phylogenetic research
-(Cranston \emph{et al.} 2014).
-
-The R language (R Core Team 2014) has become an important environment
-for phylogenetics research, with over 30 packages available including
-several with over 1000 citations (`CRAN task view' 2014), though
-stand-alone programs remain an important part of this ecosystem for
-tasks such as generating phylogenetic trees from sequence data. Within
-this ecosystem a variety of different data formats for phylogenetic data
-have been developed and modified. Most but not all are based on the
-simple and popular \texttt{ape::phylo} format provided by the ape
-package (Paradis \emph{et al.} 2004).
-
-Though certain functions allow R users to switch between these formats
-within R, these formats frequently contain more information than their
-NEXUS file counterparts, and users lack a suitable alternative for
-archiving and sharing phylogenetic data outside of the R ecosystem.
-Consequently, information added to phylogenies by these R programs
-cannot easily be shared, archived, and made available to other software;
-but inherits the limitations of the NEXUS format.
-
-\texttt{RNeXML} opens up a rich world of both currently available and
-rapidly growing data sets in the NeXML format to the R user and the
-large ecosystem of R packages currently available for handling
-phylogenetic data\\R packages and R users frequently add rich
-information to the phylogenies they work with, but need a platform
-independent data exchange format to share and archive this. We
-illustrate how NeXML can be used to distribute both phylogenies and
-character trait data used in many popular phylogenetic comparative
-methods tools in R.\\\texttt{RNeXML} also provides convenient tools for
-adding metadata to NeXML files, including the automated generation of
-metadata and examples of how to extend the NeXML standard to represent
-new data features without breaking validation or backwards compatibility
-of the file type.
-
-In this paper, we introduce the reader to the concepts and use of RNeXML
-and related technologies. We begin by discussing methods to read
-phylogenetic tree data and character trait data in the NeXML format into
-R, and also how to write phylogenetic data and character trait data into
-NeXML. The heart of this section focuses on tools for reading and
-writing the kind of semantically rich metadata that most sets NeXML
-apart from other phylogenetic data formats. Starting simple, we use
-examples of increasing complexity to illustrate concepts and tools from
-biodiversity informatics that allow users to express and interpret
-complicated data structures automatically. We end with an example that
-uses metadata to extend the NeXML standard to describe stochastic
-character mapping, one of the kinds of data annotation that is currently
-available in R formats but lacks a consistent, extensible export format.
-We illustrate how NeXML files can be formally published to a data
-repository directly from R. Finally, we provide a brief description of
-how the software has been implemented, what quality control measures are
-in place, as well as information about software reuse and getting
-support.
-
-This paper has been written as a dynamic document (Xie 2013). The reader
-may paste the code blocks shown into their R terminal to follow along
-with the exmples shown.
-
-\section{Introduction to RNeXML}\label{introduction-to-rnexml}
-
-A NeXML document can be thought of as having multiple components: (1)
-phylogeny data (2) character-trait data, (3) operational taxonomic
-units, and (4) metadata. \texttt{RNeXML} seeks to provide a seamless way
-to convert these to and from their corresponding R object
-representations, such as \texttt{ape::phylo} phylogenies, character
-matrices (\texttt{data.frames}) and metadata lists (or for advanced
-users, \texttt{XMLInternalDocument} and \texttt{rrdf} triplestores).
-Metadata can provide information about the study in which the phylogeny
-was generated or applied (such as authors, citations to publications,
-timestamps, and even descriptions of methods), but can also be used to
-annotate individual data components. We illustrate how metadata
-annotations can be added to the operational taxonomic units of the
-phylogeny and can be used to extend the data representation to new uses
-such as stochastic character maps.
-
-\subsection{Getting Started}\label{getting-started}
-
-The current stable version of the \texttt{RNeXML} package can be
-installed from the CRAN library\footnote{By providing the optional
-  \texttt{dependencies} argument, R will install additional packages
-  used in some of the advanced examples shown later but which are not
-  required for basic functionality.}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{, }\DataTypeTok{dependencies=}\OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\subsection{Reading phylogenetic
-trees}\label{reading-phylogenetic-trees}
-
-We begin by loading the \texttt{RNeXML} package and reading in an
-example NeXML file provided with the package, containing two small
-phylogenetic trees.\footnote{For convenience \texttt{nexml\_read} is
-  also aliased as \texttt{read.nexml}, for users familiar with the
-  somewhat analogous \texttt{read.nexus} function.}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"RNeXML"}\NormalTok{)}
-\NormalTok{f <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"trees.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
-\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(f)}
-\end{Highlighting}
-\end{Shaded}
-
-The resulting \texttt{nex} object is an R object representation of the
-NeXML. This is a native R object, of the \texttt{nexml} class, for
-representing phylogenetic trees (as well as character traits, and
-metadata), analogous to the \texttt{ape::phylo} object from the popular
-\texttt{ape} R package (Paradis \emph{et al.} 2004). While R functions
-and packages could be written to use this \texttt{nexml} object
-directly, most users will prefer to use already existing methods
-provided by packages they know. From a \texttt{nexml} object, we can
-therefore extract any phylogenies it contains in the \texttt{ape::phylo}
-format. In this case we will get back a \texttt{multiPhylo} object,
-since the file contains two phylogenies.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{phy <-}\StringTok{ }\KeywordTok{get_trees}\NormalTok{(nex)}
-\end{Highlighting}
-\end{Shaded}
-
-We can then leverage the rich suite of methods available from the R
-phylogenetics community. For instance, here we use the standard plotting
-function from the \texttt{ape} package to plot the first tree in the
-resulting \texttt{ape::phylo} object returned by \texttt{get\_trees}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{plot}\NormalTok{(phy[[}\DecValTok{1}\NormalTok{]])}
-\end{Highlighting}
-\end{Shaded}
-
-NeXML can not only contain multiple phylogenetic trees in a single
-document, but can also group sets of trees into separate blocks (e.g.~as
-resulting from different workflows). The \texttt{get\_trees} function
-attempts to return the most sensible R object for NeXML: if only a
-single tree is available, it will return an \texttt{ape::phylo} object;
-if multiple trees are available in a single \texttt{trees} block, it
-will return an \texttt{ape::multiPhylo} object (a list of
-\texttt{ape::phylo} objects). If multiple blocks are available, it will
-return a list of \texttt{multiPhylo} blocks. While most users will find
-this behavior convenient, there are times (such as looping over many
-NeXML files) when having a variable return type is undesirable. To
-always get back a list of \texttt{multiPhylo} objects, even when only
-one tree is available, use \texttt{get\_trees\_list()}.
-
-The ability to read in phylogenetic trees in the NeXML format opens up a
-wide and rapidly growing array of resources to the R user. For instance,
-a user can access all the phylogenies available in TreeBASE through the
-\texttt{nexml} format. \texttt{RNeXML} can read directly from a URL:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Previously, this was possible using the \texttt{treebase} package
-{[}Boettiger \& {Temple Lang} (2012)``{]} for R, which relied on the
-NEXUS parser provided in the \texttt{ape} package. Due to
-inconsistencies present in the NEXUS standard (Vos \emph{et al.} 2012),
-a significant number of TreeBASE NEXUS files could not be read in this
-way. Using the more precisely defined NeXML standard, every valid NeXML
-document on TreeBASE can now successfully be parsed to R objects.
-
-\subsection{Reading character data}\label{reading-character-data}
-
-Like NEXUS files, NeXML documents can also contain character data. While
-most researchers will be familiar with the use of character matrices in
-NEXUS files as a container for molecular sequence (or morphological)
-data used to infer the phylogenetic tree, we find this ideal for
-character data used in comparative methods popular in R. Methods to
-extract character data work much like those for the phylogenetic data.
-Here we read in an example NeXML file that contains both a phylogeny and
-character data. \texttt{get\_characters} returns a \texttt{data.frame}
-with columns as characters and rows as taxa:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{comp_analysis <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"examples"}\NormalTok{, }\StringTok{"comp_analysis.xml"}\NormalTok{, }\DataTypeTok{package=}\StringTok{"RNeXML"}\NormalTok{)}
-\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(comp_analysis)}
+Users of the popular statistical and mathematical computing platform R
+(R Core Team 2014) enjoy a wealth of readily installable comparative
+phylogenetic methods and tools (O'Meara 2014). Exploiting the
+opportunities arising from this wealth for complex and integrative
+comparative research questions relies on the ability to reuse and
+integrate previously generated or published data and metadata. The
+expanding data exchange needs of the evolutionary research community are
+rapidly outpacing the capabilities of most current and widely used data
+exchange standards (Vos \emph{et al.} 2012), which were all developed a
+decade or more ago. This has resulted in a radiation of different data
+representations and exchange standard ``flavors'' that are no longer
+interoperable at the very time when the growth of available data and
+methods has made that interoperability most valuable. In response to the
+unmet needs for standardized data exchange in phylogenetics, a modern
+XML-based exchange standard, called NeXML, has recently been developed
+(Vos \emph{et al.} 2012). NeXML comprehensively supports current data
+exchange needs, is predictably machine-readable, and is forward
+compatible.
+
+The exchange problem for phylogenetic data is particularly acute in
+light of the challenges in finding and sharing phylogenetic data without
+the otherwise common loss of most data and metadata semantics (Stoltzfus
+\emph{et al.} 2012; Drew \emph{et al.} 2013; Cranston \emph{et al.}
+2014). For example, the still popular NEXUS file format (Maddison
+\emph{et al.} 1997) cannot consistently represent horizontal gene
+transfer or ambiguity in reading a character (such as a DNA sequence
+base pair). This and other limitations have led to modifications of
+NEXUS in different ways for different needs, with the unfortunate result
+that NEXUS files generated by one program can be incompatible with
+another (Vos \emph{et al.} 2012). Without a formal grammar, software
+based on NEXUS files may also make inconsistent assumptions about
+tokens, quoting, or element lengths. Vos et al. (2012) estimates that as
+many as 15\% of the NEXUS files in the CIPRES portal contain
+unrecoverable but hard to diagnose errors.
+
+A detailed account of how the NeXML standard addresses these and other
+relevant challenges can be found in Vos et al. (2012). In brief, NeXML
+was designed with the following important properties. First, NeXML is
+defined by a precise grammar that can be programmatically
+\textbf{validated}; i.e., it can be verified whether a file precisely
+follows this grammar, and therefore whether it can be read (parsed)
+without errors by software that uses the NeXML grammar (e.g.~RNeXML) is
+predictable. Second, NeXML is \textbf{extensible}: a user can define
+representations of new, previously unanticipated information (as we will
+illustrate) without violating its defining grammar. Third and most
+importantly, NeXML is rich in \textbf{computable semantics}: it is
+designed for expressing metadata such that machines can understand their
+meaning and make inferences from it. For example, OTUs in a tree or
+character matrix for frog species can be linked to concepts in a
+formally defined hierarchy of taxonomic concepts such as the Vertebrate
+Taxonomy Ontology (Midford \emph{et al.} 2013), which enables a machine
+to infer that a query for amphibia is to include the frog data in what
+is returned. (For a more broader discussion of the value of such
+capabilities for evolutionary and biodiversity science we refer the
+reader to Parr et al. (2011).)
+
+To make the capabilities of NeXML available to R users in an easy-to-use
+form, and to lower the hurdles to adoption of the standard, we present
+RNeXML, an R package that aims to provide easy programmatic access to
+reading and writing NeXML documents, tailored for the kinds of use-cases
+that will be common for users and developers of the wealth of
+evolutionary analysis methods within the R ecosystem.
+
+\section{The RNeXML package}\label{the-rnexml-package}
+
+The \texttt{RNeXML} package is written entirely in R and available under
+a Creative Commons Zero public domain waiver. The current development
+version can be found on Github at
+\href{}{\url{https://github.com/ropensci/RNeXML}}, and the stable
+version can be installed from the CRAN repository. \texttt{RNeXML} is
+part of the rOpenSci project. Users of \texttt{RNeXML} are encouraged to
+submit bug reports or feature requests in the issues log on Github, or
+the phylogenetics R users group list at
+\texttt{r-sig-phylo@r-project.org} for help. Vignettes with more
+detailed examples of specific features of RNeXML are distributed with
+the R package and serve as a supplement to this manuscript. Each of the
+vignettes can be found at
+\href{}{\url{http://ropensci.github.io/RNeXML/}}.
+
+\subsection{Representation of NeXML documents in
+R}\label{representation-of-nexml-documents-in-r}
+
+Conceptually, a NeXML document has the following components: (1)
+phylogeny topology and branch length data, (2) character or trait data
+in matrix form, (3) operational taxonomic units (OTUs), and (4)
+metadata. To represent the contents of a NeXML document (currently in
+memory), \texttt{RNeXML} defines the \texttt{nexml} object type. This
+type therefore holds phylogenetic trees as well as character or trait
+matrices, and all metadata, which is similar to the phylogenetic data
+object types defined in the \texttt{phylobase} package (NESCENT R
+Hackathon Team 2014), but contrasts with the more widely used ones
+defined in the \texttt{ape} package (Paradis \emph{et al.} 2004), which
+represents trees alone.
+
+When reading and writing NeXML documents, \texttt{RNeXML} aims to map
+their components to and from, respectively, their most widely used
+representations in R. As a result, the types of objects accepted or
+returned by the package's methods are the \texttt{phylo} and
+\texttt{multiPhylo} objects from the \texttt{ape} package (Paradis
+\emph{et al.} 2004) for phylogenies, and R's native \texttt{data.frame}
+list structure for data matrices.
+
+\subsection{Reading phylogenies and character
+data}\label{reading-phylogenies-and-character-data}
+
+The method \texttt{nexml\_read()} reads NeXML files, either from a local
+file, or from a remote location via its URL, and returns an object of
+type \texttt{nexml}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"components/trees.xml"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The method \texttt{get\_trees\_list()} can be used to extract the
+phylogenies as an \texttt{ape::multiPhylo} object, which can be treated
+as a list of \texttt{ape::phylo} objects:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{phy <-}\StringTok{ }\KeywordTok{get_trees_list}\NormalTok{(nex)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{get\_trees\_list()} method is designed for use in scripts,
+providing a consistent and predictable return type regardless of the
+number of phylogenies a NeXML document contains. For greater convenience
+in interactive use, the method \texttt{get\_trees()} returns the R
+object most intuitive given the arrangement of phylogeny data in the
+source NeXML document. For example, the method returns an
+\texttt{ape::phylo} object if the NeXML document contains a single
+phylogeny, an \texttt{ape::multiPhylo} object if it contains multiple
+phylogenies arranged in a single \texttt{trees} block, and a list of
+\texttt{ape::multiPhylo} objects if it contains multiple \texttt{trees}
+blocks (the capability for which NeXML inherits from NEXUS).
+
+If the location parameter with which the \texttt{nexml\_read()} method
+is invoked is recognized as a URL, the method will automatically
+download the document to the local working directory and read it from
+there. This gives convenient and rapid access to phylogenetic data
+published in NeXML format on the web, such as the content of the
+phylogenetic data repository TreeBASE (Piel \emph{et al.} 2002, 2009).
+For example, the following plots a tree in TreeBASE (using ape's plot
+function):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{tb_nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}
+\StringTok{"https://raw.github.com/TreeBASE/supertreebase/master/data/treebase/S100.xml"}\NormalTok{)}
+\NormalTok{tb_phy <-}\StringTok{ }\KeywordTok{get_trees_list}\NormalTok{(tb_nex)}
+\KeywordTok{plot}\NormalTok{(tb_phy[[}\DecValTok{1}\NormalTok{]]) }
+\end{Highlighting}
+\end{Shaded}
+
+The method \texttt{get\_characters()} obtains character data matrices
+from a \texttt{nexml} object, and returns them as a standard
+\texttt{data.frame} R object with columns as characters and rows as
+taxa:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nex <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"components/comp_analysis.xml"}\NormalTok{)}
 \KeywordTok{get_characters}\NormalTok{(nex)}
@@
          log snout-vent length reef-dwelling
-taxon_8                 -3.278             0
-taxon_9                  2.096             1
-taxon_10                 3.137             0
-taxon_1                  4.753             1
-taxon_2                 -2.762             0
-taxon_3                  2.105             0
-taxon_4                 -4.950             0
-taxon_5                  1.271             1
-taxon_6                  6.259             1
-taxon_7                  0.910             1
+taxon_8             -3.2777799             0
+taxon_9              2.0959433             1
+taxon_10             3.1373971             0
+taxon_1              4.7532824             1
+taxon_2             -2.7624146             0
+taxon_3              2.1049413             0
+taxon_4             -4.9504770             0
+taxon_5              1.2714718             1
+taxon_6              6.2593966             1
+taxon_7              0.9099634             1
 \end{verbatim}
 
-As this example shows, continuous and discrete characters are
-represented as separate blocks in NeXML files, but will be combined as
-separate columns of a single data frame by \texttt{get\_characters()} if
-they correspond to the same taxa. The \texttt{get\_characters\_list()}
-function will consistently receive a list of data frames corresponding
-to each character matrix block of the NeXML.
-
-\subsection{Writing phylogenetic
-trees}\label{writing-phylogenetic-trees}
-
-Generating NeXML files from R is likewise straightforward. Here we write
-a phylogeny from the \texttt{ape::phylo} format out to a NeXML file:
+A NeXML data matrix can be of molecular (for molecular sequence
+alignments), discrete (for most morphological character data), or
+continuous type (for many trait data). To enable strict validation of
+data types NeXML allows, and if their data types differ requires
+multiple data matrices to be separated into different ``blocks''. Since
+the \texttt{data.frame} data structure in R has no such constraints, the
+\texttt{get\_characters()} method combines such blocks as separate
+columns into a single \texttt{data.frame} object, provided they
+correspond to the same taxa. Otherwise, a list of \texttt{data.frame}s
+is returned, with list elements corresponding to characters blocks.
+Similar to the methods for obtaining trees, there is also a method
+\texttt{get\_characters\_list()}, which always returns a list of
+\texttt{data.frame}s, one for each character block.
+
+\subsection{Writing phylogenies and character
+data}\label{writing-phylogenies-and-character-data}
+
+The method \texttt{nexml\_write()} generates a NeXML file from its input
+parameters. In its simplest invocation, the method writes a tree to a
+file:
 
@@
 
-\begin{verbatim}
-[1] "birds.xml"
-\end{verbatim}
-
-The first argument to \texttt{nexml\_write} can be a \texttt{ape::phylo}
-phylogeny, \texttt{multiPhylo} list of phylogenies, or a \texttt{nexml}
-object -- anything that can be coerced into the \texttt{nexml} class. As
-we have seen, NeXML files can contain multiple phylogenetic trees, as
-well as character matrices. We can add both trees and character data
-using \texttt{nexml\_write},
+The first argument to \texttt{nexml\_write()} is either an object of
+type \texttt{nexml}, or any object that can be coerced to it, such as in
+the above example an \texttt{ape::phylo} phylogeny. Alternatively,
+passing a \texttt{multiPhylo} object would write a list of phylogenies
+to the file.
+
+In addition to trees, the \texttt{nexml\_write()} method also allows to
+specify character data as another parameter. The following example uses
+data from the comparative phylogenetics R package \texttt{geiger}
+(Pennell \emph{et al.} 2014).
 
@@
 \KeywordTok{data}\NormalTok{(geospiza)}
-\KeywordTok{nexml_write}\NormalTok{(}\DataTypeTok{trees =} \NormalTok{geospiza$phy, }\DataTypeTok{characters =} \NormalTok{geospiza$dat, }\DataTypeTok{file=}\StringTok{"geospiza.xml"}\NormalTok{)}
+\KeywordTok{nexml_write}\NormalTok{(}\DataTypeTok{trees =} \NormalTok{geospiza$phy, }
+            \DataTypeTok{characters =} \NormalTok{geospiza$dat,}
+            \DataTypeTok{file=}\StringTok{"geospiza.xml"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Note that the NeXML format is well-suited for incomplete data: for
+instance, here it does not assume the character matrix has data for
+every tip in the tree.
+
+\subsection{Validating NeXML}\label{validating-nexml}
+
+File validation is a central feature of the NeXML format which ensures
+that any properly implemented NeXML parser will always be able to read
+the NeXML file. The function takes the path to any NeXML file and
+returns \texttt{TRUE} to indicate a valid file, or \texttt{FALSE}
+otherwise, along with a display of any error messages generated by the
+validator.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{nexml_validate}\NormalTok{(}\StringTok{"geospiza.xml"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{verbatim}
-[1] "geospiza.xml"
+[1] TRUE
 \end{verbatim}
 
-These examples use data from the popular phylogenetics software
-\texttt{geiger} (Harmon \emph{et al.} 2008).
-
-NeXML files can also be built up bit by bit using the
-\texttt{add\_characters}, \texttt{add\_trees}, and \texttt{add\_meta}.
-Each of these functions can add data to an existing \texttt{nexml}
-object or create a new nexml object if none is provided. Here we first
-create a \texttt{nexml} object containing the phylogeny data, and then
-add the character data to it:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(geospiza$phy)}
-\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_characters}\NormalTok{(geospiza$dat, geiger_nex)}
-\end{Highlighting}
-\end{Shaded}
-
-The data need not share the same taxa. We can append additional
-phylogenies and characters corresponding to different taxa:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{data}\NormalTok{(primates)}
-\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(primates$phy, geiger_nex)}
-\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_characters}\NormalTok{(primates$dat, geiger_nex)}
-\end{Highlighting}
-\end{Shaded}
-
-This allows us to include multiple trees and character matrices to a
-single NeXML file.
-
-\section{Metadata}\label{metadata}
-
-At the heart of the NeXML standard is the ability to encode arbitrary
-metadata in precisely specified, machine-readable formats that maintain
-backwards compatibility with any existing NeXML parser. NeXML uses the
-``Resource Description Framework in Annotations'', or RDFa, as described
-in Vos et al. (2012) or Prud'hommeaux (2014). These annotations can be
-used to add information to any particular element of the data, (a tree,
-node, edge, or taxonomic unit; a character matrix or specific trait, or
-even another metadata annotation or the NeXML data file as a whole).
-\texttt{RNeXML} provides a range of tools for accessing metadata
-embedded in a NeXML file.
-
-\subsection{Writing NeXML metadata}\label{writing-nexml-metadata}
-
-Researchers cannot take advantage of NeXML metadata manipulation if
-there is no metadata there. The burden of adding good metadata remains a
-significant barrier to increased sharing and reuse of scientific
-research (Tenopir \emph{et al.} 2011). \texttt{RNeXML} seeks to reduce
-the burden of adding high quality, machine-readable metadata to NeXML
-files. This occurs on several levels:
-
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Automatic annotation
-\item
-  Adding common metadata fields (\texttt{add\_basic\_metadata})
-\item
-  Taxonomic identifiers (\texttt{taxize\_nexml})
-\item
-  Generic metadata (\texttt{add\_meta})
-\item
-  Extending the NeXML standard using RDFa
-\end{itemize}
-
-\texttt{RNeXML} automatically adds basic metadata by default such as a
-time-stamp of when the tree was produced and an open data license (CC0,
-compatible with deposition on Dryad data repository), which can be
-configured as needed.
-
-To adjust these and other common metadata fields that should be part of
-any data deposition, \texttt{RNeXML} provides the
-\texttt{add\_basic\_meta()} function. This function does not assume any
-familiarity with the concepts of namespaces or other more advanced
-metadata tools discussed later. The \texttt{add\_basic\_meta} takes an
-existing \texttt{nexml} object (or will create a new one if none is
-provided, like the other \texttt{add\_} functions), and any of the
-following fields: \texttt{title}, \texttt{description},
-\texttt{creator}, \texttt{pubdate}, \texttt{rights}, \texttt{publisher},
-\texttt{citation}. Additional fields may be added to subsequent
-versions, see \texttt{?add\_basic\_metadata} for the most up-to-date
-information. Here we create a new \texttt{nexml} object with the basic
-metadata for the \texttt{bird.orders} data provided in the \texttt{ape}
-package:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{data}\NormalTok{(bird.orders)}
-\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(bird.orders)}
-\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(birds,}
-  \DataTypeTok{title =} \StringTok{"Phylogeny of the Orders of Birds From Sibley and Ahlquist"}\NormalTok{,}
-
-  \DataTypeTok{description =} \StringTok{"This data set describes the phylogenetic relationships of the}
-\StringTok{     orders of birds as reported by Sibley and Ahlquist (1990). Sibley}
-\StringTok{     and Ahlquist inferred this phylogeny from an extensive number of}
-\StringTok{     DNA/DNA hybridization experiments. The ``tapestry'' reported by}
-\StringTok{     these two authors (more than 1000 species out of the ca. 9000}
-\StringTok{     extant bird species) generated a lot of debates.}
-
-\StringTok{     The present tree is based on the relationships among orders. The}
-\StringTok{     branch lengths were calculated from the values of Delta T50H as}
-\StringTok{     found in Sibley and Ahlquist (1990, fig. 353)."}\NormalTok{,}
-
-  \DataTypeTok{citation =} \StringTok{"Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and}
-\StringTok{     classification of birds: a study in molecular evolution. New}
-\StringTok{     Haven: Yale University Press."}\NormalTok{,}
-
-  \DataTypeTok{creator =} \StringTok{"Sibley, C. G. and Ahlquist, J. E."}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Citations can also be added using R's \texttt{bibentry} type, which
-allows a user to add citations to R packages,
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{birds <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(birds, }\DataTypeTok{citation =} \KeywordTok{citation}\NormalTok{(}\StringTok{"ape"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-or to papers by simply using the object's DOI (using Boettiger (2014)).
-For instance to add the citation information of the paper that generated
-the \texttt{geospiza} phylogeny included in the \texttt{geiger} package:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"knitcitations"}\NormalTok{)}
-\NormalTok{geiger_nex <-}\StringTok{ }\KeywordTok{add_basic_meta}\NormalTok{(geiger_nex, }\DataTypeTok{citation =} \KeywordTok{bib_metadata}\NormalTok{(}\StringTok{"10.2307/2408428"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\subsection{Taxonomic identifiers}\label{taxonomic-identifiers}
-
-\texttt{RNeXML} provides various functions to assist in automatically
-adding other more specific metadata to various elements of the NeXML
-data. One common use case is that of taxonomic identifiers. Misspellings
-or inconsistent use of taxonomic names in phylogenetic data are a common
-source of frustration in reusing phylogenetic trees, especially when
-integrating data from different sources: trees might come from one place
-that uses one taxonomy, character data and occurrence data might come
-from other places that use different taxonomies. Hence, not just
-misspellings, but also synonyms, homonyms and incompatible taxon
-concepts might plague the end user. By associating the taxonomic labels
-against a specific, centralized database we can catch some of these
-inconsistencies and indicate more precisely which classification we are
-appealing to in the event of conflicting taxonomic definitions. In this
-example, we indicate that we intend the bird orders listed in the tree
-to correspond to the NCBI definitions.
-
-The \texttt{taxize\_nexml} function uses the R package \texttt{taxize}
-(Chamberlain \& Sz{ö}cs 2013) to check each taxon label against the NCBI
-database. If a unique match is found, a metadata annotation is added to
-the taxon providing the NCBI identification number to the taxonomic
-unit. If no match is found, the user is warned to check for possible
-typographic errors in the taxonomic labels provided.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{birds <-}\StringTok{ }\KeywordTok{taxize_nexml}\NormalTok{(birds, }\StringTok{"NCBI"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\subsection{Custom metadata
-extensions}\label{custom-metadata-extensions}
-
-So far our examples have relied on functions that have a built-in
-understanding of the kind of metadata being provided, and can thus
-generate the appropriate metadata with little further guidance. In order
-to be machine-readable and understandable, NeXML requires all metadata
-annotations follow the RDFa format (see Vos et al. (2012)) which
-explicitly declares the meaning of the annotation.
-
-Some basic definitions about the XML syntax used to express NeXML will
-be helpful for this discussion, though a more comprehensive introduction
-can be found in Vos et al. (2012). For instance, consider how the title
-we gave to the \texttt{bird.orders} phylogeny appears in the NeXML file
-using the XML format:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{<meta}\OtherTok{ id=}\StringTok{"m2"}\OtherTok{ property=}\StringTok{"dc:title"}\OtherTok{ datatype=}\StringTok{"xsd:string"}
-\OtherTok{      content=}\StringTok{"Phylogeny of the Orders of Birds From Sibley and Ahlquist"}\KeywordTok{/>} 
-\end{Highlighting}
-\end{Shaded}
-
-The format begins with an XML ``tag'' and the name ``meta'', indicated
-by the opening angle bracket; \texttt{\textless{}meta}. A closing tag
-\texttt{\textbackslash{}\textgreater{}} indicates the end of this XML
-``element'' . This tag contains several ``attributes'', named ``id'',
-``property'', etc., that an R user can think of as names in a
-\texttt{list}. What follows the equals sign after the attribute name is
-called the attribute ``value'', (e.g.~the \texttt{id} attribute has
-value ``m2'').
-
-The name of any element (like ``meta'') or attribute (like ``id'') in
-XML must be associated with a particular dictionary called a
-``namespace'' and specified using a ``prefix'' of an abbreviation
-followed by a colon appearing before the name. If no prefix is given, as
-is the case for all attributes here: \texttt{id}, \texttt{property},
-\texttt{datatype}, and \texttt{content}, then a default prefix specified
-at the beginning of the document applies, in this case, the NeXML
-namespace. This allows users to combine different XML that have
-different definitions of terms like \texttt{id} or \texttt{property}.
-
-The example element above also uses namespaces on some of the attribute
-values themselves. This is a consequence of RDFa annotation, and not
-strictly part of XML, though it works much the same way. While we hope
-the following examples will make these concepts clearer, users do not
-need to understand XML and namespaces to use most of the basic
-annotation features provided by \texttt{RNeXML}.
-
-So much for the structure; we can now briefly explain the meaning of
-this element. We see the title appears as the value of the
-\texttt{content} attribute of this \texttt{meta} element. The
-\texttt{id} attribute provides us with a unique way to refer to this
-meta element in this document. The \texttt{datatype} tells us that the
-content is a text string (as opposed to, say, a numerical or boolean
-value). The \texttt{property} is the attribute that most interests us
-here; it tells us that the content is a title. The prefix \texttt{dc}
-indicates the vocabulary (here identified as an XML namespace) to which
-the word \texttt{title} belongs. The prefix is defined elsewhere in the
-NeXML file, as we can see with:
-
-\begin{Shaded}
-\begin{Highlighting}[]
+The \texttt{nexml\_validate()} function performs this validation using
+the online NeXML validator (when a network connection is available),
+which performs additional checks not expressed in the NeXML schema
+itself (Vos \emph{et al.} 2012). If a network connection is not
+available, the function falls back on the schema validation method from
+the \texttt{XML} package (Lang 2013).
+
+\subsection{Creating and populating \texttt{nexml}
+objects}\label{creating-and-populating-nexml-objects}
+
+Instead of packaging the various components for a NeXML file at the time
+of writing the file, \texttt{RNeXML} also allows users to create and
+iteratively populate in-memory \texttt{nexml} objects. The methods to do
+this are \texttt{add\_characters()}, \texttt{add\_trees()}, and
+\texttt{add\_meta()}, for adding characters, trees, and metadata,
+respectively. Each of these functions will automatically create a new
+nexml object if not supplied with an existing one as the last (optional)
+argument.
+
+For example, here we use \texttt{add\_trees()} to first create a
+\texttt{nexml} object with the phylogeny data, and then add the
+character data to it:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nexObj <-}\StringTok{ }\KeywordTok{add_trees}\NormalTok{(geospiza$phy)}
+\NormalTok{nexObj <-}\StringTok{ }\KeywordTok{add_characters}\NormalTok{(geospiza$dat, nexObj)}
+\end{Highlighting}
+\end{Shaded}
+
+The data with which a \texttt{nexml} object is populated need not share
+the same OTUs. \texttt{RNeXML} automatically adds new, separate OTU
+blocks into the NeXML file for each data matrix and tree that uses a
+different set of OTUs.
+
+Other than storage size, there is no limit to the number of phylogenies
+and character matrices that can be included in a single NeXML document.
+This allows, for example, to capture samples from a posterior
+probability distribution of inferred or simulated phylogenies and
+character states in a single NeXML file.
+
+\subsection{Data documentation and annotation with built-in
+metadata}\label{data-documentation-and-annotation-with-built-in-metadata}
+
+NeXML allows attaching (``\emph{annotating}'') metadata to any data
+element, and even to metadata themselves. Whether at the level of the
+document as a whole or an individual data matrix or phylogeny, metadata
+can provide bibliographic and provenance information, for example about
+the study as part of which the phylogeny was generated or applied, which
+data matrix and which methods were used to generate it. Metadata can
+also be attached to very specific elements of the data, such as specific
+traits, individual OTUs, nodes, or even edges of the phylogeny.
+
+As described in Vos et al. (2012), to encode metadata annotations NeXML
+uses the ``Resource Description Framework in Annotations'' (RDFa)
+(Prud'hommeaux 2014). This standard provides for a strict
+machine-readable format yet enables future backwards compatibility with
+compliant NeXML parsers (and thus \texttt{RNeXML}), because the capacity
+of a tool to \emph{parse} annotations is not predicated on
+\emph{understanding} the meaning of annotations it has not seen before.
+
+To lower the barriers to sharing well-documented phylogenetic data,
+\texttt{RNeXML} aims to make recording useful and machine-readable
+metadata easier at several levels.
+
+First, when writing a NeXML file the package adds certain basic metadata
+automatically if they are absent, using default values consistent with
+recommended best practices (Cranston \emph{et al.} 2014). Currently,
+this includes naming the software generating the NeXML, a time-stamp of
+when a tree was produced, and an open data license. These are merely
+default arguments to \texttt{add\_basic\_meta()} and can be configured.
+
+Second, \texttt{RNeXML} provides a simple method, called
+\texttt{add\_basic\_metadata()}, to set metadata attributes commonly
+recommended for inclusion with data to be publicly archived or shared
+(Cranston \emph{et al.} 2014). The currently accepted parameters include
+\texttt{title}, \texttt{description}, \texttt{creator},
+\texttt{pubdate}, \texttt{rights}, \texttt{publisher}, and
+\texttt{citation}. Behind the scenes the method automatically anchors
+these attributes in common vocabularies (such as Dublin Core).
+
+Third, \texttt{RNeXML} integrates with the R package \texttt{taxize}
+(Chamberlain \& Sz{ö}cs 2013) to mitigate one of the most common
+obstacles to reuse of phylogenetic data, namely the misspellings and
+inconsistent taxonomic naming with which OTU labels are often fraught.
+The \texttt{taxize\_nexml()} method in \texttt{RNeXML} uses
+\texttt{taxize} to match OTU labels against the NCBI database, and,
+where a unique match is found, it annotates the respective OTU with the
+matching NCBI identifier.
+
+\subsection{Data annotation with custom
+metadata}\label{data-annotation-with-custom-metadata}
+
+The \texttt{RNeXML} interface described above for built-in metadata
+allows users to create precise and semantically rich annotations without
+confronting any of the complexity of namespaces and ontologies.
+Nevertheless, advanced users may desire the explicit control over these
+semantic tools that takes full advantage of the flexibility and
+extensibility of the NeXML specification (Parr \emph{et al.} 2011; Vos
+\emph{et al.} 2012). In this section we detail how to accomplish these
+more complex uses in RNeXML.
+
+Using a vocabulary or ontology terms rather than simple text strings to
+describe data is crucial for allowing machines to not only parse but
+also interpret and potentially reason over their semantics. To achieve
+this benefit for custom metadata extensions, the user necessarily needs
+to handle certain technical details from which the \texttt{RNeXML}
+interface shields her otherwise, in particular the globally unique
+identifiers (normally HTTP URIs) of metadata terms and vocabularies. To
+be consistent with XML terminology, \texttt{RNeXML} calls vocabulary
+URIs \emph{namespaces}, and their abbreviations \emph{prefixes}. For
+example, the namespace for the Dublin Core Metadata Terms vocabulary is
+``\url{http://purl.org/dc/elements/1.1/}''. Using its common
+abbreviation ``dc'', a metadata property ``dc:title'' expands to the
+identifier ``\url{http://purl.org/dc/elements/1.1/title}''. This URI
+resolves to a human and machine-readable (depending on access)
+definition of precisely what the term \texttt{title} in Dublin Core
+means. In contrast, just using the text string ``title'' could also mean
+the title of a person, a legal title, the verb title, etc. URI
+identifiers of metadata vocabularies and terms are not mandated to
+resolve, but if machines are to derive the maximum benefit from them,
+they should resolve to a definition of their semantics in RDF.
+
+\texttt{RNeXML} includes methods to obtain and manipulate metadata
+properties, values, identifiers, and namespaces. The
+\texttt{get\_namespaces()} method accepts a \texttt{nexml} object and
+returns a named list of namespace prefixes and their corresponding
+identifiers known to the object:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{birds <-}\StringTok{ }\KeywordTok{nexml_read}\NormalTok{(}\StringTok{"birds.xml"}\NormalTok{)}
 \NormalTok{prefixes <-}\StringTok{ }\KeywordTok{get_namespaces}\NormalTok{(birds)}
@@
 
-This URL links us to a web page providing both a human and
-machine-readable version of precisely what the term \texttt{title}
-means, showing us that it refers to \texttt{title} in the sense of a
-creative work, rather than the title of a person, a legal title, the
-verb title, etc. This precision can allow us to perform precise computer
-reasoning on metadata, without relying on heuristics or assumptions, as
-we illustrate later. For the moment though, the need to refer to an
-explicit controlled vocabulary (or ontology) with a url, saying
-``dc:title'' rather than just saying the property is ``title'', is
-something of a burden in creating metadata. The previous examples
-involve functions that already know what kind of metadata is being
-provided (a title, a data creator, etc), and so can provide the
-appropriate vocabulary automatically. The main advantage of linking to
-an external vocabulary, however, is the ability to express concepts that
-are not already ``built in'' to the data model. \texttt{RNeXML} provides
-tools to generate arbitrary metadata annotations given an external
-vocabulary, such as we illustrate here.
-
-From the list of namespaces in \texttt{prefixes}, we can follow links a
-handful of established vocabularies and find terms that are not included
-in \texttt{add\_basic\_meta}. For instance, by resolving the location
-identified by the \texttt{prism} prefix,
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{prefixes[}\StringTok{"prism"}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-                                           prism 
-"http://prismstandard.org/namespaces/1.2/basic/" 
-\end{verbatim}
-
-we will find that there is a term for \texttt{modificationDate} for a
-file. We create a \texttt{meta} element containing this annotation using
-the \texttt{meta} function:
+The \texttt{get\_metadata()} method returns, as a named list, the
+metadata annotations for a given \texttt{nexml} object at a given level,
+with the whole NeXML document being the default level (\texttt{"all"}
+extracts all metadata objects):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds) }
+\NormalTok{otu_meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds, }\DataTypeTok{level=}\StringTok{"otu"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The returned list does not include the data elements to which the
+metadata are attached. Therefore, a different approach, documented in
+the metadata vignette, is recommended for accessing the metadata
+attached to data elements.
+
+The \texttt{meta()} method creates a new metadata object from a property
+name and content (value). For example, the following creates a
+modification date metadata object, using a property in the PRISM
+vocabulary:
 
@@
 
-We can add this annotation to our existing \texttt{birds} NeXML file
-using the \texttt{add\_meta} function. Because we do not specify a
-level, it is added to the root node, referring to the NeXML file as a
-whole.
+Metadata annotations in \texttt{NeXML} can be nested within another
+annotation, which the \texttt{meta()} method accommodates by accepting a
+parameter \texttt{children}, with the list of nested metadata objects
+(which can themselves be nested) as value.
+
+The \texttt{add\_meta()} function adds metadata objects as annotations
+to a \texttt{nexml} object at a specified level, with the default level
+being the NeXML document as a whole:
 
@@
 
-The built-in vocabularies are just the tip of the iceberg of established
-vocabularies. Here we add an annotation from the \texttt{skos} namespace
-which describes the history of where the data comes from:
+If the prefix used by the metadata property is not among the built-in
+ones (which can be obtained using \texttt{get\_namespaces()}), it has to
+be provided along with its URI as the \texttt{namespaces} parameter. For
+example, the following uses the
+``\href{http://www.w3.org/TR/skos-reference/}{Simple Knowledge
+Organization System}'' (SKOS) vocabulary to add a note to the trees in
+the \texttt{nexml} object:
 
@@
   \DataTypeTok{content =} \StringTok{"Mapped from the bird.orders data in the ape package using RNeXML"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Because \texttt{skos} is not in the current namespace list, we add it
-with a url when adding this meta element. We also specify that this
-annotation be placed at the level of the \texttt{trees} sub-node in the
-NeXML file.
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \NormalTok{birds <-}\StringTok{ }\KeywordTok{add_meta}\NormalTok{(history, }
@@
 
-For finer control of the level at which a \texttt{meta} element is
-added, we will manipulate the \texttt{nexml} R object directly using S4
-sub-setting, as shown in the supplement.
-
-Much richer metadata annotation is possible. Later we illustrate how
-metadata annotation can be used to extend the base NeXML format to
-represent new forms of data while maintaining compatibility with any
-NeXML parser. The \texttt{RNeXML} package can be easily extended to
-support helper functions such as \texttt{taxize\_nexml} to add
-additional metadata without imposing a large burden on the user.
-
-\subsection{Reading NeXML metadata}\label{reading-nexml-metadata}
-
-A call to the \texttt{nexml} object prints some metadata summarizing the
-data structure:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{birds}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-A nexml object representing:
-     1 phylogenetic tree blocks, where: 
-     block 1 contains 1 phylogenetic trees 
-     44 meta elements 
-     0 character matrices 
-     23 taxonomic units 
- Taxa:   Struthioniformes, Tinamiformes, Craciformes, Galliformes, Anseriformes, Turniciformes ...
-
- NeXML generated by RNeXML using schema version: 0.9 
- size: 370.7 Kb 
-\end{verbatim}
-
-We can extract all metadata pertaining to the NeXML document as a whole
-(annotations of the XML root node,
-\texttt{\textless{}nexml\textgreater{}}) with the command
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds) }
-\end{Highlighting}
-\end{Shaded}
-
-This returns a named list of available metadata. We can see the kinds of
-metadata recorded from the names (showing the first 4):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{names}\NormalTok{(meta)[}\DecValTok{1}\NormalTok{:}\DecValTok{4}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-[1] "dc:title"                      "dc:creator"                   
-[3] "dc:description"                "dcterms:bibliographicCitation"
-\end{verbatim}
-
-and can ask for a particular element using the standard list sub-setting
-mechanism (i.e.~either the name of an element or its numeric position),
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{meta[[}\StringTok{"dc:title"}\NormalTok{]]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-[1] "Phylogeny of the Orders of Birds From Sibley and Ahlquist"
-\end{verbatim}
-
-All metadata terms must belong to an explicit \emph{namespace} or
-vocabulary that allows a computer to interpret the term precisely. The
-prefix (before the \texttt{:}) indicates to which vocabulary the term
-belongs, e.g. \texttt{dc} in this case. The \texttt{get\_namespaces}
-function tells us the definition of the vocabulary using a link:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{prefixes <-}\StringTok{ }\KeywordTok{get_namespaces}\NormalTok{(birds)}
-\NormalTok{prefixes[}\StringTok{"dc"}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-                                dc 
-"http://purl.org/dc/elements/1.1/" 
-\end{verbatim}
-
-Common metadata can be accessed with a few dedicated functions:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{get_citation}\NormalTok{(birds)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
-     classification of birds: a study in molecular evolution. New
-     Haven: Yale University Press. Paradis E, Claude J and Strimmer K (2004). "APE: analyses of
-phylogenetics and evolution in R language." _Bioinformatics_,
-*20*, pp. 289-290.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{get_taxa}\NormalTok{(birds)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
- [1] "Struthioniformes" "Tinamiformes"     "Craciformes"     
- [4] "Galliformes"      "Anseriformes"     "Turniciformes"   
- [7] "Piciformes"       "Galbuliformes"    "Bucerotiformes"  
-[10] "Upupiformes"      "Trogoniformes"    "Coraciiformes"   
-[13] "Coliiformes"      "Cuculiformes"     "Psittaciformes"  
-[16] "Apodiformes"      "Trochiliformes"   "Musophagiformes" 
-[19] "Strigiformes"     "Columbiformes"    "Gruiformes"      
-[22] "Ciconiiformes"    "Passeriformes"   
-\end{verbatim}
-
-Which returns text from the otu element labels, typically used to define
-taxonomic names, rather than text from explicit meta elements.
-
-We can also access metadata at a specific level (or use
-\texttt{level=all} to extract all meta elements in a list). Here we show
-only the first few results:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{otu_meta <-}\StringTok{ }\KeywordTok{get_metadata}\NormalTok{(birds, }\DataTypeTok{level=}\StringTok{"otu"}\NormalTok{)}
-\NormalTok{otu_meta[}\DecValTok{1}\NormalTok{:}\DecValTok{4}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-$`tc:toTaxon`
-[1] "http://ncbi.nlm.nih.gov/taxonomy/8798"
-
-$`tc:toTaxon`
-[1] "http://ncbi.nlm.nih.gov/taxonomy/8802"
-
-$`tc:toTaxon`
-[1] "http://ncbi.nlm.nih.gov/taxonomy/8976"
-
-$`tc:toTaxon`
-[1] "http://ncbi.nlm.nih.gov/taxonomy/8976"
-\end{verbatim}
-
-This approach does not preserve the relationship between the metadata
-element and the \texttt{otu} it was annotating, and as such is most
-useful for quick visual exploration. Metadata values for specific
-elements or sets of elements can always be accessed using the S4
-sub-setting mechanisms (see section below).
-
-\subsection{RDF manipulation and SPARQL
-queries}\label{rdf-manipulation-and-sparql-queries}
-
-The most powerful way to take advantage of the machine-readable nature
-of the metadata involves the construction of SPARQL queries. This
-approach can exploit the reasoning and logical deduction capacity of
-formal vocabularies (ontologies). For a proper introduction to SPARQL
-queries, see Prud'hommeaux (2014) or Hartig (2012)).
-
-The real expressiveness of SPARQL is better illustrated through more
-complex examples such as the one shown in the supplementary materials
-(which can also be run from R using the command:
-\texttt{demo("sparql", "RNeXML")}). This demonstration constructs a
-cladogram based on the taxonomic ranks provided in the NeXML. SPARQL
-allows us to make explicit use of ontological concepts such as
-manipulating sub-classes (i.e.~that ``Genus'' is a subclass of
-``Family''), or joining multiple predicates (i.e.~retrieve the NCBI
-identifier of the OTU that has taxon rank of ``Order'').
-
-\subsection{Extending the NeXML standard through metadata
-annotation.}\label{extending-the-nexml-standard-through-metadata-annotation.}
-
-No data standard can anticipate all the uses and extensions researchers
-may one day need. Researchers have repeatedly adapted the existing NEXUS
-file format to accommodate encoding new kinds of information, but in the
-process break compatibility between different flavors of NEXUS files,
-with few reliable methods to distinguish between these flavors. NeXML
-provides a solution to this problem through the use of metadata
-annotation.
-
-Here we illustrate this process using the example of stochastic
-character mapping (Huelsenbeck \emph{et al.} 2003). A stochastic
-character map is simply an annotation of the branches on a phylogeny,
-assigning each section of each branch to a particular ``state''
-(typically of a morphological characteristic).
-
-Bollback (2006) provides a widely used stand-alone software
-implementation of this method in the software \texttt{simmap}, which
-modified the standard Newick tree format to express this additional
-information. This can break compatibility with other software,\footnote{By
-  using the commenting mechanism of the Newick format, it is possible
-  that other software that doesn't also use the comment mechanism for
-  some other such purpose would be able to successfully parse the tree.
-  However there is no way to guarantee that this is the case or for the
-  data format to describe its use.} and creates a format that cannot be
-interpreted without additional information describing this convention.
-By contrast, the NeXML extension is not only backwards compatible but
-contains a precise and machine-readable description of what it is
-encoding.
-
-In this example, we illustrate how the additional information required
-to define a stochastic character mapping (a \texttt{simmap} mapping) in
-NeXML.
-
-Revell (2012) describes the \texttt{phytools} package for R, which
-includes utilities for reading, manipulating, and writing
-\texttt{simmap} files in R. In this example, we also show how to define
-\texttt{RNeXML} functions that map the R representation used by Revell
-(an extension of the \texttt{ape} class) into the NeXML extension we
-have defined by using \texttt{RNeXML} functions.
-
-Since a stochastic character map simply assigns different states to
-parts of a branch (or edge) on the phylogenetic tree, we can create a
-NeXML representation by annotating the \texttt{edge} elements with
-appropriate \texttt{meta} elements. These elements need to describe the
-character state being assigned and the duration (in terms of
-branch-length) that the edge spends in that state (Stochastic character
-maps are specific to time-calibrated or ultrametric trees).
-
-NeXML already defines the \texttt{characters} element to handle discrete
-character traits (\texttt{nex:char}) and the states they can assume
-(\texttt{nex:state}). We will thus reuse the \texttt{characters} element
-for this purpose, referring to both the character trait and the states
-by the ids assigned to them in that element. (NeXML's convention of
-referring to everything by id permits a single canonical definition of
-each term, making it clear where additional annotation belongs). For
-each edge, we need to indicate:
-
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  That our annotation contains a stochastic character mapping
-  reconstruction
-\item
-  Since many reconstructions are possible for a single edge, we give
-  each reconstruction an id
-\item
-  We indicate for which character trait we are defining the
-  reconstruction
-\item
-  We then indicate which states the character assumes on that edge. For
-  each state realized on the edge, that involves stating:
-
-  \begin{itemize}
-  \itemsep1pt\parskip0pt\parsep0pt
-  \item
-    the state assignment
-  \item
-    the duration (length of time) for which the edge spends in the given
-    state
-  \item
-    the order in which the state changes happen (Though we could just
-    assume state transitions are listed chronologically, NeXML suggests
-    making all data explicit, rather than relying on the structure of
-    the data file to convey information).
-  \end{itemize}
-\end{itemize}
-
-Thus the annotation for an edge that switches from state \texttt{s2} to
-state \texttt{s1} of character \texttt{cr1} would be constructed like
-this:
-
-\begin{Shaded}
-\begin{Highlighting}[]
- \NormalTok{m <-}\StringTok{ }\KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:reconstructions"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
-        \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:reconstruction"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
-
-          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:char"}\NormalTok{, }\StringTok{"cr1"}\NormalTok{),}
-          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:stateChange"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:order"}\NormalTok{, }\DecValTok{1}\NormalTok{),}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:length"}\NormalTok{, }\StringTok{"0.2030"}\NormalTok{),}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:state"}\NormalTok{, }\StringTok{"s2"}\NormalTok{))),}
-
-          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:char"}\NormalTok{, }\StringTok{"cr1"}\NormalTok{),}
-          \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:stateChange"}\NormalTok{, }\DataTypeTok{children =} \KeywordTok{c}\NormalTok{(}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:order"}\NormalTok{, }\DecValTok{2}\NormalTok{),}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:length"}\NormalTok{, }\StringTok{"0.0022"}\NormalTok{),}
-            \KeywordTok{meta}\NormalTok{(}\StringTok{"simmap:state"}\NormalTok{, }\StringTok{"s1"}\NormalTok{)))}
-          \NormalTok{))))}
-\end{Highlighting}
-\end{Shaded}
-
-Of course writing out such a definition manually becomes tedious
-quickly. Because these are just R commands, we can easily define a
-function that can loop over an assignment like this for each edge,
-extracting the appropriate order, length and state from an existing R
-object such as that provided in the \texttt{phytools}
-package.\\Likewise, it is straightforward to define a function that
-reads this data using the \texttt{RNeXML} utilities and converts it back
-to the \texttt{phytools} package. The full implementation of this
-mapping can be seen in the \texttt{simmap\_to\_nexml()} and the
-\texttt{nexml\_to\_simmap()} functions provided in the \texttt{RNeXML}
-package.
-
-As the code indicates, the key step is simply to define the data in meta
-elements. In so doing, we have defined a custom namespace,
-\texttt{simmap}, to hold our variables. This allows us to provide a URL
-with more detailed descriptions of what each of these elements mean:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nex <-}\StringTok{ }\KeywordTok{add_namespaces}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DataTypeTok{simmap =} \StringTok{"https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-At that URL we have posted a simple description of each term.
-
-Using this convention we can generate NeXML files containing
-\texttt{simmap} data, read those files into R, and convert them back
-into the \texttt{phytools} package format. These simple functions serve
-as further illustration of how \texttt{RNeXML} can be used to extend the
-NeXML standard. We illustrate their use briefly here, starting with
-loading a \texttt{nexml} object containing a \texttt{simmap}
-reconstruction into R:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{data}\NormalTok{(simmap_ex)}
-\end{Highlighting}
-\end{Shaded}
-
-The \texttt{get\_trees} function can be used to return an
-\texttt{ape::phylo} tree as usual. \texttt{RNeXML} automatically detects
-the \texttt{simmap} reconstruction data and returns includes this in a
-\texttt{maps} element of the \texttt{ape::phylo} object, for use with
-other \texttt{phytools} functions.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{phy <-}\StringTok{ }\KeywordTok{nexml_to_simmap}\NormalTok{(simmap_ex)}
-\end{Highlighting}
-\end{Shaded}
-
-We can then use various functions from \texttt{phytools} designed for
-\texttt{simmap} objects (Revell 2012), such as the plotting function:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"phytools"}\NormalTok{)}
-\KeywordTok{plotSimmap}\NormalTok{(phy)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-no colors provided. using the following legend:
-       A        B        C 
- "black"    "red" "green3" 
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-Figure1.pdf}
-\caption{Stochastic character mapping on a phylogeny, as generated by
-the phytools package after parsing the simmap-extended NeXML.}
-\end{figure}
-
-Likewise, we can convert the object back in the NeXML format and write
-it out to file to be read by other users.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nex <-}\StringTok{ }\KeywordTok{simmap_to_nexml}\NormalTok{(phy) }
-\KeywordTok{nexml_write}\NormalTok{(nex, }\StringTok{"simmap.xml"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-[1] "simmap.xml"
-\end{verbatim}
-
-Though other NeXML parsers (for instance, for Perl or Python) have not
-been written explicitly to express \texttt{simmap} data, those parsers
-will nonetheless be able to successfully parse this file and expose the
-\texttt{simmap} data to the user.
+Alternatively, additional namespaces can also be added in batch using
+the \texttt{add\_namespaces()} method.
+
+By virtue of subsetting the S4 \texttt{nexml} object, \texttt{RNeXML}
+also offers fine control of where a \texttt{meta} element is added, for
+which the package vignette on S4 subsetting of \texttt{nexml} contains
+examples.
+
+Because NeXML expresses all metadata using the RDF standard, and stores
+them compliant with RDFa, they can be extracted as an RDF graph,
+queried, analyzed, and mashed up with other RDF data, local or on the
+web, using a wealth of off-the-shelf tools for working with RDF (see
+Prud'hommeaux (2014) or Hartig (2012)). Examples for these possibilities
+are included in the \texttt{RNeXML} SPARQL vignette (a recursive acronym
+for SPARQL Protocol and RDF Query Language, see
+\url{http://www.w3.org/TR/rdf-sparql-query/}), and the package also
+comes with a demonstration that can be run from R using the following
+command: \texttt{demo("sparql", "RNeXML")}).
+
+\subsection{Using metadata to extend the NeXML
+standard}\label{using-metadata-to-extend-the-nexml-standard}
+
+NeXML was designed to prevent the need for future non-interoperable
+``flavors'' of the standard in response to new research directions. Its
+solution to this inevitable problem is a highly flexible metadata system
+without sacrificing strict validation of syntax and structure.
+
+Here we illustrate how \texttt{RNeXML}'s interface to NeXML's metadata
+system can be used to record and share a type of phylogenetic data not
+taken into account when NeXML was designed, in this case stochastic
+character maps (Huelsenbeck \emph{et al.} 2003). Such data assign
+certain parts (corresponding to time) of each branch in a
+time-calibrated phylogeny to a particular ``state'' (typically of a
+morphological characteristic). The current de-facto format for sharing
+stochastic character maps, created by \texttt{simmap} (Bollback 2006), a
+widely used tool for creating such maps, is a non-interoperable
+modification of the standard Newick tree format. This means that
+computer programs designed to read Newick or NEXUS formats may fail when
+trying to read in a phylogeny that includes \texttt{simmap} annotations.
+
+In contrast, by allowing new data types to be added as --- sometimes
+complex --- metadata annotations NeXML can accommodate data extensions
+without compromise to its grammar and thus syntax In NeXML. To
+illustrate how RNeXML facilitates extending the NeXML standard in this
+way, we have implemented two functions in the package,
+\texttt{nexml\_to\_simmap} and \texttt{simmap\_to\_nexml}. These
+functions show how simmap data can be represented as \texttt{meta}
+annotations on the branch length elements of a NeXML tree, and provide
+routines to convert between this NeXML representation and the extended
+\texttt{ape::phylo} representation of a \texttt{simmap} tree in R that
+was introduced by Revell (2012). We encourage readers interested in this
+capability to consult the example code in \texttt{simmap\_to\_nexml} to
+see how this is implemented.
+
+Extensions to NeXML must also be defined in the file's namespace in
+order to valid. This provides a way to ensure that a URI providing
+documentation of the extension is always included. Our examples here use
+the prefix, \texttt{simmap}, to group the newly introduced metadata
+properties in a vocabulary, for which the \texttt{add\_namespace()}
+method can be used to give a URI as an identifier:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nex <-}\StringTok{ }\KeywordTok{add_namespaces}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DataTypeTok{simmap =} 
+  \StringTok{"https://github.com/ropensci/RNeXML/tree/master/inst/simmap.md"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Here the URI does not resolve to a fully machine-readable definition of
+the terms and their semantics, but it can nonetheless be used to provide
+at least a human-readable informal definition of the terms.
 
@@
 
-Data archiving is increasingly required by journals in evolutionary
-biology and biodiversity (e.g. Rausher et al. (2010)). The burden of
-preparing and submitting properly annotated data to archives continues
-to be a significant barrier (Tenopir \emph{et al.} 2011; Stodden 2014),
-and many phylogenetic trees upon which studies are based are
-inaccessible or lost to the research community (Drew \emph{et al.}
-2013). \texttt{RNeXML} seeks to lower these barriers by providing
-functions to immediately archive NeXML files in scientific repositories.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{doi <-}\StringTok{ }\KeywordTok{nexml_publish}\NormalTok{(birds, }\DataTypeTok{visibility =} \StringTok{"public"}\NormalTok{, }\DataTypeTok{repository=}\StringTok{"figshare"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Robust data archiving need not involve immediate release of the data.
-The \texttt{figshare} repository supports the option of
-\texttt{visibility = "private"}, allowing secure publication that
-provides a way to backup and share data with collaborators prior to
-public release.
-
-\section{Software reuse and support}\label{software-reuse-and-support}
-
-The \texttt{RNeXML} package is available through the CRAN repository
-under a CC0 license. At the time of writing, the current version is
-1.1.3. The current development branch can be found on Github at
-{[}\url{https://github.com/ropensci/RNeXML}{]}. See the NEWS file for
-changes in recent versions. Users are encouraged to submit bug reports
-or feature requests in the issues log. Contact
-\texttt{ropensci-discuss@googlegroups.com} or the package maintainer for
-help, or email the phylogenetics R users group list at
-\texttt{r-sig-phylo@r-project.org}.
-
-There are many promising directions for further development of the
-\texttt{RNeXML} package. The project's
-\href{https://github.com/ropensci/RNeXML/issues/milestones}{Milestones
-page} page provides a list of steps currently planned or under
-consideration. In particular, these milestones include increased support
-for interactive mode of entering data (wizards), conversions to and from
-other existing R phylogenetics formats such those found in the
-\texttt{phylobase} package (al. 2014), and support for integration of
-semantics defined through the Web Ontology language (OWL), as discussed
-in the Challenges section.
+Data archiving is increasingly required by scientific journals,
+including in evolutionary biology, ecology, and biodiversity (e.g.
+Rausher et al. (2010)). The effort involved with preparing and
+submitting properly annotated data to archives remains a notable barrier
+to the broad adoption of data archiving and sharing as a normal part of
+the scholarly publication workflow (Tenopir \emph{et al.} 2011; Stodden
+2014). In particular, the majority of phylogenetic trees published in
+the scholarly record are inaccessible or lost to the research community
+(Drew \emph{et al.} 2013).
+
+One of \texttt{RNeXML}'s aims is to promote the archival of
+well-documented phylogenetic data in scientific data repositories, in
+the form of NeXML files. To this end, the method
+\texttt{nexml\_publish()} provides an API directly from within R that
+allows data archival to become a step programmed into data management
+scripts. Initially, the method supports the data repository Figshare
+(\url{http://figshare.com}):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{doi <-}\StringTok{ }\KeywordTok{nexml_publish}\NormalTok{(birds, }\DataTypeTok{repository=}\StringTok{"figshare"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+This method reserves a permanent identifier (DOI) on the figshare
+repository that can later be made public through the figshare web
+interface. This also acts as a secure backup of the data to a repository
+and a way to share with collaborators prior to public release.
+
+\section{Conclusions and future
+directions}\label{conclusions-and-future-directions}
+
+\texttt{RNeXML} allows R's ecosystem to read and write data in the NeXML
+format through an interface that is no more involved than reading or
+writing data from other phylogenetic data formats. It also carries
+immediate benefits for its users compared to other formats. For example,
+comparative analysis R packages and users frequently add their own
+metadata annotations to the phylogenies they work with, such as
+annotations of species, stochastic character maps, trait values, model
+estimates and parameter values. \texttt{RNeXML} affords R the capability
+to harness machine-readable semantics and an extensible metadata schema
+to capture, preserve, and share these and other kinds of information,
+all through an API instead of having to understand in detail the schema
+underlying the NeXML standard. To assist users in meeting the rising bar
+for best practices in data sharing in phylogenetic research (Cranston
+\emph{et al.} 2014), \texttt{RNeXML} captures metadata information from
+the R environment to the extent possible, and applies reasonable
+defaults.
+
+The goals for continued development of \texttt{RNeXML} revolve primarily
+around better interoperability with other existing phylogenetic data
+representations in R, such as those found in the \texttt{phylobase}
+package (NESCENT R Hackathon Team 2014); and better integration of the
+rich metadata semantics found in ontologies defined in the Web Ontology
+Language (OWL), including programmatic access to machine reasoning with
+such metadata.
 
@@
 
-.
-
-This project was supported in part by a grant from the Alfred P Sloan
-Foundation Grant 2013-6-22 (CB and SC), NSF Grant DBI-1306697 (CB),
-\ldots{} We thank F. Michonneau for helpful comments on an earlier
-version of this manuscript. \texttt{RNeXML} is part of the rOpenSci
-project, {[}\url{http://ropensci.org}{]}.
-
-al., R.H. et. (2014). \emph{phylobase: Base package for phylogenetic
-structures and comparative data}. Retrieved from
-\url{http://CRAN.R-project.org/package=phylobase}
-
-Boettiger, C. (2014). \emph{knitcitations: Citations for knitr markdown
-files}. Retrieved from \url{https://github.com/cboettig/knitcitations}
-
-Boettiger, C. \& {Temple Lang}, D. (2012). Treebase: an r package for
-discovery, access and manipulation of online phylogenies (L. Harmon,
-Ed.). \emph{Methods Ecol Evol}, \textbf{3}, 1060--1066. Retrieved from
-\url{http://dx.doi.org/10.1111/j.2041-210x.2012.00247.x}
+This project was supported in part by the National Evolutionary
+Synthesis Center (NESCent) (NSF \#EF-0905606), and grants from the
+National Science Foundation (DBI-1306697) and the Alfred P Sloan
+Foundation (Grant 2013-6-22). \texttt{RNeXML} started as a project idea
+for the Google Summer of Code(TM), and we thank Kseniia Shumelchyk for
+taking the first steps to implement it. We are grateful to F. Michonneau
+for helpful comments on an earlier version of this manuscript.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 
@@
 
-Chamberlain, S.A. \& Sz{ö}cs, E. (2013). taxize: taxonomic search and
+Chamberlain, S.A. \& Sz{ö}cs, E. (2013). Taxize: Taxonomic search and
 retrieval in r. \emph{F1000Research}. Retrieved from
 \url{http://dx.doi.org/10.12688/f1000research.2-191.v2}
-
-CRAN task view: Phylogenetics, especially comparative methods. (2014).
-Retrieved from
-\url{http://cran.r-project.org/web/views/Phylogenetics.html}
 
@@
 
-Harmon, L., Weir, J., Brock, C., Glor, R. \& Challenger, W. (2008).
-GEIGER: investigating evolutionary radiations. \emph{Bioinformatics},
-\textbf{24}, 129--131.
-
 Hartig, O. (2012). An introduction to sPARQL and queries over linked
@@
 
-Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: analyses of
+Lang, D.T. (2013). \emph{XML: Tools for parsing and generating xML
+within r and s-plus.} Retrieved from
+\url{http://CRAN.R-project.org/package=XML}
+
+Maddison, D., Swofford, D. \& Maddison, W. (1997). NEXUS: An extensible
+file format for systematic information. \emph{Syst. Biol.}, \textbf{46},
+590--621. Retrieved from
+\url{http://www.ncbi.nlm.nih.gov/pubmed/11975335}
+
+Midford, P., Dececchi, T., Balhoff, J., Dahdul, W., Ibrahim, N., Lapp,
+H., Lundberg, J., Mabee, P., Sereno, P., Westerfield, M., Vision, T. \&
+Blackburn, D. (2013). The vertebrate taxonomy ontology: A framework for
+reasoning across model organism and species phenotypes. \emph{J. Biomed.
+Semantics}, \textbf{4}, 34. Retrieved from
+\url{http://dx.doi.org/10.1186/2041-1480-4-34}
+
+NESCENT R Hackathon Team. (2014). \emph{Phylobase: Base package for
+phylogenetic structures and comparative data}. Retrieved from
+\url{http://CRAN.R-project.org/package=phylobase}
+
+O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
+comparative methods. Retrieved from
+\url{http://cran.r-project.org/web/views/Phylogenetics.html}
+
+Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: Analyses of
 phylogenetics and evolution in R language. \emph{Bioinformatics},
 \textbf{20}, 289--290.
+
+Parr, C.S., Guralnick, R., Cellinese, N. \& Page, R.D.M. (2011).
+Evolutionary informatics: unifying knowledge about the diversity of
+life. \emph{Trends in ecology \& evolution}, \textbf{27}, 94--103.
+Retrieved from \url{http://www.ncbi.nlm.nih.gov/pubmed/22154516}
+
+Pennell, M.W., Eastman, J.M., Slater, G.J., Brown, J.W., Uyeda, J.C.,
+Fitzjohn, R.G., Alfaro, M.E. \& Harmon, L.J. (2014). Geiger v2.0: An
+expanded suite of methods for fitting macroevolutionary models to
+phylogenetic trees. \emph{Bioinformatics}, \textbf{30}, 2216--2218.
+
+Piel, W.H., Chan, L., Dominus, M.J., Ruan, J., Vos, R.A. \& Tannen, V.
+(2009). TreeBASE v. 2: A database of phylogenetic knowledge. Retrieved
+from \url{http://www.e-biosphere09.org}
+
+Piel, W.H., Donoghue, M.J. \& Sanderson, M.J. (2002). TreeBASE: A
+database of phylogenetic information. \emph{The interoperable `catalog
+of life'} (eds J. Shimura, K.L. Wilson \& D. Gordon), pp. 41--47.
+Research report. National Institute for Environmental Studies, Tsukuba,
+Japan. Retrieved from
+\url{http://donoghuelab.yale.edu/sites/default/files/124_piel_shimura02.pdf}
 
@@
 
-Revell, L.J. (2012). phytools: An r package for phylogenetic comparative
+Revell, L.J. (2012). Phytools: An r package for phylogenetic comparative
 biology (and other things). \emph{Methods in Ecology and Evolution},

### 2015-05-25 manuscripts/submit.tex
tex intermediate files for records

--- 2015-05-25 manuscripts/manuscript.tex
@@
 95064, USA}    
+    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, Durham,
+NC, USA and National Evolutionary Synthesis Center and Duke University,
+Durham, NC, USA}    
     \address[ropensci]{University of California, Berkeley, CA, USA}    
     \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}
-    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, and
-National Evolutionary Synthesis Center, Durham, NC, USA}
 

### 2015-05-29 README.Rmd
adds Data Accessibility section as requested before going to proofs

--- 2014-05-02 manuscripts/tutorial.Rmd
@@
 
+[![DOI](https://zenodo.org/badge/doi/10.5281/zenodo.13131.svg)](http://dx.doi.org/10.5281/zenodo.13131)
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
@@
 
-An R package for reading, writing, integrating and publishing data using the Ecological Metadata Language (EML) format.
-
-* **Note:** This package is still in active development and not yet submitted to CRAN.  Functions and documentation may be incomplete and subject to change.  
 * Maintainer: Carl Boettiger
@@
 
+```{r compile-settings, include=FALSE}
+## Set the paths for cache and figure
+library(methods)
+library(knitr)
+opts_chunk$set(cache = 1)
+basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
+opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
+               cache.path = paste("cache/", basename, "/", sep=""))
+
 ```{r echo=FALSE}
@@
 opts_knit$set(upload.fun = imgur_upload)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, comment = NA, verbose = TRUE)
 ```
@@
 
-Once we have created the `meta` elements, we can pass them to our `nexml_write` function, along with definitions of the namespaces.
+For this kind of richer annotation, it is best to build up our NeXML object sequentially. Frist
+we will add `bird.orders` phylogeny to a new phylogenetic object, and then we will add the metadata
+elements created above to this object. Finally, we will write the object out as an XML file:
 
 ```{r}
-  nexml_write(bird.orders, 
-              file = "example.xml", 
-              meta = list(history, modified, website), 
+  birds <- add_trees(bird.orders)
+  birds <- add_meta(meta = list(history, modified, website),
                     namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
-                             foaf = "http://xmlns.com/foaf/0.1/"))
+                                   foaf = "http://xmlns.com/foaf/0.1/"),
+                    nexml=birds)
+  nexml_write(birds, 
+              file = "example.xml")
 
@@
 Add taxonomic identifier metadata to the OTU elements:
+<!-- This block relies on a robust internet connection that can occassionally fail.  Also it's a bit slow, so don't run it. After all, this command is tested in the unit tests.-->
 
-```{r}
+```{r eval=FALSE}
 nex <- add_trees(bird.orders)
@@
 library(geiger)
-fitContinuous(tree, traits[1])
-fitDiscrete(tree, traits[2])
+fitContinuous(tree, traits[1], ncores=1)
+fitDiscrete(tree, traits[2], ncores=1)
 ```
@@
 ```
+
+-----
+
+[![ropensci footer](http://ropensci.org/public_images/github_footer.png)](http://ropensci.org)

### 2015-05-29 manuscripts/manuscript.Rmd
adds Data Accessibility section as requested before going to proofs

@@
 the first steps to implement it. We are grateful to F. Michonneau for
-helpful comments on an earlier version of this manuscript.
+helpful comments on an earlier version of this manuscript, and reviews
+by Matthew Pennell, Associate Editor Richard FitzJohn, and an anonymous
+reviewer. At their behest, the reviews of FitzJohn and Pennell can be found in this
+project's GitHub page at [github.com/ropensci/RNeXML/issues/120](https://github.com/ropensci/RNeXML/issues/121) and [github.com/ropensci/RNeXML/issues/120](https://github.com/ropensci/RNeXML/issues/120), together with our replies and a record of our revisions.
+
+## Data Accessibility
+
+All software, scripts and data used in this paper can be found in the permanent 
+data archive Zenodo under the digital object identifier
+ doi:10.5281/zenodo.13131 [@zenodo]. This DOI corresponds to a snapshot of the GitHub repository
+at [github.com/ropensci/RNeXML](https://github.com/ropensci/RNeXML).
 

### 2015-05-29 manuscripts/manuscript.tex
committing submitted manuscript.tex file (automatically generated from .Rmd) and submit.tex (slightly modified by hand for ScholarOne compatibility)

@@
 taking the first steps to implement it. We are grateful to F. Michonneau
-for helpful comments on an earlier version of this manuscript.
+for helpful comments on an earlier version of this manuscript, and
+reviews by Matthew Pennell, Associate Editor Richard FitzJohn, and an
+anonymous reviewer. At their behest, the reviews of FitzJohn and Pennell
+can be found in this project's GitHub page at
+\href{https://github.com/ropensci/RNeXML/issues/121}{github.com/ropensci/RNeXML/issues/120}
+and
+\href{https://github.com/ropensci/RNeXML/issues/120}{github.com/ropensci/RNeXML/issues/120},
+together with our replies and a record of our revisions.
+
+\subsection{Data Accessibility}\label{data-accessibility}
+
+All software, scripts and data used in this paper can be found in the
+permanent data archive Zenodo under the digital object identifier
+\url{doi:10.5281/zenodo.13131} (Boettiger \emph{et al.} 2014). This DOI
+corresponds to a snapshot of the GitHub repository at
+\href{https://github.com/ropensci/RNeXML}{github.com/ropensci/RNeXML}.
 
@@
 \addcontentsline{toc}{section}{References}
+
+Boettiger, C., Vos, R., Chamberlain, S. \& Lapp, H. (2014). RNeXML
+v2.0.0. Retrieved from \url{http://dx.doi.org/10.5281/zenodo.13131}
 

### 2015-05-29 manuscripts/submit.tex
committing submitted manuscript.tex file (automatically generated from .Rmd) and submit.tex (slightly modified by hand for ScholarOne compatibility)

--- 2015-05-25 manuscripts/manuscript.tex
@@
-\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
+\documentclass[author-year, review, 11pt]{elsarticle} %review=doublespace preprint=single 5p=2 column
 %%% Begin My package additions %%%%%%%%%%%%%%%%%%%

### 2015-09-23 manuscripts/manuscript.Rmd
omit rrdf

@@
   - name: Carl Boettiger
-    affiliation: cstar
+    affiliation: ucb
     email: cboettig(at)gmail.com
@@
 address: 
-  - code: cstar
+  - code: ucb 
     address: |
-      Center for Stock Assessment Research, 
-      Department of Applied Math and Statistics, 
-      University of California, Mail Stop SOE-2,
-      Santa Cruz, CA 95064, USA
+      University of California, Berkeley,
+      130 Mulford Hall \#3114,
+      Berkeley, CA 94720-3114, USA 
   - code: ropensci

### 2015-10-27 inst/examples/merge_data.Rmd
Example with merging data tables

---
output:
 md_document:
   variant: markdown_github
---

```{r}
library("RNeXML")
library("dplyr")
library("geiger")
```

Let's generate a `NeXML` file using the tree and trait data from the `geiger` package's "primates" data:

```{r}
data("primates")
add_trees(primates$phy) %>% 
  add_characters(primates$dat, ., append=TRUE) %>% 
  taxize_nexml() -> nex 
```

(Note that we've used `dplyr`'s cute pipe syntax, but unfortunately our `add_` methods take the `nexml` object as the _second_
argument instead of the first, so this isn't as elegant since we need the stupid `.` to show where the piped output should go...)

We now read in the three tables of interest.  Note that we tell `get_characters` to give us species labels as there own column, rather than as rownames.  The latter is the default only because this plays more nicely with the default format for character matrices that is expected by `geiger` and other phylogenetics packages, but is in general a silly choice for data manipulation.

```{r}
otu_meta <- get_metadata(nex, "otu")
taxa <- get_taxa(nex)
char <- get_characters(nex, rownames_as_col = TRUE)
```

Now that we have nice `data.frame` objects for all our data, it's easy to join them into the desired table with a few obvious `dplyr` commands:

```{r}
taxa %>% 
  left_join(char, by = c("label" = "taxa")) %>% 
  left_join(otu_meta, by = c("id" = "parent_id")) %>%
  select(id, label, x, href)
```

### 2015-10-27 R/add_basic_meta.R

@@ -60,15 +60,19 @@ add_basic_meta <- function(title = NULL,
                            nexml = new("nexml")
                            ){
 
-  m <- get_metadata(nexml)
-
+  mymeta <- get_metadata(nexml)
+  m <- mymeta$content
+  names(m) <- mymeta$property
+  
+  
   if(!is.null(title)) 
     nexml <- add_meta(meta("dc:title", title), nexml)
   if(!is.null(creator) || creator == "") 
     nexml <- add_meta(meta("dc:creator", format(creator)), nexml)
   if(!is.null(pubdate)) 
-    if(is.null(m["dc:pubdate"]) | is.na(m["dc:pubdate"]))
-    nexml <- add_meta(meta("dc:pubdate", format(pubdate)), nexml)
+    if(!is.null(m)) 
+      if(is.null(m["dc:pubdate"]) | is.na(m["dc:pubdate"]))
+        nexml <- add_meta(meta("dc:pubdate", format(pubdate)), nexml)
   if(!is.null(description)) 
     nexml <- add_meta(meta("dc:description", description), nexml)
   if(!is.null(rights)){

### 2015-10-27 R/get_metadata.R

@@ -30,10 +30,10 @@ get_metadata <- function(nex, level = "nexml", depth = 1){
   
   if(is(nodes, "list"))
     out <- dplyr::bind_rows(lapply(nodes, function(x){
-      cbind(nexlist_as_data_frame(slot(x, "meta")), parent_id = parent_id(x))
+      nexlist_as_data_frame(slot(x, "meta"), append = parent_id(x))
     }))
   else
-    out <- cbind(nexlist_as_data_frame(slot(nodes, "meta")), parent_id = parent_id(nodes))
+    out <- nexlist_as_data_frame(slot(nodes, "meta"), append = parent_id(nodes))
   
   out
 }
@@ -42,20 +42,25 @@ parent_id <- function(node){
   if("id" %in% slotNames(node))
     slot(node, "id")[[1]]
   else
-    as.character(class(node))
+    NULL
 }
 
 
 ## Assumes meta elements do not have child meta elements
-nexlist_as_data_frame <- function(nexlist, skip = c("meta", "children")){
-  dplyr::bind_rows(lapply(nexlist, 
-    function(node){
+nexlist_as_data_frame <- function(nexlist, append = NULL, skip = c("meta", "children")){
+  if(length(nexlist) > 0){ 
+    out <- dplyr::bind_rows(lapply(nexlist, function(node){
       who <- slotNames(node)
       who <- who[-which(who %in% skip)]
       tmp <- sapply(who, function(x) slot(node, x))
       tmp[sapply(tmp,length) < 1] <- NA
-      data.frame(as.list(tmp))
-  }))
+      data.frame(as.list(tmp), stringsAsFactors=FALSE)
+    }))
+    if(!is.null(append)) out <- data.frame(out, parent_id = append, stringsAsFactors = FALSE)
+  out
+  } else { 
+    NULL
+  }
 }
 
 

### 2015-10-27 R/get_taxa.R

@@ -11,12 +11,11 @@
 #' nex <- nexml_read(comp_analysis)
 #' get_taxa(nex)
 #' @seealso  \code{\link{get_item}}
-get_taxa <-  
-          function(nexml){
-            out <- lapply(nexml@otus, function(otus)
-              sapply(otus@otu, function(otu) otu@label))
-            unname(unlist(out, recursive = FALSE))
-          }
+get_taxa <- function(nexml){
+  dplyr::bind_rows(lapply(nexml@otus, function(otus){
+    nexlist_as_data_frame(otus@otu, append = parent_id(otus))
+  }))
+}
 get_otu <- get_taxa
 
 #' get_taxa_list

### 2015-10-27 R/nexml_methods.R

@@ -41,7 +41,7 @@ setMethod("summary", signature("nexml"), function(object){
               "\t", nmeta, "meta elements", "\n",
               "\t", ncharacters, "character matrices", "\n",
               "\t", notu, "taxonomic units", "\n",
-              "Taxa: \t", paste(head(get_taxa(object)), collapse = ", "), "...", "\n\n",
+              "Taxa: \t", paste(head(get_taxa(object)$label), collapse = ", "), "...", "\n\n",
               "NeXML generated by", object@generator, "using",
               "schema version:", object@version, "\n", 
               "size:", capture.output(print(object.size(object), units="auto")), "\n"))

### 2015-10-27 inst/examples/merge_data.Rmd
merge examples
Much more general and flexible method for get_metadata

@@
 library("geiger")
+knitr::opts_chunk(message = FALSE, comment = NA)
 ```
@@
 ```{r}
-otu_meta <- get_metadata(nex, "otu")
+otu_meta <- get_metadata(nex, "otus/otu")
 taxa <- get_taxa(nex)
 char <- get_characters(nex, rownames_as_col = TRUE)
+```
+
+We can take a peek at what the tables look like, just to orient ourselves:
+
+```{r}
+otu_meta
+taxa
+head(char)
 ```
@@
   left_join(char, by = c("label" = "taxa")) %>% 
-  left_join(otu_meta, by = c("id" = "parent_id")) %>%
+  left_join(otu_meta, by = c("id" = "otu")) %>%
   select(id, label, x, href)
 ```
+
+Because these are all from the same otus block anyway, we haven't selected that column, but were it of interest it is also available in the taxa table.

### 2015-10-27 R/get_metadata.R

@@ -5,107 +5,77 @@
 #' 
 #' get_metadata 
 #' @param nexml a nexml object
-#' @param level the name of the level of element desired, or
-#'  use "all" to access all meta elements.  
-#' @return the requested metadata
+#' @param level the name of the level of element desired, see details
+#' @return the requested metadata as a data.frame. Additional columns
+#' indicate tha parent element of the return value.
+#' @details 'level' should be either the name of a child element of a NeXML document 
+#' (e.g. "otu", "characters"), or a path to the desired element, e.g. 'trees/tree'
+#' will return the metadata for all phylogenies in all trees blocks.
 #' @import XML
 #' @examples
-#' comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
+#' comp_analysis <- system.file("examples", "primates.xml", package="RNeXML")
 #' nex <- nexml_read(comp_analysis)
 #' get_metadata(nex)
+#' get_metadata(nex, "otus/otu")
 #' @export
-get_metadata <- function(nex, level = "nexml", depth = 1){
+get_metadata <- function(nex, level = "nexml"){
   
-  # level = c("nexml", "otus", "otu", "trees", "tree", "characters", "format", "states", "all")
-  nodes <- 
-    switch(level, 
-           nexml = nex,
-           otus = slot(nex, "otus")[[depth]],
-           otu = slot(slot(nex, "otus")[[depth]], "otu"),
-           trees = slot(nex, "trees")[[depth]],
-           tree = slot(slot(nex, "trees")[[depth]], "tree"),
-           characters = slot(nex, "characters")[[depth]],
-           format = slot(nex, "format")[[depth]],
-           states = slot(nex, "states")[[depth]])
-  
-  if(is(nodes, "list"))
-    out <- dplyr::bind_rows(lapply(nodes, function(x){
-      nexlist_as_data_frame(slot(x, "meta"), append = parent_id(x))
-    }))
+#  level = c("nexml", "otus", "trees", "characters", 
+#            "otus/otu", "trees/tree", "characters/format", "characters/matrix",
+#            "characters/format/states")
+#  level <- match.arg(level)
+ 
+  ## Append the 
+  if(level == "nexml")
+    level <- "meta"
   else
-    out <- nexlist_as_data_frame(slot(nodes, "meta"), append = parent_id(nodes))
[…]

### 2015-10-27 R/get_taxa.R

@@ -12,9 +12,7 @@
 #' get_taxa(nex)
 #' @seealso  \code{\link{get_item}}
 get_taxa <- function(nexml){
-  dplyr::bind_rows(lapply(nexml@otus, function(otus){
-    nexlist_as_data_frame(otus@otu, append = parent_id(otus))
-  }))
+  get_level(nexml, "otus/otu")
 }
 get_otu <- get_taxa
 

### 2016-06-26 README.Rmd
Adds citation instructions to README.

@@
 An extensive and rapidly growing collection of richly annotated phylogenetics data is now available in the NeXML format.
-NeXML relies on state-of-the-art data exchange technology to provide a format that can be both validated and extended, providing a data quality assurance and and adaptability to the future that is lacking in other formats [Vos et al 2012](http://doi.org/10.1093/sysbio/sys025 "NeXML: Rich, Extensible, and Verifiable Representation of Comparative Data and Metadata.").
+NeXML relies on state-of-the-art data exchange technology to provide a format that can be both validated and extended, providing a data quality assurance and adaptability to the future that is lacking in other formats.
+See [Vos et al 2012](http://doi.org/10.1093/sysbio/sys025 "NeXML: Rich, Extensible, and Verifiable Representation of Comparative Data and Metadata.") for further details on the NeXML format.
+
+How to cite
+-----------
+
+RNeXML has been published in the following article:
+
+> Boettiger C, Chamberlain S, Vos R and Lapp H (2016). “RNeXML: A Package for
+> Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data
+> in R.” _Methods in Ecology and Evolution_, **7**, pp. 352-357.
+> [doi:10.1111/2041-210X.12469](http://doi.org/10.1111/2041-210X.12469)
+
+Although the published version of the article is paywalled, the source of the manuscript, and a much better rendered PDF, are included in this package (in the `manuscripts` folder).
+You can also find it [freely available on arXiv](http://arxiv.org/abs/1506.02722).
 

### 2016-06-27 manuscripts/manuscript.Rmd

@@
 reviewer. At their behest, the reviews of FitzJohn and Pennell can be found in this
-project's GitHub page at [github.com/ropensci/RNeXML/issues/120](https://github.com/ropensci/RNeXML/issues/121) and [github.com/ropensci/RNeXML/issues/120](https://github.com/ropensci/RNeXML/issues/120), together with our replies and a record of our revisions.
+project's GitHub page at [github.com/ropensci/RNeXML/issues/121](https://github.com/ropensci/RNeXML/issues/121) and [github.com/ropensci/RNeXML/issues/120](https://github.com/ropensci/RNeXML/issues/120), together with our replies and a record of our revisions.
 

### 2016-06-27 manuscripts/manuscript.tex
omit rrdf

@@
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
 \newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\ImportTok}[1]{{#1}}
 \newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
 \newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
+\newcommand{\BuiltInTok}[1]{{#1}}
+\newcommand{\ExtensionTok}[1]{{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
+\newcommand{\RegionMarkerTok}[1]{{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
 \newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\RegionMarkerTok}[1]{{#1}}
-\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
 \newcommand{\NormalTok}[1]{{#1}}
@@
 character, and trait data in R}
-    \author[cstar]{Carl Boettiger\corref{c1}}
+    \author[ucb]{Carl Boettiger\corref{c1}}
    \ead{cboettig(at)gmail.com} 
@@
 
-      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
-Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
-95064, USA}    
+      \address[ucb]{University of California, Berkeley, 130 Mulford Hall \#3114, Berkeley,
+CA 94720-3114, USA}    
     \address[ropensci]{University of California, Berkeley, CA, USA}    
@@
 based on NEXUS files may also make inconsistent assumptions about
-tokens, quoting, or element lengths. Vos et al. (2012) estimates that as
-many as 15\% of the NEXUS files in the CIPRES portal contain
+tokens, quoting, or element lengths. Vos \emph{et al.} (2012) estimates
+that as many as 15\% of the NEXUS files in the CIPRES portal contain
 unrecoverable but hard to diagnose errors.
@@
 A detailed account of how the NeXML standard addresses these and other
-relevant challenges can be found in Vos et al. (2012). In brief, NeXML
-was designed with the following important properties. First, NeXML is
-defined by a precise grammar that can be programmatically
+relevant challenges can be found in Vos \emph{et al.} (2012). In brief,
+NeXML was designed with the following important properties. First, NeXML
+is defined by a precise grammar that can be programmatically
 \textbf{validated}; i.e., it can be verified whether a file precisely
@@
 capabilities for evolutionary and biodiversity science we refer the
-reader to Parr et al. (2011).)
+reader to Parr \emph{et al.} (2011).)
 
@@
 version can be found on Github at
-\href{}{\url{https://github.com/ropensci/RNeXML}}, and the stable
-version can be installed from the CRAN repository. \texttt{RNeXML} is
-part of the rOpenSci project. Users of \texttt{RNeXML} are encouraged to
-submit bug reports or feature requests in the issues log on Github, or
-the phylogenetics R users group list at
-\texttt{r-sig-phylo@r-project.org} for help. Vignettes with more
-detailed examples of specific features of RNeXML are distributed with
-the R package and serve as a supplement to this manuscript. Each of the
-vignettes can be found at
-\href{}{\url{http://ropensci.github.io/RNeXML/}}.
+\href{}{https://github.com/ropensci/RNeXML}, and the stable version can
+be installed from the CRAN repository. \texttt{RNeXML} is part of the
+rOpenSci project. Users of \texttt{RNeXML} are encouraged to submit bug
+reports or feature requests in the issues log on Github, or the
+phylogenetics R users group list at \texttt{r-sig-phylo@r-project.org}
+for help. Vignettes with more detailed examples of specific features of
+RNeXML are distributed with the R package and serve as a supplement to
+this manuscript. Each of the vignettes can be found at
+\href{}{http://ropensci.github.io/RNeXML/}.
 
@@
          log snout-vent length reef-dwelling
-taxon_8             -3.2777799             0
-taxon_9              2.0959433             1
+taxon_1              4.7532824             1
 taxon_10             3.1373971             0
-taxon_1              4.7532824             1
 taxon_2             -2.7624146             0
@@
 taxon_7              0.9099634             1
+taxon_8             -3.2777799             0
+taxon_9              2.0959433             1
 \end{verbatim}
@@
 
-\subsection{Creating and populating \texttt{nexml}
-objects}\label{creating-and-populating-nexml-objects}
+\subsection{\texorpdfstring{Creating and populating \texttt{nexml}
+objects}{Creating and populating nexml objects}}\label{creating-and-populating-nexml-objects}
 
@@
 
-As described in Vos et al. (2012), to encode metadata annotations NeXML
-uses the ``Resource Description Framework in Annotations'' (RDFa)
+As described in Vos \emph{et al.} (2012), to encode metadata annotations
+NeXML uses the ``Resource Description Framework in Annotations'' (RDFa)
 (Prud'hommeaux 2014). This standard provides for a strict
@@
 Third, \texttt{RNeXML} integrates with the R package \texttt{taxize}
-(Chamberlain \& Sz{ö}cs 2013) to mitigate one of the most common
-obstacles to reuse of phylogenetic data, namely the misspellings and
-inconsistent taxonomic naming with which OTU labels are often fraught.
-The \texttt{taxize\_nexml()} method in \texttt{RNeXML} uses
-\texttt{taxize} to match OTU labels against the NCBI database, and,
-where a unique match is found, it annotates the respective OTU with the
-matching NCBI identifier.
+(Chamberlain \& Szöcs 2013) to mitigate one of the most common obstacles
+to reuse of phylogenetic data, namely the misspellings and inconsistent
+taxonomic naming with which OTU labels are often fraught. The
+\texttt{taxize\_nexml()} method in \texttt{RNeXML} uses \texttt{taxize}
+to match OTU labels against the NCBI database, and, where a unique match
+is found, it annotates the respective OTU with the matching NCBI
+identifier.
 
@@
 comes with a demonstration that can be run from R using the following
-command: \texttt{demo("sparql", "RNeXML")}).
+command: \texttt{demo("sparql",\ "RNeXML")}).
 
@@
 including in evolutionary biology, ecology, and biodiversity (e.g.
-Rausher et al. (2010)). The effort involved with preparing and
+Rausher \emph{et al.} (2010)). The effort involved with preparing and
 submitting properly annotated data to archives remains a notable barrier
@@
 
+\hypertarget{refs}{}
+\hypertarget{ref-zenodo}{}
 Boettiger, C., Vos, R., Chamberlain, S. \& Lapp, H. (2014). RNeXML
@@
 
+\hypertarget{ref-Bollbackux5f2006}{}
 Bollback, J. (2006).\emph{BMC Bioinformatics}, \textbf{7}, 88. Retrieved
@@
 
-Chamberlain, S.A. \& Sz{ö}cs, E. (2013). Taxize: Taxonomic search and
+\hypertarget{ref-Chamberlainux5f2013}{}
+Chamberlain, S.A. \& Szöcs, E. (2013). Taxize: Taxonomic search and
 retrieval in r. \emph{F1000Research}. Retrieved from
@@
 
+\hypertarget{ref-Cranstonux5f2014}{}
 Cranston, K., Harmon, L.J., O'Leary, M.A. \& Lisle, C. (2014). Best
@@
 
+\hypertarget{ref-Drewux5f2013}{}
 Drew, B.T., Gazis, R., Cabezas, P., Swithers, K.S., Deng, J., Rodriguez,
@@
 
+\hypertarget{ref-Hartigux5f2012}{}
 Hartig, O. (2012). An introduction to sPARQL and queries over linked
@@
 
+\hypertarget{ref-Huelsenbeckux5f2003}{}
 Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
@@
 
+\hypertarget{ref-XML}{}
 Lang, D.T. (2013). \emph{XML: Tools for parsing and generating xML
@@
 
+\hypertarget{ref-Maddisonux5f1997}{}
 Maddison, D., Swofford, D. \& Maddison, W. (1997). NEXUS: An extensible
@@
 
+\hypertarget{ref-Midford2013}{}
 Midford, P., Dececchi, T., Balhoff, J., Dahdul, W., Ibrahim, N., Lapp,
@@
 
+\hypertarget{ref-phylobase}{}
 NESCENT R Hackathon Team. (2014). \emph{Phylobase: Base package for
@@
 
+\hypertarget{ref-taskview}{}
 O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
@@
 
+\hypertarget{ref-Paradisux5f2004}{}
 Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: Analyses of
@@
 
+\hypertarget{ref-Parr2011}{}
 Parr, C.S., Guralnick, R., Cellinese, N. \& Page, R.D.M. (2011).
@@
 
+\hypertarget{ref-Pennell2014}{}
 Pennell, M.W., Eastman, J.M., Slater, G.J., Brown, J.W., Uyeda, J.C.,
@@
 
+\hypertarget{ref-Pielux5f2009}{}
 Piel, W.H., Chan, L., Dominus, M.J., Ruan, J., Vos, R.A. \& Tannen, V.
@@
 
+\hypertarget{ref-Pielux5f2002}{}
 Piel, W.H., Donoghue, M.J. \& Sanderson, M.J. (2002). TreeBASE: A
@@
 
+\hypertarget{ref-W3Cux5f2014}{}
 Prud'hommeaux, E. (2014). SPARQL query language for rDF. \emph{W3C}.
@@
 
+\hypertarget{ref-R}{}
 R Core Team. (2014). \emph{R: A language and environment for statistical
@@
 
+\hypertarget{ref-Rausherux5f2010}{}
 Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
@@
 
+\hypertarget{ref-Revellux5f2012}{}
 Revell, L.J. (2012). Phytools: An r package for phylogenetic comparative
@@
 
+\hypertarget{ref-Stoddenux5f2014}{}
 Stodden, V. (2014). The scientific method in practice: Reproducibility
@@
 
+\hypertarget{ref-Stoltzfusux5f2012}{}
 Stoltzfus, A., O'Meara, B., Whitacre, J., Mounce, R., Gillespie, E.L.,
@@
 
+\hypertarget{ref-Tenopirux5f2011}{}
 Tenopir, C., Allard, S., Douglass, K., Aydinoglu, A.U., Wu, L., Read,
@@
 
+\hypertarget{ref-Vosux5f2012}{}
 Vos, R.A., Balhoff, J.P., Caravas, J.A., Holder, M.T., Lapp, H.,

### 2016-06-28 README.Rmd
use devtools-style README.Rmd

@@
-<!--
-%\VignetteEngine{knitr::knitr}
-%\VignetteIndexEntry{An Introduction to the RNeXML package}
--->
+---
+output:
+  md_document:
+    variant: markdown_github
+---
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>",
+  fig.path = "README-"
+)
+```
 
@@
 ```{r compile-settings, include=FALSE}
-## Set the paths for cache and figure
 library(methods)
-library(knitr)
-opts_chunk$set(cache = 1)
-basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile')) 
-opts_chunk$set(fig.path = paste("figure/", basename, "-", sep=""),
-               cache.path = paste("cache/", basename, "/", sep=""))
-
-```{r echo=FALSE}
-library(knitr)
-opts_knit$set(upload.fun = imgur_upload)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, comment = NA, verbose = TRUE)
 ```
@@
 
-The development version of RNeXML is [available on Github](https://github.com/ropensci/RNeXML).  With the `devtools` package installed on your system, RNeXML can be installed using:
+The latest stable release of RNeXML is on CRAN, and can be installed with the usual `install.packages("RNeXML")` command.
+Some of the more specialized functionality described in the Vignettes (such as RDF manipulation) requires additional packages which can be installed using:
+
+```{r eval=FALSE}
+install.packages("RNeXML", deps=TRUE, repos=c("https://cran.rstudio.com", "http://packages.ropensci.org"))
+```
+
+which will also install the development version of the RNeXML package.
+For most common tasks such as shown here, those additional packages are not required.
+The development version of RNeXML is also [available on Github](https://github.com/ropensci/RNeXML).
+With the `devtools` package installed on your system, RNeXML can be installed using:
 
@@
 library(devtools)
-install_github("RNeXML", "ropensci")
+install_github("ropensci/RNeXML")
 library(RNeXML)

### 2017-02-11 README.Rmd
Update README.Rmd

@@
 [![DOI](https://zenodo.org/badge/doi/10.5281/zenodo.13131.svg)](http://dx.doi.org/10.5281/zenodo.13131)
+[![codecov](https://codecov.io/gh/ropensci/RNeXML/branch/master/graph/badge.svg?branch=master)](https://codecov.io/gh/ropensci/RNeXML?branch=master)
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)

### 2017-02-12 README.Rmd
Update README.Rmd (#157)
* Update README.Rmd

* Update README.Rmd

--- 2016-06-28 README.Rmd
@@
 [![DOI](https://zenodo.org/badge/doi/10.5281/zenodo.13131.svg)](http://dx.doi.org/10.5281/zenodo.13131)
+[![codecov.io](https://codecov.io/github/ropensci/RNeXML/coverage.svg?branch=master)](https://codecov.io/github/ropensci/RNeXML?branch=master)
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)

### 2017-11-18 README.Rmd
Change fig.path in README.Rmd so output is in tools (#162)
This addresses a CRAN comment and allows CRAN to render the README.md as HTML for their purposes

Closes #160 on upstream
update doi badge

--- 2016-06-28 README.Rmd
@@
   comment = "#>",
-  fig.path = "README-"
+  fig.path = "tools/README-"
 )
@@
 
-[![DOI](https://zenodo.org/badge/doi/10.5281/zenodo.13131.svg)](http://dx.doi.org/10.5281/zenodo.13131)
+[![DOI](https://zenodo.org/badge/11856817.svg)](https://zenodo.org/badge/latestdoi/11856817)
+[![codecov.io](https://codecov.io/github/ropensci/RNeXML/coverage.svg?branch=master)](https://codecov.io/github/ropensci/RNeXML?branch=master)
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)

### 2018-04-21 R/get_rdf.R
- replace Sxslt with xslt - replace rrdf with librdf - fix sparql vignette

@@ -7,6 +7,7 @@
 #'   SPARQL queries from the rrdf R package.  
 #' @export
 #' @import httr XML
+#' @importFrom xml2 read_xml
 # @import Sxslt # not yet
 #' @examples \dontrun{
 #' f <- system.file("examples", "meta_example.xml", package="RNeXML")
@@ -23,13 +24,22 @@
 #' }
 get_rdf <- function(file){
 
+  
+  ## Soft dependency on xslt
+  if (!requireNamespace("xslt", quietly = TRUE)) {
+    stop("xslt package required to convert to rdf",
+         call. = FALSE)
+  }
+  xml_xslt <- getExportedValue("xslt", "xml_xslt")
+  
     if(is(file, "nexml")){
       who <- tempfile()
       nexml_write(x=file, file=who)
       file <- who
     }
     to_rdf <- system.file("examples", "RDFa2RDFXML.xsl", package="RNeXML")
-    rdf <- Sxslt::xsltApplyStyleSheet(file, to_rdf)
+    rdf <- 
+    rdf <- xml_xslt(xml2::read_xml(file), xml2::read_xml(to_rdf))
 
   rdf  
 }

### 2018-04-21 R/taxize_nexml.R

@@ -7,14 +7,23 @@
 #' @param nexml a nexml object
 #' @param type the name of the identifier to use
 #' @param ... additional arguments (not implemented yet)
-#' @import taxize
 #' @export 
 #' @examples \dontrun{
 #' data(bird.orders)
 #' birds <- add_trees(bird.orders)
 #' birds <- taxize_nexml(birds, "NCBI")
 #' }
+# @importFrom taxize get_uid
 taxize_nexml <- function(nexml, type = c("NCBI"), ...){
+  
+  ## Soft dependency on taxize
+  if (!requireNamespace("taxize", quietly = TRUE)) {
+    stop("taxize package required to convert look up taxonomic ids",
+         call. = FALSE)
+  }
+  get_uid <- getExportedValue("taxize", "get_uid")
+  
+  
           type <- match.arg(type)
           if(type == "NCBI"){
             for(j in 1:length(nexml@otus)){

### 2018-04-21 R/get_rdf.R
update docs

@@ -8,20 +8,6 @@
 #' @export
 #' @import httr XML
 #' @importFrom xml2 read_xml
-# @import Sxslt # not yet
-#' @examples \dontrun{
-#' f <- system.file("examples", "meta_example.xml", package="RNeXML")
-#' rdf <- get_rdf(f)
-#'
-#' ## Write to a file and read in with rrdf
-#' tmp <- tempfile()
-#' saveXML(rdf, tmp)
-#' library(rrdf)
-#' lib <- load.rdf(tmp)
-#' 
-#' ## Perform a SPARQL query:
-#' sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-#' }
 get_rdf <- function(file){
 
   

### 2018-04-22 README.Rmd
taxize as opt in  (#167)
* - replace Sxslt with xslt
- replace rrdf with librdf
- fix sparql vignette

* update docs

* update NEWS

* patch travis

* additional repositories not needed

@@
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
+[![Coverage Status](https://img.shields.io/codecov/c/github/ropensci/RNeXML/master.svg)](https://codecov.io/github/ropensci/RNeXML?branch=master)
 

### 2018-04-22 R/get_rdf.R

@@ -7,29 +7,25 @@
 #'   SPARQL queries from the rrdf R package.  
 #' @export
 #' @import httr XML
-# @import Sxslt # not yet
-#' @examples \dontrun{
-#' f <- system.file("examples", "meta_example.xml", package="RNeXML")
-#' rdf <- get_rdf(f)
-#'
-#' ## Write to a file and read in with rrdf
-#' tmp <- tempfile()
-#' saveXML(rdf, tmp)
-#' library(rrdf)
-#' lib <- load.rdf(tmp)
-#' 
-#' ## Perform a SPARQL query:
-#' sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-#' }
+#' @importFrom xml2 read_xml
 get_rdf <- function(file){
 
+  
+  ## Soft dependency on xslt
+  if (!requireNamespace("xslt", quietly = TRUE)) {
+    stop("xslt package required to convert to rdf",
+         call. = FALSE)
+  }
+  xml_xslt <- getExportedValue("xslt", "xml_xslt")
+  
     if(is(file, "nexml")){
       who <- tempfile()
       nexml_write(x=file, file=who)
       file <- who
     }
     to_rdf <- system.file("examples", "RDFa2RDFXML.xsl", package="RNeXML")
-    rdf <- Sxslt::xsltApplyStyleSheet(file, to_rdf)
+    rdf <- 
+    rdf <- xml_xslt(xml2::read_xml(file), xml2::read_xml(to_rdf))
 
   rdf  
 }

### 2018-04-22 tests/testthat/test_rdf.R

@@ -1,29 +1,39 @@
 context("rdf")
 
+library(rdflib)
+
 test_that("we can extract rdf-xml", {
 
-  if(require("Sxslt")){
-    f <- system.file("examples", "meta_example.xml", package="RNeXML")
-    rdf <- get_rdf(f)
-    expect_is(rdf, "XMLInternalXSLTDocument")
-  }
+  rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+  tmp <- tempfile()  # so we must write the XML out first
+  xml2::write_xml(rdf, tmp) 
+  
+  graph <- rdf_parse(tmp)
+  
+
+  expect_is(graph, "rdf")
+  
+  rdf_free(graph)
 })
 
 test_that("we can perform sparql queries with rrdf", {
   skip_on_travis()
 
-  if(require("Sxslt")){
-    f <- system.file("examples", "meta_example.xml", package="RNeXML")
-    rdf <- get_rdf(f)
-
-## Write to a file and read in with rrdf
-    saveXML(rdf, "rdf_meta.xml")
-    success <- require(rrdf)
-    if(success){
-      lib <- load.rdf("rdf_meta.xml")
-## Perform a SPARQL query:
-      out <- sparql.rdf(lib, "SELECT ?title WHERE { ?x <http://purl.org/dc/elements/1.1/title> ?title}")
-    }
-    unlink("rdf_meta.xml")
-  }
+  rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
+  tmp <- tempfile()  # so we must write the XML out first
+  xml2::write_xml(rdf, tmp) 
+  
+  graph <- rdf_parse(tmp)
+  
+  
+  root <- rdf_query(graph, 
+                    "SELECT ?uri WHERE { 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#rank> <http://rs.tdwg.org/ontology/voc/TaxonRank#Order> . 
+    ?id <http://rs.tdwg.org/ontology/voc/TaxonConcept#toTaxon> ?uri    
+}")
+  
+  expect_is(root, "data.frame")
+  
+  rdf_free(graph)
+  
 })

### 2018-05-05 tests/testthat/test_rdf.R
address failing tests on solaris

@@ -1,9 +1,12 @@
 context("rdf")
 
-library(rdflib)
+has_rdf <- require(rdflib)
 
 test_that("we can extract rdf-xml", {
 
+  skip_if_not(has_rdf)
+  skip_on_os("solaris")
+  
   rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
   tmp <- tempfile()  # so we must write the XML out first
   xml2::write_xml(rdf, tmp) 
@@ -18,7 +21,8 @@ test_that("we can extract rdf-xml", {
 
 test_that("we can perform sparql queries with rrdf", {
   skip_on_travis()
-
+  skip_on_os("solaris")
+  
   rdf <- get_rdf(system.file("examples/primates.xml", package="RNeXML"))
   tmp <- tempfile()  # so we must write the XML out first
   xml2::write_xml(rdf, tmp) 

### 2018-05-05 README.Rmd
update README

--- 2017-11-18 README.Rmd
@@
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
+[![Coverage Status](https://img.shields.io/codecov/c/github/ropensci/RNeXML/master.svg)](https://codecov.io/github/ropensci/RNeXML?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/RNeXML)](https://cran.r-project.org/package=RNeXML)
+![downloads](http://cranlogs.r-pkg.org/badges/grand-total/codemetar)
 

### 2018-05-05 README.Rmd
update README

--- 2017-11-18 README.Rmd
@@
 [![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
+[![Coverage Status](https://img.shields.io/codecov/c/github/ropensci/RNeXML/master.svg)](https://codecov.io/github/ropensci/RNeXML?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/RNeXML)](https://cran.r-project.org/package=RNeXML)
+![downloads](http://cranlogs.r-pkg.org/badges/grand-total/RNeXML)
 

### 2018-08-10 tests/testthat/test_taxonomy.R
fix for CRAN crul issues

@@ -1,6 +1,9 @@
 context("taxonomy")
 
-
+testthat::test_that("taxize", {
+  
+  testthat::skip_on_cran()
+  
 data(bird.orders)
 birdorders_small <- drop.tip(bird.orders, tip = 1:10)
 birds <- add_trees(birdorders_small)
@@ -14,7 +17,16 @@ chiroptera_super_small <- drop.tip(chiroptera, tip = 1:911)
 chir_super_small <- add_trees(chiroptera_super_small)
 chir_super_small <- taxize_nexml(chir_super_small, "NCBI")
 
+
+})
+
+
+
 test_that("taxize_nexml correctly collects ncbi identifiers", {
+  
+  testthat::skip_on_cran()
+  
+  
   expect_is(birds@otus, "ListOfotus")
   expect_is(birds@otus@.Data[[1]]@otu, "ListOfotu")
   expect_is(birds@otus@.Data[[1]]@otu[[1]], "otu")
@@ -41,6 +53,10 @@ test_that("we can extract taxonomy data from the object", {
 
 
 test_that("taxize_nexml throws appropriate warnings", {
+  
+  testthat::skip_on_cran()
+  
+  
   chir1 <- drop.tip(chiroptera, tip = 1:910)
   chir1 <- add_trees(chir1)
   expect_warning(taxize_nexml(chir1, "NCBI"))

### 2018-11-16 R/simmap.R
Replaces the simmap_ex saved data file with source NeXML
The simmap examples and vignette loaded a saved nexml-simmap object with
unclear (really, unknown) provenance, and which would be certain to break
without a path to repair for any change to the S4 class structure or
definition. This replaces the saved data file with NeXML source data file
that should result in the equivalent object.

I obtained this file in the following way:
- Read in the simmap_ex saved data file at a codepoint with compatible
  class structure and definition.
- Wrote to NeXML using write.nexml().
- Corrected the erron

@@ -6,10 +6,9 @@
 ## FIXME support writing multiphylos, list of multiphylos to nexml 
 
 
-#' simmap_to_nexml
+#' Convert phylo with attached simmap to nexml object
 #' 
-#' simmap_to_nexml
-#' @param phy a phy object containing simmap phy$maps element,
+#' @param phy a phylo object containing simmap `phy$maps`` element,
 #'  from the phytools package
 #' @param state_ids a named character vector giving the state 
 #'  names corresponding to the ids used to refer to each state
@@ -19,7 +18,7 @@
 #' @export 
 #' @import XML
 #' @examples
-#' data(simmap_ex)
+#' simmap_ex <- read.nexml(system.file("examples","simmap_ex.xml", package="RNeXML"))
 #' phy <- nexml_to_simmap(simmap_ex)
 #' nex <- simmap_to_nexml(phy) 
 simmap_to_nexml <- function(phy, state_ids = NULL){
@@ -110,17 +109,13 @@ simmap_edge_annotations <- function(maps, nexml, state_ids = NULL, char_id = "si
 
 ## Returns list of multiPhylo ...
 
-#' nexml_to_simmap
+#' Convert nexml object with simmap to phylo
 #'
-#' nexml_to_simmap
 #' @param nexml a nexml object
-#' @return a simmap object (phylo object with a $maps element 
+#' @return a simmap object (phylo object with a `$maps` element
 #'  for use in phytools functions).
 #' @export 
-#' @examples
-#' data(simmap_ex)
-#' phy <- nexml_to_simmap(simmap_ex)
-#' nex <- simmap_to_nexml(phy) 
+#' @describeIn simmap_to_nexml Convert nexml object with simmap to phylo
 nexml_to_simmap <- function(nexml){
 
   ## Get the statemap, if available
@@ -208,22 +203,6 @@ tree_to_simmap <- function(tree, otus, state_maps = NULL){
   ## Return phylo object
   phy
 }
-
-
-
-#' @name simmap_ex 
-#' @title A nexml class R object that includes simmap annotations
-#' @description A nexml object with simmap stochastic character mapping
-#'  annotations added to the edges, for use with the RNeXML package
-#'  parsing and serializing NeXML into formats that work with the ape and
-#'  phytools packages. 
-#' @docType data
-#' @usage simmap_ex
-#' @format a \code{nexml} instance
[…]

### 2018-11-28 README.Rmd
Adds comparing simplify = FALSE for metadata table
Also regenerated the README.md.

@@
 data(bird.orders)
-nexml_write(bird.orders, "test.xml")
+nexml_write(bird.orders, "test.xml", creator = "Carl Boettiger")
 ```
@@
 get_taxa(birds)
+get_metadata(birds, simplify = FALSE)
+```
+
+By default, the table of metadata annotations is simplified by combining
+`property` and `rel` into `property`, and the identifiers into `Meta`:
+```{r } 
 get_metadata(birds)
@@
 This next block defines a resource (link), described by the `rel` attribute as a homepage, a term in the `foaf` vocabulalry.
-Becuase `foaf` is not a default namespace, we will have to provide its URL in the full definition below.
+Because `foaf` is not a default namespace, we will have to provide its URL in the full definition below.
 
@@
 
-For this kind of richer annotation, it is best to build up our NeXML object sequentially. Frist
+For this kind of richer annotation, it is best to build up our NeXML object sequentially. First
 we will add `bird.orders` phylogeny to a new phylogenetic object, and then we will add the metadata

### 2018-11-30 README.Rmd
add appveyor, rebuild README

--- 2018-05-05 README.Rmd
@@
 ---
-output:
-  md_document:
-    variant: markdown_github
+output: github_document
 ---
@@
 [![DOI](https://zenodo.org/badge/11856817.svg)](https://zenodo.org/badge/latestdoi/11856817)
+[![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
+[![Windows build status](https://ci.appveyor.com/api/projects/status/dhiwp5blx2ns2yba/branch/master?svg=true)](https://ci.appveyor.com/project/cboettig/rnexml/branch/master)
+[![CRAN status](https://www.r-pkg.org/badges/version/RNeXML)](https://cran.r-project.org/package=RNeXML)
 [![codecov.io](https://codecov.io/github/ropensci/RNeXML/coverage.svg?branch=master)](https://codecov.io/github/ropensci/RNeXML?branch=master)
-[![Build Status](https://api.travis-ci.org/ropensci/RNeXML.png)](https://travis-ci.org/ropensci/RNeXML)
-[![Coverage Status](https://img.shields.io/codecov/c/github/ropensci/RNeXML/master.svg)](https://codecov.io/github/ropensci/RNeXML?branch=master)
-[![CRAN status](https://www.r-pkg.org/badges/version/RNeXML)](https://cran.r-project.org/package=RNeXML)
 ![downloads](http://cranlogs.r-pkg.org/badges/grand-total/RNeXML)

### 2018-12-06 R/taxize_nexml.R
swap out taxize for taxald

@@ -6,7 +6,7 @@
 #' containing the corresponding identifier. 
 #' @param nexml a nexml object
 #' @param type the name of the identifier to use
-#' @param ... additional arguments (not implemented yet)
+#' @param ... additional arguments to `[taxald::get_ids()]`
 #' @export 
 #' @examples \dontrun{
 #' data(bird.orders)
@@ -14,30 +14,39 @@
 #' birds <- taxize_nexml(birds, "NCBI")
 #' }
 # @importFrom taxize get_uid
-taxize_nexml <- function(nexml, type = c("NCBI"), ...){
+taxize_nexml <- function(nexml, type = c("ncbi", "itis", "col", "tpl",
+                                         "gbif", "wd"), ...){
   
   ## Soft dependency on taxize
-  if (!requireNamespace("taxize", quietly = TRUE)) {
-    stop("taxize package required to convert look up taxonomic ids",
+  if (!requireNamespace("taxald", quietly = TRUE)) {
+    stop("taxald package required to convert look up taxonomic ids",
          call. = FALSE)
   }
-  get_uid <- getExportedValue("taxize", "get_uid")
+  get_ids <- getExportedValue("taxald", "get_ids")
   
-  
-          type <- match.arg(type)
-          if(type == "NCBI"){
-            for(j in 1:length(nexml@otus)){
-              for(i in 1:length(nexml@otus[[j]]@otu)){
-                id <- get_uid(nexml@otus[[j]]@otu[[i]]@label)
-                if(is.na(id))
-                  warning(paste("ID for otu", nexml@otus[[j]]@otu[[i]]@label, "not found. Consider checking the spelling or alternate classification"))
-                else 
-                  nexml@otus[[j]]@otu[[i]]@meta <- New("ListOfmeta", list(
-                                 meta(href = paste0("http://ncbi.nlm.nih.gov/taxonomy/", id),
-                                      rel = "tc:toTaxon")))
+  type <- tolower(type)
+  type <- match.arg(type)
 
-              }
-            }
-          }
+  for(j in 1:length(nexml@otus)){
+    
+    # Resolve all ids at once
+    labels <- unname(vapply(nexml@otus[[j]]@otu, slot, character(1L), "label"))
+    taxa_ids <- get_ids(labels, type, format = "uri", ...)
+    
+    for(i in 1:length(taxa_ids)){
+      id <- taxa_ids[[i]]
+      if(is.na(id))
+        warning(paste("ID for otu", 
[…]

### 2018-12-08 README.Rmd

@@
 
-Getting Started
+Installation
 ---------------
@@
 ```{r eval=FALSE}
-install.packages("RNeXML", deps=TRUE, repos=c("https://cran.rstudio.com", "http://packages.ropensci.org"))
+install.packages("RNeXML", deps = TRUE)
 ```
 
-which will also install the development version of the RNeXML package.
-For most common tasks such as shown here, those additional packages are not required.
-The development version of RNeXML is also [available on Github](https://github.com/ropensci/RNeXML).
-With the `devtools` package installed on your system, RNeXML can be installed using:
+The development version can be installed using:
 
 ```{r eval=FALSE}
-library(devtools)
-install_github("ropensci/RNeXML")
-library(RNeXML)
+remotes::install_github("ropensci/RNeXML")
 ```
 
-```{r echo=FALSE}
-library(RNeXML)
-```
+Getting Started
+----------------
 
-Read in a `nexml` file into the `ape::phylo` format:
+See the vignettes below for both a general quick start and an overview of more specialized features.
 
-```{r}
-f <- system.file("examples", "comp_analysis.xml", package="RNeXML")
-nexml <- nexml_read(f)
-tr <- get_trees(nexml) # or: as(nexml, "phylo")
-plot(tr)
-```
-
-Write an `ape::phylo` tree into the `nexml` format:
-
-```{r}
-data(bird.orders)
-nexml_write(bird.orders, "test.xml")
-```
-
-A key feature of NeXML is the ability to formally validate the construction of the data file against the standard (the lack of such a feature in nexus files had lead to inconsistencies across different software platforms, and some files that cannot be read at all).
-While it is difficult to make an invalid NeXML file from `RNeXML`, it never hurts to validate just to be sure:
-
-```{r}
-nexml_validate("test.xml")
-```
-
-Extract metadata from the NeXML file:
-
-```{r } 
-birds <- nexml_read("test.xml")
-get_taxa(birds)
-get_metadata(birds) 
-```
-
---------------------------------------------
-
-Add basic additional metadata:
-
-```{r}
-  nexml_write(bird.orders, file="meta_example.xml",
-              title = "My test title",
-              description = "A description of my test",
-              creator = "Carl Boettiger <cboettig@gmail.com>",
-              publisher = "unpublished data",
-              pubdate = "2012-04-01")
-
-```
-By default, `RNeXML` adds certain metadata, including the NCBI taxon id numbers for all named taxa.
-This acts a check on the spelling and definitions of the taxa as well as providing a link to additional metadata about each taxonomic unit described in the dataset.
-
-### Advanced annotation
-
-We can also add arbitrary metadata to a NeXML tree by define `meta` objects:
-
-```{r}
-modified <- meta(property = "prism:modificationDate",
-                 content = "2013-10-04")
-```
-
-Advanced use requires specifying the namespace used.
-Metadata follows the RDFa conventions.
-Here we indicate the modification date using the prism vocabulary.
-This namespace is included by default, as it is used for some of the basic metadata shown in the previous example.
-We can see from this list:
-
-```{r}
-RNeXML:::nexml_namespaces
-```
-
-This next block defines a resource (link), described by the `rel` attribute as a homepage, a term in the `foaf` vocabulalry.
-Becuase `foaf` is not a default namespace, we will have to provide its URL in the full definition below.
-
-```{r}
-website <- meta(href = "http://carlboettiger.info", 
-                rel = "foaf:homepage")
-```
-
-Here we create a history node using the `skos` namespace.  We can also add id values to any metadata element to make the element easier to reference externally:
-
-```{r }
-  history <- meta(property = "skos:historyNote", 
-                  content = "Mapped from the bird.orders data in the ape package using RNeXML",
-                  id = "meta123")
-```
-
-For this kind of richer annotation, it is best to build up our NeXML object sequentially. Frist
-we will add `bird.orders` phylogeny to a new phylogenetic object, and then we will add the metadata
-elements created above to this object. Finally, we will write the object out as an XML file:
-
-```{r}
-  birds <- add_trees(bird.orders)
-  birds <- add_meta(meta = list(history, modified, website),
-                    namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#",
-                                   foaf = "http://xmlns.com/foaf/0.1/"),
-                    nexml=birds)
-  nexml_write(birds, 
-              file = "example.xml")
-
-```
-
-### Taxonomic identifiers
-
-Add taxonomic identifier metadata to the OTU elements:
-<!-- This block relies on a robust internet connection that can occassionally fail.  Also it's a bit slow, so don't run it. After all, this command is tested in the unit tests.-->
-
-```{r eval=FALSE}
-nex <- add_trees(bird.orders)
-nex <- taxize_nexml(nex)
-```
-
-## Working with character data
-
-NeXML also provides a standard exchange format for handling character data.
-The R platform is particularly popular in the context of phylogenetic comparative methods, which consider both a given phylogeny and a set of traits.
-NeXML provides an ideal tool for handling this metadata.
-
-### Extracting character data
-
-We can load the library, parse the NeXML file and extract both the characters and the phylogeny.
-
-```{r}
-library(RNeXML)
-nexml <- read.nexml(system.file("examples", "comp_analysis.xml", package="RNeXML"))
-traits <- get_characters(nexml)
-tree <- get_trees(nexml)
-```
-
-(Note that `get_characters` would return both discrete and continuous characters together in the same data.frame, but we use `get_characters_list` to get separate data.frames for the continuous `characters` block and the discrete `characters` block).
-
-We can then fire up `geiger` and fit, say, a Brownian motion model the continuous data and a Markov transition matrix to the discrete states:
-
-```{r}
-library(geiger)
-fitContinuous(tree, traits[1], ncores=1)
-fitDiscrete(tree, traits[2], ncores=1)
-```
-
-```{r include=FALSE}
-unlink("example.xml")
-unlink("meta_example.xml")
-unlink("test.xml")
-```
+- [A Brief Introduction to RNeXML](https://ropensci.github.com/RNeXML/articles/intro)
+- [RNeXML: A Package for Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data in R](https://ropensci.github.com/RNeXML/articles/manuscript.pdf) (published in MEE).
+- [Handling Metadata in RNeXML](https://ropensci.github.com/RNeXML/articles/metadata)
+- [The `nexml` S4 Object](https://ropensci.github.com/RNeXML/articles/S4)
+- [Semantic data & SPARQL with RNeXML](https://ropensci.github.com/RNeXML/articles/SPARQL)
+- [Extending NeXML: an example based on simmap](https://ropensci.github.com/RNeXML/articles/simmap)
 

### 2019-01-16 README.Rmd
update pkgdown with candidate hex
spellcheck
The dev version of the `spelling` package now spellchecks README and NEWS

@@
 
+# RNeXML  <img src="man/figures/logo.svg" align="right" alt="" width="120" />
+
 [![DOI](https://zenodo.org/badge/11856817.svg)](https://zenodo.org/badge/latestdoi/11856817)
@@
 ![downloads](http://cranlogs.r-pkg.org/badges/grand-total/RNeXML)
-
-RNeXML: The next-generation phylogenetics format comes to R
-=====================================================
 
@@
 
-Although the published version of the article is paywalled, the source of the manuscript, and a much better rendered PDF, are included in this package (in the `manuscripts` folder).
+Although the published version of the article is pay-walled, the source of the manuscript, and a much better rendered PDF, are included in this package (in the `manuscripts` folder).
 You can also find it [freely available on arXiv](http://arxiv.org/abs/1506.02722).

### 2019-06-10 R/nexml_methods.R
Rewrites summary() to provide an actual summary object (a list)
The summary object is a list of counts of key elements and metadata
that comprise the nexml object.

This closes #232, and partially addresses #233.

@@ -16,35 +16,157 @@ setMethod("show", signature("nexml"), function(object){
 })
 
 
-# FIXME: consider showing author/title/citation information if available?  
+# FIXME: consider showing author/title/citation information if available?
+#' Summary method for nexml objects
+#'
+#' Generates a list of various counts of the major elements that comprise a
+#' [nexml][nexml-class] object, such as number of different kinds of blocks,
+#' characters, states, OTUs (taxa), etc.
+#'
+#' @return A list with the following elements:
+#'  - `nblocks` the number of trees, otus, and characters blocks
+#'  - `ncharacters` the number of characters in each characters block
+#'  - `nstates` summary statistics of the number of character states per state set
+#'       defined for each characters block
+#'  - `nnonstdstatedefs` the number of polymorphic and uncertain states defined
+#'       for each character block
+#'  - `nmatrixrows` the number of rows in the matrix for each character block
+#'  - `ntrees` the number of trees contained in each trees block
+#'  - `notus` the number of OTUs defined in each OTUs block
+#'  - `nmeta` a list of the number of the number of metadata annotations at
+#'       several levels, specifically:
+#'       * `nexml` at the top (nexml) level
+#'       * `otu` at the OTU level, for each OTUs block
+#'       * `char` at the character level, for each characters block
+#'       * `state` at the character state level, for each characters block
+#' @examples
+#' nex <- nexml_read(system.file("examples", "comp_analysis.xml", package = "RNeXML"))
+#' s <- summary(nex)
+#' # number of major blocks:
+#' s$nblocks
+#'
+#' # each characters block defines 1 character:
+#' s$ncharacters
+#'
+#' # summary stats of states per character (for morphological matrices there is
+#' # typically one state set per character)
+#' s$nstates # note that first block is of continuous type, so no stats there
+#' @name summary.nexml
 setMethod("summary", signature("nexml"), function(object){
-          doc <- xmlParse(write.nexml(object))
-          nmeta <- length(getNodeSet(doc, "//x:meta", namespaces="x"))
-          ntree_blocks <- length(object@trees)
-          n_per_block <- sapply(unname(object@trees), function(x)  length(x@tree))
-          ntrees <- length(getNodeSet(doc, "//x:tree", namespaces="x"))
[…]

### 2019-06-10 R/nexml_methods.R
Rewrites the show() method to pretty-print the nexml summary
Also adds demonstration to Intro vignette, Partially addresses and
closes #233.

@@ -12,7 +12,89 @@
 #          cat(txt, "\n")
 #})
 setMethod("show", signature("nexml"), function(object){
-          summary(object)
+  s <- summary(object)
+
+  tr_block_counts <- ""
+  if (s$nblocks["trees"] > 0) {
+    tr_block_counts <-
+      paste0(sapply(1:s$nblocks["trees"],
+                    function(i)
+                      paste("\t   block", i, "contains",
+                            s$ntrees[i], "phylogenetic tree(s)")),
+             collapse = "\n")
+    tr_block_counts <- paste0("where:\n", tr_block_counts)
+  }
+
+  m_block_counts <- ""
+  if (s$nblocks["characters"] > 0) {
+    m_block_counts <-
+      paste0(
+        sapply(1:s$nblocks["characters"],
+               function(i) {
+                 p1 <- paste("\t   block", i, "defines", s$ncharacters[i])
+                 if (endsWith(object@characters[[i]]@`xsi:type`, "ContinuousCells"))
+                   p1 <- paste(p1, "continuous character(s)")
+                 else {
+                   p1 <- paste(p1, "standard/discrete character(s),")
+                   if (is.na(s$nstates[1, i]))
+                     p1 <- paste(p1, "no state definitions")
+                   else {
+                     p1 <- paste(p1, "with")
+                     if (s$nstates["Min.", i] == s$nstates["Max.", i])
+                       p1 <- paste(p1, s$nstates["Max.", i], "states each")
+                     else
+                       p1 <- paste(p1, s$nstates["Min.", i], "to", s$nstates["Max.", i], "states")
+                     p1 <- paste(p1,
+                                 "\n\t    and ",
+                                 sum(s$nnonstdstatedefs[i,]),
+                                 "polymorphic or uncertain state(s) defined")
+                   }
+                 }
+                 paste(p1,
+                       "\n\t   matrix", i, "has", s$nmatrixrows[i], "row(s)")
+               }),
+        collapse = "\n")
+    m_block_counts <- paste0("where:\n", m_block_counts)
+  }
+
+  cat(paste("A nexml object representing:\n",
+            "\t", s$nblocks["trees"], "phylogenetic tree block(s),",
+            tr_block_counts, "\n",
+            "\t", s$nblocks["characters"], "character block(s),",
+            m_block_counts, "\n",
+            "\t", sum(s$notus), "taxonomic units in", s$nblocks["otus"], "block(s)",
[…]

### 2019-08-22 README.Rmd
docs are built by ropensci directly now, so drop pkgdown

@@
 
-- [A Brief Introduction to RNeXML](https://ropensci.github.com/RNeXML/articles/intro)
-- [RNeXML: A Package for Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data in R](https://ropensci.github.com/RNeXML/articles/manuscript.pdf) (published in MEE).
-- [Handling Metadata in RNeXML](https://ropensci.github.com/RNeXML/articles/metadata)
-- [The `nexml` S4 Object](https://ropensci.github.com/RNeXML/articles/S4)
-- [Semantic data & SPARQL with RNeXML](https://ropensci.github.com/RNeXML/articles/SPARQL)
-- [Extending NeXML: an example based on simmap](https://ropensci.github.com/RNeXML/articles/simmap)
+- [A Brief Introduction to RNeXML](https://docs.ropensci.org/RNeXML/articles/intro)
+- [RNeXML: A Package for Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data in R](https://docs.ropensci.org/RNeXML/articles/manuscript.pdf) (published in MEE).
+- [Handling Metadata in RNeXML](https://docs.ropensci.org/RNeXML/articles/metadata)
+- [The `nexml` S4 Object](https://docs.ropensci.org/RNeXML/articles/S4)
+- [Semantic data & SPARQL with RNeXML](https://docs.ropensci.org/RNeXML/articles/sparql)
+- [Extending NeXML: an example based on simmap](https://docs.ropensci.org/RNeXML/articles/simmap)
 

### 2019-08-22 README.Rmd
attempt release of v2.4.0

@@
 - [A Brief Introduction to RNeXML](https://docs.ropensci.org/RNeXML/articles/intro)
-- [RNeXML: A Package for Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data in R](https://docs.ropensci.org/RNeXML/articles/manuscript.pdf) (published in MEE).
+- [RNeXML: A Package for Reading and Writing Richly Annotated Phylogenetic, Character, and Trait Data in R](https://github.com/ropensci/RNeXML/tree/master/manuscripts) (published in MEE).
 - [Handling Metadata in RNeXML](https://docs.ropensci.org/RNeXML/articles/metadata)

### 2020-01-28 tests/testthat/test_02_summary.R
precompute vignettes and skip slow tests because CRAN wants check to run FASTER

@@ -1,6 +1,10 @@
 testthat::context("summary for nexml objects")
 
 test_that("summary() doesn't error on various inputs", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   files <- list.files(path = system.file("examples", package = "RNeXML"),
                       pattern = ".*\\.xml$", full.names = TRUE)
   for (f in files) {

### 2020-01-28 tests/testthat/test_ape.R

@@ -2,11 +2,18 @@ context("ape")
 
 
 test_that("From ape::phylo to RNeXML::nexml object", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
    data(bird.orders)
    expect_is(as(bird.orders, "nexml"), class="nexml") 
 })
 
 test_that("We can go from various orderings of ape::phylo to RNeXML::nexml", {
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   data(bird.orders)
   nexml <- as(bird.orders, "nexml")
   phy <- as(nexml, "phylo")
@@ -19,6 +26,8 @@ test_that("We can go from various orderings of ape::phylo to RNeXML::nexml", {
 
 test_that("From nexml to multiPhylo", {
 
+  ## tests are too slow for CRAN
+  skip_on_cran()
 
   # part of base testing, could be replaced with higher level, but why 
   f <- system.file("examples", "trees.xml", package="RNeXML")
@@ -35,6 +44,10 @@ test_that("From nexml to multiPhylo", {
 
 ## This unit test is really not testing ape functions but just the higher-level nexml_write function...
 test_that("We can serialize the various versions of the ape format", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   data(bird.orders)
   nexml <- as(bird.orders, "nexml")
   nexml_write(nexml, file = "test.xml")
@@ -47,6 +60,10 @@ test_that("We can serialize the various versions of the ape format", {
 
 
 test_that("We can read and write NeXML to phylo and back without edge.lengths", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
            s <- "owls(((Strix_aluco,Asio_otus),Athene_noctua),Tyto_alba);"
            cat(s, file = "ex.tre", sep = "\n")
            owls <- read.tree("ex.tre")
@@ -61,6 +78,10 @@ test_that("We can read and write NeXML to phylo and back without edge.lengths", 
 
 
 test_that("Rooted trees remain rooted on conversions", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
           expect_true(is.rooted(bird.orders))
           expect_true(is.rooted(as(as(bird.orders, "nexml"), "phylo")))
           write.nexml(bird.orders, file = "tmp.xml")
@@ -70,6 +91,10 @@ test_that("Rooted trees remain rooted on conversions", {
 
[…]

### 2020-01-28 tests/testthat/test_comp_analysis.R

@@ -2,6 +2,9 @@ context("Comparative analysis")
 
 test_that("We can extract tree and trait data to run fitContinuous and fitDiscrete", {
   skip_if_not_installed("geiger")
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
 
   nexml <- read.nexml(system.file("examples", "comp_analysis.xml", package="RNeXML"))
   traits <- get_characters(nexml)
@@ -13,6 +16,9 @@ test_that("We can extract tree and trait data to run fitContinuous and fitDiscre
 
 
 test_that("We can serialize tree and trait data for a comparative analysis", {
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   skip_if_not_installed("geiger")
 
   require("geiger")

### 2020-01-28 tests/testthat/test_meta.R

@@ -4,6 +4,10 @@ context("meta")
 data(bird.orders)
 
 test_that("We can add additional metadata", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   ## The short version using an RNeXML API
   nex <- add_basic_meta(
               title = "My test title",
@@ -22,6 +26,10 @@ test_that("We can add additional metadata", {
 
 
 test_that("We can add R bibentry type metadata", {
+  
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
   ## The short version using an RNeXML API
 
   nex <- add_trees(bird.orders)
@@ -35,6 +43,10 @@ test_that("We can add R bibentry type metadata", {
 })
 
 test_that("Adding meta data has some basic error checking", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   testthat::expect_error(add_meta(NULL, level = "nexml"))
   m <- meta("foo-rel", "a test")
   testthat::expect_error(add_meta(list(m, "not a meta")))
@@ -42,6 +54,10 @@ test_that("Adding meta data has some basic error checking", {
 })
 
 test_that("ResourceMeta maps rel to property for simplicity of API", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   m <- meta(rel = "foo-rel", href="http://example.com/")
   testthat::expect_is(m, "ResourceMeta")
   testthat::expect_error(m@property)
@@ -54,6 +70,10 @@ test_that("ResourceMeta maps rel to property for simplicity of API", {
 })
 
 test_that("ResourceMeta maps meta to children for simplicity of API", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   nested <- c(meta(property = "foo-prop", content = "foo"),
               meta(property = "bar-prop", content = "bar"))
   m <- meta(rel = "foo-rel", children = nested)
@@ -75,6 +95,10 @@ test_that("ResourceMeta maps meta to children for simplicity of API", {
 })
 
 test_that("Citation BibEntry objects are transformed to structured metadata", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   nexml_cit <- nexml_citation(citation("RNeXML"))
   testthat::expect_is(nexml_cit, "list")
   testthat::expect_length(nexml_cit, 1)
@@ -92,6 +116,10 @@ test_that("Citation BibEntry objects are transformed to structured metadata", {
 })
 
 test_that("We can add additional metadata", {
[…]

### 2020-01-28 tests/testthat/test_parsing.R

@@ -3,6 +3,10 @@ context("parsing")
 # More lower-level parsing tests in inheritance
 
 test_that("We can parse a NeXML file to an S4 RNeXML::tree object", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   f <- system.file("examples", "trees.xml", package="RNeXML")
   doc <- xmlParse(f)
   root <- xmlRoot(doc)
@@ -13,7 +17,9 @@ test_that("We can parse a NeXML file to an S4 RNeXML::tree object", {
 
 
 test_that("We preserve existing namespace", {
-
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   f <- system.file("examples", "biophylo.xml", package="RNeXML")
   nex <- nexml_read(f)
 
@@ -43,6 +49,10 @@ test_that("We preserve existing namespace", {
 })
 
 test_that("base namespace gets added if not present", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   doc <- xmlParse(system.file("examples", "no-base-ns.xml", package="RNeXML"))
   xmlroot <- xmlRoot(doc)
 
@@ -59,6 +69,10 @@ test_that("base namespace gets added if not present", {
 
 
 test_that("files with rootedge can be parsed and roundtripped", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   f <- system.file("examples", "coal.xml", package = "RNeXML")
   nex <- read.nexml(f)
 

### 2020-01-28 tests/testthat/test_serializing.R

@@ -4,6 +4,10 @@ context("serializing")
 ## More tests at lower-level serializing from S4 to XML in inheritance.R
 
 test_that("We can serialize ape to S4 RNeXML into valid NeXML",{
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   data(bird.orders)
 
 
@@ -35,6 +39,10 @@ test_that("We can serialize ape to S4 RNeXML into valid NeXML",{
 
 
 test_that("We can serialize parsed NeXML to S4 RNeXML into valid NeXML",{
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   root <- xmlRoot(xmlParse(system.file("examples", "trees.xml", package="RNeXML")))
   tree <- as(root, "nexml")
   nexml_write(tree, file="test.xml")
@@ -48,6 +56,10 @@ test_that("We can serialize parsed NeXML to S4 RNeXML into valid NeXML",{
   })
 
 test_that("We can correctly serialize XML literals as metadata", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   nex1 <- read.nexml(system.file("examples", "phenex.xml", package="RNeXML"))
   m_xml1 <- get_metadata(nex1, "characters/format/states/state") %>%
     dplyr::filter(property == "ps:describesPhenotype")
@@ -66,6 +78,10 @@ test_that("We can correctly serialize XML literals as metadata", {
 })
 
 test_that("we can serialize NeXML with rootedge to a valid NeXML file", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   f <- system.file("examples", "coal.xml", package = "RNeXML")
   nex <- read.nexml(f)
   write.nexml(nex, file = "test.xml")

### 2020-01-28 tests/testthat/test_taxonomy.R

@@ -17,14 +17,16 @@ chir_super_small <- add_trees(chiroptera_super_small)
 
 test_that("taxize_nexml correctly collects ncbi identifiers", {
   
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   testthat::skip_on_cran()
   testthat::skip_if_not_installed("taxadb")
-  
-  birds <- taxize_nexml(birds, "NCBI")
-  
-  chir <- taxize_nexml(chir, "NCBI")
-  chir_super_small <- taxize_nexml(chir_super_small, "NCBI")
-  
+  suppressWarnings({
+    birds <- taxize_nexml(birds, "NCBI")
+    chir <- taxize_nexml(chir, "NCBI")
+    chir_super_small <- taxize_nexml(chir_super_small, "NCBI")
+  })
   
   expect_is(birds@otus, "ListOfotus")
   
@@ -44,6 +46,9 @@ test_that("taxize_nexml correctly collects ncbi identifiers", {
 })
 
 test_that("we can extract taxonomy data from the object", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
   
   expect_is(get_metadata(birds, "otus/otu"), "data.frame")
   expect_is(get_metadata(chir_super_small, 'otus/otu'), "data.frame")

### 2020-01-28 tests/testthat/test_toplevel_api.R

@@ -3,7 +3,8 @@ context("top level API")
 
 test_that("read.nexml works", {
   ## The short version using an RNeXML API
-
+  ## tests are too slow for CRAN
+  skip_on_cran()
   f <- system.file("examples", "trees.xml", package="RNeXML")
   nex <- read.nexml(f) # check alias
   expect_is(nex, "nexml")
@@ -12,7 +13,8 @@ test_that("read.nexml works", {
 
 test_that("write.nexml works (from ape::phylo)", {
   ## The short version using an RNeXML API
-
+  ## tests are too slow for CRAN
+  skip_on_cran()
   data(bird.orders)
   nexml_write(bird.orders, file="example.xml")
   write.nexml(bird.orders, file="example.xml") # check alias too
@@ -27,15 +29,19 @@ test_that("write.nexml works (from ape::phylo)", {
 
 
 test_that("write.nexml can write multiple trees at once ", {
+  ## tests are too slow for CRAN
+  skip_on_cran()
   f <- system.file("examples", "trees.xml", package="RNeXML")
   nex <- nexml_read(f)
   trees <- get_trees(nex)
 
   ##  We can write a listOfmultiPhylo if the argument is named
+  
   nexml_write(trees = trees, file="example.xml")
   expect_true_or_null(nexml_validate("example.xml"))
 
   # we can write a multiPhylo (or phylo) by attempting coercion on the first argument instead:  
+  
   nexml_write(trees[[1]], file="example.xml")
   expect_true_or_null(nexml_validate("example.xml"))
 
@@ -47,6 +53,8 @@ test_that("write.nexml can write multiple trees at once ", {
 
 
 test_that("We can get the right level of lists of trees ", {
+  ## tests are too slow for CRAN
+  skip_on_cran()
 
   f <- system.file("examples", "trees.xml", package="RNeXML")
   nex <- nexml_read(f)

### 2020-01-28 tests/testthat/test_validate.R

@@ -1,10 +1,17 @@
 context("Online validator tool")
 test_that("example file validates", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
   f <- system.file("examples", "trees.xml", package="RNeXML")
   expect_true_or_null(nexml_validate(f)) # null if we cannot perform validation, don't fail
 })
 
 test_that("RNeXML-generated file validates", {
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
+  
   data(bird.orders)
   f <- nexml_write(bird.orders, file="test.xml") 
   o <- nexml_validate(f)
@@ -18,8 +25,15 @@ test_that("RNeXML-generated file validates", {
 
 
 test_that("Validation can fail gracefully", {
+  
+  ## tests are too slow for CRAN
+  skip_on_cran()
+  
   f <- system.file("examples/sparql.newick", package="RNeXML")
+  
+  testthat::expect_warning(
   o <- nexml_validate(f)
+  )
   if(!is.null(o)) { 
     expect_false(o)
   } else {

### 2020-01-28 tests/test-all.R
try try again

@@ -1,5 +1,6 @@
 library("testthat")
 Sys.setenv("_R_CHECK_LENGTH_1_CONDITION_" = TRUE)
+
 test_check("RNeXML")
 
 

### 2022-02-08 README.Rmd
update url redirects

@@
 NeXML relies on state-of-the-art data exchange technology to provide a format that can be both validated and extended, providing a data quality assurance and adaptability to the future that is lacking in other formats.
-See [Vos et al 2012](http://doi.org/10.1093/sysbio/sys025 "NeXML: Rich, Extensible, and Verifiable Representation of Comparative Data and Metadata.") for further details on the NeXML format.
+See [Vos et al 2012](https://doi.org/10.1093/sysbio/sys025 "NeXML: Rich, Extensible, and Verifiable Representation of Comparative Data and Metadata.") for further details on the NeXML format.
 
@@
 > in R.” _Methods in Ecology and Evolution_, **7**, pp. 352-357.
-> [doi:10.1111/2041-210X.12469](http://doi.org/10.1111/2041-210X.12469)
+> [doi:10.1111/2041-210X.12469](https://doi.org/10.1111/2041-210X.12469)
 
 Although the published version of the article is pay-walled, the source of the manuscript, and a much better rendered PDF, are included in this package (in the `manuscripts` folder).
-You can also find it [freely available on arXiv](http://arxiv.org/abs/1506.02722).
+You can also find it [freely available on arXiv](https://arxiv.org/abs/1506.02722).
 
@@
 
-[![ropensci footer](http://ropensci.org/public_images/github_footer.png)](http://ropensci.org)
+[![ropensci footer](https://ropensci.org//public_images/github_footer.png)](https://ropensci.org/)

### 2022-05-09 R/add_characters.R
redoc

@@ -16,9 +16,11 @@
 #' files from scratch with both characters and trees.  
 #' @include classes.R
 #' @examples
+#' if( require("geiger") ) {
 #' library("geiger")
 #' data(geospiza)
 #' geiger_nex <- add_characters(geospiza$dat)
+#' }
 #' @export 
 add_characters <- function(x, 
                            nexml = new("nexml"), 
@@ -59,7 +61,8 @@ add_character_nodes <- function(nexml, x){
          characters <- nexml.characters(
              id = uid,
              about = paste0("#", uid))
-         if(class(x[[i]][[1]]) == "numeric")  ## Should be numeric but not integer!
+         cls <- class(x[[i]][[1]])
+         if(cls == "numeric")  ## Should be numeric but not integer!
            type <- "ContinuousCells"
          else ## Should be integer!
            type <- "StandardCells"

### 2022-05-09 R/add_trees.R

@@ -13,9 +13,11 @@ NULL
 #' @return a nexml object containing the phy in nexml format. 
 #' @export 
 #' @examples 
-#' library("geiger")
+#' 
+#' if( require("geiger") ) {
 #' data(geospiza)
 #' geiger_nex <- add_trees(geospiza$phy)
+#' }
 add_trees <- function(phy, 
                       nexml=new("nexml"), 
                       append_to_existing_otus=FALSE){

### 2022-05-09 R/classes.R

@@ -103,7 +103,7 @@ setMethod("toNeXML",
           function(object, parent){
             parent <- callNextMethod()
             cnt <- object@content
-            if (length(cnt) > 0 && class(cnt) == "XMLString") {
+            if (length(cnt) > 0 && is(cnt, "XMLString")) {
               # need to add back in the namespace definitions used by the
               # XML value to avoid a stream of warnings, even if this will
               # probably be redundant with the doc root

### 2022-05-09 R/nexml_add.R

@@ -12,10 +12,12 @@
 #' @seealso \code{\link{add_trees}} \code{\link{add_characters}} \code{\link{add_meta}} \code{\link{add_namespaces}}    
 #' @export 
 #' @examples
+#' #' if( require("geiger") ) {
 #' library("geiger")
 #' data(geospiza)
 #' geiger_nex <- nexml_add(geospiza$phy, type="trees")
 #' geiger_nex <- nexml_add(geospiza$dat, nexml = geiger_nex, type="characters")
+#' }
 nexml_add <- function(x, 
                       nexml = new("nexml"), 
                       type = c("trees", 

### 2022-05-09 R/nexml_write.R

@@ -25,7 +25,7 @@
 #' 
 #' \dontrun{ # takes > 5s
 #'  ## Assemble a nexml section by section and then write to file:
-#'  library(geiger)
+#' if( require("geiger") ) {
 #'  data(geospiza)
 #'  nexml <- add_trees(geospiza$phy) # creates new nexml
 #'  nexml <- add_characters(geospiza$dat, nexml = nexml) # pass the nexml obj to append character data
@@ -50,7 +50,7 @@
 #'              title = "My title", 
 #'              creator = "Carl Boettiger",
 #'              file = ex)
-#' 
+#' }
 #' }
 nexml_write <- function(x = nexml(),
                         file = NULL,

### 2022-05-09 tests/testthat/helper-RNeXML.R

@@ -8,5 +8,4 @@ expect_true_or_null <- function(o){
   }
 }
 
-
 library("XML")

### 2022-05-09 tests/testthat/test_global_ids.R

@@ -3,10 +3,11 @@ context("Set global (uuid) identifiers")
 test_that("We can generate valid EML with uuid ids on all elements", {
 
   skip_if_not_installed("geiger")
-
+  skip_if_not_installed("uuid")
+  
   if(require("uuid")){
     options(uuid = TRUE)
-
+    require(geiger)
     data(geospiza)
     add_trees(geospiza$phy)
     nexml <- add_characters(geospiza$dat)

### 2022-05-09 tests/testthat/test_publish.R

@@ -14,7 +14,6 @@ if(is(status, "try-error") || (is(status, "response") && status$status_code != 2
 } else {
 
 
-
 ## Create example file
   library(geiger)
   data(geospiza)
